### 2019-03-13 paper.tex
first commit

%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For double-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For single-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission, w/ required CCS and ACM Reference
%\documentclass[acmsmall]{acmart}\settopmatter{}

%% Journal information
%% Supplied to authors by publisher for camera-ready submission;
%% use defaults for review submission.
\acmJournal{PACMPL}
\acmVolume{1}
\acmNumber{CONF} % CONF = POPL or ICFP or OOPSLA
\acmArticle{1}
\acmYear{2018}
\acmMonth{1}
\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}

%% Copyright information
%% Supplied to authors (based on authors' rights management selection;
%% see authors.acm.org) by publisher for camera-ready submission;
%% use 'none' for review submission.
\setcopyright{none}
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\copyrightyear{2018}           %% If different from \acmYear

%% Bibliography style
\bibliographystyle{ACM-Reference-Format}
%% Citation style
%% Note: author/year citations are required for papers published as an
%% issue of PACMPL.
\citestyle{acmauthoryear}   %% For author/year citations

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Note: Authors migrating a paper from PACMPL format to traditional
%% SIGPLAN proceedings format must update the '\documentclass' and
%% topmatter commands above; see 'acmart-sigplanproc-template.tex'.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% Some recommended packages.
\usepackage{booktabs}   %% For formal tables:
                        %% http://ctan.org/pkg/booktabs
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
                        %% http://ctan.org/pkg/subcaption

\begin{document}

%% Title information
\title{Compile Time Verification of Deterministic properties for Sequential Programs}         %% [Short Title] is optional;
                                        %% when present, will be used in
                                        %% header instead of Full Title.
%\titlenote{with title note}             %% \titlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'
%\subtitle{Subtitle}                     %% \subtitle is optional
%\subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'

%% Author information
%% Contents and number of authors suppressed with 'anonymous'.
%% Each author should be introduced by \author, followed by
%% \authornote (optional), \orcid (optional), \affiliation, and
%% \email.
%% An author may have multiple affiliations and/or emails; repeat the
%% appropriate command.
%% Many elements are not rendered, but should be provided for metadata
%% extraction tools.

%% Author with single affiliation.
\author{First1 Last1}
\authornote{with author1 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
  \position{Position1}
  \department{Department1}              %% \department is recommended
  \institution{Institution1}            %% \institution is required
  \streetaddress{Street1 Address1}
  \city{City1}
  \state{State1}
  \postcode{Post-Code1}
  \country{Country1}                    %% \country is recommended
}
\email{first1.last1@inst1.edu}          %% \email is recommended

%% Author with two affiliations and emails.
\author{First2 Last2}
\authornote{with author2 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
  \position{Position2a}
  \department{Department2a}             %% \department is recommended
  \institution{Institution2a}           %% \institution is required
  \streetaddress{Street2a Address2a}
  \city{City2a}
  \state{State2a}
  \postcode{Post-Code2a}
  \country{Country2a}                   %% \country is recommended
}
\email{first2.last2@inst2a.com}         %% \email is recommended
\affiliation{
  \position{Position2b}
  \department{Department2b}             %% \department is recommended
  \institution{Institution2b}           %% \institution is required
  \streetaddress{Street3b Address2b}
  \city{City2b}
  \state{State2b}
  \postcode{Post-Code2b}
  \country{Country2b}                   %% \country is recommended
}
\email{first2.last2@inst2b.org}         %% \email is recommended

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
Text of abstract \ldots.
\end{abstract}

%% 2012 ACM Computing Classification System (CSS) concepts
%% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10011007.10011006.10011008</concept_id>
<concept_desc>Software and its engineering~General programming languages</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10003456.10003457.10003521.10003525</concept_id>
<concept_desc>Social and professional topics~History of programming languages</concept_desc>
<concept_significance>300</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Software and its engineering~General programming languages}
\ccsdesc[300]{Social and professional topics~History of programming languages}
%% End of generated code

%% Keywords
%% comma separated list
\keywords{keyword1, keyword2, keyword3}  %% \keywords are mandatory in final camera-ready submission

%% \maketitle
%% Note: \maketitle command must come after title commands, author
%% commands, abstract environment, Computing Classification System
%% environment and commands, and keywords command.
\maketitle

\section{Introduction}

Text of paper \ldots

%% Acknowledgments
\begin{acks}                            %% acks environment is optional
                                        %% contents suppressed with 'anonymous'
  %% Commands \grantsponsor{<sponsorID>}{<name>}{<url>} and
  %% \grantnum[<url>]{<sponsorID>}{<number>} should be used to
  %% acknowledge financial support and will be used by metadata
  %% extraction tools.
  This material is based upon work supported by the
  \grantsponsor{GS100000001}{National Science
    Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
  No.~\grantnum{GS100000001}{nnnnnnn} and Grant
  No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
  conclusions or recommendations expressed in this material are those
  of the author and do not necessarily reflect the views of the
  National Science Foundation.
\end{acks}

%% Bibliography
%\bibliography{bibfile}

%% Appendix
\appendix
\section{Appendix}

Text of appendix \ldots

\end{document}

### 2019-03-14 paper.tex
Intro start

@@
 
-Text of paper \ldots
+Writing correct programs is hard.
+Programmers often write correctness specifications for their code in the form of assertions.
+They write test cases that cover several usage scenarios to strengthen their belief that
+the code works as intended. Despite all the testing effort, it often happens that a test suite produces
+diverging results for different users or for the same user when the test suite is run again. When a test suite fails
+nondeterministically, it  reduces the programmers' confidence in the program. Additionally, reproducing
+such bugs is difficult making debugging an onerous task. Examples of the underlying causes of such flakiness 
+could be OS scheduling in case of concurrent programs, use of an API that has nondeterministic specification,
+accessing system properties such as the file system or environment variables. In this paper, we focus on
+nondeterminism arising from the usage of APIs that have nondeterministic specifications in java programs.
+Instances of such APIs in the JDK include Date(), Random(), iterating over a HashSet. Note that mere calls
+to these APIs is not problematic. But if the programmer intends the results of these calls to be deterministic,
+it could result in undesirable results and we aim to capture this behavior and report it. The high level goal
+of our work is to provide programmers with a tool for specifying  deterministic properties ina program
+and be able to verify them at compile time.
+
+Ideally with any program, we expect the output to be the same.
+Example sources of nondeterminism
+
+Test suites, production environemnt, Flakiness, recent work in this area
+
+Dependency on APIs , Talk about specifications
+
+Approach is to build a type system
+
+Focus on collections
+
+Ease of extensibility of our approach
+
+We make the folowing contributions through this paper:
+\begin{enumerate}
+	\item Design and implement a type system for expressing deterministic properties in sequential programs.
+	\item Annotate several classes in JDK. By doing this, we have translated the specification for these APIs from
+	human readable comments to machine readable and verifyable documentation.
+	\item Perform a case study of a real work Java program - Randoop (loc, etc) which revealed bugs.
+	\item Compare our technique to flaky test detectors and present results.
+\end{enumerate}
+
+\section{Design}
 

### 2019-03-16 introduction.tex
intro changes

\section{Introduction}

Writing correct programs is hard. Programmers often write correctness specifications for their code in the form of assertions. They write test cases that cover several usage scenarios to strengthen their belief that
the code works as intended. Despite all the testing effort, it often happens that a test suite produces
diverging results for different users or for the same user when the test suite is run again. When a test suite fails
nondeterministically, it  reduces the programmers' confidence in the program. Additionally, reproducing
such bugs is difficult making debugging an onerous task. Examples of the underlying causes of such flakiness 
could be OS scheduling in case of concurrent programs, use of an API that has nondeterministic specification,
accessing system properties such as the file system or environment variables. In this paper, we focus on
nondeterminism arising from the use of APIs that have nondeterministic specifications in java programs.
Instances of such APIs in the JDK include Date(), Random(), and HashSet iteration. Note that mere calls
to these APIs is not problematic. But if the programmer intends the results of these calls to be deterministic,
it could result in undesirable results. We aim to capture this behavior and report it. The high level goal
of our work is to provide programmers with a tool for specifying  deterministic properties in a program
and be able to verify them at compile time.

Recent work in this area~\cite{nondex,deflaker} has focussed its efforts on identifying flaky test cases.
NONDEX~\cite{nondex} is a tool that manually identifies methods in the JDK that have nondeterministic specifications, builds models for
these methods that behave nondeterministically, and  use sa modified JVM that runs the test suite multiple times.
Diverging test results in this setting indicate a flaky test. DeFlaker~\cite{deflaker} looks at a range of commit versions
of a code, and marks a test as flaky if it doesn't execute any modified code but still fails in the newer version. These techniques
have been able to identify issues in real world programs, some of which have been fixed by the developers. This shows that nondeterminism
is an important problem and solving it at an earlier phase in the software development lifecycle would be greatly beneficial to 
developers.

In this paper, we propose to build a lightweight compile time checker that can verify deterministic properties. We design a
type system that allows programmers to express deterministic properties as types in their programs. We use
java's annotation support to build our type system. Since our approach is based on compile time type checking, we provide
soundness guarantees. Another advantage of our approach
is that the analysis designer doesn't need to make changes to the JVM or have to rerun the test suite multiple times in order
to detect flakiness. Like with any static time checker, our approach is prone to producing false positives. There is also increased programmer effort involved in writing type annotations while developing.
We believe (give numbers here) that the benefits of having a verified system far exceed the programmer efforts.

Our type system hierarchy consists of three types: a) @Det which represents values that will be the same across executions,
b) @OrderNonDet which represents collections that are guaranteed to contain the same elements albeit in possibly
different iteration order, and c) @NonDet which provides no determism guarantees.  We found this type hierarchy 
is expressive enough to identify nondeterministic behaviors in production software. We annotated all the collection
classes (Lists, Sets and Maps) in the JDK with our new type system. Our tool is therefore guaranteed to report any violation
of deterministic properties in a client program caused by methods in these classes.

We make the folowing contributions through this paper:
\begin{enumerate}
	\item We provide a design and implemention of a type system for expressing deterministic properties in sequential programs.
	\item We annotated several classes in JDK. By doing this, we have translated the specification for these APIs from
	human readable comments to machine readable and verifyable documentation.
	\item We performed a case study of a real world java program Randoop (loc, etc) which revealed bugs.
	\item We perform a comparative study of our technique with state of the art flaky test detectors and present results.
\end{enumerate}

### 2019-03-18 design.tex
design sketch

\section{Design}
\subsection{Type checking background}
type hierarchy, subtype checking, Dataflow analysis rules, polymorphism

\subsection{Determinism type hierarchy}
Diagram of type hierarchy and explanation
Examples of errors produced when subtyping rules are violated (here or in the section above this)?

\subsection{Additional typing rules for collection elements}
Lists, set and maps - element types have to be subtypes of collection type 
Explain iterator and extracting elements of an ordernondet collection here
Examples of valid and invalid types

\subsection{Avoid aliasing unsoundness}
Give an example of unsoundness 
Recall CLIMB-to-top rule and defaulting choice for local collections (to avoid too many warnings)
Required user annotations for collections and arrays

\subsection{Special cases of polymorphism}
Explain modifiers to Polymorphic annotations
Poly up and down for ordernondetermistic types
Poly use for avoiding polluting det/ordernondet collections
Subtype hierarchy among all variants of Poly
Possibility of type refinement?

\subsection{Discussion on defaulting rules}
Choice of defaults on class type, constructors and method params and return types

\subsection{Additional hardcoded rules for precision}
Example: equals on HashSet
Discussion on System properties - env, line, file and path separator.

\subsection{Annotating more classes/providing more specifications}
Ease of annotating more JDk classes
User provided stub files to override existing behavior

### 2019-03-19 introduction.tex
Edit first half of introduction

@@
 
-Writing correct programs is hard.
-Programmers often write correctness specifications for their code in the form of assertions.
-They write test cases that cover several usage scenarios to strengthen their belief that
-the code works as intended. Despite all the testing effort, it often happens that a test suite produces
-diverging results for different users or for the same user when the test suite is run again. When a test suite fails
-nondeterministically, it  reduces the programmers' confidence in the program. Additionally, reproducing
-such bugs is difficult making debugging an onerous task. Examples of the underlying causes of such flakiness 
-could be OS scheduling in case of concurrent programs, use of an API that has nondeterministic specification,
-accessing system properties such as the file system or environment variables. In this paper, we focus on
-nondeterminism arising from the use of APIs that have nondeterministic specifications in java programs.
-Instances of such APIs in the JDK include Date(), Random(), and HashSet iteration. Note that mere calls
-to these APIs is not problematic. But if the programmer intends the results of these calls to be deterministic,
-it could result in undesirable results. We aim to capture this behavior and report it. The high level goal
-of our work is to provide programmers with a tool for specifying  deterministic properties in a program
-and be able to verify them at compile time.
+A nondeterministic program may produce different output on different runs
+when provided with the same input.
+Nondeterminism is a serious problem for software developers.
+\todo{Add lots of citations in the below.}
+\begin{itemize}
+\item
+  Nondeterminism makes a program difficult to \textbf{test}, because test
+  oracles must account for all possible behaviors while still enforcing
+  correct behaviors.  Test oracles that are too strict lead to flaky tests
+  that sometimes pass and sometimes fail.  Flaky tests must be re-run, or
+  developers ignore them; in either case, their utility to detect defects
+  is limited.
+\item
+  Nondeterminism makes it difficult to \textbf{compare} two runs of a
+  program on different data, or to compare a run of a slightly modified
+  program to an original program.  This hinders debugging and maintenance,
+  and prevents use of techniques such as Delta Debugging~\cite{Zeller1999}.
+\item
+  Nondeterminism reduces users' and developers' \textbf{trust} in a program's output.
+\end{itemize}
 
-Recent work in this area~\cite{nondex,deflaker} has focussed its efforts on identifying flaky test cases.
-NONDEX~\cite{nondex} is a tool that manually identifies methods in the JDK that have nondeterministic specifications, builds models for
-these methods that behave nondeterministically, and  use sa modified JVM that runs the test suite multiple times.
+Two well-known sources of nondeterminism are concurrency
+% (due to OS scheduling or message delivery over a network)
+and coin-flipping
+(calls to a \<random> API).
+It may be surprising that nondeterminism is common even in sequential
+programs that do not flip coins.
+For example, any program that iterates over a set or map, and that uses the
+results to affect the output, may suffer from nondeterminism.
+Other nondeterministic APIs include date-and-time functions and
+accessing system properties such as the file system or environment variables.
+
+We have created an analysis that detects nondeterminism so that
+programmers can eliminate it.
+The analysis permits programmers to specify which parts of their program
+are nondeterministic, and it verifies that the remainder is deterministic.
+%
+The analysis is sound:  if it issues no warnings, then the program does not
+suffer nondeterminism of the given types.  (If the program uses concurrent,
+it might still be nondeterministic.)
+%
+The analysis works at compile time, giving a guarantee over every possible
+execution of the program, by contrast to unsound dynamic tools that attempt
+to discover when a program has exhibited nondeterministic behavior on
+specific runs.
+%
+The analysis differentiates between values that can differ arbitrarily on
+different runs, and collections that will contain the same values, but
+possibly in a different order.
+%
+The analysis is precise enough for practical use.  It permits calls to
+nondeterministic APIs, and only issues a warning if they are used in ways
+that may lead to nondeterministic output observed by a users.  Like any
+sound analysis, it can issue false positive warnings.
+
+The high level goal of our work is to provide programmers with a tool for
+specifying deterministic properties in a program and be able to verify them
+at compile time.
+
+Previous work in program analysis for nondeterminism has focused on unsound dynamic
+approaches that identify flaky test cases.
+NonDex~\cite{nondex} is a tool that manually identifies methods in the JDK that have nondeterministic specifications, builds models for
+these methods that behave nondeterministically, and  use a modified JVM that runs the test suite multiple times.
 Diverging test results in this setting indicate a flaky test. DeFlaker~\cite{deflaker} looks at a range of commit versions
@@
 b) @OrderNonDet which represents collections that are guaranteed to contain the same elements albeit in possibly
-different iteration order, and c) @NonDet which provides no determism guarantees.  We found this type hierarchy 
+different iteration order, and c) @NonDet which provides no determinism guarantees.  We found this type hierarchy 
 is expressive enough to identify nondeterministic behaviors in production software. We annotated all the collection
@@
 
-We make the folowing contributions through this paper:
+We make the following contributions through this paper:
 \begin{enumerate}
-	\item We provide a design and implemention of a type system for expressing deterministic properties in sequential programs.
+	\item We provide a design and implementation of a type system for expressing deterministic properties in sequential programs.
 	\item We annotated several classes in JDK. By doing this, we have translated the specification for these APIs from
-	human readable comments to machine readable and verifyable documentation.
+	human readable comments to machine readable and verifiable documentation.
 	\item We performed a case study of a real world java program Randoop (loc, etc) which revealed bugs.
@@
 \end{enumerate}
+
+% LocalWords:  NONDEX NonDex DeFlaker

### 2019-03-19 paper.tex
bib file
design sketch
Add macros, change title
Use more bibliographies
Add todo fork keywords

--- 2019-03-13 paper.tex
@@
 
+\usepackage{url}
+
+\usepackage{color}
+%% Comment out one of these two definitions.
+% \newcommand{\todo}[1]{\relax}
+\newcommand{\todo}[1]{{\color{red}\bfseries [[#1]]}}
+
+% \|name| or \mathid{name} denotes identifiers and slots in formulas
+\def\|#1|{\mathid{#1}}
+\newcommand{\mathid}[1]{\ensuremath{\mathit{#1}}}
+% \<name> or \codeid{name} denotes computer code identifiers
+\def\<#1>{\codeid{#1}}
+\protected\def\codeid#1{\ifmmode{\mbox{\sf{#1}}}\else{\sf #1}\fi}
+\protected\def\codeid#1{\ifmmode{\mbox{\ttfamily{#1}}}\else{\ttfamily #1}\fi}
+\protected\def\codeid#1{\ifmmode{\mbox{\smaller\ttfamily{#1}}}\else{\smaller\ttfamily #1}\fi}
+
 \begin{document}
@@
 %% Title information
-\title{Compile Time Verification of Deterministic properties for Sequential Programs}         %% [Short Title] is optional;
+% \title{Compile Time Verification of Deterministic properties for Sequential Programs}
+\title{Making Sequential Programs Deterministic}
+         %% [Short Title] is optional;
                                         %% when present, will be used in
@@
 \begin{abstract}
-Text of abstract \ldots.
+\todo{Text of abstract \ldots.}
 \end{abstract}
@@
 %% comma separated list
-\keywords{keyword1, keyword2, keyword3}  %% \keywords are mandatory in final camera-ready submission
+\keywords{\todo{keyword1}, \todo{keyword2}, \todo{keyword3}}  %% \keywords are mandatory in final camera-ready submission
 
@@
 
-\section{Introduction}
-
-Text of paper \ldots
+\input{introduction.tex}
+\input{design.tex}
 
@@
 %% Bibliography
-%\bibliography{bibfile}
+%\bibliographystyle{ACM-Reference-Format.bst}
+\bibliography{paper,bibstring-unabbrev,testing,crossrefs}
 
 %% Appendix
-\appendix
-\section{Appendix}
-
-Text of appendix \ldots
+%\appendix
+%\section{Appendix}
+%
+%Text of appendix \ldots
 

### 2019-03-19 introduction.tex
Edits in second half of introduction

@@
 when provided with the same input.
-Nondeterminism is a serious problem for software developers.
+Nondeterminism is a serious problem for software developers and users.
 \todo{Add lots of citations in the below.}
@@
 \item
+  Nondeterminism may may outputs incompatible with previous outputs relied
+  on by users or by other systems.\todo{Citations about bugs related to this?}
+\item
   Nondeterminism reduces users' and developers' \textbf{trust} in a program's output.
@@
 and coin-flipping
-(calls to a \<random> API).
+(calls to a \<random> API\@).
 It may be surprising that nondeterminism is common even in sequential
@@
 results to affect the output, may suffer from nondeterminism.
+So may any program that uses default formatting, such as Java's
+\<Object.toString()> which includes an address or hash code.
 Other nondeterministic APIs include date-and-time functions and
@@
 
-We have created an analysis that detects nondeterminism so that
-programmers can eliminate it.
+We have created an analysis that detects nondeterminism or verifies its
+absence.
 The analysis permits programmers to specify which parts of their program
@@
 The analysis is sound:  if it issues no warnings, then the program does not
-suffer nondeterminism of the given types.  (If the program uses concurrent,
+suffer nondeterminism of the given types.  (If the program uses concurrency,
 it might still be nondeterministic.)
@@
 %
-The analysis differentiates between values that can differ arbitrarily on
-different runs, and collections that will contain the same values, but
-possibly in a different order.
+The analysis handles collections that will contain the same values, but
+possibly in a different order, on different runs.
 %
@@
 nondeterministic APIs, and only issues a warning if they are used in ways
-that may lead to nondeterministic output observed by a users.  Like any
+that may lead to nondeterministic output observed by a user.  Like any
 sound analysis, it can issue false positive warnings.
+\todo{Another advantage of our approach
+is that the analysis designer doesn't need to make changes to the JVM or have to rerun the test suite multiple times in order
+to detect flakiness.}
 
@@
 approaches that identify flaky test cases.
-NonDex~\cite{nondex} is a tool that manually identifies methods in the JDK that have nondeterministic specifications, builds models for
+NonDex~\cite{nondex} is a tool that manually\todo{This is confusing:  the
+  tool can't manually identify.  Do you mean that the developers manually
+  identified, then the tool uses those models?}
+identifies methods in the JDK that have nondeterministic specifications, builds models for
 these methods that behave nondeterministically, and  use a modified JVM that runs the test suite multiple times.
@@
 of a code, and marks a test as flaky if it doesn't execute any modified code but still fails in the newer version. These techniques
-have been able to identify issues in real world programs, some of which have been fixed by the developers. This shows that nondeterminism
-is an important problem and solving it at an earlier phase in the software development lifecycle would be greatly beneficial to 
+have been able to identify issues in real-world programs, some of which
+have been fixed by the developers. We believe that identifying and
+resolving nondeterminism
+at an earlier phase in the software development lifecycle would be greatly beneficial to
 developers.
 
-In this paper, we propose to build a lightweight compile time checker that can verify deterministic properties. We design a
-type system that allows programmers to express deterministic properties as types in their programs. We use
-java's annotation support to build our type system. Since our approach is based on compile time type checking, we provide
-soundness guarantees. Another advantage of our approach
-is that the analysis designer doesn't need to make changes to the JVM or have to rerun the test suite multiple times in order
-to detect flakiness.
-Like with any static time checker, our approach is prone to producing false positives.
-There is also increased programmer effort involved in writing type annotations while developing.
-We believe (give numbers here) that the benefits of having a verified system far exceed the programmer efforts.
+We have defined a type system with three types\todo{reference a figure}:
+\<Det> represents values that will be the same across executions,
+\<OrderNonDet> represents collections that are guaranteed to contain the same elements albeit in possibly
+different iteration order, and
+\<NonDet> provides no determinism guarantees:  it represents values that
+might differ from run to run.
+Programmers can write these types to specify their program's behavior.
+Our type system contains a variety of other features that increase
+expressiveness and reduce false positive alarms.\todo{reference a section}
 
-Our type system hierarchy consists of three types: a) @Det which represents values that will be the same across executions,
-b) @OrderNonDet which represents collections that are guaranteed to contain the same elements albeit in possibly
-different iteration order, and c) @NonDet which provides no determinism guarantees.  We found this type hierarchy 
-is expressive enough to identify nondeterministic behaviors in production software. We annotated all the collection
-classes (Lists, Sets and Maps) in the JDK with our new type system. Our tool is therefore guaranteed to report any violation
-of deterministic properties in a client program caused by methods in these classes.
+We have implemented this type system in the context of Java.
+We annotated all the collection
+classes (Lists, Sets and Maps) in the JDK with our new type system.
 
-We make the following contributions through this paper:
+We have run experiments\todo{describe them}.
+We found this type hierarchy
+is expressive enough to identify nondeterministic behaviors in production software.
+
+This paper makes the following contributions:
 \begin{enumerate}

### 2019-03-19 design.tex
adding figures

@@
 \subsection{Type checking background}
+A type system consists of two major component: 1) \textit{types}, and 2) \textit{subtyping rules}.
+A \textit{type} serves as an abstraction for the set of acceptable values for any expression. A \textit{type system} consists
+\textit{types} that form a lattice of finite height. The hierarchy of types in this lattice defines subtyping relationships among them.
+In our type system, we require every type hierarchy to define a unique \textit{@Top} and a \textit{@Bottom} element. This ensures that
+any given pair of types has a \textit{least upper bound} and a \textit{greatest lower bound}.
+Throughout this paper, we use the notation $@A <: @B$ to denote that type $@A$ is a subtype of type $@B$.
+As an illustration of the type hierarchy, consider the lattice of types as shown in Figure~\ref{fig-example-lattice}.
+
+\begin{figure}
+	\begin{center}
+		\includegraphics[scale=0.15]{lattice}
+	\end{center}
+	\caption{Example type hierarchy}
+	\label{fig-example-lattice}
+\end{figure}
+
+example lattice
+assignment, return examples
+method overriding
+collections type parameters
+covariant, contravariant, and invariant subtyping
+
 type hierarchy, subtype checking, Dataflow analysis rules, polymorphism
@@
 Diagram of type hierarchy and explanation
-Examples of errors produced when subtyping rules are violated (here or in the section above this)?
+Examples of errors produced when subtyping rules are violated (here)
+
+The Determinism type system uses the following type qualifiers (see Figure~\ref{fig-determinism-hierarchy}):
+\begin{itemize}
+	\item $@NonDet$ indicates
+	that the expression might have different values in two different executions on the same machine.
+	\item $@OrderNonDet$ indicates that
+	a collection (i.e any class that is a subtype of java.util.Collection or java.util.Iterator) or an array will have the same elements in every execution, but in a
+	possibly different order.  $@OrderNonDet$ may only be written on
+	collections and arrays.
+	\item $@Det$ indicates that
+	the expression evaluates to the same value (with respect to $.equals()$) in all
+	executions on the same machine; for a collection, iteration also yields the values in the same
+	order.
+	This is the default qualifier.
+\end{itemize}
+
+\begin{figure}
+	\begin{center}
+		\includegraphics[scale=0.5]{determinism}
+	\end{center}
+	\caption{Determinism type hierarchy}
+	\label{fig-determinism-hierarchy}
+\end{figure}
 
@@
 \subsection{Special cases of polymorphism}
+Explain polymorphism first
 Explain modifiers to Polymorphic annotations

### 2019-03-20 design.tex
desing background

@@
 \subsection{Type checking background}
-A type system consists of two major component: 1) \textit{types}, and 2) \textit{subtyping rules}.
-A \textit{type} serves as an abstraction for the set of acceptable values for any expression. A \textit{type system} consists
-\textit{types} that form a lattice of finite height. The hierarchy of types in this lattice defines subtyping relationships among them.
-In our type system, we require every type hierarchy to define a unique \textit{@Top} and a \textit{@Bottom} element. This ensures that
+The major components of a type system include: 1) \textit{types}, 2) \textit{subtyping rules}, and 3) \textit{dataflow analysis}.
+A \textit{type} serves as an abstraction for the set of acceptable values for any expression. The types in a type system form a lattice of finite height. 
+The hierarchy of types in this lattice defines subtyping relationships among them.
+In our framework, we require every type hierarchy to define a unique \textit{@Top} and a \textit{@Bottom} element. This ensures that
 any given pair of types has a \textit{least upper bound} and a \textit{greatest lower bound}.
-Throughout this paper, we use the notation $@A <: @B$ to denote that type $@A$ is a subtype of type $@B$.
-As an illustration of the type hierarchy, consider the lattice of types as shown in Figure~\ref{fig-example-lattice}.
-
+Throughout this paper, we use the notation \<@A> <: \<@B> to denote that type \<@A> is a subtype of type \<@B>.
+As an illustration of the type hierarchy, consider the lattice of types shown in Figure~\ref{fig-example-lattice}.
 \begin{figure}
@@
 \end{figure}
+In this type system, \<@C> is a subtype of \<@A> as well as \<@B> and types \<@A> and \<@B> are incomparable.
+The type checker now performs additional type checking (similar to javac's type checking) with respect to this type hierarchy
+and reports any violations. The following code snippet would result in an error at the assignment statements:
+\begin{verbatim}
+@A int x; @B int y; @C int z;
+x = y;    // Because types @A and @B are incomparable.
+z = x;    // Because @C is a subtype of @A.
+\end{verbatim}
+Similarly, if the return type of a method is annotated as \<@C> but it returns a type that has either \<@A> or \<@B>, the
+checker would report an error at compile time.
 
-example lattice
-assignment, return examples
-method overriding
-collections type parameters
-covariant, contravariant, and invariant subtyping
+%Method overriding and example
+For method overriding, the usual rules apply for parameters and arguments i.e covariant subtyping for parameters
+and contravariant subtyping for return types. Consider the following example where method \<foo> of \<class X> is overridden
+in \<class Y>.
+\begin{verbatim}
+class X {
+	@B int foo (@C int param) {}
+}
+class Y extends X {
+	@C int foo (@B int param) {}
+}
+\end{verbatim}
+The overriding method is invalid for two reasons: type of the parameter (\<@B>) is not a subtype of the parameter type in the overriden method (\<@C>) and the return type (\<@C>) is not a supertype of the return type of the overriden method (\<@B>).
 
-type hierarchy, subtype checking, Dataflow analysis rules, polymorphism
+%collections type parameters - invariant
+Collection types are invariant. So, if a \<List> x is declared as \<@A List><\<@A String>> \<x;> and another \<List y> is declared
+as  \<@A List><\<@B String>> \<y;>, writing \<x = y> would result in an error.
+
+%covariant array types
+Arrays follow covariant subtyping rules in java. Therefore, the assignment statement \<@A int @A[] x = y;> where \<y> was 
+declared as \<@B int @A[] y> would type check without any warning.
 
@@
 \begin{itemize}
-	\item $@NonDet$ indicates
-	that the expression might have different values in two different executions on the same machine.
-	\item $@OrderNonDet$ indicates that
+	\item \<@NonDet> indicates
+	that the expression might have different values in two different executions.
+	\item \<@OrderNonDet> indicates that
 	a collection (i.e any class that is a subtype of java.util.Collection or java.util.Iterator) or an array will have the same elements in every execution, but in a
-	possibly different order.  $@OrderNonDet$ may only be written on
+	possibly different order.  \<@OrderNonDet> may only be written on
 	collections and arrays.
-	\item $@Det$ indicates that
-	the expression evaluates to the same value (with respect to $.equals()$) in all
+	\item \<@Det> indicates that
+	the expression evaluates to the same value (with respect to \<.equals()>) in all
 	executions on the same machine; for a collection, iteration also yields the values in the same

### 2019-03-20 design.tex
determinism type hierarchy

@@
 The hierarchy of types in this lattice defines subtyping relationships among them.
-In our framework, we require every type hierarchy to define a unique \textit{@Top} and a \textit{@Bottom} element. This ensures that
+In our framework, we require every type hierarchy to define a unique\<@Top> and a \<@Bottom> element. This ensures that
 any given pair of types has a \textit{least upper bound} and a \textit{greatest lower bound}.
@@
 The type checker now performs additional type checking (similar to javac's type checking) with respect to this type hierarchy
-and reports any violations. The following code snippet would result in an error at the assignment statements:
+and reports any violations. The snippet in Figure~\ref{code-invalid1} would result in an error at the assignment statements:
+\begin{figure}
 	\begin{verbatim}
@@
 	\end{verbatim}
+\caption{Example: Invalid assingment}
+\label{code-invalid1}
+\end{figure}
+
 Similarly, if the return type of a method is annotated as \<@C> but it returns a type that has either \<@A> or \<@B>, the
@@
 For method overriding, the usual rules apply for parameters and arguments i.e covariant subtyping for parameters
-and contravariant subtyping for return types. Consider the following example where method \<foo> of \<class X> is overridden
+and contravariant subtyping for return types. Consider the example in Figure~\ref{code-invalid2} where method \<foo> of \<class X> is overridden
 in \<class Y>.
+\begin{figure}
 	\begin{verbatim}
@@
 	\end{verbatim}
+	\caption{Example: Invalid Method override}
+	\label{code-invalid2}
+\end{figure}
+
 The overriding method is invalid for two reasons: type of the parameter (\<@B>) is not a subtype of the parameter type in the overriden method (\<@C>) and the return type (\<@C>) is not a supertype of the return type of the overriden method (\<@B>).
@@
 \subsection{Determinism type hierarchy}
-Diagram of type hierarchy and explanation
-Examples of errors produced when subtyping rules are violated (here)
-
-The Determinism type system uses the following type qualifiers (see Figure~\ref{fig-determinism-hierarchy}):
+The Determinism type system uses the following type qualifiers (see Figure~\ref{fig-determinism-hierarchy}) (We have omitted the
+\<@Top> and \<@Bottom> types for brevity):
 \begin{itemize}
@@
 	\item \<@OrderNonDet> indicates that
-	a collection (i.e any class that is a subtype of java.util.Collection or java.util.Iterator) or an array will have the same elements in every execution, but in a
+	a collection (i.e any class that is a subtype of java.util.Collection, java.util.Iterator, or java.util.Map) or an array will have the same elements in every execution, but in a
 	possibly different order.  \<@OrderNonDet> may only be written on
@@
 	the expression evaluates to the same value (with respect to \<.equals()>) in all
-	executions on the same machine; for a collection, iteration also yields the values in the same
+	executions; for a collection, iteration also yields the values in the same
 	order.
-	This is the default qualifier.
 \end{itemize}
@@
 	\label{fig-determinism-hierarchy}
+\end{figure}
+
+In Figure~\ref{code-determinism}, we present some of the JDK methods that we have annotated with our determinism types and give examples of
+client code that would produce errors at compile time.
+\begin{figure}
+	\begin{verbatim}
+	// Annotated JDK methods
+	public class Random implements java.io.Serializable {
+	    public @NonDet Random() {}
+	}
+	public class PrintStream extends FilterOutputStream 
+	    implements Appendable, Closeable {
+	    public void println(@Det Object x) {}
+	}
+
+	// Client code
+	class Client {
+	    void test() {
+	        @NonDet Random rnd = new Random(); // No error.
+	        System.out.println(rnd);                    // Error - println takes @Det arguments.
+	        @Det Random rnd1 = new Random();  // Error - subtyping rules violated.
+	    }
+	}
+	\end{verbatim}
+	\caption{Example: Errors detected by determinism checker}
+	\label{code-determinism}
 \end{figure}
@@
 \subsection{Discussion on defaulting rules}
+Det - default, but @Poly is the default for parameters and return types
 Choice of defaults on class type, constructors and method params and return types

### 2019-03-20 design.tex
polymorphism

@@
 %collections type parameters - invariant
-Collection types are invariant. So, if a \<List> x is declared as \<@A List><\<@A String>> \<x;> and another \<List y> is declared
-as  \<@A List><\<@B String>> \<y;>, writing \<x = y> would result in an error.
+Collection types are invariant. So, if a \<List> x is declared as \codeid{@A List<@A String> x;} and another \<List y> is declared
+as  \codeid{@A List<@B String> y;}, writing \<x = y> would result in an error.
 
@@
 \subsection{Additional typing rules for collection elements}
-Lists, set and maps - element types have to be subtypes of collection type 
-Explain iterator and extracting elements of an ordernondet collection here
-Examples of valid and invalid types
+The (determinism) type of a Collection or Iterator must be a supertype or equal to
+the type of the type parameter (see Figure~\ref{fig-determinism-collections}).
+If the Collection is \<@NonDet>, then the type parameter may not be
+\<@Det> or \<@OrderNonDet>. Although such types could exist, they are
+disallowed to prevent the following bug:
 
-\subsection{Avoid aliasing unsoundness}
-Give an example of unsoundness 
-Recall CLIMB-to-top rule and defaulting choice for local collections (to avoid too many warnings)
-Required user annotations for collections and arrays
+\begin{verbatim}
+public static void add(@NonDet List<@Det String> list, @NonDet int index, @Det String s) {
+    list.add(index, s);
+}
 
-\subsection{Special cases of polymorphism}
-Explain polymorphism first
-Explain modifiers to Polymorphic annotations
-Poly up and down for ordernondetermistic types
-Poly use for avoiding polluting det/ordernondet collections
-Subtype hierarchy among all variants of Poly
-Possibility of type refinement?
+public static void f(@Det List<@Det String> list, @NonDet int index, @Det String s) {
+    add(list, index, s);
+}
+\end{verbatim}
+
+The possibility of mutation allows us to add to the \<@Det List> at a
+\<@NonDet> index, which is unsound.
+
+Some examples of valid types are:
+\begin{itemize}
+	\item \codeid{@NonDet\ \ \ \ \ \ List<@NonDet\ \ \ \ \ \ Integer>}
+	\item \codeid{@Det\ \ \ \ \ \ \ \ \ List<@Det\ \ \ \ \ \ \ \ \ Integer>}
+	\item \codeid{@OrderNonDet Set <@OrderNonDet Set<...>\relax >}
+	\item \codeid{@OrderNonDet Set <@Det String>}
+	\item \codeid{@Det\ \ \ \ \ \ \ \ \ Map <@Det String, @Det Integer>}
+	\item \codeid{@NonDet\ \ \ \ \ \ Set <T extends @NonDet Object>}
+\end{itemize}
+
+These types are invalid:
+\begin{itemize}
+	\item \codeid{@OrderNonDet Set <@NonDet\ \ \ \ \ \ Integer>}
+	\item \codeid{@NonDet\ \ \ \ \ \ List<@Det\ \ \ \ \ \ \ \ \ Integer>}
+	\item \codeid{@NonDet\ \ \ \ \ \ List<@OrderNonDet Set<...>\relax >}
+	\item \codeid{@Det\ \ \ \ \ \ \ \ \ List<@NonDet\ \ \ \ \ \ Integer>}
+	\item \codeid{@Det\ \ \ \ \ \ \ \ \ List<@OrderNonDet Set<...>\relax >}
+	\item \codeid{@Det\ \ \ \ \ \ \ \ \ Map <@Det String, @OrderNonDet List<@Det String>>}
+	\item \codeid{@Det\ \ \ \ \ \ \ \ \ List<T extends @NonDet Object>}
+	\item \codeid{@NonDet\ \ \ \ \ \ Set\ <T extends @Det Object>}
+\end{itemize}
+
+Similarly, the (determinism) type of an array must be a supertype or equal to
+the type of the component type (see Figure~\ref{fig-determinism-collections}).
+As with collections, \codeid{@NonDet} arrays of \codeid{@Det} or \codeid{@OrderNonDet}
+elements are not allowed.
+
+For example, these types are valid:
+\begin{itemize}
+	\item \codeid{@NonDet int @NonDet \ \ \ \ []}
+	\item \codeid{@Det \ \ \ int @Det \ \ \ \ \ \ \ []}
+	\item \codeid{@Det \ \ \ int @OrderNonDet[] @OrderNonDet[]}
+\end{itemize}
+
+These types are invalid:
+\begin{itemize}
+	\item \codeid{@NonDet int @OrderNonDet[]}
+	\item \codeid{@NonDet int @Det \ \ \ \ \ \ \ []}
+	\item \codeid{@Det \ \ \ int @NonDet \ \ \ \ []}
+\end{itemize}
+
+\begin{figure}
+	\centering
+	\begin{tabular}{|l|l|l|l|l|}
+		\cline{3-5}
+		\multicolumn{2}{c|}{~}  &  \multicolumn{3}{c|}{Type Argument (or array component type)} \\ \cline{3-5}
+		\multicolumn{2}{c|}{~}  & NonDet     & OrderNonDet & Det \\ \hline
+		& NonDet      &   valid    &  invalid    & invalid  \\ \cline{2-5}
+		Collection (or array)   & OrderNonDet &   invalid  &  valid      & valid  \\ \cline{2-5}
+		& Det         &   invalid  &  invalid    & valid      \\ \hline
+	\end{tabular}
+	\caption{Valid Collection (and array) declarations.  The Collection's (or array's) type qualifier
+		must be a supertype or equal to the type argument (or array component type).}
+	\label{fig-determinism-collections}
+\end{figure}
+
+A collection (\<List, Set, or a Map>) annotated as \<@OrderNonDet> has the following properties:
+\begin{enumerate}
+	\item The individual elements extracted by iterating over the collection get the type \<@NonDet>.
+	\item Accessing properties sich as \<size()> or querying if the collection \<isEmpty()> will return types
+	that are annotated as \<@Det> as they do not depend on the iteration order. 
+\end{enumerate}
+
+A user defined type may be annotated as \<@OrderNonDet> if and only if it
+is a subtype of Collection or Iterator.
+For example, if a user defines a type as
+\begin{verbatim}
+public class TestUserCollection<E> extends ArrayList<E> {...}
+\end{verbatim}
+Writing \codeid{@OrderNonDet TestUserCollection<@Det Integer>} is valid.\\
+Writing \codeid{@OrderNonDet TestUserCollection<@NonDet Integer>} is invalid
+because the type of the type parameter (\codeid{@NonDet}) is not a subtype
+of the user defined Collection type (\codeid{@OrderNonDet}).
+
+User defined types that are not a subtype of Collection or Iterator
+may not be annotated as \codeid{@OrderNonDet}.
+
+\subsection{Polymorphic annotations and special cases}
+The underlying checker-framework on which we build our determinism checker has support for
+polymorphic annotations. When a user writes a polymorphic annotation on a method signature or a type parameter,
+it indicates that it could resolve to any type in the type system depending on how it is used.
+In the determinism checker, we define a polymorphic annotation \<@PolyDet>.
+One of the most common locations for a polymorphic annotation is at a method signature.
+Consider the following method declaration annotated with polymorphic annotations
+\begin{verbatim}
+@PolyDet int foo (@PolyDet String param1, @PolyDet boolean param2) { }
+\end{verbatim}
+This indicates that method \<foo> can be called with arguments having any of the \codeid{@Det or @NonDet} type annotations 
+(\<@OrderNonDet> is not allowed because it is invalid on primitive types). \<@PolyDet> resolves to the least upper bound of
+the actual types on arguments. For instance, if this method is called as \codeid{foo(@Det String arg1, @NonDet boolean arg2)}, the 
+checker framework resolves the method declaration as \codeid{@NonDet int foo(@NonDet String param1, @NonDet boolean param2)}
+causing the return type to have the type annotation \<@NonDet>.
+
+Recall that the determinism checker does not permit \<@OrderNonDet> on a type that isn't a collection, iterator, or an array.
+As a consequence, resolution of polymorphic type annotations at method signatures could have undesirable effects.
+Suppose a method declaration is annotated as \codeid{@PolyDet int size(@PolyDet List<T> list)}. Invoking this method
+as \codeid{size(@OrderNonDet List<@Det String>)} will result in \<@PolyDet> getting resolved to \<@OrderNonDet> which is invalid
+on the return type \<int>. We circumvent this scenario by introducing two variants to the \<@PolyDet> annotation, \<@PolyDet("up")>
+and \<@PolyDet("down")>. The semantics of these variants are as follows:
+\begin{itemize}
+	\item \<@PolyDet("up")>: Same as \<@PolyDet> except when it resolves to \<@OrderNonDet>, it is replaced with \<@NonDet>.
+	\item \<@PolyDet("down")>: Same as \<@PolyDet> except when it resolves to \<@OrderNonDet>, it is replaced with \<@Det>.
+\end{itemize} 
+The modifiers "up" and "down" are indicative of the subtyping relationship of \<@OrderNonDet> with \<@NonDet> and \<@Det> respectively
+in the determinism type hierarchy.
+Annotating \<size()> as \codeid{@PolyDet("down") int size(@PolyDet List<T> list)} will only allow the following type resolutions:
+\begin{itemize}
+	\item \codeid{@Det int size(@Det List<T> list)}
+	\item \codeid{@Det int size(@OrderNonDet List<T> list)}
+	\item \codeid{@NonDet int size(@NonDet List<T> list)}
+\end{itemize}
+
+In order to provide soundness guarantees, we must ensure that a collection that is
+declared as \<@Det> is not side-effected in non-deterministic ways. Consider the following annotation 
+on the List add method: \codeid{public void add(@PolyDet ArrayList<E> this, @PolyDet int index, E element)}.
+Suppose a client calls this method as follows:
+\begin{verbatim}
+@Det ArrayList<@Det String> list;
+@NonDet int random;
+@Det String str;
+list.add(random, str);
+\end{verbatim}
+This will result in the instantion \codeid{public void add(@NonDet ArrayList<E> this, @NonDet int index, E element)} and
+the method invocation will type check. This is problematic because even though \<list> was declared to be \<@Det>,
+the type checker allows the addition of an element at a \<@NonDet> location violating our determinism guarantees.
+We prevent this behavior by introducing another valiant of \<@PolyDet>, \<@PolyDet("use")> that does not affect
+type instantiations. This is especially useful in preventing methods from non-deterministically modifying the state
+of a deterministic receiver. To have this effect, the library method \<ArrayList.add> is annotated as:
+\codeid{public void add(@PolyDet ArrayList<E> this, @PolyDet("use") int index, E element)}.
+Since \<@PolyDet("use")> has no effect on type instantiations, calling this method on a \<@Det List>
+will result in the method declaration being resolved as \codeid{public void add(@Det ArrayList<E> this, @Det int index, E element)}
+irrespective of the type of \<index>, thereby preventing undesired side-effects.
+
+\subsection{Dataflow analysis}
+
+\subsection{Type Refinement}
 
@@
 Choice of defaults on class type, constructors and method params and return types
+Defaults on hashsets, hashmap, etc
+
+\subsection{Avoid aliasing unsoundness}
+Recall CLIMB-to-top rule and defaulting choice for local collections (to avoid too many warnings)
+Required user annotations for collections and arrays
 
@@
 User provided stub files to override existing behavior
+
+\subsection{Maps discussion}
+@Det --> keySet iteration is @Det
+get(): can return o-nd collections
+
+\subsection{About generics and upper bounds?}

### 2019-03-22 design.tex
bullet pointing dataflow analysis and defaulting subsections.

@@
 
-\subsection{Dataflow analysis}
+\subsection{Dataflow analysis and type refinement}
+The checker framework performs automatic type inference for local types within method bodies.
+It also performs dataflow analysis and refines the types of these locals, thereby reducing the annotation burden 
+on the programmer.
 
-\subsection{Type Refinement}
+\begin{itemize}
+	\item Example code with explicit annotations on local variable, showing flow of values into this local variable
+	\item Explain that CF refines the type of this local variable; inbuild dataflow analysis
+	\item Constrast this with a type on a method argument that won't be type refined.
+	\item CF allows type system developers to define additional refinement rules.
+	\item Determinism checker performs type refinement for the receivers of sort() methods in Collection classes.
+	\item Example client code where this type refinement results in precise results
+\end{itemize}
+
+\subsubsection{CLIMB-to-top and collection, array locals}
+
+\begin{itemize}
+	\item Every type system has a default annotation - the type that will be applied to unannotated locations.
+	\item CF defaults types at CLIMB locations to top and type refines them. (steal some text from manual).
+	\item CLIMB locations do not include type parameters. So this is problematic for determinism checker
+	\item Example: Unannotated collection; its default would be @NonDet<@Det> which is invalid.
+	\item So we default locals collections to @NonDet<@NonDet>. 
+	\item This is sound but could produce a lot of FPs. Example assignment: @NonDet<@NonDet> collection = collection1.
+	Type of collection1 is @Det<@Det> and it is a method argument.
+	\item To avoid this, we require users to explicitly annotate local collections.
+	\item Similarly for local arrays.
+\end{itemize}
 
 \subsection{Discussion on defaulting rules}
-Det - default, but @Poly is the default for parameters and return types
-Choice of defaults on class type, constructors and method params and return types
-Defaults on hashsets, hashmap, etc
 
-\subsection{Avoid aliasing unsoundness}
-Recall CLIMB-to-top rule and defaulting choice for local collections (to avoid too many warnings)
-Required user annotations for collections and arrays
+\begin{itemize}
+	\item Determinism checker has @Det as the usual default. Reason being we expect the program to be 
+	deterministic unless specified by programmer or nondeterminism from libraries that we have annotated (Random, Collections, etc).
+	\item All method parameters and return types (including constructors) are @PolyDet.
+	\item Reason: We want to be able to report nondeterministic behavior only where it is invalid.
+	\item Example of a method with params and return types annotated as @PolyDet
+	This method called from 2 places - one harmless(@Det instantiations), and another that call with @NonDet args and prints the retun type.
+	\item Exception to this defaulting is the construction of HashSets and HashMaps because they can't be @Det.
+	We explicitly annotate the return type on constructors as @OrderNonDet and disallow constructing @Det HashSet or HashMap.
+\end{itemize}
 
-\subsection{Additional hardcoded rules for precision}
+\subsection{Additional hardcoded rules in the determinism checker  for precision}
 Example: equals on HashSet
 Discussion on System properties - env, line, file and path separator.
-
-\subsection{Annotating more classes/providing more specifications}
-Ease of annotating more JDk classes
-User provided stub files to override existing behavior
 
@@
 
-\subsection{About generics and upper bounds?}
+\subsection{Annotating more classes/providing more specifications}
+Ease of annotating more JDk classes
+User provided stub files to override existing behavior

### 2019-03-22 design.tex
bulletpointing next subsection

@@
 \subsection{Additional hardcoded rules in the determinism checker  for precision}
-Example: equals on HashSet
-Discussion on System properties - env, line, file and path separator.
+
+\begin{itemize}
+	\item Set.equals() is annotated as @Poly("up"). This is sound but imprecise.
+	\item Example1: @OrderNonDet Set<@Det List>.equals(@OrderNonDet Set<@Det List>) is deterministic
+	\item Example2: @OrderNonDet Set<@OrderNonDet List>.equals(@OrderNonDet Set<@OrderNonDet List>) is not deterministic.
+	\item Explain the rules under which we make Set equality deterministic.
+	\item We annotate System.getProperty as @NonDet because the results could vary across machines.
+	\item Exceptions: line.Sepaator, file.Separator. and path.Separator.
+\end{itemize}
 

### 2019-03-22 design.tex
bullet points for design

@@
 \subsection{Maps discussion}
-@Det --> keySet iteration is @Det
-get(): can return o-nd collections
+
+\begin{itemize}
+	\item A map is deterministic if the keySet() is deterministic.
+	\item the iteration order of a keyset is the same as the iteration order of entrySet and Values.
+	\item HashMap - can be @OrderNonDet or @NonDet.
+	\item LinkedHashMap - can be @Det, @OrderNonDet, ot @NonDet.
+	\item TreeMap - can be @Det or @NonDet.
+	\item Problm with Map.get() : Annotating as @PolyDet is imprecise.
+	\item We circumvent this by hardcoding rules to make it precise. (Describe the rule here)
+	\item It was probably a bad design choice to make HashMap a subclass of Map (talk about how
+	every method is overriden). Similarly for HashSet.
+\end{itemize}
 
 \subsection{Annotating more classes/providing more specifications}
-Ease of annotating more JDk classes
-User provided stub files to override existing behavior
+\begin{itemize}
+	\item Type system developer has to annotate methods in the JDK or any other library.
+	\item Note: Another source of Nondeterminism we haven't analyzed: File system.
+	\item Users can also provide stubs instead. (Steal some text from manual here)
+\end{itemize}

### 2019-03-22 design.tex
design subsection: dataflow analysis

@@
 It also performs dataflow analysis and refines the types of these locals, thereby reducing the annotation burden 
-on the programmer.
+on the programmer. Consider the example below:
+\begin{verbatim}
+void foo(@Det int x) {
+    @NonDet int y = x;    // Type of y gets refined to @Det.
+    @Det int z = y;       //  No error here.
+}
+\end{verbatim}
+The checker framework performs dataflow analysis within method bodies and refines the types of local variables.
+In the example above, even though the local variable \codeid{y} was declared to be \codeid{@NonDet}, it gets
+type refined to \codeid{@Det} after the first assignment statement. As the result, the assignment of \codeid{y}
+to a \codeid{@Det} variable \codeid{z} is sound. In the follwing example,
+\begin{verbatim}
+void foo(@NonDet Object x, @Det Object y) {
+    x = y;                // No type refinement for x.
+    @Det Object z = x;    // Error.
+}
+\end{verbatim}
+the checker reports a warning at the second assignment statement of \codeid{foo}. Type refinement only applies to locals. Refining non-locals can result in side effects making the analysis unsound.
 
-\begin{itemize}
-	\item Example code with explicit annotations on local variable, showing flow of values into this local variable
-	\item Explain that CF refines the type of this local variable; inbuild dataflow analysis
-	\item Constrast this with a type on a method argument that won't be type refined.
-	\item CF allows type system developers to define additional refinement rules.
-	\item Determinism checker performs type refinement for the receivers of sort() methods in Collection classes.
-	\item Example client code where this type refinement results in precise results
-\end{itemize}
+The determinism checker has additional type refinement rules for the sorting methods in collections. 
+The checker type refines a receiver of type \codeid{@OrderNonDet List} to a \codeid{@Det List} if all the
+type argument of this \codeid{List} have the type annotation \codeid{@Det}. Similar rules apply for \codeid{Arrays.sort()} and
+\codeid{Collections.sort()}.
+Under this if \codeid{sort()} is called on a receiver of type \codeid{@OrderNonDet List<@Det Set<@Det String>>}, it will
+be type refined to \codeid{@Det List<@Det Set<@Det String>>}. However a receiver of type \codeid{@OrderNonDet List<@OrderNonDet Set<@Det String>>} will not be type refined.
 

### 2019-03-22 design.tex
Formatting tweaks

@@
 	\item CLIMB locations do not include type parameters. So this is problematic for determinism checker
-	\item Example: Unannotated collection; its default would be @NonDet<@Det> which is invalid.
-	\item So we default locals collections to @NonDet<@NonDet>. 
-	\item This is sound but could produce a lot of FPs. Example assignment: @NonDet<@NonDet> collection = collection1.
-	Type of collection1 is @Det<@Det> and it is a method argument.
+	\item Example: Unannotated collection; its default would be \codeid{@NonDet<@Det>} which is invalid.
+	\item So we default locals collections to \codeid{@NonDet<@NonDet>}. 
+	\item This is sound but could produce a lot of FPs. Example assignment: \codeid{@NonDet<@NonDet> collection = collection1}.
+	Type of collection1 is \codeid{@Det<@Det>} and it is a method argument.
 	\item To avoid this, we require users to explicitly annotate local collections.
@@
 \begin{itemize}
-	\item Determinism checker has @Det as the usual default. Reason being we expect the program to be 
+	\item Determinism checker has \<@Det> as the usual default. Reason being we expect the program to be 
 	deterministic unless specified by programmer or nondeterminism from libraries that we have annotated (Random, Collections, etc).
-	\item All method parameters and return types (including constructors) are @PolyDet.
+	\item All method parameters and return types (including constructors) are \<@PolyDet>.
 	\item Reason: We want to be able to report nondeterministic behavior only where it is invalid.
-	\item Example of a method with params and return types annotated as @PolyDet
-	This method called from 2 places - one harmless(@Det instantiations), and another that call with @NonDet args and prints the retun type.
-	\item Exception to this defaulting is the construction of HashSets and HashMaps because they can't be @Det.
-	We explicitly annotate the return type on constructors as @OrderNonDet and disallow constructing @Det HashSet or HashMap.
+	\item Example of a method with params and return types annotated as \<@PolyDet>.
+	This method called from 2 places - one harmless(@Det instantiations), and another that call with \<@NonDet> args and prints the retun type.
+	\item Exception to this defaulting is the construction of HashSets and HashMaps because they can't be \<@Det>.
+	We explicitly annotate the return type on constructors as
+        \<@OrderNonDet> and disallow constructing \<@Det HashSet> or \<@Det
+        HashMap>.
 \end{itemize}
@@
 \begin{itemize}
-	\item Set.equals() is annotated as @Poly("up"). This is sound but imprecise.
-	\item Example1: @OrderNonDet Set<@Det List>.equals(@OrderNonDet Set<@Det List>) is deterministic
-	\item Example2: @OrderNonDet Set<@OrderNonDet List>.equals(@OrderNonDet Set<@OrderNonDet List>) is not deterministic.
+	\item Set.equals() is annotated as \<@Poly("up")>. This is sound but imprecise.
+	\item Example1: \codeid{@OrderNonDet Set<@Det List>.equals(@OrderNonDet Set<@Det List>)} is deterministic
+	\item Example2: \codeid{@OrderNonDet Set<@OrderNonDet List>.equals(@OrderNonDet Set<@OrderNonDet List>)} is not deterministic.
 	\item Explain the rules under which we make Set equality deterministic.
-	\item We annotate System.getProperty as @NonDet because the results could vary across machines.
+	\item We annotate System.getProperty as \<@NonDet> because the results could vary across machines.
 	\item Exceptions: line.Sepaator, file.Separator. and path.Separator.
@@
 	\item the iteration order of a keyset is the same as the iteration order of entrySet and Values.
-	\item HashMap - can be @OrderNonDet or @NonDet.
-	\item LinkedHashMap - can be @Det, @OrderNonDet, ot @NonDet.
-	\item TreeMap - can be @Det or @NonDet.
-	\item Problm with Map.get() : Annotating as @PolyDet is imprecise.
+	\item \<HashMap> - can be \<@OrderNonDet> or \<@NonDet>.
+	\item \<LinkedHashMap> - can be \<@Det>, \<@OrderNonDet>, ot \<@NonDet>.
+	\item \<TreeMap> - can be \<@Det> or \<@NonDet>.
+	\item Problem with \<Map.get()> : Annotating as \<@PolyDet> is imprecise.
 	\item We circumvent this by hardcoding rules to make it precise. (Describe the rule here)

### 2019-03-22 design.tex
More tweaks

@@
 \begin{itemize}
-	\item Set.equals() is annotated as \<@Poly("up")>. This is sound but imprecise.
+	\item \<Set.equals()> is annotated as \<@Poly("up")>. This is sound but imprecise.
 	\item Example1: \codeid{@OrderNonDet Set<@Det List>.equals(@OrderNonDet Set<@Det List>)} is deterministic
@@
 \begin{itemize}
-	\item A map is deterministic if the keySet() is deterministic.
+	\item A map is deterministic iff the \<entrySet()> is
+          deterministic; equivalently, iff the \<keySet()> and \<values()>
+          are both deterministic
 	\item the iteration order of a keyset is the same as the iteration order of entrySet and Values.

### 2019-03-22 design.tex
Move text, add new section header

@@
 	class X {
-	    @B int foo (@C int param) {}
+	    @B int foo(@C int param) {}
 	}
 	class Y extends X {
-	    @C int foo (@B int param) {}
+	    @C int foo(@B int param) {}
 	}
@@
 \end{figure}
+
+\subsection{Determinism}
+
+The underlying checker-framework on which we build our determinism checker has support for
+polymorphic annotations. When a user writes a polymorphic annotation on a method signature or a type parameter,
+it indicates that it could resolve to any type in the type system depending on how it is used.
+In the determinism checker, we define a polymorphic annotation \<@PolyDet>.
+One of the most common locations for a polymorphic annotation is at a method signature.
+Consider the following method declaration annotated with polymorphic annotations
+\begin{verbatim}
+@PolyDet int foo(@PolyDet String param1, @PolyDet boolean param2) { }
+\end{verbatim}
+This indicates that method \<foo> can be called with arguments having any of the \codeid{@Det or @NonDet} type annotations 
+(\<@OrderNonDet> is not allowed because it is invalid on primitive types). \<@PolyDet> resolves to the least upper bound of
+the actual types on arguments. For instance, if this method is called as \codeid{foo(@Det String arg1, @NonDet boolean arg2)}, the 
+checker framework resolves the method declaration as \codeid{@NonDet int foo(@NonDet String param1, @NonDet boolean param2)}
+causing the return type to have the type annotation \<@NonDet>.
 
@@
 
-\subsection{Polymorphic annotations and special cases}
-The underlying checker-framework on which we build our determinism checker has support for
-polymorphic annotations. When a user writes a polymorphic annotation on a method signature or a type parameter,
-it indicates that it could resolve to any type in the type system depending on how it is used.
-In the determinism checker, we define a polymorphic annotation \<@PolyDet>.
-One of the most common locations for a polymorphic annotation is at a method signature.
-Consider the following method declaration annotated with polymorphic annotations
-\begin{verbatim}
-@PolyDet int foo (@PolyDet String param1, @PolyDet boolean param2) { }
-\end{verbatim}
-This indicates that method \<foo> can be called with arguments having any of the \codeid{@Det or @NonDet} type annotations 
-(\<@OrderNonDet> is not allowed because it is invalid on primitive types). \<@PolyDet> resolves to the least upper bound of
-the actual types on arguments. For instance, if this method is called as \codeid{foo(@Det String arg1, @NonDet boolean arg2)}, the 
-checker framework resolves the method declaration as \codeid{@NonDet int foo(@NonDet String param1, @NonDet boolean param2)}
-causing the return type to have the type annotation \<@NonDet>.
+\subsection{Polymorphism extensions}
 
@@
 \end{itemize}
+
+\subsection{Differentiating binding and use}
 

### 2019-03-22 design.tex
Don't use tab characters

@@
         void test() {
-	        @NonDet Random rnd = new Random(); // No error.
-	        System.out.println(rnd);                    // Error - println takes @Det arguments.
-	        @Det Random rnd1 = new Random();  // Error - subtyping rules violated.
+            @Det double d = Math.random();  // Error - subtyping rules violated.
+            @NonDet double nd = Math.random(); // No error.
+            System.out.println(nd);                    // Error - println takes @Det arguments.
         }

### 2019-03-22 design.tex
change subsection

--- 2019-03-22 design.tex
@@
 
-\subsubsection{CLIMB-to-top and collection, array locals}
+\subsection{CLIMB-to-top and collection, array locals}
+
+Every type system in the checker framework must define a default type annotation. This annotation will be applied
+to every location that isn't explicitly annotated.
 
 \begin{itemize}
-	\item Every type system has a default annotation - the type that will be applied to unannotated locations.
 	\item CF defaults types at CLIMB locations to top and type refines them. (steal some text from manual).
 	\item CLIMB locations do not include type parameters. So this is problematic for determinism checker
-	\item Example: Unannotated collection; its default would be \codeid{@NonDet<@Det>} which is invalid.
-	\item So we default locals collections to \codeid{@NonDet<@NonDet>}. 
-	\item This is sound but could produce a lot of FPs. Example assignment: \codeid{@NonDet<@NonDet> collection = collection1}.
-	Type of collection1 is \codeid{@Det<@Det>} and it is a method argument.
+	\item Example: Unannotated collection; its default would be @NonDet<@Det> which is invalid.
+	\item So we default locals collections to @NonDet<@NonDet>. 
+	\item This is sound but could produce a lot of FPs. Example assignment: @NonDet<@NonDet> collection = collection1.
+	Type of collection1 is @Det<@Det> and it is a method argument.
 	\item To avoid this, we require users to explicitly annotate local collections.
@@
 \begin{itemize}
-	\item Determinism checker has \<@Det> as the usual default. Reason being we expect the program to be 
+	\item Determinism checker has @Det as the usual default. Reason being we expect the program to be 
 	deterministic unless specified by programmer or nondeterminism from libraries that we have annotated (Random, Collections, etc).
-	\item All method parameters and return types (including constructors) are \<@PolyDet>.
+	\item All method parameters and return types (including constructors) are @PolyDet.
 	\item Reason: We want to be able to report nondeterministic behavior only where it is invalid.
-	\item Example of a method with params and return types annotated as \<@PolyDet>.
-	This method called from 2 places - one harmless(@Det instantiations), and another that call with \<@NonDet> args and prints the retun type.
-	\item Exception to this defaulting is the construction of HashSets and HashMaps because they can't be \<@Det>.
-	We explicitly annotate the return type on constructors as
-        \<@OrderNonDet> and disallow constructing \<@Det HashSet> or \<@Det
-        HashMap>.
+	\item Example of a method with params and return types annotated as @PolyDet
+	This method called from 2 places - one harmless(@Det instantiations), and another that call with @NonDet args and prints the retun type.
+	\item Exception to this defaulting is the construction of HashSets and HashMaps because they can't be @Det.
+	We explicitly annotate the return type on constructors as @OrderNonDet and disallow constructing @Det HashSet or HashMap.
 \end{itemize}
@@
 \begin{itemize}
-	\item \<Set.equals()> is annotated as \<@Poly("up")>. This is sound but imprecise.
-	\item Example1: \codeid{@OrderNonDet Set<@Det List>.equals(@OrderNonDet Set<@Det List>)} is deterministic
-	\item Example2: \codeid{@OrderNonDet Set<@OrderNonDet List>.equals(@OrderNonDet Set<@OrderNonDet List>)} is not deterministic.
+	\item Set.equals() is annotated as @Poly("up"). This is sound but imprecise.
+	\item Example1: @OrderNonDet Set<@Det List>.equals(@OrderNonDet Set<@Det List>) is deterministic
+	\item Example2: @OrderNonDet Set<@OrderNonDet List>.equals(@OrderNonDet Set<@OrderNonDet List>) is not deterministic.
 	\item Explain the rules under which we make Set equality deterministic.
-	\item We annotate System.getProperty as \<@NonDet> because the results could vary across machines.
+	\item We annotate System.getProperty as @NonDet because the results could vary across machines.
 	\item Exceptions: line.Sepaator, file.Separator. and path.Separator.
@@
 \begin{itemize}
-	\item A map is deterministic iff the \<entrySet()> is
-          deterministic; equivalently, iff the \<keySet()> and \<values()>
-          are both deterministic
+	\item A map is deterministic if the keySet() is deterministic.
 	\item the iteration order of a keyset is the same as the iteration order of entrySet and Values.
-	\item \<HashMap> - can be \<@OrderNonDet> or \<@NonDet>.
-	\item \<LinkedHashMap> - can be \<@Det>, \<@OrderNonDet>, ot \<@NonDet>.
-	\item \<TreeMap> - can be \<@Det> or \<@NonDet>.
-	\item Problem with \<Map.get()> : Annotating as \<@PolyDet> is imprecise.
+	\item HashMap - can be @OrderNonDet or @NonDet.
+	\item LinkedHashMap - can be @Det, @OrderNonDet, ot @NonDet.
+	\item TreeMap - can be @Det or @NonDet.
+	\item Problm with Map.get() : Annotating as @PolyDet is imprecise.
 	\item We circumvent this by hardcoding rules to make it precise. (Describe the rule here)

### 2019-03-22 design.tex

--- 2019-03-22 design.tex
@@
 
-\subsubsection{CLIMB-to-top and collection, array locals}
+\subsection{CLIMB-to-top and collection, array locals}
+
+Every type system in the checker framework must define a default type annotation. This annotation will be applied
+to every location that isn't explicitly annotated.
 
 \begin{itemize}
-    \item Every type system has a default annotation - the type that will be applied to unannotated locations.
 	\item CF defaults types at CLIMB locations to top and type refines them. (steal some text from manual).
 	\item CLIMB locations do not include type parameters. So this is problematic for determinism checker
-    \item Example: Unannotated collection; its default would be \codeid{@NonDet<@Det>} which is invalid.
-    \item So we default locals collections to \codeid{@NonDet<@NonDet>}. 
-    \item This is sound but could produce a lot of FPs. Example assignment: \codeid{@NonDet<@NonDet> collection = collection1}.
-    Type of collection1 is \codeid{@Det<@Det>} and it is a method argument.
+	\item Example: Unannotated collection; its default would be @NonDet<@Det> which is invalid.
+	\item So we default locals collections to @NonDet<@NonDet>. 
+	\item This is sound but could produce a lot of FPs. Example assignment: @NonDet<@NonDet> collection = collection1.
+	Type of collection1 is @Det<@Det> and it is a method argument.
 	\item To avoid this, we require users to explicitly annotate local collections.

### 2019-03-22 introduction.tex
Edits on page 1

@@
 \item
-  Nondeterminism may may outputs incompatible with previous outputs relied
+  Nondeterminism may make outputs incompatible with previous outputs relied
   on by users or by other systems.\todo{Citations about bugs related to this?}
@@
 For example, any program that iterates over a set or map, and that uses the
-results to affect the output, may suffer from nondeterminism.
+results to affect the output, may produce different output on different runs.
 So may any program that uses default formatting, such as Java's
@@
 absence.
-The analysis permits programmers to specify which parts of their program
-are nondeterministic, and it verifies that the remainder is deterministic.
+Our analysis permits programmers to specify which parts of their program
+are intentionally nondeterministic, and it verifies that the remainder is deterministic.
 %
-The analysis is sound:  if it issues no warnings, then the program does not
+Our analysis is sound:  if it issues no warnings, then the program does not
 suffer nondeterminism of the given types.  (If the program uses concurrency,
@@
 %
-The analysis works at compile time, giving a guarantee over every possible
+Our analysis works at compile time, giving a guarantee over every possible
 execution of the program, by contrast to unsound dynamic tools that attempt
 to discover when a program has exhibited nondeterministic behavior on
-specific runs.
+specific runs.  There is no need for a special JVM nor rerunning a program
+multiple times.
 %
-The analysis handles collections that will contain the same values, but
+Our analysis handles collections that will contain the same values, but
 possibly in a different order, on different runs.
 %
-The analysis is precise enough for practical use.  It permits calls to
+Our analysis is precise enough for practical use.  It permits calls to
 nondeterministic APIs, and only issues a warning if they are used in ways
@@
 sound analysis, it can issue false positive warnings.
-\todo{Another advantage of our approach
-is that the analysis designer doesn't need to make changes to the JVM or have to rerun the test suite multiple times in order
-to detect flakiness.}
 
-The high level goal of our work is to provide programmers with a tool for
-specifying deterministic properties in a program and be able to verify them
+The high-level goal of our work is to provide programmers with a tool for
+specifying deterministic properties in a program and verifing them
 at compile time.
 
+Other researchers have also recognized the importance of this problem.
 Previous work in program analysis for nondeterminism has focused on unsound dynamic

### 2019-03-22 introduction.tex
Edits in introduction

@@
 The high-level goal of our work is to provide programmers with a tool for
-specifying deterministic properties in a program and verifing them
+specifying deterministic properties in a program and verifying them
 at compile time.
@@
 
-We have defined a type system with three types\todo{reference a figure}:
-\<Det> represents values that will be the same across executions,
-\<OrderNonDet> represents collections that are guaranteed to contain the same elements albeit in possibly
-different iteration order, and
-\<NonDet> provides no determinism guarantees:  it represents values that
+Our analysis uses three main abstractions, or approximations to run-time values:
+\begin{description}
+\item[\<Det>] represents values that will be the same across executions.
+\item[\<OrderNonDet>] represents collections that are guaranteed to contain the same elements albeit in possibly
+different iteration order.
+\item[\<NonDet>] represents values that
 might differ from run to run.
-Programmers can write these types to specify their program's behavior.
+\end{description}
+\noindent
+Programmers can write these to specify their program's behavior.
 Our type system contains a variety of other features that increase
@@
 
-We have implemented this type system in the context of Java.
-We annotated all the collection
-classes (Lists, Sets and Maps) in the JDK with our new type system.
+We have implemented this analysis as a pluggable type system for Java.
 
@@
 \begin{enumerate}
-	\item We provide a design and implementation of a type system for expressing deterministic properties in sequential programs.
-	\item We annotated several classes in JDK. By doing this, we have translated the specification for these APIs from
-	human readable comments to machine readable and verifiable documentation.
-	\item We performed a case study of a real world java program Randoop (loc, etc) which revealed bugs.
-	\item We perform a comparative study of our technique with state of the art flaky test detectors and present results.
+  \item We designed a type type system for expressing determinism properties.
+
+  \item We implemented the type system for Java.
+
+  \item In a case study, we annotated several libraries, including the
+    relevant parts of the JDK (Java's standard library).  This provides a
+    formal, machine-readable specification for the libraries, and it
+    demonstrates the richness of our type system.
+
+  \item In another case study, we ran our analysis on a 24 KLOC project that
+    developers had spent weeks of effort to make deterministic.  Our system
+    discovered nondeterminism that the developers had overlooked, and they
+    accepted our patches.
+
+  \item We compared our tool against state-of-the-art flaky test
+    detectors.\todo{Say something about the results.}
 \end{enumerate}
 
-% LocalWords:  NONDEX NonDex DeFlaker
+% LocalWords:  NONDEX NonDex DeFlaker Det OrderNonDet NonDet

### 2019-03-22 design.tex
Comment out a section

@@
 \section{Design}
-\subsection{Type checking background}
-The major components of a type system include: 1) \textit{types}, 2) \textit{subtyping rules}, and 3) \textit{dataflow analysis}.
-A \textit{type} serves as an abstraction for the set of acceptable values for any expression. The types in a type system form a lattice of finite height. 
-The hierarchy of types in this lattice defines subtyping relationships among them.
-In our framework, we require every type hierarchy to define a unique\<@Top> and a \<@Bottom> element. This ensures that
-any given pair of types has a \textit{least upper bound} and a \textit{greatest lower bound}.
-Throughout this paper, we use the notation \<@A> <: \<@B> to denote that type \<@A> is a subtype of type \<@B>.
-As an illustration of the type hierarchy, consider the lattice of types shown in Figure~\ref{fig-example-lattice}.
-\begin{figure}
-    \begin{center}
-        \includegraphics[scale=0.15]{lattice}
-    \end{center}
-    \caption{Example type hierarchy}
-    \label{fig-example-lattice}
-\end{figure}
-In this type system, \<@C> is a subtype of \<@A> as well as \<@B> and types \<@A> and \<@B> are incomparable.
-The type checker now performs additional type checking (similar to javac's type checking) with respect to this type hierarchy
-and reports any violations. The snippet in Figure~\ref{code-invalid1} would result in an error at the assignment statements:
-\begin{figure}
-    \begin{verbatim}
-    @A int x; @B int y; @C int z;
-    x = y;    // Because types @A and @B are incomparable.
-    z = x;    // Because @C is a subtype of @A.
-    \end{verbatim}
-\caption{Example: Invalid assingment}
-\label{code-invalid1}
-\end{figure}
 
-Similarly, if the return type of a method is annotated as \<@C> but it returns a type that has either \<@A> or \<@B>, the
-checker would report an error at compile time.
-
-%Method overriding and example
-For method overriding, the usual rules apply for parameters and arguments i.e covariant subtyping for parameters
-and contravariant subtyping for return types. Consider the example in Figure~\ref{code-invalid2} where method \<foo> of \<class X> is overridden
-in \<class Y>.
-\begin{figure}
-    \begin{verbatim}
-    class X {
-        @B int foo(@C int param) {}
-    }
-    class Y extends X {
-        @C int foo(@B int param) {}
-    }
-    \end{verbatim}
-    \caption{Example: Invalid Method override}
-    \label{code-invalid2}
-\end{figure}
-
-The overriding method is invalid for two reasons: type of the parameter (\<@B>) is not a subtype of the parameter type in the overriden method (\<@C>) and the return type (\<@C>) is not a supertype of the return type of the overriden method (\<@B>).
-
-%collections type parameters - invariant
-Collection types are invariant. So, if a \<List> x is declared as \codeid{@A List<@A String> x;} and another \<List y> is declared
-as  \codeid{@A List<@B String> y;}, writing \<x = y> would result in an error.
-
-%covariant array types
-Arrays follow covariant subtyping rules in java. Therefore, the assignment statement \<@A int @A[] x = y;> where \<y> was 
-declared as \<@B int @A[] y> would type check without any warning.
+%% This section is too basic, and therefore is boring because most readers
+%% will already know it.  It also delays the interesting parts of the paper
+%% for too long.  We can introduce concepts as they are needed in the rest
+%% of the paper.
+% \subsection{Type checking background}
+% The major components of a type system include: 1) \textit{types}, 2) \textit{subtyping rules}, and 3) \textit{dataflow analysis}.
+% A \textit{type} serves as an abstraction for the set of acceptable values for any expression. The types in a type system form a lattice of finite height. 
+% The hierarchy of types in this lattice defines subtyping relationships among them.
+% In our framework, we require every type hierarchy to define a unique\<@Top> and a \<@Bottom> element. This ensures that
+% any given pair of types has a \textit{least upper bound} and a
+% \textit{greatest lower bound}.\todo{It guarantees that any pair of types
+%   has an upper and a lower bound, but not that a unique least upper bound
+%   and greatest lower bound exist.}
+% Throughout this paper, we use the notation \<@A> <: \<@B> to denote that type \<@A> is a subtype of type \<@B>.
+% As an illustration of the type hierarchy, consider the lattice of types shown in Figure~\ref{fig-example-lattice}.
+% \begin{figure}
+%     \begin{center}
+%         \includegraphics[scale=0.15]{lattice}
+%     \end{center}
+%     \caption{Example type hierarchy}
+%     \label{fig-example-lattice}
+% \end{figure}
+% In this type system, \<@C> is a subtype of \<@A> as well as \<@B> and types \<@A> and \<@B> are incomparable.
+% The type checker now performs additional type checking (similar to javac's type checking) with respect to this type hierarchy
+% and reports any violations. The snippet in Figure~\ref{code-invalid1} would result in an error at the assignment statements:
+% \begin{figure}
+%     \begin{verbatim}
+%     @A int x; @B int y; @C int z;
+%     x = y;    // Because types @A and @B are incomparable.
+%     z = x;    // Because @C is a subtype of @A.
+%     \end{verbatim}
+% \caption{Example: Invalid assingment}
+% \label{code-invalid1}
+% \end{figure}
+% 
+% Similarly, if the return type of a method is annotated as \<@C> but it returns a type that has either \<@A> or \<@B>, the
+% checker would report an error at compile time.
+% 
+% %Method overriding and example
+% For method overriding, the usual rules apply for parameters and arguments i.e covariant subtyping for parameters
+% and contravariant subtyping for return types. Consider the example in Figure~\ref{code-invalid2} where method \<foo> of \<class X> is overridden
+% in \<class Y>.
+% \begin{figure}
+%     \begin{verbatim}
+%     class X {
+%         @B int foo(@C int param) {}
+%     }
+%     class Y extends X {
+%         @C int foo(@B int param) {}
+%     }
+%     \end{verbatim}
+%     \caption{Example: Invalid Method override}
+%     \label{code-invalid2}
+% \end{figure}
+% 
+% The overriding method is invalid for two reasons: type of the parameter (\<@B>) is not a subtype of the parameter type in the overriden method (\<@C>) and the return type (\<@C>) is not a supertype of the return type of the overriden method (\<@B>).
+% 
+% %collections type parameters - invariant
+% Collection types are invariant. So, if a \<List> x is declared as \codeid{@A List<@A String> x;} and another \<List y> is declared
+% as  \codeid{@A List<@B String> y;}, writing \<x = y> would result in an error.
+% 
+% %covariant array types
+% Arrays follow covariant subtyping rules in java. Therefore, the assignment statement \<@A int @A[] x = y;> where \<y> was 
+% declared as \<@B int @A[] y> would type check without any warning.
 

### 2019-03-22 design.tex
more text

--- 2019-03-22 design.tex
@@
-\section{Design}
-\subsection{Type checking background}
+\section{Design}\label{design}
+\subsection{Type checking background}\label{type-checking}
 The major components of a type system include: 1) \textit{types}, 2) \textit{subtyping rules}, and 3) \textit{dataflow analysis}.
@@
 
-\subsection{Determinism type hierarchy}
+\subsection{Determinism type hierarchy}\label{type-hierarchy}
 The Determinism type system uses the following type qualifiers (see Figure~\ref{fig-determinism-hierarchy}) (We have omitted the
@@
 
-\subsection{Determinism}
+\subsection{Determinism}\label{determinism}
 
@@
 
-\subsection{Additional typing rules for collection elements}
+\subsection{Additional typing rules for collection elements}\label{collection-rules}
 The (determinism) type of a Collection or Iterator must be a supertype or equal to
@@
 
-\subsection{Polymorphism extensions}
+\subsection{Polymorphism extensions}\label{polymorphism}
 
@@
 
-\subsection{Differentiating binding and use}
+\subsection{Differentiating binding and use}\label{bindings-uses}
 
@@
 
-\subsection{Dataflow analysis and type refinement}
+\subsection{Dataflow analysis and type refinement}\label{dataflow}
 The checker framework performs automatic type inference for local types within method bodies.
@@
 
-\subsection{CLIMB-to-top and collection, array locals}
+\subsection{CLIMB-to-top and collection, array locals}\label{climb-rules}
 
 Every type system in the checker framework must define a default type annotation. This annotation will be applied
-to every location that isn't explicitly annotated.
+to every location that isn't explicitly annotated. In our determinism checker, we chose \codeid{@Det} as the default annotation.
+The reason for this design choice is that we expect the program to be 
+deterministic unless specified by programmer or the program calls nondeterministic library methods that we have annotated 
+(Random, Collections, etc).
+The framework applies the CLIMB-to-top rule. This rule states that the \textit{top} qualifier 
+in the hierarchy is the default for the CLIMB locations: Casts, Locals, Instanceof, and (some) iMplicit Bounds. The rationale
+for this rule is as follows:
 
 \begin{itemize}
-	\item CF defaults types at CLIMB locations to top and type refines them. (steal some text from manual).
-	\item CLIMB locations do not include type parameters. So this is problematic for determinism checker
-	\item Example: Unannotated collection; its default would be @NonDet<@Det> which is invalid.
+	\item Local variables are defaulted to top because type refinement is applied to local variables. If a local variable starts as the 
+	\textit{top} type, then the Checker Framework refines it to the best (most specific) possible type based on assignments to it. As a 
+	result, a programmer rarely writes an explicit annotation on any of those locations.
+	Variables defaulted to top include \textit{local variables}, resource variables in the \textit{try-with-resources} construct, variables 
+	in \textit{for} statements, and \textit{catch} arguments (known as exception parameters in the Java Language Specification). 
+	\textit{Exception parameters} need to have the \textit{top} type because exceptions of arbitrary qualified types can be thrown 
+	and the Checker Framework does not provide runtime checks.
+	\item \textit{Cast} and \textit{instanceof} types are given the same type as their argument expression. This has the same effect as 
+	if they were given the \textit{top} type and then flow-sensitively refined to the type of their argument.
+	\item Implicit upper bounds are defaulted to top to allow them to be instantiated in any way. If a user declared class \codeid{C<T> 
+		{ ... }}, then the Checker Framework assumes that the user intended to allow any instantiation of the class, and the declaration
+	is interpreted as class \codeid{C<T extends @NonDet Object> { ...
+}} rather than as class \codeid{C<T extends @Det Object> { ...
+}}.
+The latter would forbid instantiations such as \codeid{C<@Det String>}, or would require rewriting of code.
+On the other hand, if a user writes
+	an explicit bound such as class \codeid{C<T extends D> { ... }}, then the user intends some restriction on instantiation and can write a 
+	qualifier on the upper bound as desired.
+	This rule means that the upper bound of class \codeid{C<T>} is defaulted differently than the upper bound of class \codeid{C<T extends Object>}. 
+	This may seem confusing, but it is the least bad option. The more confusing alternative would be for \codeid{Object} to be defaulted 
+	differently in class \codeid{C<T extends Object>} and in an instantiation \codeid{C<Object>}, and for the upper bounds to be defaulted differently 
+	in class \codeid{C<T extends Object>} and class \codeid{C<T extends Date>}.
+	\item \textit{Implicit lower bounds} are defaulted to the \textit{bottom} type, again to allow maximal instantiation.
+Note that Java does not allow a programmer to express both the upper and lower bounds of a type, but the Checker Framework allows the programmer to specify either or both.
+\end{itemize}
+With the above rules, the checker framework will automatically annotate a local collection with type \<@NonDet> and its
+type arguments will get the type \<@Det>. For example, a locally declared \codeid{List} of \codeid{Strings} will be automatically annotated
+as \codeid{@NonDet List<@Det String>}. Recall from~\ref{collection-rules} that the determinism checker does not allow this type on
+any collection. While this is still sound, it will result in the determinism checker reporting a lot of errors because every locally declared 
+collection gets annotated with this invalid type. To avoid this situation, we make an exception to the defaulting rules for local collections
+and annotate their type parameters also with \<@NonDet>.
+
+\begin{itemize}
 	\item So we default locals collections to @NonDet<@NonDet>. 
@@
 
-\subsection{Discussion on defaulting rules}
+\subsection{Discussion on defaulting rules}\label{defaulting}
 
 \begin{itemize}
-    \item Determinism checker has \<@Det> as the usual default. Reason being we expect the program to be 
-    deterministic unless specified by programmer or nondeterminism from libraries that we have annotated (Random, Collections, etc).
     \item All method parameters and return types (including constructors) are \<@PolyDet>.
@@
 
-\subsection{Additional hardcoded rules in the determinism checker  for precision}
+\subsection{Determinism checker specific rules for precision}
 
@@
     \item Exceptions: line.Sepaator, file.Separator. and path.Separator.
+    \item discuss type on Exceptions and unsoundness due to this.
 \end{itemize}

### 2019-03-23 design.tex
design: climb to top

@@
 
-The overriding method is invalid for two reasons: type of the parameter (\<@B>) is not a subtype of the parameter type in the overriden method (\<@C>) and the return type (\<@C>) is not a supertype of the return type of the overriden method (\<@B>).
+The overriding method is invalid for two reasons: type of the parameter (\<@B>) is not a subtype of the parameter type in the overridden method (\<@C>) and the return type (\<@C>) is not a supertype of the return type of the overridden method (\<@B>).
 
@@
 type refined to \codeid{@Det} after the first assignment statement. As the result, the assignment of \codeid{y}
-to a \codeid{@Det} variable \codeid{z} is sound. In the follwing example,
+to a \codeid{@Det} variable \codeid{z} is sound. In the following example,
 \begin{verbatim}
@@
 and annotate their type parameters also with \<@NonDet>.
-
-\begin{itemize}
-	\item So we default locals collections to @NonDet<@NonDet>. 
-	\item This is sound but could produce a lot of FPs. Example assignment: @NonDet<@NonDet> collection = collection1.
-	Type of collection1 is @Det<@Det> and it is a method argument.
-	\item To avoid this, we require users to explicitly annotate local collections.
-	\item Similarly for local arrays.
-\end{itemize}
+So a local list of Strings gets the default annotation of \codeid{@NonDet List<@NonDet String>}.
+While this defaulting eliminates the possibility of automatically annotating locals collections with an invalid
+type, it would could in a high number of false positives. In the following code snippet,
+\begin{verbatim}
+void test(@Det List<@Det String> argList) {
+            List<String> localList = argList;
+}    
+\end{verbatim}
+the checker would report an error at the assignment statement since the inferred type for \codeid{localList} is 
+\codeid{@NonDet List<@NonDet> String} and since collections types are invariant.
+We therefore recommend that programmers explicitly annotate
+all local variables that are collections to reduce false positives. The same reasoning and recommendation applies to arrays.
 

### 2019-03-23 design.tex
more text

@@
 \subsection{Discussion on defaulting rules}\label{defaulting}
+In our determinism checker, we chose \<@PolyDet> as the default for all method parameters and return types including constructors.
+The rationale for this design choice was to not make our type checker too restrictive by making everything \<@Det>.  This implies that
+the checker still allows some non-determinism as long as it doesn't break any determinism specifications explicitly written by the programmer.
 
 \begin{itemize}
-    \item All method parameters and return types (including constructors) are \<@PolyDet>.
-    \item Reason: We want to be able to report nondeterministic behavior only where it is invalid.
     \item Example of a method with params and return types annotated as \<@PolyDet>.
-    This method called from 2 places - one harmless(@Det instantiations), and another that call with \<@NonDet> args and prints the retun type.
+    This method called from 2 places - one harmless(@NonDet instantiations but no printing), and another that call with \<@NonDet> args and prints the return type.
     \item Exception to this defaulting is the construction of HashSets and HashMaps because they can't be \<@Det>.

### 2019-03-23 design.tex
Todo comments, preparatory to editing

--- 2019-03-22 design.tex
@@
-\section{Design}\label{design}
-\subsection{Type checking background}\label{type-checking}
-The major components of a type system include: 1) \textit{types}, 2) \textit{subtyping rules}, and 3) \textit{dataflow analysis}.
-A \textit{type} serves as an abstraction for the set of acceptable values for any expression. The types in a type system form a lattice of finite height. 
-The hierarchy of types in this lattice defines subtyping relationships among them.
-In our framework, we require every type hierarchy to define a unique\<@Top> and a \<@Bottom> element. This ensures that
-any given pair of types has a \textit{least upper bound} and a \textit{greatest lower bound}.
-Throughout this paper, we use the notation \<@A> <: \<@B> to denote that type \<@A> is a subtype of type \<@B>.
-As an illustration of the type hierarchy, consider the lattice of types shown in Figure~\ref{fig-example-lattice}.
-\begin{figure}
-    \begin{center}
-        \includegraphics[scale=0.15]{lattice}
-    \end{center}
-    \caption{Example type hierarchy}
-    \label{fig-example-lattice}
-\end{figure}
-In this type system, \<@C> is a subtype of \<@A> as well as \<@B> and types \<@A> and \<@B> are incomparable.
-The type checker now performs additional type checking (similar to javac's type checking) with respect to this type hierarchy
-and reports any violations. The snippet in Figure~\ref{code-invalid1} would result in an error at the assignment statements:
-\begin{figure}
-    \begin{verbatim}
-    @A int x; @B int y; @C int z;
-    x = y;    // Because types @A and @B are incomparable.
-    z = x;    // Because @C is a subtype of @A.
-    \end{verbatim}
-\caption{Example: Invalid assingment}
-\label{code-invalid1}
-\end{figure}
+\section{Design}
 
-Similarly, if the return type of a method is annotated as \<@C> but it returns a type that has either \<@A> or \<@B>, the
-checker would report an error at compile time.
+%% This section is too basic, and therefore is boring because most readers
+%% will already know it.  It also delays the interesting parts of the paper
+%% for too long.  We can introduce concepts as they are needed in the rest
+%% of the paper.
+% \subsection{Type checking background}
+% The major components of a type system include: 1) \textit{types}, 2) \textit{subtyping rules}, and 3) \textit{dataflow analysis}.
+% A \textit{type} serves as an abstraction for the set of acceptable values for any expression. The types in a type system form a lattice of finite height. 
+% The hierarchy of types in this lattice defines subtyping relationships among them.
+% In our framework, we require every type hierarchy to define a unique\<@Top> and a \<@Bottom> element. This ensures that
+% any given pair of types has a \textit{least upper bound} and a
+% \textit{greatest lower bound}.\todo{It guarantees that any pair of types
+%   has an upper and a lower bound, but not that a unique least upper bound
+%   and greatest lower bound exist.}
+% Throughout this paper, we use the notation \<@A> <: \<@B> to denote that type \<@A> is a subtype of type \<@B>.
+% As an illustration of the type hierarchy, consider the lattice of types shown in Figure~\ref{fig-example-lattice}.
+% \begin{figure}
+%     \begin{center}
+%         \includegraphics[scale=0.15]{lattice}
+%     \end{center}
+%     \caption{Example type hierarchy}
+%     \label{fig-example-lattice}
+% \end{figure}
+% In this type system, \<@C> is a subtype of \<@A> as well as \<@B> and types \<@A> and \<@B> are incomparable.
+% The type checker now performs additional type checking (similar to javac's type checking) with respect to this type hierarchy
+% and reports any violations. The snippet in Figure~\ref{code-invalid1} would result in an error at the assignment statements:
+% \begin{figure}
+%     \begin{verbatim}
+%     @A int x; @B int y; @C int z;
+%     x = y;    // Because types @A and @B are incomparable.
+%     z = x;    // Because @C is a subtype of @A.
+%     \end{verbatim}
+% \caption{Example: Invalid assingment}
+% \label{code-invalid1}
+% \end{figure}
+% 
+% Similarly, if the return type of a method is annotated as \<@C> but it returns a type that has either \<@A> or \<@B>, the
+% checker would report an error at compile time.
+% 
+% %Method overriding and example
+% For method overriding, the usual rules apply for parameters and arguments i.e covariant subtyping for parameters
+% and contravariant subtyping for return types. Consider the example in Figure~\ref{code-invalid2} where method \<foo> of \<class X> is overridden
+% in \<class Y>.
+% \begin{figure}
+%     \begin{verbatim}
+%     class X {
+%         @B int foo(@C int param) {}
+%     }
+%     class Y extends X {
+%         @C int foo(@B int param) {}
+%     }
+%     \end{verbatim}
+%     \caption{Example: Invalid Method override}
+%     \label{code-invalid2}
+% \end{figure}
+% 
+% The overriding method is invalid for two reasons: type of the parameter (\<@B>) is not a subtype of the parameter type in the overriden method (\<@C>) and the return type (\<@C>) is not a supertype of the return type of the overriden method (\<@B>).
+% 
+% %collections type parameters - invariant
+% Collection types are invariant. So, if a \<List> x is declared as \codeid{@A List<@A String> x;} and another \<List y> is declared
+% as  \codeid{@A List<@B String> y;}, writing \<x = y> would result in an error.
+% 
+% %covariant array types
+% Arrays follow covariant subtyping rules in java. Therefore, the assignment statement \<@A int @A[] x = y;> where \<y> was 
+% declared as \<@B int @A[] y> would type check without any warning.
 
-%Method overriding and example
-For method overriding, the usual rules apply for parameters and arguments i.e covariant subtyping for parameters
-and contravariant subtyping for return types. Consider the example in Figure~\ref{code-invalid2} where method \<foo> of \<class X> is overridden
-in \<class Y>.
-\begin{figure}
-    \begin{verbatim}
-    class X {
-        @B int foo(@C int param) {}
-    }
-    class Y extends X {
-        @C int foo(@B int param) {}
-    }
-    \end{verbatim}
-    \caption{Example: Invalid Method override}
-    \label{code-invalid2}
-\end{figure}
+\todo{Need some background around here about what a type qualifier is,
+  how it is represented in Java, and how to read a type \<@Q BaseType>.}
 
-The overriding method is invalid for two reasons: type of the parameter (\<@B>) is not a subtype of the parameter type in the overriden method (\<@C>) and the return type (\<@C>) is not a supertype of the return type of the overriden method (\<@B>).
-
-%collections type parameters - invariant
-Collection types are invariant. So, if a \<List> x is declared as \codeid{@A List<@A String> x;} and another \<List y> is declared
-as  \codeid{@A List<@B String> y;}, writing \<x = y> would result in an error.
-
-%covariant array types
-Arrays follow covariant subtyping rules in java. Therefore, the assignment statement \<@A int @A[] x = y;> where \<y> was 
-declared as \<@B int @A[] y> would type check without any warning.
-
-\subsection{Determinism type hierarchy}\label{type-hierarchy}
-The Determinism type system uses the following type qualifiers (see Figure~\ref{fig-determinism-hierarchy}) (We have omitted the
-\<@Top> and \<@Bottom> types for brevity):
+\subsection{Determinism type hierarchy}
+The core of the determinism type system is the following type qualifiers (see Figure~\ref{fig-determinism-hierarchy}):
+\todo{introduce just @Det and @NonDet first, and then @OrderNonDet after polymorphism.}
 \begin{itemize}
@@
     \item \<@OrderNonDet> indicates that
-    a collection (i.e any class that is a subtype of java.util.Collection, java.util.Iterator, or java.util.Map) or an array will have the same elements in every execution, but in a
+    a collection or an array will have the same elements in every execution, but in a
     possibly different order.  \<@OrderNonDet> may only be written on
     collections and arrays.
+    A collection is any subtype of \<java.util.Collection>,
+    \<java.util.Iterator>, or \<java.util.Map>, including user-defined types.
     \item \<@Det> indicates that
@@
 
-\subsection{Determinism}\label{determinism}
+\subsection{Polymorphism}
 
@@
 
-\subsection{Additional typing rules for collection elements}\label{collection-rules}
+\subsection{Additional typing rules for collection elements}
+
+\todo{It is very important to express the whole section as formal type
+  jugdments, or we will get laughed
+  out of the room as amateurs, during the program committee meeting.  Also,
+  the formalism will help make the ideas precise.
+
+  Type well-formedness rules:\todo{format nicely in \LaTeX}
+
+  antecedent: $\Gamma : \kappa_e <: \kappa_c$
+
+  consequent: $\Gamma : \kappa_c \ \codeid{List}\langle\kappa_e\ 
+  \codeid{Object}\rangle$
+  }
+
 The (determinism) type of a Collection or Iterator must be a supertype or equal to
@@
 \<@NonDet> index, which is unsound.
+
+\todo{Fewer examples, less variation.  Use \<List> and \<String>.}
 
@@
 
-\subsection{Polymorphism extensions}\label{polymorphism}
+\subsection{Polymorphism extensions}
 
@@
 
-\subsection{Differentiating binding and use}\label{bindings-uses}
+\subsection{Differentiating binding and use}
 
@@
 
-\subsection{Dataflow analysis and type refinement}\label{dataflow}
+\subsection{Dataflow analysis and type refinement}
 The checker framework performs automatic type inference for local types within method bodies.
@@
 
-\subsection{CLIMB-to-top and collection, array locals}\label{climb-rules}
+\subsection{CLIMB-to-top and collection, array locals}
 
 Every type system in the checker framework must define a default type annotation. This annotation will be applied
-to every location that isn't explicitly annotated. In our determinism checker, we chose \codeid{@Det} as the default annotation.
-The reason for this design choice is that we expect the program to be 
-deterministic unless specified by programmer or the program calls nondeterministic library methods that we have annotated 
-(Random, Collections, etc).
-The framework applies the CLIMB-to-top rule. This rule states that the \textit{top} qualifier 
-in the hierarchy is the default for the CLIMB locations: Casts, Locals, Instanceof, and (some) iMplicit Bounds. The rationale
-for this rule is as follows:
+to every location that isn't explicitly annotated.
 
 \begin{itemize}
-	\item Local variables are defaulted to top because type refinement is applied to local variables. If a local variable starts as the 
-	\textit{top} type, then the Checker Framework refines it to the best (most specific) possible type based on assignments to it. As a 
-	result, a programmer rarely writes an explicit annotation on any of those locations.
-	Variables defaulted to top include \textit{local variables}, resource variables in the \textit{try-with-resources} construct, variables 
-	in \textit{for} statements, and \textit{catch} arguments (known as exception parameters in the Java Language Specification). 
-	\textit{Exception parameters} need to have the \textit{top} type because exceptions of arbitrary qualified types can be thrown 
-	and the Checker Framework does not provide runtime checks.
-	\item \textit{Cast} and \textit{instanceof} types are given the same type as their argument expression. This has the same effect as 
-	if they were given the \textit{top} type and then flow-sensitively refined to the type of their argument.
-	\item Implicit upper bounds are defaulted to top to allow them to be instantiated in any way. If a user declared class \codeid{C<T> 
-		{ ... }}, then the Checker Framework assumes that the user intended to allow any instantiation of the class, and the declaration
-	is interpreted as class \codeid{C<T extends @NonDet Object> { ...
-}} rather than as class \codeid{C<T extends @Det Object> { ...
-}}.
-The latter would forbid instantiations such as \codeid{C<@Det String>}, or would require rewriting of code.
-On the other hand, if a user writes
-	an explicit bound such as class \codeid{C<T extends D> { ... }}, then the user intends some restriction on instantiation and can write a 
-	qualifier on the upper bound as desired.
-	This rule means that the upper bound of class \codeid{C<T>} is defaulted differently than the upper bound of class \codeid{C<T extends Object>}. 
-	This may seem confusing, but it is the least bad option. The more confusing alternative would be for \codeid{Object} to be defaulted 
-	differently in class \codeid{C<T extends Object>} and in an instantiation \codeid{C<Object>}, and for the upper bounds to be defaulted differently 
-	in class \codeid{C<T extends Object>} and class \codeid{C<T extends Date>}.
-	\item \textit{Implicit lower bounds} are defaulted to the \textit{bottom} type, again to allow maximal instantiation.
-Note that Java does not allow a programmer to express both the upper and lower bounds of a type, but the Checker Framework allows the programmer to specify either or both.
-\end{itemize}
-With the above rules, the checker framework will automatically annotate a local collection with type \<@NonDet> and its
-type arguments will get the type \<@Det>. For example, a locally declared \codeid{List} of \codeid{Strings} will be automatically annotated
-as \codeid{@NonDet List<@Det String>}. Recall from~\ref{collection-rules} that the determinism checker does not allow this type on
-any collection. While this is still sound, it will result in the determinism checker reporting a lot of errors because every locally declared 
-collection gets annotated with this invalid type. To avoid this situation, we make an exception to the defaulting rules for local collections
-and annotate their type parameters also with \<@NonDet>.
-
-\begin{itemize}
+  \item CF defaults types at CLIMB locations to top and type refines them. (steal some text from manual).
+  \item CLIMB locations do not include type parameters. So this is problematic for determinism checker
+  \item Example: Unannotated collection; its default would be @NonDet<@Det> which is invalid.
   \item So we default locals collections to @NonDet<@NonDet>.
@@
 
-\subsection{Discussion on defaulting rules}\label{defaulting}
+\subsection{Discussion on defaulting rules}
 
 \begin{itemize}
+    \item Determinism checker has \<@Det> as the usual default. Reason being we expect the program to be 
+    deterministic unless specified by programmer or nondeterminism from libraries that we have annotated (Random, Collections, etc).
     \item All method parameters and return types (including constructors) are \<@PolyDet>.
@@
 
-\subsection{Determinism checker specific rules for precision}
+\subsection{Additional hardcoded rules in the determinism checker  for precision}
 
@@
     \item Exceptions: line.Sepaator, file.Separator. and path.Separator.
-    \item discuss type on Exceptions and unsoundness due to this.
 \end{itemize}

### 2019-03-23 design.tex

--- 2019-03-23 design.tex
@@
 \section{Design}\label{design}
-\subsection{Type checking background}\label{type-checking}
-The major components of a type system include: 1) \textit{types}, 2) \textit{subtyping rules}, and 3) \textit{dataflow analysis}.
-A \textit{type} serves as an abstraction for the set of acceptable values for any expression. The types in a type system form a lattice of finite height. 
-The hierarchy of types in this lattice defines subtyping relationships among them.
-In our framework, we require every type hierarchy to define a unique\<@Top> and a \<@Bottom> element. This ensures that
-any given pair of types has a \textit{least upper bound} and a \textit{greatest lower bound}.
-Throughout this paper, we use the notation \<@A> <: \<@B> to denote that type \<@A> is a subtype of type \<@B>.
-As an illustration of the type hierarchy, consider the lattice of types shown in Figure~\ref{fig-example-lattice}.
-\begin{figure}
-    \begin{center}
-        \includegraphics[scale=0.15]{lattice}
-    \end{center}
-    \caption{Example type hierarchy}
-    \label{fig-example-lattice}
-\end{figure}
-In this type system, \<@C> is a subtype of \<@A> as well as \<@B> and types \<@A> and \<@B> are incomparable.
-The type checker now performs additional type checking (similar to javac's type checking) with respect to this type hierarchy
-and reports any violations. The snippet in Figure~\ref{code-invalid1} would result in an error at the assignment statements:
-\begin{figure}
-    \begin{verbatim}
-    @A int x; @B int y; @C int z;
-    x = y;    // Because types @A and @B are incomparable.
-    z = x;    // Because @C is a subtype of @A.
-    \end{verbatim}
-\caption{Example: Invalid assingment}
-\label{code-invalid1}
-\end{figure}
 
-Similarly, if the return type of a method is annotated as \<@C> but it returns a type that has either \<@A> or \<@B>, the
-checker would report an error at compile time.
+%% This section is too basic, and therefore is boring because most readers
+%% will already know it.  It also delays the interesting parts of the paper
+%% for too long.  We can introduce concepts as they are needed in the rest
+%% of the paper.
+% \subsection{Type checking background}\label{type-checking}
+% The major components of a type system include: 1) \textit{types}, 2) \textit{subtyping rules}, and 3) \textit{dataflow analysis}.
+% A \textit{type} serves as an abstraction for the set of acceptable values for any expression. The types in a type system form a lattice of finite height. 
+% The hierarchy of types in this lattice defines subtyping relationships among them.
+% In our framework, we require every type hierarchy to define a unique\<@Top> and a \<@Bottom> element. This ensures that
+% any given pair of types has a \textit{least upper bound} and a \textit{greatest lower bound}.
+% Throughout this paper, we use the notation \<@A> <: \<@B> to denote that type \<@A> is a subtype of type \<@B>.
+% As an illustration of the type hierarchy, consider the lattice of types shown in Figure~\ref{fig-example-lattice}.
+% \begin{figure}
+%     \begin{center}
+%         \includegraphics[scale=0.15]{lattice}
+%     \end{center}
+%     \caption{Example type hierarchy}
+%     \label{fig-example-lattice}
+% \end{figure}
+% In this type system, \<@C> is a subtype of \<@A> as well as \<@B> and types \<@A> and \<@B> are incomparable.
+% The type checker now performs additional type checking (similar to javac's type checking) with respect to this type hierarchy
+% and reports any violations. The snippet in Figure~\ref{code-invalid1} would result in an error at the assignment statements:
+% \begin{figure}
+%     \begin{verbatim}
+%     @A int x; @B int y; @C int z;
+%     x = y;    // Because types @A and @B are incomparable.
+%     z = x;    // Because @C is a subtype of @A.
+%     \end{verbatim}
+% \caption{Example: Invalid assingment}
+% \label{code-invalid1}
+% \end{figure}
+% 
+% %Method overriding and example
+% For method overriding, the usual rules apply for parameters and arguments i.e covariant subtyping for parameters
+% and contravariant subtyping for return types. Consider the example in Figure~\ref{code-invalid2} where method \<foo> of \<class X> is overridden
+% in \<class Y>.
+% \begin{figure}
+%     \begin{verbatim}
+%     class X {
+%         @B int foo(@C int param) {}
+%     }
+%     class Y extends X {
+%         @C int foo(@B int param) {}
+%     }
+%     \end{verbatim}
+%     \caption{Example: Invalid Method override}
+%     \label{code-invalid2}
+% \end{figure}
+% 
+% The overriding method is invalid for two reasons: type of the parameter (\<@B>) is not a subtype of the parameter type in the overridden method (\<@C>) and the return type (\<@C>) is not a supertype of the return type of the overridden method (\<@B>).
+% 
+% %collections type parameters - invariant
+% Collection types are invariant. So, if a \<List> x is declared as \codeid{@A List<@A String> x;} and another \<List y> is declared
+% as  \codeid{@A List<@B String> y;}, writing \<x = y> would result in an error.
+% 
+% %covariant array types
+% Arrays follow covariant subtyping rules in java. Therefore, the assignment statement \<@A int @A[] x = y;> where \<y> was 
+% declared as \<@B int @A[] y> would type check without any warning.
 
-%Method overriding and example
-For method overriding, the usual rules apply for parameters and arguments i.e covariant subtyping for parameters
-and contravariant subtyping for return types. Consider the example in Figure~\ref{code-invalid2} where method \<foo> of \<class X> is overridden
-in \<class Y>.
-\begin{figure}
-    \begin{verbatim}
-    class X {
-        @B int foo(@C int param) {}
-    }
-    class Y extends X {
-        @C int foo(@B int param) {}
-    }
-    \end{verbatim}
-    \caption{Example: Invalid Method override}
-    \label{code-invalid2}
-\end{figure}
-
-The overriding method is invalid for two reasons: type of the parameter (\<@B>) is not a subtype of the parameter type in the overridden method (\<@C>) and the return type (\<@C>) is not a supertype of the return type of the overridden method (\<@B>).
-
-%collections type parameters - invariant
-Collection types are invariant. So, if a \<List> x is declared as \codeid{@A List<@A String> x;} and another \<List y> is declared
-as  \codeid{@A List<@B String> y;}, writing \<x = y> would result in an error.
-
-%covariant array types
-Arrays follow covariant subtyping rules in java. Therefore, the assignment statement \<@A int @A[] x = y;> where \<y> was 
-declared as \<@B int @A[] y> would type check without any warning.
+\todo{Need some background around here about what a type qualifier is,
+  how it is represented in Java, and how to read a type \<@Q BaseType>.}
 
 \subsection{Determinism type hierarchy}\label{type-hierarchy}
-The Determinism type system uses the following type qualifiers (see Figure~\ref{fig-determinism-hierarchy}) (We have omitted the
-\<@Top> and \<@Bottom> types for brevity):
+The core of the determinism type system is the following type qualifiers (see Figure~\ref{fig-determinism-hierarchy}):
+\todo{introduce just @Det and @NonDet first, and then @OrderNonDet after polymorphism.}
 \begin{itemize}
@@
 
-\subsection{Determinism}\label{determinism}
+\subsection{Polymorphism}\label{polymorphism}
 
@@
 \subsection{Additional typing rules for collection elements}\label{collection-rules}
+
+\todo{It is very important to express the whole section as formal type
+  jugdments, or we will get laughed
+  out of the room as amateurs, during the program committee meeting.  Also,
+  the formalism will help make the ideas precise.
+
+  Type well-formedness rules:\todo{format nicely in \LaTeX}
+
+  antecedent: $\Gamma : \kappa_e <: \kappa_c$
+
+  consequent: $\Gamma : \kappa_c \ \codeid{List}\langle\kappa_e\ 
+  \codeid{Object}\rangle$
+  }
+
 The (determinism) type of a Collection or Iterator must be a supertype or equal to

### 2019-03-24 introduction.tex
intro todo

@@
 approaches that identify flaky test cases.
-NonDex~\cite{nondex} is a tool that manually\todo{This is confusing:  the
-  tool can't manually identify.  Do you mean that the developers manually
-  identified, then the tool uses those models?}
-identifies methods in the JDK that have nondeterministic specifications, builds models for
-these methods that behave nondeterministically, and  use a modified JVM that runs the test suite multiple times.
+The authors of~\cite{nondex} manually identify methods in the JDK that have nondeterministic specifications and build models for
+these methods. They  modified that JVM so that it runs the test suite multiple times.
+%NonDex~\cite{nondex} is a tool that manually\todo{This is confusing:  the
+%  tool can't manually identify.  Do you mean that the developers manually
+%  identified, then the tool uses those models?}
+%identifies 
 Diverging test results in this setting indicate a flaky test. DeFlaker~\cite{deflaker} looks at a range of commit versions

### 2019-03-24 design.tex
Avoid label duplication
design background todo

@@
 
-\todo{Need some background around here about what a type qualifier is,
-  how it is represented in Java, and how to read a type \<@Q BaseType>.}
+%\todo{Need some background around here about what a type qualifier is,
+%  how it is represented in Java, and how to read a type \<@Q BaseType>.}
+
+A \textit{type} serves as an abstraction for the set of acceptable values for any expression.
+A \textit{type qualifier} on a \textit{type} adds additional constraints on the values represented by that \textit{type}.
+A compile time type checker can verify the correctness of this \textit{type} with its dataflow analysis and subtyping rules.
+In Java, a type qualifier is written as an annotation. The type \<Integer> represents positive and negative integers besides zero.
+A program analysis developer could design a type system that differentiates positive integers from the negative ones with type qualifiers
+\<@Positive> and \<@Negative>.  The type \<@Positive Integer> now represents only positive values and the dataflow analysis ensures that
+any negative integer value flowing into this type is reported as an error.
 
@@
 
-\subsection{Polymorphism extensions}\label{polymorphism}
+\subsection{Polymorphism extensions}\label{polymorphism-extensions}
 

### 2019-03-25 design.tex
moving text around

@@
 
+Subtyping rule:
+
+$\infer{\Gamma \vdash x : \tau_2}{\Gamma \vdash x : \tau_1 & \tau_1 <: \tau_2}$
+
+Typing rule for method calls:
+\todo{typesetting}
+
+$\forall i : i >= 0 \infer{method(\overline{a_i}) : \tau_r}{\tau_r \  method(\overline{\tau_i \  p_i}) & \overline{a_i : \tau_i}}$
+
 \subsection{Determinism type hierarchy}\label{type-hierarchy}
 The core of the determinism type system is the following type qualifiers (see Figure~\ref{fig-determinism-hierarchy}):
-\todo{introduce just @Det and @NonDet first, and then @OrderNonDet after polymorphism.}
+%\todo{introduce just @Det and @NonDet first, and then @OrderNonDet after polymorphism.}
 \begin{itemize}
@@
     that the expression might have different values in two different executions.
-    \item \<@OrderNonDet> indicates that
-    a collection or an array will have the same elements in every execution, but in a
-    possibly different order.  \<@OrderNonDet> may only be written on
-    collections and arrays.
-    A collection is any subtype of \<java.util.Collection>,
-    \<java.util.Iterator>, or \<java.util.Map>, including user-defined types.
     \item \<@Det> indicates that
@@
 \end{itemize}
+
+In this type system, \<@Det> is a subtype of \<@NonDet> denoted as \codeid{@Det <: @NonDet}.
 
@@
 One of the most common locations for a polymorphic annotation is at a method signature.
-Consider the following method declaration annotated with polymorphic annotations
+Consider the following method declaration annotated with polymorphic annotations:
 \begin{verbatim}
@@
 \end{verbatim}
-This indicates that method \<foo> can be called with arguments having any of the \codeid{@Det or @NonDet} type annotations 
-(\<@OrderNonDet> is not allowed because it is invalid on primitive types). \<@PolyDet> resolves to the least upper bound of
+There are two valid instantiations for this declaration:
+\begin{enumerate}
+    \item @NonDet int foo(@NonDet String param1, @NonDet boolean param2) { }
+    \item @Det int foo(@Det String param1, @Det boolean param2) { }
+\end{enumerate}
+This indicates that method \<foo> can be called with arguments having any of the \codeid{@Det or @NonDet} type annotations.
+%   (\<@OrderNonDet> is not allowed because it is invalid on primitive types).
+ \<@PolyDet> resolves to the least upper bound of
 the actual types on arguments. For instance, if this method is called as \codeid{foo(@Det String arg1, @NonDet boolean arg2)}, the 
@@
 causing the return type to have the type annotation \<@NonDet>.
+
+\subsection{Handling order nondeterminism}
+In addition to the type qualifiers described in~\ref{type-hierarchy}, the determinism checker 
+defines the type qualifier \<@OrderNonDet>. This qualifier represents
+a collection or an array that will have the same elements in every execution, but in a
+possibly different order.  \<@OrderNonDet> may only be written on
+collections and arrays.
+A collection is any subtype of \<java.util.Collection>,
+\<java.util.Iterator>, or \<java.util.Map>, including user-defined types. The relationship
+of \<@OrderNonDet> with respect the the other two type qualifiers is \codeid{@Det <: @OrderNonDet <: @NonDet} as shown in Figure~\ref{fig-determinism-hierarchy}.
 

### 2019-03-25 experiments.tex
moving text around

P1 :
[ERROR] /Users/rashmi/caseStudies/nondex/reflectasm/src/com/esotericsoftware/reflectasm/FieldAccess.java:[102,87] [assignment.type.incompatible] incompatible types in assignment.
[ERROR]   found   : @Det Field @OrderNonDet []
[ERROR]   required: @Det Field @Det []

P2:

		if (object instanceof Map) {
    emitter.emit(new MappingStartEvent(anchor, tag, !showTag, false));
    @Det Set<@Det Object> items = ((Map)object).entrySet();
    @Det Iterator<@Det Object> itemIterator = items.iterator();
    while (itemIterator.hasNext()) {
        Object item = itemIterator.next();
        Entry entry = (Entry)item;
        writeValue(entry.getKey(), null, null, null);
        if (isRoot && !config.writeConfig.writeRootElementTags) elementType = entry.getValue().getClass();
        writeValue(entry.getValue(), elementType, null, null);
    }
    //			for (@Det Object item : ((Map)object).entrySet()) {
        //				Entry entry = (Entry)item;
        //				writeValue(entry.getKey(), null, null, null);
        //				if (isRoot && !config.writeConfig.writeRootElementTags) elementType = entry.getValue().getClass();
        //				writeValue(entry.getValue(), elementType, null, null);
        //			}
    emitter.emit(Event.MAPPING_END);
    return;
}

### 2019-03-25 design.tex
type validity rules
fewer examples

@@
 
-\todo{It is very important to express the whole section as formal type
-  jugdments, or we will get laughed
-  out of the room as amateurs, during the program committee meeting.  Also,
-  the formalism will help make the ideas precise.
+%\todo{It is very important to express the whole section as formal type
+%  jugdments, or we will get laughed
+%  out of the room as amateurs, during the program committee meeting.  Also,
+%  the formalism will help make the ideas precise.
+%
+%  Type well-formedness rules:\todo{format nicely in \LaTeX}
+%
+%  antecedent: $\Gamma : \kappa_e <: \kappa_c$
+%
+%  consequent: $\Gamma : \kappa_c \ \codeid{List}\langle\kappa_e\ 
+%  \codeid{Object}\rangle$
+%  }
+Type validity rules:
 
-  Type well-formedness rules:\todo{format nicely in \LaTeX}
+$\infer[ANY TYPE]{\vdash : \kappa \  \tau}{\kappa : @Det / @NonDet}$
 
-  antecedent: $\Gamma : \kappa_e <: \kappa_c$
+%$\infer[COLLECTION TYPE]{\vdash : \kappa \  \tau}{\kappa : @OrderNonDet & \tau : Collection/Map/Iterator/array}$
 
-  consequent: $\Gamma : \kappa_c \ \codeid{List}\langle\kappa_e\ 
-  \codeid{Object}\rangle$
-  }
+$\infer[ORDERNONDET \ COLLECTION]{\vdash : \kappa_c <\kappa_e \ \tau_e> \tau_c}{\kappa_c : @OrderNonDet/@Det & \kappa_e <: \kappa_c & \tau_c <: Collection/Map/Iterator}$
+
+$\infer[ORDERNONDET \ ARRAY]{\vdash : \kappa_c [\kappa_e \ \tau_e] \tau_c}{\kappa_c : @OrderNonDet/@Det & \kappa_e <: \kappa_c }$
+
+$\infer[NONDET \ COLLECTION]{\vdash : \kappa_c <\kappa_e \ \tau_e> \tau_c}{\kappa_c : @NonDet & \kappa_e : @NonDet & \tau_c <: Collection/Map/Iterator}$
+
+$\infer[NONDET \ ARRAY]{\vdash : \kappa_c [\kappa_e \ \tau_e] \tau_c}{\kappa_c : @NonDet & \kappa_e : @NonDet }$
 
@@
 
-\todo{Fewer examples, less variation.  Use \<List> and \<String>.}
+%\todo{Fewer examples, less variation.  Use \<List> and \<String>.}
 
@@
     \item \codeid{@Det\ \ \ \ \ \ \ \ \ List<@Det\ \ \ \ \ \ \ \ \ Integer>}
-    \item \codeid{@OrderNonDet Set <@OrderNonDet Set<...>\relax >}
-    \item \codeid{@OrderNonDet Set <@Det String>}
-    \item \codeid{@Det\ \ \ \ \ \ \ \ \ Map <@Det String, @Det Integer>}
-    \item \codeid{@NonDet\ \ \ \ \ \ Set <T extends @NonDet Object>}
+    \item \codeid{@OrderNonDet List <@OrderNonDet Set<...>\relax >}
+    \item \codeid{@OrderNonDet List <@Det String>}
 \end{itemize}
@@
     \item \codeid{@NonDet\ \ \ \ \ \ List<@Det\ \ \ \ \ \ \ \ \ Integer>}
-    \item \codeid{@NonDet\ \ \ \ \ \ List<@OrderNonDet Set<...>\relax >}
-    \item \codeid{@Det\ \ \ \ \ \ \ \ \ List<@NonDet\ \ \ \ \ \ Integer>}
     \item \codeid{@Det\ \ \ \ \ \ \ \ \ List<@OrderNonDet Set<...>\relax >}
-    \item \codeid{@Det\ \ \ \ \ \ \ \ \ Map <@Det String, @OrderNonDet List<@Det String>>}
-    \item \codeid{@Det\ \ \ \ \ \ \ \ \ List<T extends @NonDet Object>}
-    \item \codeid{@NonDet\ \ \ \ \ \ Set\ <T extends @Det Object>}
 \end{itemize}
@@
 elements are not allowed.
-
-For example, these types are valid:
-\begin{itemize}
-    \item \codeid{@NonDet int @NonDet \ \ \ \ []}
-    \item \codeid{@Det \ \ \ int @Det \ \ \ \ \ \ \ []}
-    \item \codeid{@Det \ \ \ int @OrderNonDet[] @OrderNonDet[]}
-\end{itemize}
-
-These types are invalid:
-\begin{itemize}
-    \item \codeid{@NonDet int @OrderNonDet[]}
-    \item \codeid{@NonDet int @Det \ \ \ \ \ \ \ []}
-    \item \codeid{@Det \ \ \ int @NonDet \ \ \ \ []}
-\end{itemize}
 

### 2019-03-25 design.tex
poly resolution rules
small fix
fixes

@@
 on the return type \<int>. We circumvent this scenario by introducing two variants to the \<@PolyDet> annotation, \<@PolyDet("up")>
-and \<@PolyDet("down")>. The semantics of these variants are as follows:
+and \<@PolyDet("down")>.
+
+Valid locations for \codeid{@PolyDet("up")} and  \codeid{@PolyDet("down")}:
+
+$\infer{\vdash : \kappa \  \tau}{isParam(\tau) / isReturn(\tau) & \kappa : @PolyDet("up") / @PolyDet("down")}$
+
+Resolution of \<@PolyDet("up")>:\todo{typesetting}
+
+$\infer{\Gamma \vdash x : @NonDet \ \tau_a}{\Gamma \vdash x : \kappa_a \ \tau_a &  \kappa_a : @OrderNonDet/@NonDet & \kappa_p : @PolyDet("up") & \kappa_p \ \tau_p : declaration \ type & \kappa_a \ \tau_a : resolved \ type}$
+
+Resolution of \<@PolyDet("down")>:
+
+$\infer{\Gamma \vdash x : @Det \  \tau_a}{\Gamma \vdash x :  \kappa_a \tau_a & \kappa_a : @OrderNonDet/@Det & \kappa_p : @PolyDet("down") & \kappa_p \ \tau_p : declaration \ type & \kappa_a \ \tau_a : resolved \ type}$
+
+The semantics of these variants are as follows:
 \begin{itemize}
@@
 \end{verbatim}
-The checker framework performs dataflow analysis within method bodies and refines the types of local variables.
 In the example above, even though the local variable \codeid{y} was declared to be \codeid{@NonDet}, it gets
@@
 
-The determinism checker has additional type refinement rules for the sorting methods in collections. 
+The determinism checker has additional type refinement rule for the sorting methods in collections.
+
+$\infer{}{}$
+
 The checker type refines a receiver of type \codeid{@OrderNonDet List} to a \codeid{@Det List} if all the

### 2019-03-25 design.tex
typesetting

@@
 
-$\infer{\Gamma \vdash x : \tau_2}{\Gamma \vdash x : \tau_1 & \tau_1 <: \tau_2}$
+$\infer{\Gamma \vdash x : \tau_2}{\Gamma \vdash x : \tau_1 & \tau_1 \sqsubseteq \tau_2}$
 
@@
 
-$\forall i : i >= 0 \infer{method(\overline{a_i}) : \tau_r}{\tau_r \  method(\overline{\tau_i \  p_i}) & \overline{a_i : \tau_i}}$
+$\infer{\mathit{method}(\overline{a_i}) : \tau_r}{\tau_r \  \mathit{method}(\overline{\tau_i \  p_i}) & \overline{a_i : \tau_i}}$
 
@@
 \begin{itemize}
-    \item \<@NonDet> indicates
+    \item \<NonDet> indicates
     that the expression might have different values in two different executions.
-    \item \<@Det> indicates that
+    \item \<Det> indicates that
     the expression evaluates to the same value (with respect to \<.equals()>) in all
@@
 
-In this type system, \<@Det> is a subtype of \<@NonDet> denoted as \codeid{@Det <: @NonDet}.
+In this type system, \<Det> is a subtype of \<NonDet> denoted as \codeid{Det $\sqsubseteq$ NonDet}.
 
@@
 it indicates that it could resolve to any type in the type system depending on how it is used.
-In the determinism checker, we define a polymorphic annotation \<@PolyDet>.
+In the determinism checker, we define a polymorphic annotation \<PolyDet>.
 One of the most common locations for a polymorphic annotation is at a method signature.
@@
 In addition to the type qualifiers described in~\ref{type-hierarchy}, the determinism checker 
-defines the type qualifier \<@OrderNonDet>. This qualifier represents
+defines the type qualifier \<OrderNonDet>. This qualifier represents
 a collection or an array that will have the same elements in every execution, but in a
-possibly different order.  \<@OrderNonDet> may only be written on
+possibly different order.  \<OrderNonDet> may only be written on
 collections and arrays.
@@
 \<java.util.Iterator>, or \<java.util.Map>, including user-defined types. The relationship
-of \<@OrderNonDet> with respect the the other two type qualifiers is \codeid{@Det <: @OrderNonDet <: @NonDet} as shown in Figure~\ref{fig-determinism-hierarchy}.
+of \<OrderNonDet> with respect the the other two type qualifiers is \codeid{Det $\sqsubseteq$ OrderNonDet $\sqsubseteq$ NonDet} as shown in Figure~\ref{fig-determinism-hierarchy}.
 
@@
 
-$\infer[ANY TYPE]{\vdash : \kappa \  \tau}{\kappa : @Det / @NonDet}$
+$\infer[\sc{ANY TYPE}]{\vdash : \kappa \  \beta}{\kappa : \mathit{Det / NonDet}}$
 
-%$\infer[COLLECTION TYPE]{\vdash : \kappa \  \tau}{\kappa : @OrderNonDet & \tau : Collection/Map/Iterator/array}$
+%$\infer[COLLECTION TYPE]{\vdash : \kappa \  \beta}{\kappa : OrderNonDet & \beta : Collection/Map/Iterator/array}$
 
-$\infer[ORDERNONDET \ COLLECTION]{\vdash : \kappa_c <\kappa_e \ \tau_e> \tau_c}{\kappa_c : @OrderNonDet/@Det & \kappa_e <: \kappa_c & \tau_c <: Collection/Map/Iterator}$
+$\infer[\sc{ORDERNONDET \ COLLECTION}]{\vdash : \kappa_c \langle\kappa_e \ \beta_e\rangle \beta_c}{\kappa_c : \mathit{OrderNonDet/Det} & \kappa_e \sqsubseteq \kappa_c & \beta_c \sqsubseteq \mathit{Collection/Map/Iterator}}$
 
-$\infer[ORDERNONDET \ ARRAY]{\vdash : \kappa_c [\kappa_e \ \tau_e] \tau_c}{\kappa_c : @OrderNonDet/@Det & \kappa_e <: \kappa_c }$
+$\infer[\sc{ORDERNONDET \ ARRAY}]{\vdash : \kappa_c [\kappa_e \ \beta_e] \beta_c}{\kappa_c : \mathit{OrderNonDet/Det} & \kappa_e \sqsubseteq \kappa_c }$
 
-$\infer[NONDET \ COLLECTION]{\vdash : \kappa_c <\kappa_e \ \tau_e> \tau_c}{\kappa_c : @NonDet & \kappa_e : @NonDet & \tau_c <: Collection/Map/Iterator}$
+$\infer[\sc{NONDET \ COLLECTION}]{\vdash : \kappa_c \langle\kappa_e \ \beta_e\rangle \beta_c}{\kappa_c : \mathit{NonDet} & \kappa_e : \mathit{NonDet} & \beta_c \sqsubseteq \mathit{Collection/Map/Iterator}}$
 
-$\infer[NONDET \ ARRAY]{\vdash : \kappa_c [\kappa_e \ \tau_e] \tau_c}{\kappa_c : @NonDet & \kappa_e : @NonDet }$
+$\infer[\sc{NONDET \ ARRAY}]{\vdash : \kappa_c [\kappa_e \ \beta_e] \beta_c}{\kappa_c : \mathit{NonDet} & \kappa_e : \mathit{NonDet} }$
 
@@
 the type of the type parameter (see Figure~\ref{fig-determinism-collections}).
-If the Collection is \<@NonDet>, then the type parameter may not be
-\<@Det> or \<@OrderNonDet>. Although such types could exist, they are
+If the Collection is \<NonDet>, then the type parameter may not be
+\<Det> or \<OrderNonDet>. Although such types could exist, they are
 disallowed to prevent the following bug:
@@
 \begin{enumerate}
-    \item The individual elements extracted by iterating over the collection get the type \<@NonDet>.
+    \item The individual elements extracted by iterating over the collection get the type \<NonDet>.
     \item Accessing properties sich as \<size()> or querying if the collection \<isEmpty()> will return types
-    that are annotated as \<@Det> as they do not depend on the iteration order. 
+    that are annotated as \<Det> as they do not depend on the iteration order. 
 \end{enumerate}
 
-A user defined type may be annotated as \<@OrderNonDet> if and only if it
+A user defined type may be annotated as \<OrderNonDet> if and only if it
 is a subtype of Collection or Iterator.
@@
 
-$\infer{\vdash : \kappa \  \tau}{isParam(\tau) / isReturn(\tau) & \kappa : @PolyDet("up") / @PolyDet("down")}$
+$\infer{\vdash : \kappa \  \beta}{isParam(\beta) / isReturn(\beta) & \kappa : @PolyDet("up") / @PolyDet("down")}$
 
@@
 
-$\infer{\Gamma \vdash x : @NonDet \ \tau_a}{\Gamma \vdash x : \kappa_a \ \tau_a &  \kappa_a : @OrderNonDet/@NonDet & \kappa_p : @PolyDet("up") & \kappa_p \ \tau_p : declaration \ type & \kappa_a \ \tau_a : resolved \ type}$
+$\infer{\Gamma \vdash x : @NonDet \ \beta_a}{\Gamma \vdash x : \kappa_a \ \beta_a &  \kappa_a : @OrderNonDet/@NonDet & \kappa_p : @PolyDet("up") & \kappa_p \ \beta_p : declaration \ type & \kappa_a \ \beta_a : resolved \ type}$
 
@@
 
-$\infer{\Gamma \vdash x : @Det \  \tau_a}{\Gamma \vdash x :  \kappa_a \tau_a & \kappa_a : @OrderNonDet/@Det & \kappa_p : @PolyDet("down") & \kappa_p \ \tau_p : declaration \ type & \kappa_a \ \tau_a : resolved \ type}$
+$\infer{\Gamma \vdash x : @Det \  \beta_a}{\Gamma \vdash x :  \kappa_a \beta_a & \kappa_a : @OrderNonDet/@Det & \kappa_p : @PolyDet("down") & \kappa_p \ \beta_p : declaration \ type & \kappa_a \ \beta_a : resolved \ type}$
 
@@
 The determinism checker has additional type refinement rule for the sorting methods in collections.
-
-$\infer{}{}$
 

### 2019-03-25 design.tex
more typesetting
map validity rules

@@
 
-Subtyping rule:
+\begin{figure}
+    $\infer[\sc{\mathit{SUBTYPING}}]{\Gamma \vdash x : \tau_2}{\Gamma \vdash x : \tau_1 & \tau_1 \sqsubseteq \tau_2}$
 
-$\infer{\Gamma \vdash x : \tau_2}{\Gamma \vdash x : \tau_1 & \tau_1 \sqsubseteq \tau_2}$
+    \bigskip
 
-Typing rule for method calls:
-\todo{typesetting}
+    $\infer[\sc{\mathit{METHOD \ CALLS}}]{\mathit{method}(\overline{a_i}) : \tau_r}{\tau_r \  \mathit{method}(\overline{\tau_i \  p_i}) & \overline{a_i : \tau_i}}$
 
-$\infer{\mathit{method}(\overline{a_i}) : \tau_r}{\tau_r \  \mathit{method}(\overline{\tau_i \  p_i}) & \overline{a_i : \tau_i}}$
+    \caption{Type checking rules}
+    \label{typecheck-rules}
+\end{figure}
 
@@
 %  }
-Type validity rules:
+\begin{figure}
+    $\infer[\sc{\mathit{ANY\ TYPE}}]{\vdash : \kappa \  \beta}{\kappa : \mathit{Det / NonDet}}$
 
-$\infer[\sc{ANY TYPE}]{\vdash : \kappa \  \beta}{\kappa : \mathit{Det / NonDet}}$
+    \bigskip
 
-%$\infer[COLLECTION TYPE]{\vdash : \kappa \  \beta}{\kappa : OrderNonDet & \beta : Collection/Map/Iterator/array}$
+    $\infer[\sc{\mathit{ORDERNONDET \ COLLECTION}}]{\vdash : \kappa_c \  \beta_c \langle\kappa_e \ \beta_e\rangle}{\kappa_c : \mathit{OrderNonDet/Det} & \kappa_e \sqsubseteq \kappa_c & \beta_c \sqsubseteq \mathit{Collection/Iterator}}$
 
-$\infer[\sc{ORDERNONDET \ COLLECTION}]{\vdash : \kappa_c \langle\kappa_e \ \beta_e\rangle \beta_c}{\kappa_c : \mathit{OrderNonDet/Det} & \kappa_e \sqsubseteq \kappa_c & \beta_c \sqsubseteq \mathit{Collection/Map/Iterator}}$
+    \bigskip
 
-$\infer[\sc{ORDERNONDET \ ARRAY}]{\vdash : \kappa_c [\kappa_e \ \beta_e] \beta_c}{\kappa_c : \mathit{OrderNonDet/Det} & \kappa_e \sqsubseteq \kappa_c }$
+    $\infer[\sc{\mathit{NONDET \ COLLECTION}}]{\vdash : \mathit{NonDet} \ \beta_c\langle\mathit{NonDet} \ \beta_e\rangle}{\beta_c \sqsubseteq \mathit{Collection/Iterator}}$
 
-$\infer[\sc{NONDET \ COLLECTION}]{\vdash : \kappa_c \langle\kappa_e \ \beta_e\rangle \beta_c}{\kappa_c : \mathit{NonDet} & \kappa_e : \mathit{NonDet} & \beta_c \sqsubseteq \mathit{Collection/Map/Iterator}}$
+    \bigskip
 
-$\infer[\sc{NONDET \ ARRAY}]{\vdash : \kappa_c [\kappa_e \ \beta_e] \beta_c}{\kappa_c : \mathit{NonDet} & \kappa_e : \mathit{NonDet} }$
+     $\infer[\sc{\mathit{ORDERNONDET \ ARRAY}}]{\vdash : \kappa_e \ \beta_e \ \  \kappa_a []}{\kappa_a : \mathit{OrderNonDet/Det} & \kappa_e \sqsubseteq \kappa_a }$
+
+    \bigskip
+
+    $\infer[\sc{\mathit{NONDET \ ARRAY}}]{\vdash : \mathit{NonDet} \ \beta_e \ \  \mathit{NonDet} []}{}$
+
+    \bigskip
+
+    $\infer[\sc{\mathit{ORDERNONDET \ MAP}}]{\vdash : \kappa_m \  \beta_m \langle\kappa_k \ \beta_k, \kappa_v \ \beta_v\rangle}{\kappa_m: \mathit{OrderNonDet/Det} & \kappa_k \sqsubseteq \kappa_m & \kappa_v \sqsubseteq \kappa_m & \beta_m \sqsubseteq \mathit{Map}}$
+
+    \bigskip
+
+    $\infer[\sc{\mathit{NONDET \ MAP}}]{\vdash : \mathit{NonDet} \ \beta_m\langle\mathit{NonDet} \ \beta_k, \mathit{NonDet} \ \beta_v\rangle}{\beta_m \sqsubseteq \mathit{Map}}$
+
+    \caption{Type well formedness}
+    \label{type-validity}
+\end{figure}
 
@@
 
-\subsection{CLIMB-to-top and collection, array locals}\label{climb-rules}
-
-Every type system in the checker framework must define a default type annotation. This annotation will be applied
-to every location that isn't explicitly annotated. In our determinism checker, we chose \codeid{@Det} as the default annotation.
-The reason for this design choice is that we expect the program to be 
-deterministic unless specified by programmer or the program calls nondeterministic library methods that we have annotated 
-(Random, Collections, etc).
-The framework applies the CLIMB-to-top rule. This rule states that the \textit{top} qualifier 
-in the hierarchy is the default for the CLIMB locations: Casts, Locals, Instanceof, and (some) iMplicit Bounds. The rationale
-for this rule is as follows:
-
-\begin{itemize}
-	\item Local variables are defaulted to top because type refinement is applied to local variables. If a local variable starts as the 
-	\textit{top} type, then the Checker Framework refines it to the best (most specific) possible type based on assignments to it. As a 
-	result, a programmer rarely writes an explicit annotation on any of those locations.
-	Variables defaulted to top include \textit{local variables}, resource variables in the \textit{try-with-resources} construct, variables 
-	in \textit{for} statements, and \textit{catch} arguments (known as exception parameters in the Java Language Specification). 
-	\textit{Exception parameters} need to have the \textit{top} type because exceptions of arbitrary qualified types can be thrown 
-	and the Checker Framework does not provide runtime checks.
-	\item \textit{Cast} and \textit{instanceof} types are given the same type as their argument expression. This has the same effect as 
-	if they were given the \textit{top} type and then flow-sensitively refined to the type of their argument.
-	\item Implicit upper bounds are defaulted to top to allow them to be instantiated in any way. If a user declared class \codeid{C<T> 
-		{ ... }}, then the Checker Framework assumes that the user intended to allow any instantiation of the class, and the declaration
-	is interpreted as class \codeid{C<T extends @NonDet Object> { ...
-}} rather than as class \codeid{C<T extends @Det Object> { ...
-}}.
-The latter would forbid instantiations such as \codeid{C<@Det String>}, or would require rewriting of code.
-On the other hand, if a user writes
-	an explicit bound such as class \codeid{C<T extends D> { ... }}, then the user intends some restriction on instantiation and can write a 
-	qualifier on the upper bound as desired.
-	This rule means that the upper bound of class \codeid{C<T>} is defaulted differently than the upper bound of class \codeid{C<T extends Object>}. 
-	This may seem confusing, but it is the least bad option. The more confusing alternative would be for \codeid{Object} to be defaulted 
-	differently in class \codeid{C<T extends Object>} and in an instantiation \codeid{C<Object>}, and for the upper bounds to be defaulted differently 
-	in class \codeid{C<T extends Object>} and class \codeid{C<T extends Date>}.
-	\item \textit{Implicit lower bounds} are defaulted to the \textit{bottom} type, again to allow maximal instantiation.
-Note that Java does not allow a programmer to express both the upper and lower bounds of a type, but the Checker Framework allows the programmer to specify either or both.
-\end{itemize}
-With the above rules, the checker framework will automatically annotate a local collection with type \<@NonDet> and its
-type arguments will get the type \<@Det>. For example, a locally declared \codeid{List} of \codeid{Strings} will be automatically annotated
-as \codeid{@NonDet List<@Det String>}. Recall from~\ref{collection-rules} that the determinism checker does not allow this type on
-any collection. While this is still sound, it will result in the determinism checker reporting a lot of errors because every locally declared 
-collection gets annotated with this invalid type. To avoid this situation, we make an exception to the defaulting rules for local collections
-and annotate their type parameters also with \<@NonDet>.
-So a local list of Strings gets the default annotation of \codeid{@NonDet List<@NonDet String>}.
-While this defaulting eliminates the possibility of automatically annotating locals collections with an invalid
-type, it would could in a high number of false positives. In the following code snippet,
-\begin{verbatim}
-void test(@Det List<@Det String> argList) {
-            List<String> localList = argList;
-}    
-\end{verbatim}
-the checker would report an error at the assignment statement since the inferred type for \codeid{localList} is 
-\codeid{@NonDet List<@NonDet> String} and since collections types are invariant.
-We therefore recommend that programmers explicitly annotate
-all local variables that are collections to reduce false positives. The same reasoning and recommendation applies to arrays.
-
 \subsection{Discussion on defaulting rules}\label{defaulting}
@@
 
-\subsection{Determinism checker specific rules for precision}
+%\subsection{CLIMB-to-top and collection, array locals}\label{climb-rules}
+%
+%Every type system in the checker framework must define a default type annotation. This annotation will be applied
+%to every location that isn't explicitly annotated. In our determinism checker, we chose \codeid{@Det} as the default annotation.
+%The reason for this design choice is that we expect the program to be 
+%deterministic unless specified by programmer or the program calls nondeterministic library methods that we have annotated 
+%(Random, Collections, etc).
+%The framework applies the CLIMB-to-top rule. This rule states that the \textit{top} qualifier 
+%in the hierarchy is the default for the CLIMB locations: Casts, Locals, Instanceof, and (some) iMplicit Bounds. The rationale
+%for this rule is as follows:
+%
+%\begin{itemize}
+%    \item Local variables are defaulted to top because type refinement is applied to local variables. If a local variable starts as the 
+%    \textit{top} type, then the Checker Framework refines it to the best (most specific) possible type based on assignments to it. As a 
+%    result, a programmer rarely writes an explicit annotation on any of those locations.
+%    Variables defaulted to top include \textit{local variables}, resource variables in the \textit{try-with-resources} construct, variables 
+%    in \textit{for} statements, and \textit{catch} arguments (known as exception parameters in the Java Language Specification). 
+%    \textit{Exception parameters} need to have the \textit{top} type because exceptions of arbitrary qualified types can be thrown 
+%    and the Checker Framework does not provide runtime checks.
+%    \item \textit{Cast} and \textit{instanceof} types are given the same type as their argument expression. This has the same effect as 
+%    if they were given the \textit{top} type and then flow-sensitively refined to the type of their argument.
+%    \item Implicit upper bounds are defaulted to top to allow them to be instantiated in any way. If a user declared class \codeid{C<T> 
+%        { ... }}, then the Checker Framework assumes that the user intended to allow any instantiation of the class, and the declaration
+%    is interpreted as class \codeid{C<T extends @NonDet Object> { ...
+}} rather than as class \codeid{C<T extends @Det Object> { ...
+}}.
+The latter would forbid instantiations such as \codeid{C<@Det String>}, or would require rewriting of code.
+On the other hand, if a user writes
+%    an explicit bound such as class \codeid{C<T extends D> { ... }}, then the user intends some restriction on instantiation and can write a 
+%    qualifier on the upper bound as desired.
+%    This rule means that the upper bound of class \codeid{C<T>} is defaulted differently than the upper bound of class \codeid{C<T extends Object>}. 
+%    This may seem confusing, but it is the least bad option. The more confusing alternative would be for \codeid{Object} to be defaulted 
+%    differently in class \codeid{C<T extends Object>} and in an instantiation \codeid{C<Object>}, and for the upper bounds to be defaulted differently 
+%    in class \codeid{C<T extends Object>} and class \codeid{C<T extends Date>}.
+%    \item \textit{Implicit lower bounds} are defaulted to the \textit{bottom} type, again to allow maximal instantiation.
+Note that Java does not allow a programmer to express both the upper and lower bounds of a type, but the Checker Framework allows the programmer to specify either or both.
+%\end{itemize}
+%With the above rules, the checker framework will automatically annotate a local collection with type \<@NonDet> and its
+%type arguments will get the type \<@Det>. For example, a locally declared \codeid{List} of \codeid{Strings} will be automatically annotated
+%as \codeid{@NonDet List<@Det String>}. Recall from~\ref{collection-rules} that the determinism checker does not allow this type on
+%any collection. While this is still sound, it will result in the determinism checker reporting a lot of errors because every locally declared 
+%collection gets annotated with this invalid type. To avoid this situation, we make an exception to the defaulting rules for local collections
+%and annotate their type parameters also with \<@NonDet>.
+So a local list of Strings gets the default annotation of \codeid{@NonDet List<@NonDet String>}.
+While this defaulting eliminates the possibility of automatically annotating locals collections with an invalid
+%type, it would could in a high number of false positives. In the following code snippet,
+%\begin{verbatim}
+%void test(@Det List<@Det String> argList) {
+%List<String> localList = argList;
+%}    
+%\end{verbatim}
+%the checker would report an error at the assignment statement since the inferred type for \codeid{localList} is 
+%\codeid{@NonDet List<@NonDet> String} and since collections types are invariant.
+%We therefore recommend that programmers explicitly annotate
+%all local variables that are collections to reduce false positives. The same reasoning and recommendation applies to arrays.
+
+\subsection{Determinism checker specific rules for precision}\label{precision}
 
@@
 
-\subsection{Maps discussion}
+\subsection{Maps discussion}\label{maps}
 
@@
 
-\subsection{Annotating more classes/providing more specifications}
+\subsection{Annotating more classes/providing more specifications}\label{annotating-jdk}
 \begin{itemize}

### 2019-03-26 design.tex
type validity section

@@
 
+Figure~\ref{type-validity} shows the type validity rules of our type system. A type $\tau$ is represented as $\kappa \ \beta$
+where $\kappa$ is a type qualifier from the determinism type hierarchy and $\beta$ is a Java type.
+As stated in the rule $\sc{\mathit{ANY\ TYPE}}$, the types \<Det> and \<NonDet> can be written on any Java type.
 The (determinism) type of a Collection or Iterator must be a supertype or equal to
-the type of the type parameter (see Figure~\ref{fig-determinism-collections}).
+the type of the type parameter (Rule $\sc{\mathit{ORDERNONDET \ COLLECTION}}$) (see Figure~\ref{fig-determinism-collections}).
 If the Collection is \<NonDet>, then the type parameter may not be
-\<Det> or \<OrderNonDet>. Although such types could exist, they are
+\<Det> or \<OrderNonDet> (Rule $\sc{\mathit{NONDET \ COLLECTION}}$ in Figure~\ref{type-validity}). Although such types could exist, they are
 disallowed to prevent the following bug:
@@
 As with collections, \codeid{@NonDet} arrays of \codeid{@Det} or \codeid{@OrderNonDet}
-elements are not allowed.
+elements are not allowed (Rules $\sc{\mathit{ORDERNONDET \ ARRAY}}$ and $\sc{\mathit{NONDET \ ARRAY}}$ in Figure~\ref{type-validity}).
 

### 2019-03-26 design.tex
poly resolution table
poly resolution
type refinement

@@
 
-Valid locations for \codeid{@PolyDet("up")} and  \codeid{@PolyDet("down")}:
-
-$\infer{\vdash : \kappa \  \beta}{isParam(\beta) / isReturn(\beta) & \kappa : @PolyDet("up") / @PolyDet("down")}$
-
-Resolution of \<@PolyDet("up")>:\todo{typesetting}
-
-$\infer{\Gamma \vdash x : @NonDet \ \beta_a}{\Gamma \vdash x : \kappa_a \ \beta_a &  \kappa_a : @OrderNonDet/@NonDet & \kappa_p : @PolyDet("up") & \kappa_p \ \beta_p : declaration \ type & \kappa_a \ \beta_a : resolved \ type}$
-
-Resolution of \<@PolyDet("down")>:
-
-$\infer{\Gamma \vdash x : @Det \  \beta_a}{\Gamma \vdash x :  \kappa_a \beta_a & \kappa_a : @OrderNonDet/@Det & \kappa_p : @PolyDet("down") & \kappa_p \ \beta_p : declaration \ type & \kappa_a \ \beta_a : resolved \ type}$
+% Not true: We allow Poly(up) and down at every location where Poly is allowed.
+%Valid locations for \codeid{@PolyDet("up")} and  \codeid{@PolyDet("down")}:
+%$\infer{\vdash : \kappa \  \beta}{isParam(\beta) / isReturn(\beta) & \kappa : @PolyDet("up") / @PolyDet("down")}$
+%
+%Resolution of \<@PolyDet("up")>:\todo{typesetting}
+%
+%$\infer{\Gamma \vdash x : @NonDet \ \beta_a}{\Gamma \vdash x : \kappa_a \ \beta_a &  \kappa_a : @OrderNonDet/@NonDet & \kappa_p : @PolyDet("up") & \kappa_p \ \beta_p : declaration \ type & \kappa_a \ \beta_a : resolved \ type}$
+%
+%Resolution of \<@PolyDet("down")>:
+%
+%$\infer{\Gamma \vdash x : @Det \  \beta_a}{\Gamma \vdash x :  \kappa_a \beta_a & \kappa_a : @OrderNonDet/@Det & \kappa_p : @PolyDet("down") & \kappa_p \ \beta_p : declaration \ type & \kappa_a \ \beta_a : resolved \ type}$
 
@@
 \end{itemize}
+
+\begin{figure}
+    $\infer[\sc{\mathit{POLY RESOLUTION}}]{\pi_r \ \beta_r method(\overline{\pi_i \ \beta_i})\{\ \}}{\forall K \in \{\mathit{NonDet, OrderNonDet, Det\}} : K \ \beta_r method(\overline{K \ \beta_i})\{\ \} & \pi_r, \pi_i \in \{\mathit{PolyDet}, \mathit{PolyDet}\uparrow, \mathit{PolyDet}\downarrow    \}}$
+\end{figure}
+
+\begin{table}[]
+    \begin{tabular}{|l|l|l|}
+        \hline
+        \textbf{\<PolyDet>} & \textbf{\<PolyDet$\uparrow$>} & \textbf{\<PolyDet$\downarrow$>} \\ \hline
+        \<NonDet> & \<NonDet> &  \<NonDet>\\ \hline
+        \<OrderNonDet> & \<NonDet> &  \<Det>\\ \hline
+        \<Det> & \<Det> &  \<Det>\\ \hline
+    \end{tabular}
+\caption{Polymorphic resolution}
+\label{tab-poly-resolutions}
+\end{table}
 
@@
 
-The determinism checker has additional type refinement rule for the sorting methods in collections.
+The determinism checker has additional type refinement rule for the sorting methods as shown in Figure~\ref{fig-sorting}.
+
+\begin{figure}
+    $\infer[\sc{\mathit{LIST \  SORT}}]{\Gamma \vdash x : \mathit{Det \ List}\langle\mathit{Det} \ \beta\rangle}{\Gamma \vdash x : \mathit{OrderNonDet \ List}\langle\mathit{Det} \ \beta\rangle & \mathit{x.sort()}}$
+
+    \bigskip
+
+    $\infer[\sc{\mathit{COLLECTIONS \  SORT}}]{\Gamma \vdash x : \mathit{Det \ List}\langle\mathit{Det} \ \beta\rangle}{\Gamma \vdash x : \mathit{OrderNonDet \ List}\langle\mathit{Det} \ \beta\rangle & \mathit{Collections.sort.sort(x)}}$
+
+    \bigskip
+
+    $\infer[\sc{\mathit{ARRAYS \  SORT}}]{\Gamma \vdash x : \mathit{Det \ \beta \ Det[]} }{\Gamma \vdash x : \mathit{Det \ \beta \ OrderNonDet[]} & \mathit{Arrays.sort( \ x)}}$
+
+    \caption{Sorting type refinement rules}
+    \label{fig-sorting}
+\end{figure}
 
 The checker type refines a receiver of type \codeid{@OrderNonDet List} to a \codeid{@Det List} if all the
-type argument of this \codeid{List} have the type annotation \codeid{@Det}. Similar rules apply for \codeid{Arrays.sort()} and
-\codeid{Collections.sort()}.
-Under this if \codeid{sort()} is called on a receiver of type \codeid{@OrderNonDet List<@Det Set<@Det String>>}, it will
+type argument of this \codeid{List} have the type annotation \codeid{@Det} (Rule LIST SORT in Figure~\ref{fig-sorting}). Similar rules apply for \codeid{Arrays.sort()} and \codeid{Collections.sort()}.
+As per these rules, if \codeid{sort()} is called on a receiver of type \codeid{@OrderNonDet List<@Det Set<@Det String>>}, it will
 be type refined to \codeid{@Det List<@Det Set<@Det String>>}. However a receiver of type \codeid{@OrderNonDet List<@OrderNonDet Set<@Det String>>} will not be type refined.

### 2019-03-26 design.tex
defaulting and type refinement

@@
 
+\subsection{Discussion on defaulting rules}\label{defaulting}
+Every type system in the checker framework must define a default type annotation. This annotation will be applied
+to every location that isn't explicitly annotated. In our determinism checker, we chose \<@PolyDet> as the default for all 
+method parameters and return types including constructors.
+The rationale for this design choice was to not make our type checker too restrictive by making everything \<@Det>.  This implies that
+the checker still allows some non-determinism as long as it doesn't break any determinism specifications explicitly written by the programmer.
+
+\begin{verbatim}
+@PolyDet int foo(@PolyDet int x) { }
+void bar(@NonDet int random, @Det int value) {
+@NonDet int x = foo(random);      // No error.
+@Det int y = foo(value);          // No error.
+System.out.println(foo(random));  // Error.
+}
+\end{verbatim}
+In the example above, the method \codeid{foo} is annotated with the \<PolyDet> type qualifier which allows
+the caller method \codeid{bar} to invoke it with both \<Det> and \<NonDet> arguments. The checker flags an error 
+only when the returned value is used in a way that breaks any deterministic specification (Printing a \<NonDet> value
+breaks our determinism guarantees).
+
 \subsection{Dataflow analysis and type refinement}\label{dataflow}
@@
 the checker reports a warning at the second assignment statement of \codeid{foo}. Type refinement only applies to locals. Refining non-locals can result in side effects making the analysis unsound.
+
+With the above rules, the checker framework will automatically annotate a local collection with type \<@NonDet> and its
+type arguments will get the type \<@Det>. For example, a locally declared \codeid{List} of \codeid{Strings} will be automatically annotated
+as \codeid{@NonDet List<@Det String>}. Recall from~\ref{collection-rules} that the determinism checker does not allow this type on
+any collection. While this is still sound, it will result in the determinism checker reporting a lot of errors because every locally declared 
+collection gets annotated with this invalid type. To avoid this situation, we make an exception to the defaulting rules for local collections
+and annotate their type parameters also with \<@NonDet>.
+So a local list of Strings gets the default annotation of \codeid{@NonDet List<@NonDet String>}.
+While this defaulting eliminates the possibility of automatically annotating locals collections with an invalid
+type, it would could in a high number of false positives. In the following code snippet,
+\begin{verbatim}
+void test(@Det List<@Det String> argList) {
+List<String> localList = argList;
+}    
+\end{verbatim}
+the checker would report an error at the assignment statement since the inferred type for \codeid{localList} is 
+\codeid{@NonDet List<@NonDet> String} and since collections types are invariant.
+We therefore recommend that programmers explicitly annotate
+all local variables that are collections to reduce false positives. The same reasoning and recommendation applies to arrays.
 
@@
 be type refined to \codeid{@Det List<@Det Set<@Det String>>}. However a receiver of type \codeid{@OrderNonDet List<@OrderNonDet Set<@Det String>>} will not be type refined.
-
-\subsection{Discussion on defaulting rules}\label{defaulting}
-In our determinism checker, we chose \<@PolyDet> as the default for all method parameters and return types including constructors.
-The rationale for this design choice was to not make our type checker too restrictive by making everything \<@Det>.  This implies that
-the checker still allows some non-determinism as long as it doesn't break any determinism specifications explicitly written by the programmer.
-
-\begin{itemize}
-    \item Example of a method with params and return types annotated as \<@PolyDet>.
-    This method called from 2 places - one harmless(@NonDet instantiations but no printing), and another that call with \<@NonDet> args and prints the return type.
-    \item Exception to this defaulting is the construction of HashSets and HashMaps because they can't be \<@Det>.
-    We explicitly annotate the return type on constructors as
-        \<@OrderNonDet> and disallow constructing \<@Det HashSet> or \<@Det
-        HashMap>.
-\end{itemize}
 
@@
 
+\begin{itemize}
+    \item Exception to this defaulting is the construction of HashSets and HashMaps because they can't be \<@Det>.
+    We explicitly annotate the return type on constructors as
+    \<@OrderNonDet> and disallow constructing \<@Det HashSet> or \<@Det
+    HashMap>.
+\end{itemize}
+
 \subsection{Annotating more classes/providing more specifications}\label{annotating-jdk}

### 2019-03-26 design.tex
formatting and language

@@
 \begin{verbatim}
-@PolyDet int foo(@PolyDet int x) { }
+@PolyDet int foo(@PolyDet int x) { ... }
+
 void bar(@NonDet int random, @Det int value) {
@@
 With the above rules, the checker framework will automatically annotate a local collection with type \<@NonDet> and its
-type arguments will get the type \<@Det>. For example, a locally declared \codeid{List} of \codeid{Strings} will be automatically annotated
+type arguments will get the type qualifier \<@Det>. For example, a locally declared \codeid{List} of \codeid{Strings} will be automatically annotated
 as \codeid{@NonDet List<@Det String>}. Recall from~\ref{collection-rules} that the determinism checker does not allow this type on
@@
 While this defaulting eliminates the possibility of automatically annotating locals collections with an invalid
-type, it would could in a high number of false positives. In the following code snippet,
+type, it could result in a high number of false positives. In the following code snippet,
 \begin{verbatim}
@@
 
-The determinism checker has additional type refinement rule for the sorting methods as shown in Figure~\ref{fig-sorting}.
-
+The determinism checker has additional type refinement rules for the sorting methods as shown in Figure~\ref{fig-sorting}. 
 \begin{figure}
@@
 \end{figure}
-
 The checker type refines a receiver of type \codeid{@OrderNonDet List} to a \codeid{@Det List} if all the

### 2019-03-26 design.tex
extending det checker

@@
 
-\subsection{Determinism checker specific rules for precision}\label{precision}
-
-\begin{itemize}
-    \item \<Set.equals()> is annotated as \<@Poly("up")>. This is sound but imprecise.
-    \item Example1: \codeid{@OrderNonDet Set<@Det List>.equals(@OrderNonDet Set<@Det List>)} is deterministic
-    \item Example2: \codeid{@OrderNonDet Set<@OrderNonDet List>.equals(@OrderNonDet Set<@OrderNonDet List>)} is not deterministic.
-    \item Explain the rules under which we make Set equality deterministic.
-    \item We annotate System.getProperty as \<@NonDet> because the results could vary across machines.
-    \item Exceptions: line.Sepaator, file.Separator. and path.Separator.
-    \item discuss type on Exceptions and unsoundness due to this.
-\end{itemize}
-
 \subsection{Maps discussion}\label{maps}
@@
 
-\subsection{Annotating more classes/providing more specifications}\label{annotating-jdk}
+\subsection{Determinism checker specific rules for precision}\label{precision}
+
 \begin{itemize}
-    \item Type system developer has to annotate methods in the JDK or any other library.
-    \item Note: Another source of Nondeterminism we haven't analyzed: File system.
-    \item Users can also provide stubs instead. (Steal some text from manual here)
+    \item \<Set.equals()> is annotated as \<@Poly("up")>. This is sound but imprecise.
+    \item Example1: \codeid{@OrderNonDet Set<@Det List>.equals(@OrderNonDet Set<@Det List>)} is deterministic
+    \item Example2: \codeid{@OrderNonDet Set<@OrderNonDet List>.equals(@OrderNonDet Set<@OrderNonDet List>)} is not deterministic.
+    \item Explain the rules under which we make Set equality deterministic.
+    \item We annotate System.getProperty as \<@NonDet> because the results could vary across machines.
+    \item Exceptions: line.Sepaator, file.Separator. and path.Separator.
+    \item discuss type on Exceptions and unsoundness due to this.
 \end{itemize}
+
+\subsection{Extending the determinism checker}\label{extending-det-checker}
+We have primarily focused on collections and annotated the corresponding classes in the JDK.
+These annotations provide machine verifiable specifications of determinism properties.
+Another source of nondeterminism that we haven't analyzed closely is the File system methods.
+A program analysis developer interested in verifying determinism guarantees with respect to 
+the File system can easily do so by annotating the parameter and return types of those methods in the JDK.
+Alternatively, a user can write these annotations in a \textit{stub} file and supply it to the type checker. A \textit{stub}
+only needs to contain method declarations along with their annotations.

### 2019-03-26 design.tex
Fixes - removing @ from types

@@
 \begin{verbatim}
-@PolyDet int foo(@PolyDet String param1, @PolyDet boolean param2) { }
+PolyDet int foo(PolyDet String param1, PolyDet boolean param2) { }
 \end{verbatim}
@@
 \begin{enumerate}
-    \item @NonDet int foo(@NonDet String param1, @NonDet boolean param2) { }
-    \item @Det int foo(@Det String param1, @Det boolean param2) { }
+    \item NonDet int foo(NonDet String param1, NonDet boolean param2) { }
+    \item Det int foo(Det String param1, Det boolean param2) { }
 \end{enumerate}
-This indicates that method \<foo> can be called with arguments having any of the \codeid{@Det or @NonDet} type annotations.
+This indicates that method \<foo> can be called with arguments having any of the \codeid{Det or NonDet} type annotations.
 %   (\<@OrderNonDet> is not allowed because it is invalid on primitive types).
- \<@PolyDet> resolves to the least upper bound of
-the actual types on arguments. For instance, if this method is called as \codeid{foo(@Det String arg1, @NonDet boolean arg2)}, the 
-checker framework resolves the method declaration as \codeid{@NonDet int foo(@NonDet String param1, @NonDet boolean param2)}
-causing the return type to have the type annotation \<@NonDet>.
+ \<PolyDet> resolves to the least upper bound of
+the actual types on arguments. For instance, if this method is called as \codeid{foo(Det String arg1, NonDet boolean arg2)}, the 
+checker framework resolves the method declaration as \codeid{NonDet int foo(NonDet String param1, NonDet boolean param2)}
+causing the return type to have the type annotation \<NonDet>.
 
@@
 
-The possibility of mutation allows us to add to the \<@Det List> at a
-\<@NonDet> index, which is unsound.
+The possibility of mutation allows us to add to the \<Det List> at a
+\<NonDet> index, which is unsound.
 
@@
 \begin{itemize}
-    \item \codeid{@NonDet\ \ \ \ \ \ List<@NonDet\ \ \ \ \ \ Integer>}
-    \item \codeid{@Det\ \ \ \ \ \ \ \ \ List<@Det\ \ \ \ \ \ \ \ \ Integer>}
-    \item \codeid{@OrderNonDet List <@OrderNonDet Set<...>\relax >}
-    \item \codeid{@OrderNonDet List <@Det String>}
+    \item \codeid{NonDet\ \ \ \ \ \ List<NonDet\ \ \ \ \ \ Integer>}
+    \item \codeid{Det\ \ \ \ \ \ \ \ \ List<Det\ \ \ \ \ \ \ \ \ Integer>}
+    \item \codeid{OrderNonDet List <OrderNonDet Set<...>\relax >}
+    \item \codeid{OrderNonDet List <Det String>}
 \end{itemize}
@@
 \begin{itemize}
-    \item \codeid{@OrderNonDet Set <@NonDet\ \ \ \ \ \ Integer>}
-    \item \codeid{@NonDet\ \ \ \ \ \ List<@Det\ \ \ \ \ \ \ \ \ Integer>}
-    \item \codeid{@Det\ \ \ \ \ \ \ \ \ List<@OrderNonDet Set<...>\relax >}
+    \item \codeid{OrderNonDet Set <NonDet\ \ \ \ \ \ Integer>}
+    \item \codeid{NonDet\ \ \ \ \ \ List<Det\ \ \ \ \ \ \ \ \ Integer>}
+    \item \codeid{Det\ \ \ \ \ \ \ \ \ List<OrderNonDet Set<...>\relax >}
 \end{itemize}
@@
 the type of the component type (see Figure~\ref{fig-determinism-collections}).
-As with collections, \codeid{@NonDet} arrays of \codeid{@Det} or \codeid{@OrderNonDet}
+As with collections, \codeid{NonDet} arrays of \codeid{Det} or \codeid{OrderNonDet}
 elements are not allowed (Rules $\sc{\mathit{ORDERNONDET \ ARRAY}}$ and $\sc{\mathit{NONDET \ ARRAY}}$ in Figure~\ref{type-validity}).
@@
 
-A collection (\<List, Set, or a Map>) annotated as \<@OrderNonDet> has the following properties:
+A collection (\<List, Set, or a Map>) annotated as \<OrderNonDet> has the following properties:
 \begin{enumerate}
@@
 \end{verbatim}
-Writing \codeid{@OrderNonDet TestUserCollection<@Det Integer>} is valid.\\
-Writing \codeid{@OrderNonDet TestUserCollection<@NonDet Integer>} is invalid
-because the type of the type parameter (\codeid{@NonDet}) is not a subtype
-of the user defined Collection type (\codeid{@OrderNonDet}).
+Writing \codeid{OrderNonDet TestUserCollection<Det Integer>} is valid.\\
+Writing \codeid{OrderNonDet TestUserCollection<NonDet Integer>} is invalid
+because the type of the type parameter (\codeid{NonDet}) is not a subtype
+of the user defined Collection type (\codeid{OrderNonDet}).
 
 User defined types that are not a subtype of Collection or Iterator
-may not be annotated as \codeid{@OrderNonDet}.
+may not be annotated as \codeid{OrderNonDet}.
 
@@
 
-Recall that the determinism checker does not permit \<@OrderNonDet> on a type that isn't a collection, iterator, or an array.
+Recall that the determinism checker does not permit \<OrderNonDet> on a type that isn't a \<Collection>, \<Iterator>, or an array.
 As a consequence, resolution of polymorphic type annotations at method signatures could have undesirable effects.
-Suppose a method declaration is annotated as \codeid{@PolyDet int size(@PolyDet List<T> list)}. Invoking this method
-as \codeid{size(@OrderNonDet List<@Det String>)} will result in \<@PolyDet> getting resolved to \<@OrderNonDet> which is invalid
-on the return type \<int>. We circumvent this scenario by introducing two variants to the \<@PolyDet> annotation, \<@PolyDet("up")>
-and \<@PolyDet("down")>.
+Suppose a method declaration is annotated as \codeid{PolyDet int size(PolyDet List<T> list)}. Invoking this method
+with an argument of type \codeid{OrderNonDet List<Det String>} will result in \<PolyDet> getting resolved to \<OrderNonDet> which is invalid
+on the return type \<int>. We circumvent this scenario by introducing two variants to the \<PolyDet> annotation, \<PolyDet$\uparrow$>
+and \<PolyDet$\downarrow$>.
 
@@
 \begin{itemize}
-    \item \<@PolyDet("up")>: Same as \<@PolyDet> except when it resolves to \<@OrderNonDet>, it is replaced with \<@NonDet>.
-    \item \<@PolyDet("down")>: Same as \<@PolyDet> except when it resolves to \<@OrderNonDet>, it is replaced with \<@Det>.
+    \item \<PolyDet$\uparrow$>: Same as \<PolyDet> except when it resolves to \<OrderNonDet>, it is replaced with \<NonDet>.
+    \item \<PolyDet$\downarrow$>: Same as \<PolyDet> except when it resolves to \<OrderNonDet>, it is replaced with \<Det>.
 \end{itemize} 
-The modifiers "up" and "down" are indicative of the subtyping relationship of \<@OrderNonDet> with \<@NonDet> and \<@Det> respectively
+The modifiers $\uparrow$ and $\downarrow$ are indicative of the subtyping relationship of \<OrderNonDet> with \<NonDet> and \<Det> respectively
 in the determinism type hierarchy.
-Annotating \<size()> as \codeid{@PolyDet("down") int size(@PolyDet List<T> list)} will only allow the following type resolutions:
+
+Figure~\ref{fig-poly-resolution} presents the rule for type checking of methods annotated with \<PolyDet> or its variants.
+
+Annotating \<size()> as \codeid{PolyDet$\downarrow$ int size(PolyDet List<T> list)} will only allow the following type resolutions:
 \begin{itemize}
-    \item \codeid{@Det int size(@Det List<T> list)}
-    \item \codeid{@Det int size(@OrderNonDet List<T> list)}
-    \item \codeid{@NonDet int size(@NonDet List<T> list)}
+    \item \codeid{Det int size(Det List<T> list)}
+    \item \codeid{Det int size(OrderNonDet List<T> list)}
+    \item \codeid{NonDet int size(NonDet List<T> list)}
 \end{itemize}
@@
     $\infer[\sc{\mathit{POLY RESOLUTION}}]{\pi_r \ \beta_r method(\overline{\pi_i \ \beta_i})\{\ \}}{\forall K \in \{\mathit{NonDet, OrderNonDet, Det\}} : K \ \beta_r method(\overline{K \ \beta_i})\{\ \} & \pi_r, \pi_i \in \{\mathit{PolyDet}, \mathit{PolyDet}\uparrow, \mathit{PolyDet}\downarrow    \}}$
+    \caption{Polymorphic resolution rule}
+    \label{fig-poly-resolution}
 \end{figure}
@@
 In order to provide soundness guarantees, we must ensure that a collection that is
-declared as \<@Det> is not side-effected in non-deterministic ways. Consider the following annotation 
-on the List add method: \codeid{public void add(@PolyDet ArrayList<E> this, @PolyDet int index, E element)}.
+declared as \<Det> is not side-effected in non-deterministic ways. Consider the following annotation 
+on the List add method: \codeid{public void add(PolyDet ArrayList<E> this, PolyDet int index, E element)}.
 Suppose a client calls this method as follows:
@@
 \end{verbatim}
-This will result in the instantion \codeid{public void add(@NonDet ArrayList<E> this, @NonDet int index, E element)} and
-the method invocation will type check. This is problematic because even though \<list> was declared to be \<@Det>,
-the type checker allows the addition of an element at a \<@NonDet> location violating our determinism guarantees.
-We prevent this behavior by introducing another valiant of \<@PolyDet>, \<@PolyDet("use")> that does not affect
+This will result in the instantion \codeid{public void add(NonDet ArrayList<E> this, NonDet int index, E element)} and
+the method invocation will type check. This is problematic because even though \<list> was declared to be \<Det>,
+the type checker allows the addition of an element at a \<NonDet> location violating our determinism guarantees.
+We prevent this behavior by introducing another valiant of \<PolyDet>, \<PolyDet("use")> that does not affect
 type instantiations. This is especially useful in preventing methods from non-deterministically modifying the state
 of a deterministic receiver. To have this effect, the library method \<ArrayList.add> is annotated as:
-\codeid{public void add(@PolyDet ArrayList<E> this, @PolyDet("use") int index, E element)}.
-Since \<@PolyDet("use")> has no effect on type instantiations, calling this method on a \<@Det List>
-will result in the method declaration being resolved as \codeid{public void add(@Det ArrayList<E> this, @Det int index, E element)}
+\codeid{public void add(PolyDet ArrayList<E> this, PolyDet("use") int index, E element)}.
+Since \<PolyDet("use")> has no effect on type instantiations, calling this method on a \<Det List>
+will result in the method declaration being resolved as \codeid{public void add(Det ArrayList<E> this, Det int index, E element)}
 irrespective of the type of \<index>, thereby preventing undesired side-effects.
@@
 Every type system in the checker framework must define a default type annotation. This annotation will be applied
-to every location that isn't explicitly annotated. In our determinism checker, we chose \<@PolyDet> as the default for all 
+to every location that isn't explicitly annotated. In our determinism checker, we chose \<PolyDet> as the default for all 
 method parameters and return types including constructors.
-The rationale for this design choice was to not make our type checker too restrictive by making everything \<@Det>.  This implies that
+The rationale for this design choice was to not make our type checker too restrictive by making everything \<Det>.  This implies that
 the checker still allows some non-determinism as long as it doesn't break any determinism specifications explicitly written by the programmer.
@@
 \end{verbatim}
-In the example above, even though the local variable \codeid{y} was declared to be \codeid{@NonDet}, it gets
-type refined to \codeid{@Det} after the first assignment statement. As the result, the assignment of \codeid{y}
-to a \codeid{@Det} variable \codeid{z} is sound. In the following example,
+In the example above, even though the local variable \codeid{y} was declared to be \codeid{NonDet}, it gets
+type refined to \codeid{Det} after the first assignment statement. As the result, the assignment of \codeid{y}
+to a \codeid{Det} variable \codeid{z} is sound. In the following example,
 \begin{verbatim}
@@
 
-With the above rules, the checker framework will automatically annotate a local collection with type \<@NonDet> and its
-type arguments will get the type qualifier \<@Det>. For example, a locally declared \codeid{List} of \codeid{Strings} will be automatically annotated
-as \codeid{@NonDet List<@Det String>}. Recall from~\ref{collection-rules} that the determinism checker does not allow this type on
+With the above rules, the checker framework will automatically annotate a local collection with type \<NonDet> and its
+type arguments will get the type qualifier \<Det>. For example, a locally declared \codeid{List} of \codeid{Strings} will be automatically annotated
+as \codeid{NonDet List<Det String>}. Recall from~\ref{collection-rules} that the determinism checker does not allow this type on
 any collection. While this is still sound, it will result in the determinism checker reporting a lot of errors because every locally declared 
 collection gets annotated with this invalid type. To avoid this situation, we make an exception to the defaulting rules for local collections
-and annotate their type parameters also with \<@NonDet>.
-So a local list of Strings gets the default annotation of \codeid{@NonDet List<@NonDet String>}.
+and annotate their type parameters also with \<NonDet>.
+So a local list of Strings gets the default annotation of \codeid{NonDet List<NonDet String>}.
 While this defaulting eliminates the possibility of automatically annotating locals collections with an invalid
@@
 the checker would report an error at the assignment statement since the inferred type for \codeid{localList} is 
-\codeid{@NonDet List<@NonDet> String} and since collections types are invariant.
+\codeid{NonDet List<NonDet> String} and since collections types are invariant.
 We therefore recommend that programmers explicitly annotate
@@
 \end{figure}
-The checker type refines a receiver of type \codeid{@OrderNonDet List} to a \codeid{@Det List} if all the
-type argument of this \codeid{List} have the type annotation \codeid{@Det} (Rule LIST SORT in Figure~\ref{fig-sorting}). Similar rules apply for \codeid{Arrays.sort()} and \codeid{Collections.sort()}.
-As per these rules, if \codeid{sort()} is called on a receiver of type \codeid{@OrderNonDet List<@Det Set<@Det String>>}, it will
-be type refined to \codeid{@Det List<@Det Set<@Det String>>}. However a receiver of type \codeid{@OrderNonDet List<@OrderNonDet Set<@Det String>>} will not be type refined.
+The checker type refines a receiver of type \codeid{OrderNonDet List} to a \codeid{Det List} if all the
+type argument of this \codeid{List} have the type annotation \codeid{Det} (Rule LIST SORT in Figure~\ref{fig-sorting}). Similar rules apply for \codeid{Arrays.sort()} and \codeid{Collections.sort()}.
+As per these rules, if \codeid{sort()} is called on a receiver of type \codeid{OrderNonDet List<Det Set<Det String>>}, it will
+be type refined to \codeid{Det List<Det Set<Det String>>}. However a receiver of type \codeid{OrderNonDet List<OrderNonDet Set<Det String>>} will not be type refined.
 

### 2019-03-26 design.tex
polymorphic extensions

@@
 Figure~\ref{fig-poly-resolution} presents the rule for type checking of methods annotated with \<PolyDet> or its variants.
+A polymorphic method type checks if every instantiation of \<PolyDet> type checks. If this method has \<PolyDet>$\uparrow$ or \<PolyDet>$\downarrow$
+in its signature then the instantiation of \<PolyDet> at that location gets replaced by the appropriate type as shown in Table~\ref{tab-poly-resolutions}.
 
@@
 the type checker allows the addition of an element at a \<NonDet> location violating our determinism guarantees.
-We prevent this behavior by introducing another valiant of \<PolyDet>, \<PolyDet("use")> that does not affect
+We prevent this behavior by introducing another variant of \<PolyDet>, \<PolyDet("use")> that does not affect
 type instantiations. This is especially useful in preventing methods from non-deterministically modifying the state

### 2019-03-26 design.tex
maps and sets discussion

@@
 \subsection{Maps discussion}\label{maps}
+A \<Map> is a set of mappings from keys to values. A \<Map> is deterministic if its \<entrySet> is deterministic.
+In other words, iterating over this \<entrySet> is guaranteed to produce the same values in the exact same order across executions.
+The \<Map> interface in Java doesn't define an \<iterator()> interface. Therefore it's sound to equate the determinism of the \<Set>
+returned by \<entrySet()> to the determinism of the \<Map>. Interestingly, we found that every implementation of the \<Map> 
+interface had the following property: if the \<Set> returned by \<entrySet()> has a determinism type $\kappa$, then
+the \<Set>s returned by \<keySet()> and \<values()> also have the exact same determinism type $\kappa$.
+
+We noticed the following properties with respect to determinism regarding the most widely used \<Map> implementations:
+\begin{itemize}
+    \item \<HashMap> can never guarantee fully deterministic iteration over its entries. A \codeid{Det HashMap} does not exist.
+    \item \<LinkedHashMap> can have any of \<Det>, \<OrderNonDet>, or \<NonDet> types. Iterating over a \<LinkedHashMap> returns
+    its entries in the order of their insertion. An \<OrderNonDet LinkedHashMap> can be created by calling its constructor with a \<HashMap>.
+    \item \<TreeMap> can either be \<Det> or \<NonDet>. An \<OrderNonDet TreeMap> doesn't exist because the entries always ordered according
+    to the natural ordering of the keys. 
+\end{itemize}
+
+We prohibit the creation of a \codeid{Det HashMap} or an \codeid{OrderNonDet TreeMap} in our determinism checker.
+Figure~\ref{fig-creation-rules} formalizes these rules. Similar constraints apply to \<HashSet>s and \<TreeSet>s which are
+also expressed by the last two rules of Figure~\ref{fig-creation-rules}.
+
+\begin{figure}
+    $\infer[\sc{\mathit{HASHMAP \ CREATION}}]{\mathit{new \ \kappa\ HashMap()}}{\kappa = \mathit{OrderDet/NonDet}}$
+
+    \bigskip
+
+    $\infer[\sc{\mathit{TREEMAP \ CREATION}}]{\mathit{new \ \kappa\ TreeMap()}}{\kappa = \mathit{Det/NonDet}}$
+
+    \bigskip
+
+    $\infer[\sc{\mathit{HASHSET \ CREATION}}]{\mathit{new \ \kappa\ HashSet()}}{\kappa = \mathit{OrderDet/NonDet}}$
+
+    \bigskip
+
+    $\infer[\sc{\mathit{TREESET \ CREATION}}]{\mathit{new \ \kappa\ TreeSet()}}{\kappa = \mathit{Det/NonDet}}$
+    \caption{\<Map> and \<Set> creation rules}
+    \label{fig-creation-rules}
+\end{figure}
+
+In the JDK, \<LinkedHashSet> is implemented as a subclass of \<HashMap>.
+\todo{Talk about this design choice. Josh Blochs book: "If the code surrounding the first declaration depended on LinkedHashSet's ordering policy,
+    then it would be incorrect to substitute HashSet for LinkedHashSet in the declaration, because HashSet makes no guarantee concerning iteration order."}
 
 \begin{itemize}
-    \item A map is deterministic iff the \<entrySet()> is
-          deterministic; equivalently, iff the \<keySet()> and \<values()>
-          are both deterministic
-    \item the iteration order of a keyset is the same as the iteration order of entrySet and Values.
-    \item \<HashMap> - can be \<@OrderNonDet> or \<@NonDet>.
-    \item \<LinkedHashMap> - can be \<@Det>, \<@OrderNonDet>, ot \<@NonDet>.
-    \item \<TreeMap> - can be \<@Det> or \<@NonDet>.
     \item Problem with \<Map.get()> : Annotating as \<@PolyDet> is imprecise.
     \item We circumvent this by hardcoding rules to make it precise. (Describe the rule here)
-    \item It was probably a bad design choice to make HashMap a subclass of Map (talk about how
-    every method is overriden). Similarly for HashSet.
-\end{itemize}
-
-\begin{itemize}
-    \item Exception to this defaulting is the construction of HashSets and HashMaps because they can't be \<@Det>.
-    We explicitly annotate the return type on constructors as
-    \<@OrderNonDet> and disallow constructing \<@Det HashSet> or \<@Det
-    HashMap>.
 \end{itemize}

### 2019-03-27 design.tex
moving text around
precision

@@
 
+\subsection{Determinism checker specific rules for precision}\label{precision}
+We annotated the \codieid{equals()} method of the \<Set> class in the JDK as
+\begin{verbatim}
+PolyDet$\uparrow$ boolean equals(PolyDet Set<E> this, PolyDet Object o)
+\end{verbatim}
+While this is sound, it is imprecise in the case of \codeid{OrderNonDet Sets}. We added the following rule shown in Figure~\ref{fig-precision-rules} to our checker to improve precision:
+
+\begin{figure}
+    \todo{linebreak}
+    $\infer{\Gamma \vdash \mathit{z : Det}}{\Gamma \vdash \mathit{x} : \mathit{OrderNonDet} \ \mathit{Set}\langle \kappa_{x} \ \beta_{x}\rangle, \mathit{y} : \mathit{OrderNonDet} \ \mathit{Set}\langle \kappa_{y} \ \beta_{y}\rangle & \mathit{z = x.equals(y)} & \langle \kappa_{x} \ \beta_{x}\rangle \nsqsubseteq \langle \mathit{OrderNonDet \ List} \rangle & \langle \kappa_{y} \ \beta_{y}\rangle \nsqsubseteq \langle \mathit{OrderNonDet \ List} \rangle }$
+
+    \caption{Improving precision}
+    \label{fig-precision-rules}
+\end{figure}
+According to this rule, if the \<Set> receiver has type \codeid{@OrderNonDet Set<@Det List} and the argument to the \codeid{equals()} method is
+also of the same type, the return type gets the qualifier \<Det>. Without this precision rule, the return type would have resolved to \<NonDet> which
+is sound but imprecise. This rule is careful enough to not annotate the return type of \<equals()> as \<Det> if a type argument to either the
+receiver or the argument happens to be an \codeid{OrderNonDet List} (An order nondeterminstic list does not provide any guarantee on the
+equality as it is dependent on the iteration order).
+
 \begin{itemize}
@@
     \item We circumvent this by hardcoding rules to make it precise. (Describe the rule here)
-\end{itemize}
 
-\subsection{Determinism checker specific rules for precision}\label{precision}
-
-\begin{itemize}
-    \item \<Set.equals()> is annotated as \<@Poly("up")>. This is sound but imprecise.
-    \item Example1: \codeid{@OrderNonDet Set<@Det List>.equals(@OrderNonDet Set<@Det List>)} is deterministic
-    \item Example2: \codeid{@OrderNonDet Set<@OrderNonDet List>.equals(@OrderNonDet Set<@OrderNonDet List>)} is not deterministic.
-    \item Explain the rules under which we make Set equality deterministic.
     \item We annotate System.getProperty as \<@NonDet> because the results could vary across machines.
-    \item Exceptions: line.Sepaator, file.Separator. and path.Separator.
+    \item Exceptions: line.Separator, file.Separator. and path.Separator.
     \item discuss type on Exceptions and unsoundness due to this.

### 2019-03-27 design.tex
precision rules

@@
 \end{verbatim}
-While this is sound, it is imprecise in the case of \codeid{OrderNonDet Sets}. We added the following rule shown in Figure~\ref{fig-precision-rules} to our checker to improve precision:
-
+While this is sound, it is imprecise in the case of \codeid{OrderNonDet Sets}. We added the following rule shown in Figure~\ref{fig-precision-rules} to our checker to improve precision.
 \begin{figure}
     \todo{linebreak}
-    $\infer{\Gamma \vdash \mathit{z : Det}}{\Gamma \vdash \mathit{x} : \mathit{OrderNonDet} \ \mathit{Set}\langle \kappa_{x} \ \beta_{x}\rangle, \mathit{y} : \mathit{OrderNonDet} \ \mathit{Set}\langle \kappa_{y} \ \beta_{y}\rangle & \mathit{z = x.equals(y)} & \langle \kappa_{x} \ \beta_{x}\rangle \nsqsubseteq \langle \mathit{OrderNonDet \ List} \rangle & \langle \kappa_{y} \ \beta_{y}\rangle \nsqsubseteq \langle \mathit{OrderNonDet \ List} \rangle }$
+    $\infer[\sc{\mathit{SET \ PRECISION}}]{\Gamma \vdash \mathit{z : Det}}{\Gamma \vdash \mathit{x} : \mathit{OrderNonDet} \ \mathit{Set}\langle \kappa_{x} \ \beta_{x}\rangle, \mathit{y} : \mathit{OrderNonDet} \ \mathit{Set}\langle \kappa_{y} \ \beta_{y}\rangle & \mathit{z = x.equals(y)} & \langle \kappa_{x} \ \beta_{x}\rangle \nsqsubseteq \langle \mathit{OrderNonDet \ List} \rangle & \langle \kappa_{y} \ \beta_{y}\rangle \nsqsubseteq \langle \mathit{OrderNonDet \ List} \rangle }$
+
+    \bigskip
+
+    $\infer[\sc{\mathit{MAP \ PRECISION}}]{\Gamma \vdash \mathit{z : Det}}{\Gamma \vdash \mathit{x : OrderNonDet \ Map\langle Det \ \beta_k, \kappa_v\ \beta_v \rangle, y : Det \ \beta} & \mathit{z = x.get(y)}}$
 
@@
 
-\begin{itemize}
-    \item Problem with \<Map.get()> : Annotating as \<@PolyDet> is imprecise.
-    \item We circumvent this by hardcoding rules to make it precise. (Describe the rule here)
+Similarly, the \codeid{get()} method in the \<Map> class is annotated with polymorphic annotations and the rule $\sc{\mathit{MAP \ PRECISION}}$ in Figure~\ref{fig-precision-rules} improves precision for \<OrderNonDet> receivers.
 
-    \item We annotate System.getProperty as \<@NonDet> because the results could vary across machines.
-    \item Exceptions: line.Separator, file.Separator. and path.Separator.
-    \item discuss type on Exceptions and unsoundness due to this.
-\end{itemize}
+\codeid{System.getProperty()} returns a \<NonDet> type since we can't provide any determinism guarantees for
+executions across machines. However, we make exceptions when the argument to this method is one of 
+\codeid{"line.separator", "path.separator", or "file.separator"} as they return OS dependent values which are guaranteed to be deterministic on that OS.
 

### 2019-03-27 design.tex
Fix LaTeX error

@@
 \subsection{Determinism checker specific rules for precision}\label{precision}
-We annotated the \codieid{equals()} method of the \<Set> class in the JDK as
+We annotated the \codeid{equals()} method of the \<Set> class in the JDK as
 \begin{verbatim}

### 2019-03-27 design.tex
Edits in section 2.1

@@
 
-A \textit{type} serves as an abstraction for the set of acceptable values for any expression.
-A \textit{type qualifier} on a \textit{type} adds additional constraints on the values represented by that \textit{type}.
-A compile time type checker can verify the correctness of this \textit{type} with its dataflow analysis and subtyping rules.
-In Java, a type qualifier is written as an annotation. The type \<Integer> represents positive and negative integers besides zero.
-A program analysis developer could design a type system that differentiates positive integers from the negative ones with type qualifiers
-\<@Positive> and \<@Negative>.  The type \<@Positive Integer> now represents only positive values and the dataflow analysis ensures that
-any negative integer value flowing into this type is reported as an error.
+In a type system,
+a \textit{type} stands for a set of values.  It abstracts or restricts the
+set of possbile run-time values that an expression may evaluate to.
+A programming language provides \emph{basetypes}, such as \<int>.
+A \textit{type qualifier} on a \textit{type} adds additional constraints;
+that is, it reduces the size of the set of values.
+A type checker can verify the correctness of this \textit{type} with its dataflow analysis and subtyping rules.
+In Java, a type qualifier is written as a type annotation with a leading \<@>, as in
+\<@Positive int>.
 
 \begin{figure}
-    $\infer[\sc{\mathit{SUBTYPING}}]{\Gamma \vdash x : \tau_2}{\Gamma \vdash x : \tau_1 & \tau_1 \sqsubseteq \tau_2}$
+    \bigskip
+
+    $\infer{\tau_1 \sqsubseteq \tau_3}{\tau_1 \sqsubseteq \tau_2 \quad \tau_2 \sqsubseteq \tau_3}
+    \quad\quad
+    \infer[\textsc{subtyping}]{\Gamma \vdash x : \tau_2}{\Gamma \vdash x : \tau_1 & \tau_1 \sqsubseteq \tau_2}$
 
@@
 
-    $\infer[\sc{\mathit{METHOD \ CALLS}}]{\mathit{method}(\overline{a_i}) : \tau_r}{\tau_r \  \mathit{method}(\overline{\tau_i \  p_i}) & \overline{a_i : \tau_i}}$
+    $\infer{\kappa_1 \beta \sqsubseteq \kappa_2 \beta}{\kappa_1 \sqsubseteq \kappa_2}
+    \quad\quad
+    \infer{\kappa \beta_1 \sqsubseteq \kappa \beta_2}{\beta_1 \sqsubseteq \beta_2}$
+
+    \bigskip
+
+    $\infer{\Det \sqsubseteq \OrderNonDet}{}
+    \quad\quad
+    \infer{\OrderNonDet \sqsubseteq \NonDet}{}$
+
+    \bigskip
+
+    $\infer[\textsc{method call}]{\mathit{method}(\overline{a_i}) : \tau_r}{\tau_r \  \mathit{method}(\overline{\tau_i \  p_i}) & \overline{a_i : \tau_i}}$
 
@@
 \subsection{Determinism type hierarchy}\label{type-hierarchy}
+
+\begin{wrapfigure}{R}{0.3\textwidth}
+    \begin{center}
+        \includegraphics[scale=0.5]{determinism}
+    \end{center}
+    \caption{Determinism type qualifier hierarchy}
+    \label{fig-determinism-hierarchy}
+\todo{Drop the at-signs (@).  The text is weirdly small for the box.
+  Probably just redraw this.}
+\end{wrapfigure}
+
 The core of the determinism type system is the following type qualifiers (see Figure~\ref{fig-determinism-hierarchy}):
@@
 
-In this type system, \<Det> is a subtype of \<NonDet> denoted as \codeid{Det $\sqsubseteq$ NonDet}.
-
-\begin{figure}
-    \begin{center}
-        \includegraphics[scale=0.5]{determinism}
-    \end{center}
-    \caption{Determinism type hierarchy}
-    \label{fig-determinism-hierarchy}
-\end{figure}
+\subsection{Handling order nondeterminism}
+In addition to the type qualifiers described in~\ref{type-hierarchy}, the determinism checker 
+defines the type qualifier \<OrderNonDet>. This qualifier represents
+a collection or an array that will have the same elements in every execution, but in a
+possibly different order.  \<OrderNonDet> may only be written on
+collections and arrays.
+A collection is any subtype of \<java.util.Collection>,
+\<java.util.Iterator>, or \<java.util.Map>, including user-defined types. The relationship
+of \<OrderNonDet> with respect the the other two type qualifiers is \codeid{Det $\sqsubseteq$ OrderNonDet $\sqsubseteq$ NonDet} as shown in Figure~\ref{fig-determinism-hierarchy}.
 
@@
 
-\subsection{Handling order nondeterminism}
-In addition to the type qualifiers described in~\ref{type-hierarchy}, the determinism checker 
-defines the type qualifier \<OrderNonDet>. This qualifier represents
-a collection or an array that will have the same elements in every execution, but in a
-possibly different order.  \<OrderNonDet> may only be written on
-collections and arrays.
-A collection is any subtype of \<java.util.Collection>,
-\<java.util.Iterator>, or \<java.util.Map>, including user-defined types. The relationship
-of \<OrderNonDet> with respect the the other two type qualifiers is \codeid{Det $\sqsubseteq$ OrderNonDet $\sqsubseteq$ NonDet} as shown in Figure~\ref{fig-determinism-hierarchy}.
-
 \subsection{Additional typing rules for collection elements}\label{collection-rules}
 
-%\todo{It is very important to express the whole section as formal type
-%  jugdments, or we will get laughed
-%  out of the room as amateurs, during the program committee meeting.  Also,
-%  the formalism will help make the ideas precise.
-%
-%  Type well-formedness rules:\todo{format nicely in \LaTeX}
-%
-%  antecedent: $\Gamma : \kappa_e <: \kappa_c$
-%
-%  consequent: $\Gamma : \kappa_c \ \codeid{List}\langle\kappa_e\ 
-%  \codeid{Object}\rangle$
-%  }
 \begin{figure}
-    $\infer[\sc{\mathit{ANY\ TYPE}}]{\vdash : \kappa \  \beta}{\kappa : \mathit{Det / NonDet}}$
+    $\infer[\textsc{any type}]{\wellformed{\kappa \  \beta}}{\kappa : \mathit{Det / NonDet}}$
 
@@
 
-    $\infer[\sc{\mathit{ORDERNONDET \ COLLECTION}}]{\vdash : \kappa_c \  \beta_c \langle\kappa_e \ \beta_e\rangle}{\kappa_c : \mathit{OrderNonDet/Det} & \kappa_e \sqsubseteq \kappa_c & \beta_c \sqsubseteq \mathit{Collection/Iterator}}$
+    $\infer[\textsc{ordernondet collection}]{\wellformed{\kappa_c \  \beta_c \langle\kappa_e \ \beta_e\rangle}}{\kappa_c : \mathit{OrderNonDet/Det} & \kappa_e \sqsubseteq \kappa_c & \beta_c \sqsubseteq \mathit{Collection/Iterator}}$
 
@@
 
-    $\infer[\sc{\mathit{NONDET \ COLLECTION}}]{\vdash : \mathit{NonDet} \ \beta_c\langle\mathit{NonDet} \ \beta_e\rangle}{\beta_c \sqsubseteq \mathit{Collection/Iterator}}$
+    $\infer[\textsc{nondet collection}]{\wellformed{\mathit{NonDet} \ \beta_c\langle\mathit{NonDet} \ \beta_e\rangle}}{\beta_c \sqsubseteq \mathit{Collection/Iterator}}$
 
@@
 
-     $\infer[\sc{\mathit{ORDERNONDET \ ARRAY}}]{\vdash : \kappa_e \ \beta_e \ \  \kappa_a []}{\kappa_a : \mathit{OrderNonDet/Det} & \kappa_e \sqsubseteq \kappa_a }$
+     $\infer[\textsc{ordernondet array}]{\wellformed{\kappa_e \ \beta_e \ \  \kappa_a []}}{\kappa_a : \mathit{OrderNonDet/Det} & \kappa_e \sqsubseteq \kappa_a }$
 
@@
 
-    $\infer[\sc{\mathit{NONDET \ ARRAY}}]{\vdash : \mathit{NonDet} \ \beta_e \ \  \mathit{NonDet} []}{}$
+    $\infer[\textsc{nondet array}]{\wellformed{\mathit{NonDet} \ \beta_e \ \  \mathit{NonDet} []}}{}$
 
@@
 
-    $\infer[\sc{\mathit{ORDERNONDET \ MAP}}]{\vdash : \kappa_m \  \beta_m \langle\kappa_k \ \beta_k, \kappa_v \ \beta_v\rangle}{\kappa_m: \mathit{OrderNonDet/Det} & \kappa_k \sqsubseteq \kappa_m & \kappa_v \sqsubseteq \kappa_m & \beta_m \sqsubseteq \mathit{Map}}$
+    $\infer[\textsc{ordernondet map}]{\wellformed{\kappa_m \  \beta_m \langle\kappa_k \ \beta_k, \kappa_v \ \beta_v\rangle}}{\kappa_m: \mathit{OrderNonDet/Det} & \kappa_k \sqsubseteq \kappa_m & \kappa_v \sqsubseteq \kappa_m & \beta_m \sqsubseteq \mathit{Map}}$
 
@@
 
-    $\infer[\sc{\mathit{NONDET \ MAP}}]{\vdash : \mathit{NonDet} \ \beta_m\langle\mathit{NonDet} \ \beta_k, \mathit{NonDet} \ \beta_v\rangle}{\beta_m \sqsubseteq \mathit{Map}}$
+    $\infer[\textsc{nondet map}]{\wellformed{\mathit{NonDet} \ \beta_m\langle\mathit{NonDet} \ \beta_k, \mathit{NonDet} \ \beta_v\rangle}}{\beta_m \sqsubseteq \mathit{Map}}$
 
@@
 where $\kappa$ is a type qualifier from the determinism type hierarchy and $\beta$ is a Java type.
-As stated in the rule $\sc{\mathit{ANY\ TYPE}}$, the types \<Det> and \<NonDet> can be written on any Java type.
+As stated in the rule $\textsc{any type}$, the types \<Det> and \<NonDet> can be written on any Java type.
 The (determinism) type of a Collection or Iterator must be a supertype or equal to
-the type of the type parameter (Rule $\sc{\mathit{ORDERNONDET \ COLLECTION}}$) (see Figure~\ref{fig-determinism-collections}).
+the type of the type parameter (Rule $\textsc{ordernondet collection}$) (see Figure~\ref{fig-determinism-collections}).
 If the Collection is \<NonDet>, then the type parameter may not be
-\<Det> or \<OrderNonDet> (Rule $\sc{\mathit{NONDET \ COLLECTION}}$ in Figure~\ref{type-validity}). Although such types could exist, they are
+\<Det> or \<OrderNonDet> (Rule $\textsc{nondet collection}$ in Figure~\ref{type-validity}). Although such types could exist, they are
 disallowed to prevent the following bug:
@@
 As with collections, \codeid{NonDet} arrays of \codeid{Det} or \codeid{OrderNonDet}
-elements are not allowed (Rules $\sc{\mathit{ORDERNONDET \ ARRAY}}$ and $\sc{\mathit{NONDET \ ARRAY}}$ in Figure~\ref{type-validity}).
+elements are not allowed (Rules $\textsc{ordernondet array}$ and $\textsc{nondet array}$ in Figure~\ref{type-validity}).
 
@@
 %Valid locations for \codeid{@PolyDet("up")} and  \codeid{@PolyDet("down")}:
-%$\infer{\vdash : \kappa \  \beta}{isParam(\beta) / isReturn(\beta) & \kappa : @PolyDet("up") / @PolyDet("down")}$
+%$\infer{\wellformed{\kappa \  \beta}}{isParam(\beta) / isReturn(\beta) & \kappa : @PolyDet("up") / @PolyDet("down")}$
 %
@@
 \begin{figure}
-    $\infer[\sc{\mathit{POLY RESOLUTION}}]{\pi_r \ \beta_r method(\overline{\pi_i \ \beta_i})\{\ \}}{\forall K \in \{\mathit{NonDet, OrderNonDet, Det\}} : K \ \beta_r method(\overline{K \ \beta_i})\{\ \} & \pi_r, \pi_i \in \{\mathit{PolyDet}, \mathit{PolyDet}\uparrow, \mathit{PolyDet}\downarrow    \}}$
+    $\infer[\textsc{poly resolution}]{\pi_r \ \beta_r method(\overline{\pi_i \ \beta_i})\{\ \}}{\forall K \in \{\mathit{NonDet, OrderNonDet, Det\}} : K \ \beta_r method(\overline{K \ \beta_i})\{\ \} & \pi_r, \pi_i \in \{\mathit{PolyDet}, \mathit{PolyDet}\uparrow, \mathit{PolyDet}\downarrow    \}}$
     \caption{Polymorphic resolution rule}
@@
 \begin{figure}
-    $\infer[\sc{\mathit{LIST \  SORT}}]{\Gamma \vdash x : \mathit{Det \ List}\langle\mathit{Det} \ \beta\rangle}{\Gamma \vdash x : \mathit{OrderNonDet \ List}\langle\mathit{Det} \ \beta\rangle & \mathit{x.sort()}}$
+    $\infer[\textsc{list sort}]{\Gamma \vdash x : \mathit{Det \ List}\langle\mathit{Det} \ \beta\rangle}{\Gamma \vdash x : \mathit{OrderNonDet \ List}\langle\mathit{Det} \ \beta\rangle & \mathit{x.sort()}}$
 
@@
 
-    $\infer[\sc{\mathit{COLLECTIONS \  SORT}}]{\Gamma \vdash x : \mathit{Det \ List}\langle\mathit{Det} \ \beta\rangle}{\Gamma \vdash x : \mathit{OrderNonDet \ List}\langle\mathit{Det} \ \beta\rangle & \mathit{Collections.sort.sort(x)}}$
+    $\infer[\textsc{collections sort}]{\Gamma \vdash x : \mathit{Det \ List}\langle\mathit{Det} \ \beta\rangle}{\Gamma \vdash x : \mathit{OrderNonDet \ List}\langle\mathit{Det} \ \beta\rangle & \mathit{Collections.sort.sort(x)}}$
 
@@
 
-    $\infer[\sc{\mathit{ARRAYS \  SORT}}]{\Gamma \vdash x : \mathit{Det \ \beta \ Det[]} }{\Gamma \vdash x : \mathit{Det \ \beta \ OrderNonDet[]} & \mathit{Arrays.sort( \ x)}}$
+    $\infer[\textsc{arrays sort}]{\Gamma \vdash x : \mathit{Det \ \beta \ Det[]} }{\Gamma \vdash x : \mathit{Det \ \beta \ OrderNonDet[]} & \mathit{Arrays.sort( \ x)}}$
 
@@
 \begin{figure}
-    $\infer[\sc{\mathit{HASHMAP \ CREATION}}]{\mathit{new \ \kappa\ HashMap()}}{\kappa = \mathit{OrderDet/NonDet}}$
+    $\infer[\textsc{hashmap creation}]{\mathit{new \ \kappa\ HashMap()}}{\kappa = \mathit{OrderDet/NonDet}}$
 
@@
 
-    $\infer[\sc{\mathit{TREEMAP \ CREATION}}]{\mathit{new \ \kappa\ TreeMap()}}{\kappa = \mathit{Det/NonDet}}$
+    $\infer[\textsc{treemap creation}]{\mathit{new \ \kappa\ TreeMap()}}{\kappa = \mathit{Det/NonDet}}$
 
@@
 
-    $\infer[\sc{\mathit{HASHSET \ CREATION}}]{\mathit{new \ \kappa\ HashSet()}}{\kappa = \mathit{OrderDet/NonDet}}$
+    $\infer[\textsc{hashset creation}]{\mathit{new \ \kappa\ HashSet()}}{\kappa = \mathit{OrderDet/NonDet}}$
 
@@
 
-    $\infer[\sc{\mathit{TREESET \ CREATION}}]{\mathit{new \ \kappa\ TreeSet()}}{\kappa = \mathit{Det/NonDet}}$
+    $\infer[\textsc{treeset creation}]{\mathit{new \ \kappa\ TreeSet()}}{\kappa = \mathit{Det/NonDet}}$
     \caption{\<Map> and \<Set> creation rules}
@@
     \todo{linebreak}
-    $\infer[\sc{\mathit{SET \ PRECISION}}]{\Gamma \vdash \mathit{z : Det}}{\Gamma \vdash \mathit{x} : \mathit{OrderNonDet} \ \mathit{Set}\langle \kappa_{x} \ \beta_{x}\rangle, \mathit{y} : \mathit{OrderNonDet} \ \mathit{Set}\langle \kappa_{y} \ \beta_{y}\rangle & \mathit{z = x.equals(y)} & \langle \kappa_{x} \ \beta_{x}\rangle \nsqsubseteq \langle \mathit{OrderNonDet \ List} \rangle & \langle \kappa_{y} \ \beta_{y}\rangle \nsqsubseteq \langle \mathit{OrderNonDet \ List} \rangle }$
+    $\infer[\textsc{set precision}]{\Gamma \vdash \mathit{z : Det}}{\Gamma \vdash \mathit{x} : \mathit{OrderNonDet} \ \mathit{Set}\langle \kappa_{x} \ \beta_{x}\rangle, \mathit{y} : \mathit{OrderNonDet} \ \mathit{Set}\langle \kappa_{y} \ \beta_{y}\rangle & \mathit{z = x.equals(y)} & \langle \kappa_{x} \ \beta_{x}\rangle \nsqsubseteq \langle \mathit{OrderNonDet \ List} \rangle & \langle \kappa_{y} \ \beta_{y}\rangle \nsqsubseteq \langle \mathit{OrderNonDet \ List} \rangle }$
 
@@
 
-    $\infer[\sc{\mathit{MAP \ PRECISION}}]{\Gamma \vdash \mathit{z : Det}}{\Gamma \vdash \mathit{x : OrderNonDet \ Map\langle Det \ \beta_k, \kappa_v\ \beta_v \rangle, y : Det \ \beta} & \mathit{z = x.get(y)}}$
+    $\infer[\textsc{map precision}]{\Gamma \vdash \mathit{z : Det}}{\Gamma \vdash \mathit{x : OrderNonDet \ Map\langle Det \ \beta_k, \kappa_v\ \beta_v \rangle, y : Det \ \beta} & \mathit{z = x.get(y)}}$
 

### 2019-03-27 experiments.tex
expts

@@
-P1 :
-[ERROR] /Users/rashmi/caseStudies/nondex/reflectasm/src/com/esotericsoftware/reflectasm/FieldAccess.java:[102,87] [assignment.type.incompatible] incompatible types in assignment.
-[ERROR]   found   : @Det Field @OrderNonDet []
-[ERROR]   required: @Det Field @Det []
-
-P2:
-
-		if (object instanceof Map) {
-    emitter.emit(new MappingStartEvent(anchor, tag, !showTag, false));
-    @Det Set<@Det Object> items = ((Map)object).entrySet();
-    @Det Iterator<@Det Object> itemIterator = items.iterator();
-    while (itemIterator.hasNext()) {
-        Object item = itemIterator.next();
-        Entry entry = (Entry)item;
-        writeValue(entry.getKey(), null, null, null);
-        if (isRoot && !config.writeConfig.writeRootElementTags) elementType = entry.getValue().getClass();
-        writeValue(entry.getValue(), elementType, null, null);
-    }
-    //			for (@Det Object item : ((Map)object).entrySet()) {
-        //				Entry entry = (Entry)item;
-        //				writeValue(entry.getKey(), null, null, null);
-        //				if (isRoot && !config.writeConfig.writeRootElementTags) elementType = entry.getValue().getClass();
-        //				writeValue(entry.getValue(), elementType, null, null);
-        //			}
-    emitter.emit(Event.MAPPING_END);
-    return;
-}
+\section{Evaluation}
+\subsection{NONDEX case study}
+\begin{table}[]
+    \begin{tabular}{|l|l|l|l|l|l|}
+        \hline
+        Benchmark & Recall? & Error message source & # annotations & Type of annotation added & Comments \\ \hline
+        P1 &  &  &  &  &  \\
+        P2 &  &  &  &  &  \\
+        P3 &  &  &  &  &  \\ 
+        P4 &  &  &  &  &  \\ 
+        P5 &  &  &  &  &  \\ 
+        P5 &  &  &  &  &  \\ 
+        P6 &  &  &  &  &  \\ 
+        P6 &  &  &  &  &  \\
+        P7 &  &  &  &  &  \\ 
+        P7 &  &  &  &  &  \\
+        P7 &  &  &  &  &  \\
+        P7 &  &  &  &  &  \\
+        P7 &  &  &  &  &  \\
+        P8 &  &  &  &  &  \\
+        P8 &  &  &  &  &  \\
+        P8 &  &  &  &  &  \\
+        P9 &  &  &  &  &  \\
+        P10 &  &  &  &  &  \\
+        P10 &  &  &  &  &  \\
+        P11 &  &  &  &  &  \\
+        P12 &  &  &  &  &  \\
+        P13 &  &  &  &  &  \\
+        P14 &  &  &  &  &  \\
+        P15 &  &  &  &  &  \\
+        P15 &  &  &  &  &  \\
+        P16 &  &  &  &  &  \\
+        P17 &  &  &  &  &  \\
+        P18 &  &  &  &  &  \\
+        P19 &  &  &  &  &  \\
+        P19 &  &  &  &  &  \\
+        P19 &  &  &  &  &  \\
+        P20 &  &  &  &  &  \\
+        P20 &  &  &  &  &  \\
+        P21 &  &  &  &  &  \\
+        P21 &  &  &  &  &  \\ \hline
+    \end{tabular}
+\end{table}

### 2019-03-27 design.tex
Edits in section 2.2

@@
 
-    $\infer[\textsc{method call}]{\mathit{method}(\overline{a_i}) : \tau_r}{\tau_r \  \mathit{method}(\overline{\tau_i \  p_i}) & \overline{a_i : \tau_i}}$
+    $\infer[\textsc{method call}]{\|method|(\overline{a_i}) : \tau_r}{\tau_r \  \|method|(\overline{\tau_i \  p_i}) & \overline{a_i : \tau_i}}$
 
@@
 
-\subsection{Determinism type hierarchy}\label{type-hierarchy}
+\subsection{Determinism types}\label{type-hierarchy}
 
@@
 \subsection{Handling order nondeterminism}
-In addition to the type qualifiers described in~\ref{type-hierarchy}, the determinism checker 
+In addition to the type qualifiers described in~\ref{type-hierarchy}, \theDeterminismChecker 
 defines the type qualifier \<OrderNonDet>. This qualifier represents
@@
     \end{verbatim}
-    \caption{Example: Errors detected by determinism checker}
+    \caption{Example: Errors detected by \theDeterminismChecker.}
+\todo{Use real examples of bugs from case studies, throughout.  Artificial
+  examples are less compelling (and lead the reader to wonder whether no
+  real example exist).  Put the first example of an error in a figure in
+  Section 1, to help with motivating the problem.}
     \label{code-determinism}
 \end{figure}
-
-\subsection{Polymorphism}\label{polymorphism}
-
-The underlying checker-framework on which we build our determinism checker has support for
-polymorphic annotations. When a user writes a polymorphic annotation on a method signature or a type parameter,
-it indicates that it could resolve to any type in the type system depending on how it is used.
-In the determinism checker, we define a polymorphic annotation \<PolyDet>.
-One of the most common locations for a polymorphic annotation is at a method signature.
-Consider the following method declaration annotated with polymorphic annotations:
-\begin{verbatim}
-PolyDet int foo(PolyDet String param1, PolyDet boolean param2) { }
-\end{verbatim}
-There are two valid instantiations for this declaration:
-\begin{enumerate}
-    \item NonDet int foo(NonDet String param1, NonDet boolean param2) { }
-    \item Det int foo(Det String param1, Det boolean param2) { }
-\end{enumerate}
-This indicates that method \<foo> can be called with arguments having any of the \codeid{Det or NonDet} type annotations.
-%   (\<@OrderNonDet> is not allowed because it is invalid on primitive types).
- \<PolyDet> resolves to the least upper bound of
-the actual types on arguments. For instance, if this method is called as \codeid{foo(Det String arg1, NonDet boolean arg2)}, the 
-checker framework resolves the method declaration as \codeid{NonDet int foo(NonDet String param1, NonDet boolean param2)}
-causing the return type to have the type annotation \<NonDet>.
 
@@
 \begin{figure}
-    $\infer[\textsc{any type}]{\wellformed{\kappa \  \beta}}{\kappa : \mathit{Det / NonDet}}$
+    $\infer[\textsc{any type}]{\wellformed{\kappa \  \beta}}{\kappa : \|Det / NonDet|}$
 
@@
 
-    $\infer[\textsc{ordernondet collection}]{\wellformed{\kappa_c \  \beta_c \langle\kappa_e \ \beta_e\rangle}}{\kappa_c : \mathit{OrderNonDet/Det} & \kappa_e \sqsubseteq \kappa_c & \beta_c \sqsubseteq \mathit{Collection/Iterator}}$
+    $\infer[\textsc{ordernondet collection}]{\wellformed{\kappa_c \  \beta_c \langle\kappa_e \ \beta_e\rangle}}{\kappa_c : \|OrderNonDet/Det| & \kappa_e \sqsubseteq \kappa_c & \beta_c \sqsubseteq \|Collection/Iterator|}$
 
@@
 
-    $\infer[\textsc{nondet collection}]{\wellformed{\mathit{NonDet} \ \beta_c\langle\mathit{NonDet} \ \beta_e\rangle}}{\beta_c \sqsubseteq \mathit{Collection/Iterator}}$
+    $\infer[\textsc{nondet collection}]{\wellformed{\|NonDet| \ \beta_c\langle\|NonDet| \ \beta_e\rangle}}{\beta_c \sqsubseteq \|Collection/Iterator|}$
 
@@
 
-     $\infer[\textsc{ordernondet array}]{\wellformed{\kappa_e \ \beta_e \ \  \kappa_a []}}{\kappa_a : \mathit{OrderNonDet/Det} & \kappa_e \sqsubseteq \kappa_a }$
+     $\infer[\textsc{ordernondet array}]{\wellformed{\kappa_e \ \beta_e \ \  \kappa_a []}}{\kappa_a : \|OrderNonDet/Det| & \kappa_e \sqsubseteq \kappa_a }$
 
@@
 
-    $\infer[\textsc{nondet array}]{\wellformed{\mathit{NonDet} \ \beta_e \ \  \mathit{NonDet} []}}{}$
+    $\infer[\textsc{nondet array}]{\wellformed{\|NonDet| \ \beta_e \ \  \|NonDet| []}}{}$
 
@@
 
-    $\infer[\textsc{ordernondet map}]{\wellformed{\kappa_m \  \beta_m \langle\kappa_k \ \beta_k, \kappa_v \ \beta_v\rangle}}{\kappa_m: \mathit{OrderNonDet/Det} & \kappa_k \sqsubseteq \kappa_m & \kappa_v \sqsubseteq \kappa_m & \beta_m \sqsubseteq \mathit{Map}}$
+    $\infer[\textsc{ordernondet map}]{\wellformed{\kappa_m \  \beta_m \langle\kappa_k \ \beta_k, \kappa_v \ \beta_v\rangle}}{\kappa_m: \|OrderNonDet/Det| & \kappa_k \sqsubseteq \kappa_m & \kappa_v \sqsubseteq \kappa_m & \beta_m \sqsubseteq \|Map|}$
 
@@
 
-    $\infer[\textsc{nondet map}]{\wellformed{\mathit{NonDet} \ \beta_m\langle\mathit{NonDet} \ \beta_k, \mathit{NonDet} \ \beta_v\rangle}}{\beta_m \sqsubseteq \mathit{Map}}$
+    $\infer[\textsc{nondet map}]{\wellformed{\|NonDet| \ \beta_m\langle\|NonDet| \ \beta_k, \|NonDet| \ \beta_v\rangle}}{\beta_m \sqsubseteq \|Map|}$
 
@@
 
-\subsection{Polymorphism extensions}\label{polymorphism-extensions}
+\subsection{Polymorphism}\label{polymorphism}
 
-Recall that the determinism checker does not permit \<OrderNonDet> on a type that isn't a \<Collection>, \<Iterator>, or an array.
-As a consequence, resolution of polymorphic type annotations at method signatures could have undesirable effects.
+\TheDeterminismChecker supports three types of polymorphism:  type
+polymorphism, basetype polymorphism, and qualifier polymorphism.
+Type polymorphism is handled by Java's generics mechanism, which
+\theDeterminismChecker fully supports, including class and method generics,
+inference, etc.
+Basetype polymorphism is enabled by writing a type qualifier on a use of a
+type variable.
+Java does not provide a syntax that can be used for qualifier polymorphism,
+so we use a special type qualifier name, \<@PolyDet>.
+A method \<m> with signature $\forall \kappa . \kappa \<int> \times \<Det boolean> \rightarrow
+\kappa \<String>$ would be written in our dialect of Java as
+\begin{Verbatim}
+@PolyDet String m(@PolyDet int, @Det boolean)
+\end{Verbatim}
+
+Each use of \<@PolyDet> stands for a use of the qualifier variable
+$\kappa$, and there is no need to declare the qualifier variable $\kappa$.
+
+The underlying checker-framework on which we build \theDeterminismChecker has support for
+polymorphic qualifiers. When a user writes a polymorphic qualifier on a method signature or a type parameter,
+it indicates that it could resolve to any type in the type system depending on how it is used.
+In \theDeterminismChecker, we define a polymorphic qualifier \<PolyDet>.
+One of the most common locations for a polymorphic qualifier is at a method signature.
+Consider the following method declaration annotated with polymorphic qualifiers:
+\begin{verbatim}
+PolyDet int foo(PolyDet String param1, PolyDet boolean param2) { }
+\end{verbatim}
+\todo{It has the standard meaning.  There are two ways to interpret this.}
+There are two valid instantiations for this declaration:
+\begin{enumerate}
+    \item @NonDet int foo(@NonDet String param1, @NonDet boolean param2) { }
+    \item @Det int foo(@Det String param1, @Det boolean param2) { }
+\end{enumerate}
+This indicates that method \<foo> can be called with arguments having any of the \codeid{Det or NonDet} type qualifiers.
+%   (\<@OrderNonDet> is not allowed because it is invalid on primitive types).
+ \<PolyDet> resolves to the least upper bound of
+the actual types on arguments. For instance, if this method is called as \codeid{foo(@Det String arg1, @NonDet boolean arg2)}, the 
+checker framework resolves the method declaration as \codeid{@NonDet int foo(@NonDet String param1, @NonDet boolean param2)}
+causing the return type to have the type qualifier \<NonDet>.
+
+\subsection{Polymorphism rules for collections}\label{polymorphism-up-down}
+
+Recall that \theDeterminismChecker does not permit \<OrderNonDet> on a type that isn't a \<Collection>, \<Iterator>, or an array.
+As a consequence, resolution of polymorphic type qualifiers at method signatures could have undesirable effects.
 Suppose a method declaration is annotated as \codeid{PolyDet int size(PolyDet List<T> list)}. Invoking this method
 with an argument of type \codeid{OrderNonDet List<Det String>} will result in \<PolyDet> getting resolved to \<OrderNonDet> which is invalid
-on the return type \<int>. We circumvent this scenario by introducing two variants to the \<PolyDet> annotation, \<PolyDet$\uparrow$>
+on the return type \<int>. We circumvent this scenario by introducing two variants to the \<PolyDet> qualifier, \<PolyDet$\uparrow$>
 and \<PolyDet$\downarrow$>.
@@
 \begin{itemize}
-    \item \codeid{Det int size(Det List<T> list)}
-    \item \codeid{Det int size(OrderNonDet List<T> list)}
-    \item \codeid{NonDet int size(NonDet List<T> list)}
+    \item \codeid{@Det int size(@Det List<T> list)}
+    \item \codeid{@Det int size(@OrderNonDet List<T> list)}
+    \item \codeid{@NonDet int size(@NonDet List<T> list)}
 \end{itemize}
@@
 \begin{figure}
-    $\infer[\textsc{poly resolution}]{\pi_r \ \beta_r method(\overline{\pi_i \ \beta_i})\{\ \}}{\forall K \in \{\mathit{NonDet, OrderNonDet, Det\}} : K \ \beta_r method(\overline{K \ \beta_i})\{\ \} & \pi_r, \pi_i \in \{\mathit{PolyDet}, \mathit{PolyDet}\uparrow, \mathit{PolyDet}\downarrow    \}}$
+    $\infer[\textsc{poly resolution}]{\pi_r \ \beta_r method(\overline{\pi_i \ \beta_i})\{\ \}}{\forall K \in \{\|NonDet, OrderNonDet, Det|\} : K \ \beta_r method(\overline{K \ \beta_i})\{\ \} & \pi_r, \pi_i \in \{\|PolyDet|, \|PolyDet|\uparrow, \|PolyDet|\downarrow    \}}$
     \caption{Polymorphic resolution rule}
@@
 
-\subsection{Differentiating binding and use}\label{bindings-uses}
+\subsubsection{Differentiating binding and use}\label{bindings-uses}
 
 In order to provide soundness guarantees, we must ensure that a collection that is
-declared as \<Det> is not side-effected in non-deterministic ways. Consider the following annotation 
+declared as \<Det> is not side-effected in non-deterministic ways. Consider the following qualifier 
 on the List add method: \codeid{public void add(PolyDet ArrayList<E> this, PolyDet int index, E element)}.
@@
 \end{verbatim}
-This will result in the instantion \codeid{public void add(NonDet ArrayList<E> this, NonDet int index, E element)} and
+This will result in the instantion \codeid{public void add(@NonDet ArrayList<E> this, @NonDet int index, E element)} and
 the method invocation will type check. This is problematic because even though \<list> was declared to be \<Det>,
@@
 Since \<PolyDet("use")> has no effect on type instantiations, calling this method on a \<Det List>
-will result in the method declaration being resolved as \codeid{public void add(Det ArrayList<E> this, Det int index, E element)}
+will result in the method declaration being resolved as \codeid{public void add(@Det ArrayList<E> this, @Det int index, E element)}
 irrespective of the type of \<index>, thereby preventing undesired side-effects.
@@
 \subsection{Discussion on defaulting rules}\label{defaulting}
-Every type system in the checker framework must define a default type annotation. This annotation will be applied
-to every location that isn't explicitly annotated. In our determinism checker, we chose \<PolyDet> as the default for all 
+Every type system in the Checker Framework must define a default type qualifier. This qualifier will be applied
+to every location that isn't explicitly annotated. In \theDeterminismChecker, we chose \<PolyDet> as the default for all 
 method parameters and return types including constructors.
 The rationale for this design choice was to not make our type checker too restrictive by making everything \<Det>.  This implies that
-the checker still allows some non-determinism as long as it doesn't break any determinism specifications explicitly written by the programmer.
+\theDeterminismChecker still allows some non-determinism as long as it doesn't break any determinism specifications explicitly written by the programmer.
 
@@
 In the example above, the method \codeid{foo} is annotated with the \<PolyDet> type qualifier which allows
-the caller method \codeid{bar} to invoke it with both \<Det> and \<NonDet> arguments. The checker flags an error 
+the caller method \codeid{bar} to invoke it with both \<Det> and \<NonDet> arguments. \TheDeterminismChecker flags an error 
 only when the returned value is used in a way that breaks any deterministic specification (Printing a \<NonDet> value
@@
 The checker framework performs automatic type inference for local types within method bodies.
-It also performs dataflow analysis and refines the types of these locals, thereby reducing the annotation burden 
+It also performs dataflow analysis and refines the types of these locals, thereby reducing the qualifier burden 
 on the programmer. Consider the example below:
@@
 \end{verbatim}
-the checker reports a warning at the second assignment statement of \codeid{foo}. Type refinement only applies to locals. Refining non-locals can result in side effects making the analysis unsound.
+\theDeterminismChecker reports a warning at the second assignment statement of \codeid{foo}.
+Type refinement only applies to locals.
+Refining non-locals can result in side effects making the analysis unsound.
 
@@
 type arguments will get the type qualifier \<Det>. For example, a locally declared \codeid{List} of \codeid{Strings} will be automatically annotated
-as \codeid{NonDet List<Det String>}. Recall from~\ref{collection-rules} that the determinism checker does not allow this type on
-any collection. While this is still sound, it will result in the determinism checker reporting a lot of errors because every locally declared 
+as \codeid{NonDet List<Det String>}. Recall from~\ref{collection-rules} that \theDeterminismChecker does not allow this type on
+any collection. While this is still sound, it will result in \theDeterminismChecker reporting a lot of errors because every locally declared 
 collection gets annotated with this invalid type. To avoid this situation, we make an exception to the defaulting rules for local collections
 and annotate their type parameters also with \<NonDet>.
-So a local list of Strings gets the default annotation of \codeid{NonDet List<NonDet String>}.
+So a local list of Strings gets the default qualifier of \codeid{NonDet List<NonDet String>}.
 While this defaulting eliminates the possibility of automatically annotating locals collections with an invalid
@@
 \end{verbatim}
-the checker would report an error at the assignment statement since the inferred type for \codeid{localList} is 
+\theDeterminismChecker would report an error at the assignment statement since the inferred type for \codeid{localList} is 
 \codeid{NonDet List<NonDet> String} and since collections types are invariant.
@@
 
-The determinism checker has additional type refinement rules for the sorting methods as shown in Figure~\ref{fig-sorting}. 
+\TheDeterminismChecker has additional type refinement rules for the sorting methods as shown in Figure~\ref{fig-sorting}. 
 \begin{figure}
-    $\infer[\textsc{list sort}]{\Gamma \vdash x : \mathit{Det \ List}\langle\mathit{Det} \ \beta\rangle}{\Gamma \vdash x : \mathit{OrderNonDet \ List}\langle\mathit{Det} \ \beta\rangle & \mathit{x.sort()}}$
+    $\infer[\textsc{list sort}]{\Gamma \vdash x : \|Det \ List|\langle\|Det| \ \beta\rangle}{\Gamma \vdash x : \|OrderNonDet \ List|\langle\|Det| \ \beta\rangle & \|x.sort()|}$
 
@@
 
-    $\infer[\textsc{collections sort}]{\Gamma \vdash x : \mathit{Det \ List}\langle\mathit{Det} \ \beta\rangle}{\Gamma \vdash x : \mathit{OrderNonDet \ List}\langle\mathit{Det} \ \beta\rangle & \mathit{Collections.sort.sort(x)}}$
+    $\infer[\textsc{collections sort}]{\Gamma \vdash x : \|Det \ List|\langle\|Det| \ \beta\rangle}{\Gamma \vdash x : \|OrderNonDet \ List|\langle\|Det| \ \beta\rangle & \|Collections.sort.sort(x)|}$
 
@@
 
-    $\infer[\textsc{arrays sort}]{\Gamma \vdash x : \mathit{Det \ \beta \ Det[]} }{\Gamma \vdash x : \mathit{Det \ \beta \ OrderNonDet[]} & \mathit{Arrays.sort( \ x)}}$
+    $\infer[\textsc{arrays sort}]{\Gamma \vdash x : \|Det \ \beta \ Det[]| }{\Gamma \vdash x : \|Det \ \beta \ OrderNonDet[]| & \|Arrays.sort( \ x)|}$
 
@@
 \end{figure}
-The checker type refines a receiver of type \codeid{OrderNonDet List} to a \codeid{Det List} if all the
-type argument of this \codeid{List} have the type annotation \codeid{Det} (Rule LIST SORT in Figure~\ref{fig-sorting}). Similar rules apply for \codeid{Arrays.sort()} and \codeid{Collections.sort()}.
+\TheDeterminismChecker type refines a receiver of type \codeid{OrderNonDet List} to a \codeid{Det List} if all the
+type argument of this \codeid{List} have the type qualifier \codeid{Det} (Rule LIST SORT in Figure~\ref{fig-sorting}). Similar rules apply for \codeid{Arrays.sort()} and \codeid{Collections.sort()}.
 As per these rules, if \codeid{sort()} is called on a receiver of type \codeid{OrderNonDet List<Det Set<Det String>>}, it will
@@
 %
-%Every type system in the checker framework must define a default type annotation. This annotation will be applied
-%to every location that isn't explicitly annotated. In our determinism checker, we chose \codeid{@Det} as the default annotation.
+%Every type system in the checker framework must define a default type qualifier. This qualifier will be applied
+%to every location that isn't explicitly annotated. In our determinism checker, we chose \codeid{@Det} as the default qualifier.
 %The reason for this design choice is that we expect the program to be 
@@
 %    \textit{top} type, then the Checker Framework refines it to the best (most specific) possible type based on assignments to it. As a 
-%    result, a programmer rarely writes an explicit annotation on any of those locations.
+%    result, a programmer rarely writes an explicit qualifier on any of those locations.
 %    Variables defaulted to top include \textit{local variables}, resource variables in the \textit{try-with-resources} construct, variables 
@@
 %type arguments will get the type \<@Det>. For example, a locally declared \codeid{List} of \codeid{Strings} will be automatically annotated
-%as \codeid{@NonDet List<@Det String>}. Recall from~\ref{collection-rules} that the determinism checker does not allow this type on
-%any collection. While this is still sound, it will result in the determinism checker reporting a lot of errors because every locally declared 
+%as \codeid{@NonDet List<@Det String>}. Recall from~\ref{collection-rules} that \theDeterminismChecker does not allow this type on
+%any collection. While this is still sound, it will result in \theDeterminismChecker reporting a lot of errors because every locally declared 
 %collection gets annotated with this invalid type. To avoid this situation, we make an exception to the defaulting rules for local collections
 %and annotate their type parameters also with \<@NonDet>.
-So a local list of Strings gets the default annotation of \codeid{@NonDet List<@NonDet String>}.
+So a local list of Strings gets the default qualifier of \codeid{@NonDet List<@NonDet String>}.
 While this defaulting eliminates the possibility of automatically annotating locals collections with an invalid
@@
 %\end{verbatim}
-%the checker would report an error at the assignment statement since the inferred type for \codeid{localList} is 
+%\theDeterminismChecker would report an error at the assignment statement since the inferred type for \codeid{localList} is 
 %\codeid{@NonDet List<@NonDet> String} and since collections types are invariant.
@@
 \begin{figure}
-    $\infer[\textsc{hashmap creation}]{\mathit{new \ \kappa\ HashMap()}}{\kappa = \mathit{OrderDet/NonDet}}$
+    $\infer[\textsc{hashmap creation}]{\|new \ \kappa\ HashMap()|}{\kappa = \|OrderDet/NonDet|}$
 
@@
 
-    $\infer[\textsc{treemap creation}]{\mathit{new \ \kappa\ TreeMap()}}{\kappa = \mathit{Det/NonDet}}$
+    $\infer[\textsc{treemap creation}]{\|new \ \kappa\ TreeMap()|}{\kappa = \|Det/NonDet|}$
 
@@
 
-    $\infer[\textsc{hashset creation}]{\mathit{new \ \kappa\ HashSet()}}{\kappa = \mathit{OrderDet/NonDet}}$
+    $\infer[\textsc{hashset creation}]{\|new \ \kappa\ HashSet()|}{\kappa = \|OrderDet/NonDet|}$
 
@@
 
-    $\infer[\textsc{treeset creation}]{\mathit{new \ \kappa\ TreeSet()}}{\kappa = \mathit{Det/NonDet}}$
+    $\infer[\textsc{treeset creation}]{\|new \ \kappa\ TreeSet()|}{\kappa = \|Det/NonDet|}$
     \caption{\<Map> and \<Set> creation rules}
@@
     \todo{linebreak}
-    $\infer[\textsc{set precision}]{\Gamma \vdash \mathit{z : Det}}{\Gamma \vdash \mathit{x} : \mathit{OrderNonDet} \ \mathit{Set}\langle \kappa_{x} \ \beta_{x}\rangle, \mathit{y} : \mathit{OrderNonDet} \ \mathit{Set}\langle \kappa_{y} \ \beta_{y}\rangle & \mathit{z = x.equals(y)} & \langle \kappa_{x} \ \beta_{x}\rangle \nsqsubseteq \langle \mathit{OrderNonDet \ List} \rangle & \langle \kappa_{y} \ \beta_{y}\rangle \nsqsubseteq \langle \mathit{OrderNonDet \ List} \rangle }$
+    $\infer[\textsc{set precision}]{\Gamma \vdash \|z : Det|}{\Gamma \vdash \|x| : \|OrderNonDet| \ \|Set|\langle \kappa_{x} \ \beta_{x}\rangle, \|y| : \|OrderNonDet| \ \|Set|\langle \kappa_{y} \ \beta_{y}\rangle & \|z = x.equals(y)| & \langle \kappa_{x} \ \beta_{x}\rangle \nsqsubseteq \langle \|OrderNonDet \ List| \rangle & \langle \kappa_{y} \ \beta_{y}\rangle \nsqsubseteq \langle \|OrderNonDet \ List| \rangle }$
 
@@
 
-    $\infer[\textsc{map precision}]{\Gamma \vdash \mathit{z : Det}}{\Gamma \vdash \mathit{x : OrderNonDet \ Map\langle Det \ \beta_k, \kappa_v\ \beta_v \rangle, y : Det \ \beta} & \mathit{z = x.get(y)}}$
+    $\infer[\textsc{map precision}]{\Gamma \vdash \|z : Det|}{\Gamma \vdash \|x : OrderNonDet \ Map\langle Det \ \beta_k, \kappa_v\ \beta_v \rangle, y : Det \ \beta| & \|z = x.get(y)|}$
 
@@
 
-Similarly, the \codeid{get()} method in the \<Map> class is annotated with polymorphic annotations and the rule $\sc{\mathit{MAP \ PRECISION}}$ in Figure~\ref{fig-precision-rules} improves precision for \<OrderNonDet> receivers.
+Similarly, the \codeid{get()} method in the \<Map> class is annotated with polymorphic qualifiers and the rule $\sc{\|MAP \ PRECISION|}$ in Figure~\ref{fig-precision-rules} improves precision for \<OrderNonDet> receivers.
 
@@
 
-\subsection{Extending the determinism checker}\label{extending-det-checker}
+\subsection{Extending \theDeterminismChecker}\label{extending-det-checker}
 We have primarily focused on collections and annotated the corresponding classes in the JDK.
-These annotations provide machine verifiable specifications of determinism properties.
+These qualifiers provide machine verifiable specifications of determinism properties.
 Another source of nondeterminism that we haven't analyzed closely is the File system methods.
@@
 the File system can easily do so by annotating the parameter and return types of those methods in the JDK.
-Alternatively, a user can write these annotations in a \textit{stub} file and supply it to the type checker. A \textit{stub}
-only needs to contain method declarations along with their annotations.
+Alternatively, a user can write these qualifiers in a \textit{stub} file and supply it to the type checker. A \textit{stub}
+only needs to contain method declarations along with their qualifiers.

### 2019-03-28 design.tex
Edits in section 2.4

@@
 a \textit{type} stands for a set of values.  It abstracts or restricts the
-set of possbile run-time values that an expression may evaluate to.
+set of possible run-time values that an expression may evaluate to.
 A programming language provides \emph{basetypes}, such as \<int>.
@@
     that the expression might have different values in two different executions.
+    \item \<OrderNonDet> indicates that the expression is a collection or
+      map (dictionary or associative array) that
+      will have the same elements in every execution, but in a possibly different order.
     \item \<Det> indicates that
@@
 
-\subsection{Handling order nondeterminism}
-In addition to the type qualifiers described in~\ref{type-hierarchy}, \theDeterminismChecker 
-defines the type qualifier \<OrderNonDet>. This qualifier represents
-a collection or an array that will have the same elements in every execution, but in a
-possibly different order.  \<OrderNonDet> may only be written on
-collections and arrays.
-A collection is any subtype of \<java.util.Collection>,
-\<java.util.Iterator>, or \<java.util.Map>, including user-defined types. The relationship
-of \<OrderNonDet> with respect the the other two type qualifiers is \codeid{Det $\sqsubseteq$ OrderNonDet $\sqsubseteq$ NonDet} as shown in Figure~\ref{fig-determinism-hierarchy}.
+\<OrderNonDet> may only be written on collections and maps.  Our implementation
+handles array types, and any class (including user-defined classes) that
+implements the \<Iterable> or \<Iterator> interfaces; this includes all
+Java collections such as \<List>s and \<Set>s.  For simplicity, and because
+the semantics and rules are similar even though the syntax varies, this
+paper uses the term ``collection'' and the type \<Collection> to represent
+all these types.
+
+Our type system largely treats a map as a collection of key--value pairs.  Like any
+collection, a map may be \<Det>, \<OrderNonDet>, or \<NonDet>.
 
@@
 
+\todo{Note that the \textsc{ordernondet} versions also handle \<Det>,
+  because $\<Det> \sqsubseteq \<OrderNonDet>$.}
+
 \begin{figure}
@@
 
-    $\infer[\textsc{ordernondet collection}]{\wellformed{\kappa_c \  \beta_c \langle\kappa_e \ \beta_e\rangle}}{\kappa_c : \|OrderNonDet/Det| & \kappa_e \sqsubseteq \kappa_c & \beta_c \sqsubseteq \|Collection/Iterator|}$
+    $\infer[\textsc{ordernondet collection}]{\wellformed{\kappa_c \  \beta_c \langle\kappa_e \ \beta_e\rangle}}{\kappa_c : \|OrderNonDet/Det| & \kappa_e \sqsubseteq \kappa_c & \beta_c \sqsubseteq \<Collection>}$
 
@@
 
-    $\infer[\textsc{nondet collection}]{\wellformed{\|NonDet| \ \beta_c\langle\|NonDet| \ \beta_e\rangle}}{\beta_c \sqsubseteq \|Collection/Iterator|}$
+    $\infer[\textsc{nondet collection}]{\wellformed{\|NonDet| \ \beta_c\langle\|NonDet| \ \beta_e\rangle}}{\beta_c \sqsubseteq \<Collection>}$
 
@@
 
-     $\infer[\textsc{ordernondet array}]{\wellformed{\kappa_e \ \beta_e \ \  \kappa_a []}}{\kappa_a : \|OrderNonDet/Det| & \kappa_e \sqsubseteq \kappa_a }$
+%      $\infer[\textsc{ordernondet array}]{\wellformed{\kappa_e \ \beta_e \ \  \kappa_a []}}{\kappa_a : \|OrderNonDet/Det| & \kappa_e \sqsubseteq \kappa_a }$
+%     
+%     \bigskip
+%     
+%     $\infer[\textsc{nondet array}]{\wellformed{\|NonDet| \ \beta_e \ \  \|NonDet| []}}{}$
+%     
+%     \bigskip
+
+    $\infer[\textsc{ordernondet map}]{\wellformed{\kappa_m \  \beta_m \langle\kappa_k \ \beta_k, \kappa_v \ \beta_v\rangle}}{\kappa_m: \|OrderNonDet/Det| & \kappa_k \sqsubseteq \kappa_m & \kappa_v \sqsubseteq \kappa_m & \beta_m \sqsubseteq \<Map>}$
 
@@
 
-    $\infer[\textsc{nondet array}]{\wellformed{\|NonDet| \ \beta_e \ \  \|NonDet| []}}{}$
-
-    \bigskip
-
-    $\infer[\textsc{ordernondet map}]{\wellformed{\kappa_m \  \beta_m \langle\kappa_k \ \beta_k, \kappa_v \ \beta_v\rangle}}{\kappa_m: \|OrderNonDet/Det| & \kappa_k \sqsubseteq \kappa_m & \kappa_v \sqsubseteq \kappa_m & \beta_m \sqsubseteq \|Map|}$
-
-    \bigskip
-
-    $\infer[\textsc{nondet map}]{\wellformed{\|NonDet| \ \beta_m\langle\|NonDet| \ \beta_k, \|NonDet| \ \beta_v\rangle}}{\beta_m \sqsubseteq \|Map|}$
+    $\infer[\textsc{nondet map}]{\wellformed{\|NonDet| \ \beta_m\langle\|NonDet| \ \beta_k, \|NonDet| \ \beta_v\rangle}}{\beta_m \sqsubseteq \<Map>}$
 
     \caption{Type well formedness}
+\todo{The \textsc{any type} rule is not correct, because not every type is
+  well-formed.  It needs an antecedent stating that $\beta$ is not a
+  collection or map.}
+
+\todo{The collection rules do not permit nested collections; similarly, the
+  (currently commented out) array rules did not permit any two-dimensional
+  array to be well-formed.  There is a bigger issue here.
+  Consider the ``nondet collection'' rule.  $\beta_c$ is a class (we need
+  to use a different symbol for it, probably $c$), but $\beta_e$ should be
+  a full type, which might contain type qualifiers within it.  (Actually,
+  $\|NonDet|\ \beta_e$ is a type, not $\beta_e$ itself.)  So, there should
+  be an antecedent stating that $\|NonDet| \beta_e$ is well-formed, and
+  for the ``ordernondet collection'' rule stating that $\kappa_e \beta_e$
+  is well-formed.  Likewise, antecedents are needed for the map rules.
+  Otherwise, these rules permit construction of $\<Map>\langle\tau'\rangle$
+  where $\tau'$ is itself not well-defined.}
+
+\todo{``$\kappa : \|Det / NonDet|$'' should be ``$\kappa = \|Det /
+  NonDet|$'' or (better)
+  ``$\kappa \in \{\|Det|, \|NonDet|\}$''.
+  You could also express $\kappa_c : \|OrderNonDet/Det|$ as just
+  $\kappa_c \sqsubseteq \|OrderNonDet|$ since $\|Det| \sqsubseteq \|OrderNonDet|$.
+}
+
   \label{type-validity}
@@
 
-Figure~\ref{type-validity} shows the type validity rules of our type system. A type $\tau$ is represented as $\kappa \ \beta$
+Figure~\ref{type-validity} shows the type validity\todo{The caption says
+  ``type well formedness''.  Choose one term and use it consistently
+  throughout.}
+rules of our type system. A type $\tau$ is represented as $\kappa \ \beta$
 where $\kappa$ is a type qualifier from the determinism type hierarchy and $\beta$ is a Java type.
 As stated in the rule $\textsc{any type}$, the types \<Det> and \<NonDet> can be written on any Java type.
-The (determinism) type of a Collection or Iterator must be a supertype or equal to
-the type of the type parameter (Rule $\textsc{ordernondet collection}$) (see Figure~\ref{fig-determinism-collections}).
+The (determinism) type of a collection must be a supertype or equal to
+its element type (Rule $\textsc{ordernondet collection}$) (see Figure~\ref{fig-determinism-collections}).
 If the Collection is \<NonDet>, then the type parameter may not be
@@
 
-Some examples of valid types are:
+Some examples of valid types are:\todo{double-check}
 \begin{itemize}
     \item \codeid{NonDet\ \ \ \ \ \ List<NonDet\ \ \ \ \ \ Integer>}
+    \item \codeid{OrderNonDet List<OrderNonDet Set<...>\relax >}
+    \item \codeid{OrderNonDet List<Det Integer>}
     \item \codeid{Det\ \ \ \ \ \ \ \ \ List<Det\ \ \ \ \ \ \ \ \ Integer>}
-    \item \codeid{OrderNonDet List <OrderNonDet Set<...>\relax >}
-    \item \codeid{OrderNonDet List <Det String>}
 \end{itemize}
@@
 \begin{itemize}
-    \item \codeid{OrderNonDet Set <NonDet\ \ \ \ \ \ Integer>}
+    \item \codeid{NonDet\ \ \ \ \ \ List<OrderNonDet Integer>}
     \item \codeid{NonDet\ \ \ \ \ \ List<Det\ \ \ \ \ \ \ \ \ Integer>}
+    \item \codeid{OrderNonDet List<NonDet\ \ \ \ \ \ Integer>}
+    \item \codeid{Det\ \ \ \ \ \ \ \ \ List<NonDet\ \ \ \ \ \ Set<...>\relax >}
     \item \codeid{Det\ \ \ \ \ \ \ \ \ List<OrderNonDet Set<...>\relax >}
 \end{itemize}
-
-Similarly, the (determinism) type of an array must be a supertype or equal to
-the type of the component type (see Figure~\ref{fig-determinism-collections}).
-As with collections, \codeid{NonDet} arrays of \codeid{Det} or \codeid{OrderNonDet}
-elements are not allowed (Rules $\textsc{ordernondet array}$ and $\textsc{nondet array}$ in Figure~\ref{type-validity}).
 
@@
         \cline{3-5}
-        \multicolumn{2}{c|}{~}  &  \multicolumn{3}{c|}{Type Argument (or array component type)} \\ \cline{3-5}
+        \multicolumn{2}{c|}{~}  &  \multicolumn{3}{c|}{element type} \\ \cline{3-5}
         \multicolumn{2}{c|}{~}  & NonDet     & OrderNonDet & Det \\ \hline
         & NonDet      &   valid    &  invalid    & invalid  \\ \cline{2-5}
-        Collection (or array)   & OrderNonDet &   invalid  &  valid      & valid  \\ \cline{2-5}
+        Collection    & OrderNonDet &   invalid  &  valid      & valid  \\ \cline{2-5}
         & Det         &   invalid  &  invalid    & valid      \\ \hline
     \end{tabular}
-    \caption{Valid Collection (and array) declarations.  The Collection's (or array's) type qualifier
-        must be a supertype or equal to the type argument (or array component type).}
+    \caption{Valid Collection declarations.  The Collection's type qualifier
+        must be a supertype or equal to the element type.}
     \label{fig-determinism-collections}
@@
 
-A collection (\<List, Set, or a Map>) annotated as \<OrderNonDet> has the following properties:
+A collection or map annotated as \<OrderNonDet> has the following properties:
+\todo{Write these as type rules.}
 \begin{enumerate}
     \item The individual elements extracted by iterating over the collection get the type \<NonDet>.
-    \item Accessing properties sich as \<size()> or querying if the collection \<isEmpty()> will return types
+    \item Accessing properties such as \<size()> or querying if the collection \<isEmpty()> will return types
     that are annotated as \<Det> as they do not depend on the iteration order. 
 \end{enumerate}
-
-A user defined type may be annotated as \<OrderNonDet> if and only if it
-is a subtype of Collection or Iterator.
-For example, if a user defines a type as
-\begin{verbatim}
-public class TestUserCollection<E> extends ArrayList<E> {...}
-\end{verbatim}
-Writing \codeid{OrderNonDet TestUserCollection<Det Integer>} is valid.\\
-Writing \codeid{OrderNonDet TestUserCollection<NonDet Integer>} is invalid
-because the type of the type parameter (\codeid{NonDet}) is not a subtype
-of the user defined Collection type (\codeid{OrderNonDet}).
-
-User defined types that are not a subtype of Collection or Iterator
-may not be annotated as \codeid{OrderNonDet}.
 
@@
 
-Recall that \theDeterminismChecker does not permit \<OrderNonDet> on a type that isn't a \<Collection>, \<Iterator>, or an array.
+Some types are invalid and must not be permitted (during polymorphic instantiation).
+
+Recall that \theDeterminismChecker does not permit \<OrderNonDet> on a type
+that isn't a collection.
 As a consequence, resolution of polymorphic type qualifiers at method signatures could have undesirable effects.
@@
 \begin{figure}
-    $\infer[\textsc{poly resolution}]{\pi_r \ \beta_r method(\overline{\pi_i \ \beta_i})\{\ \}}{\forall K \in \{\|NonDet, OrderNonDet, Det|\} : K \ \beta_r method(\overline{K \ \beta_i})\{\ \} & \pi_r, \pi_i \in \{\|PolyDet|, \|PolyDet|\uparrow, \|PolyDet|\downarrow    \}}$
+    $\infer[\textsc{poly resolution}]{\pi_r \ \beta_r\ \<m>(\overline{\pi_i \ \beta_i})\{\ \}}{\forall \kappa \in \{\|NonDet, OrderNonDet, Det|\} : \kappa \ \beta_r\ \<m>(\overline{K \ \beta_i})\{\ \} & \pi_r, \pi_i \in \{\|PolyDet|, \|PolyDet|\uparrow, \|PolyDet|\downarrow    \}}$
     \caption{Polymorphic resolution rule}
@@
 \caption{Polymorphic resolution}
+\todo{Write this as a rule rather than a table.}
 \label{tab-poly-resolutions}
@@
 \end{verbatim}
-This will result in the instantion \codeid{public void add(@NonDet ArrayList<E> this, @NonDet int index, E element)} and
+This will result in the instantiation \codeid{public void add(@NonDet ArrayList<E> this, @NonDet int index, E element)} and
 the method invocation will type check. This is problematic because even though \<list> was declared to be \<Det>,
@@
 We therefore recommend that programmers explicitly annotate
-all local variables that are collections to reduce false positives. The same reasoning and recommendation applies to arrays.
+all local variables that are collections to reduce false positives.
 
-\TheDeterminismChecker has additional type refinement rules for the sorting methods as shown in Figure~\ref{fig-sorting}. 
+\TheDeterminismChecker has additional type refinement rules for the sorting
+methods such as \<List.sort>, \<Collections.sort>, and \<Arrays.sort>:
+ as shown in Figure~\ref{fig-sorting}. 
 \begin{figure}
-    $\infer[\textsc{list sort}]{\Gamma \vdash x : \|Det \ List|\langle\|Det| \ \beta\rangle}{\Gamma \vdash x : \|OrderNonDet \ List|\langle\|Det| \ \beta\rangle & \|x.sort()|}$
+%     $\infer[\textsc{list sort}]{\Gamma \vdash x : \|Det \ List|\langle\|Det| \ \beta\rangle}{\Gamma \vdash x : \|OrderNonDet \ List|\langle\|Det| \ \beta\rangle & \|x.sort()|}$
+%     
+%     \bigskip
 
-    \bigskip
+    $\infer[\textsc{sort}]{\Gamma \vdash x : \|Det \ List|\langle\|Det| \ \beta\rangle}{\Gamma \vdash x : \|OrderNonDet \ List|\langle\|Det| \ \beta\rangle & \|Collections.sort(x)|}$
 
-    $\infer[\textsc{collections sort}]{\Gamma \vdash x : \|Det \ List|\langle\|Det| \ \beta\rangle}{\Gamma \vdash x : \|OrderNonDet \ List|\langle\|Det| \ \beta\rangle & \|Collections.sort.sort(x)|}$
-
-    \bigskip
-
-    $\infer[\textsc{arrays sort}]{\Gamma \vdash x : \|Det \ \beta \ Det[]| }{\Gamma \vdash x : \|Det \ \beta \ OrderNonDet[]| & \|Arrays.sort( \ x)|}$
+%     \bigskip
+%     
+%     $\infer[\textsc{arrays sort}]{\Gamma \vdash x : \|Det \ \beta \ Det[]| }{\Gamma \vdash x : \|Det \ \beta \ OrderNonDet[]| & \|Arrays.sort( \ x)|}$
 
@@
 \TheDeterminismChecker type refines a receiver of type \codeid{OrderNonDet List} to a \codeid{Det List} if all the
-type argument of this \codeid{List} have the type qualifier \codeid{Det} (Rule LIST SORT in Figure~\ref{fig-sorting}). Similar rules apply for \codeid{Arrays.sort()} and \codeid{Collections.sort()}.
+type argument of this \codeid{List} have the type qualifier \codeid{Det} (Rule LIST SORT in Figure~\ref{fig-sorting}).
 As per these rules, if \codeid{sort()} is called on a receiver of type \codeid{OrderNonDet List<Det Set<Det String>>}, it will
@@
 In the JDK, \<LinkedHashSet> is implemented as a subclass of \<HashMap>.
-\todo{Talk about this design choice. Josh Blochs book: "If the code surrounding the first declaration depended on LinkedHashSet's ordering policy,
+\todo{Talk about this design choice. Josh Bloch's book: "If the code surrounding the first declaration depended on LinkedHashSet's ordering policy,
     then it would be incorrect to substitute HashSet for LinkedHashSet in the declaration, because HashSet makes no guarantee concerning iteration order."}
@@
 \begin{figure}
-    \todo{linebreak}
+    \todo{add line break}
     $\infer[\textsc{set precision}]{\Gamma \vdash \|z : Det|}{\Gamma \vdash \|x| : \|OrderNonDet| \ \|Set|\langle \kappa_{x} \ \beta_{x}\rangle, \|y| : \|OrderNonDet| \ \|Set|\langle \kappa_{y} \ \beta_{y}\rangle & \|z = x.equals(y)| & \langle \kappa_{x} \ \beta_{x}\rangle \nsqsubseteq \langle \|OrderNonDet \ List| \rangle & \langle \kappa_{y} \ \beta_{y}\rangle \nsqsubseteq \langle \|OrderNonDet \ List| \rangle }$
@@
 is sound but imprecise. This rule is careful enough to not annotate the return type of \<equals()> as \<Det> if a type argument to either the
-receiver or the argument happens to be an \codeid{OrderNonDet List} (An order nondeterminstic list does not provide any guarantee on the
+receiver or the argument happens to be an \codeid{OrderNonDet List} (An order nondeterministic list does not provide any guarantee on the
 equality as it is dependent on the iteration order).
@@
 only needs to contain method declarations along with their qualifiers.
+
+% LocalWords:  basetypes NonDet OrderNonDet Det Iterable formedness param1
+% LocalWords:  PolyDet boolean param2 arg1 arg2 localList entrySet keySet
+% LocalWords:  LinkedHashMap TreeMap HashSet LinkedHashSet LinkedHashSet's

### 2019-03-28 experiments.tex

@@
 \begin{table}[]
-    \begin{tabular}{|l|l|l|l|l|l|}
+    \begin{tabular}{|p{1.5cm}|p{1cm}|p{3cm}|p{2cm}|p{2cm}|p{3cm}|}
         \hline
-        Benchmark & Recall? & Error message source & # annotations & Type of annotation added & Comments \\ \hline
-        P1 &  &  &  &  &  \\
+        Benchmark & Recall? & Error message source & \#annotations & Type of annotation added & Error msg \\ \hline
+        P1 & Yes & FieldAccess.java: [102,87] &  &  &  assignment type incompatible \\
         P2 &  &  &  &  &  \\

### 2019-03-28 design.tex
removing array rules
Tweaks in 2.0

--- 2019-03-27 design.tex
@@
 In a type system,
-a \textit{type} stands for a set of values.  It abstracts or restricts the
+a \textit{type} represents a set of values.  It abstracts or restricts the
 set of possbile run-time values that an expression may evaluate to.
@@
 that is, it reduces the size of the set of values.
-A type checker can verify the correctness of this \textit{type} with its dataflow analysis and subtyping rules.
-In Java, a type qualifier is written as a type annotation with a leading \<@>, as in
+In Java, a type qualifier is written as a type annotation (which has a leading \<@>), as in
 \<@Positive int>.
+A type checker can verify the correctness of a \textit{type} via subtyping
+rules and dataflow analysis.
 
@@
 
-     $\infer[\textsc{ordernondet array}]{\wellformed{\kappa_e \ \beta_e \ \  \kappa_a []}}{\kappa_a : \|OrderNonDet/Det| & \kappa_e \sqsubseteq \kappa_a }$
+%     $\infer[\textsc{ordernondet array}]{\wellformed{\kappa_e \ \beta_e \ \  \kappa_a []}}{\kappa_a : \|OrderNonDet/Det| & \kappa_e \sqsubseteq \kappa_a }$
+%    
+%    \bigskip
 
-    \bigskip
-
-    $\infer[\textsc{nondet array}]{\wellformed{\|NonDet| \ \beta_e \ \  \|NonDet| []}}{}$
-
-    \bigskip
+%    $\infer[\textsc{nondet array}]{\wellformed{\|NonDet| \ \beta_e \ \  \|NonDet| []}}{}$
+%    
+%    \bigskip
 
@@
 
-    \bigskip
-
-    $\infer[\textsc{arrays sort}]{\Gamma \vdash x : \|Det \ \beta \ Det[]| }{\Gamma \vdash x : \|Det \ \beta \ OrderNonDet[]| & \|Arrays.sort( \ x)|}$
+%    \bigskip
+%    
+%    $\infer[\textsc{arrays sort}]{\Gamma \vdash x : \|Det \ \beta \ Det[]| }{\Gamma \vdash x : \|Det \ \beta \ OrderNonDet[]| & \|Arrays.sort( \ x)|}$
 

### 2019-03-28 experiments.tex

--- 2019-03-27 experiments.tex
@@
 \begin{table}[]
-    \begin{tabular}{|l|l|l|l|l|l|}
+    \begin{tabular}{|p{1.5cm}|p{1cm}|p{3cm}|p{4cm}|p{2cm}|p{1.5cm}|}
         \hline
-        Benchmark & Recall? & Error message source & # annotations & Type of annotation added & Comments \\ \hline
-        P1 &  &  &  &  &  \\
-        P2 &  &  &  &  &  \\
-        P3 &  &  &  &  &  \\ 
-        P4 &  &  &  &  &  \\ 
-        P5 &  &  &  &  &  \\ 
-        P5 &  &  &  &  &  \\ 
-        P6 &  &  &  &  &  \\ 
-        P6 &  &  &  &  &  \\
-        P7 &  &  &  &  &  \\ 
-        P7 &  &  &  &  &  \\
-        P7 &  &  &  &  &  \\
-        P7 &  &  &  &  &  \\
-        P7 &  &  &  &  &  \\
-        P8 &  &  &  &  &  \\
-        P8 &  &  &  &  &  \\
-        P8 &  &  &  &  &  \\
-        P9 &  &  &  &  &  \\
-        P10 &  &  &  &  &  \\
-        P10 &  &  &  &  &  \\
-        P11 &  &  &  &  &  \\
-        P12 &  &  &  &  &  \\
-        P13 &  &  &  &  &  \\
-        P14 &  &  &  &  &  \\
-        P15 &  &  &  &  &  \\
-        P15 &  &  &  &  &  \\
-        P16 &  &  &  &  &  \\
-        P17 &  &  &  &  &  \\
-        P18 &  &  &  &  &  \\
-        P19 &  &  &  &  &  \\
-        P19 &  &  &  &  &  \\
-        P19 &  &  &  &  &  \\
-        P20 &  &  &  &  &  \\
-        P20 &  &  &  &  &  \\
-        P21 &  &  &  &  &  \\
-        P21 &  &  &  &  &  \\ \hline
+        Benchmark & Recall? & Error message source & Error msg & \#annotations & Annotation type  \\ \hline
+        P1 & Yes & FieldAccess.java: [102,87] & assignment type incompatible &  &   \\
+        P2 & MI &  &  &  &  \\
+        P3 & Yes & DateTimeZone.java: [475,15] &  return type incompatible&  &  \\ 
+        P4 & MI &  &  &  &  \\ 
+        P5 & src &  &  &  &  \\ 
+        P5 & src &  &  &  &  \\ 
+        P6 & MI &  &  &  &  \\ 
+        P6 & MI &  &  &  &  \\
+        P7(test error) & Yes & checker/ tests/ determinism/P71.java: 5 & return type incompatible &  &  \\ 
+        P7(2) & Yes & FieldUtils.java: [219,84] & assignment type incompatible &  &  \\
+        P7(2) & Yes & Reflection ToStringBuilder.java: [543,65]  & assignment type incompatible &  &  \\
+        P7(2) & Yes & HashCodeBuilder.java: [193,69]  & assignment type incompatible &  &  \\
+        P7(2) & Yes & CompareToBuilder.java: [322,65] & assignment type incompatible &  &  \\
+        P7(2) & Yes & EqualsBuilder.java: [419,69]  & assignment type incompatible &  &  \\
+        P7 & Yes & FieldUtils.java: [200,69] & assignment type incompatible & &  \\
+        P7 & Yes & FieldUtils.java: [219,84] & assignment type incompatible &  &  \\
+        P7 & Yes & FieldUtils.java: [240,86] & assignment type incompatible &  &  \\
+        P8 & RM &  &  &  &  \\
+        P8 & RM &  &  &  &  \\
+        P8 & RM &  &  &  &  \\
+        P9 & MI &  &  &  &  \\
+        P10 & crash &  &  &  &  \\
+        P10 & crash &  &  &  &  \\
+        P11 & MI &  &  &  &  \\
+        P12 & MI &  &  &  &  \\
+        P13 & crash &  &  &  &  \\
+        P14 & MI &  &  &  &  \\
+        P15 & MI &  &  &  &  \\
+        P15 & MI &  &  &  &  \\
+        P16 & Yes & MethodSorter.java: [56,61]  & assignment type incompatible &  &  \\
+        P17 & Yes &  &  &  &  \\
+        P18 & src &  &  &  &  \\
+        P19 & MI &  &  &  &  \\
+        P19 & MI &  &  &  &  \\
+        P19 & MI &  &  &  &  \\
+        P20 & MI &  &  &  &  \\
+        P20 & MI &  &  &  &  \\
+        P21 & MI &  &  &  &  \\
+        P21 & MI &  &  &  &  \\ \hline
     \end{tabular}

### 2019-03-28 design.tex

--- 2019-03-28 design.tex
@@
 In a type system,
-a \textit{type} stands for a set of values.  It abstracts or restricts the
+a \textit{type} represents a set of values.  It abstracts or restricts the
 set of possible run-time values that an expression may evaluate to.

### 2019-03-28 design.tex
More macros

@@
 
-    $\infer{\kappa_1 \beta \sqsubseteq \kappa_2 \beta}{\kappa_1 \sqsubseteq \kappa_2}
-    \quad\quad
-    \infer{\kappa \beta_1 \sqsubseteq \kappa \beta_2}{\beta_1 \sqsubseteq \beta_2}$
+    $\infer[\textsc{method call}]{\|method|(\overline{a_i}) : \tau_r}{\tau_r \  \|method|(\overline{\tau_i \  p_i}) & \overline{a_i : \tau_i}}$
 
+    \bigskip
+
+    $\infer[\textsc{invariant type param}]{\|method|(\overline{a_i}) : \tau_r}{\tau_r \  \|method|(\overline{\tau_i \  p_i}) & \overline{a_i : \tau_i}}$
+
+    \caption{A sampling of standard type checking rules.  Most standard rules are omitted
+    for brevity.  $\sqsubseteq$ is overloaded for classes and types.}
+    \label{typecheck-rules}
+\end{figure}
+
+\begin{figure}
     \bigskip
@@
 
-    $\infer[\textsc{method call}]{\|method|(\overline{a_i}) : \tau_r}{\tau_r \  \|method|(\overline{\tau_i \  p_i}) & \overline{a_i : \tau_i}}$
+    $\infer{\kappa_1 \beta \sqsubseteq \kappa_2 \beta}{\kappa_1 \sqsubseteq \kappa_2}
+    \quad\quad
+    \infer{\kappa \beta_1 \sqsubseteq \kappa \beta_2}{\beta_1 \sqsubseteq \beta_2}$
 
-    \caption{Type checking rules}
+    \caption{Type checking rules.  $\sqsubseteq$ is overloaded for classes,
+    basetypes, type qualifiers, and types.  Most standard rules are omitted
+    for brevity.}
     \label{typecheck-rules}
@@
     \item \<OrderNonDet> indicates that the expression is a collection or
-      map (dictionary or associative array) that
-      will have the same elements in every execution, but in a possibly different order.
+      map that contains the same elements in every execution, but possibly
+      in a different order.
     \item \<Det> indicates that
@@
 
+A map is a dictionary or associative array, such as a hash table.
 Our type system largely treats a map as a collection of key--value pairs.  Like any
@@
 
-    $\infer[\textsc{ordernondet collection}]{\wellformed{\kappa_c \  \beta_c \langle\kappa_e \ \beta_e\rangle}}{\kappa_c : \|OrderNonDet/Det| & \kappa_e \sqsubseteq \kappa_c & \beta_c \sqsubseteq \<Collection>}$
+    $\infer[\textsc{ordernondet collection}]{\wellformed{\kappa_c \  \beta_c \angles{\kappa_e \ \beta_e}}}{\kappa_c : \|OrderNonDet/Det| & \kappa_e \sqsubseteq \kappa_c & \beta_c \sqsubseteq \<Collection>}$
 
@@
 
-    $\infer[\textsc{nondet collection}]{\wellformed{\|NonDet| \ \beta_c\langle\|NonDet| \ \beta_e\rangle}}{\beta_c \sqsubseteq \<Collection>}$
+    $\infer[\textsc{nondet collection}]{\wellformed{\|NonDet| \ \beta_c\angles{\|NonDet| \ \beta_e}}}{\beta_c \sqsubseteq \<Collection>}$
 
@@
 
-    $\infer[\textsc{ordernondet map}]{\wellformed{\kappa_m \  \beta_m \langle\kappa_k \ \beta_k, \kappa_v \ \beta_v\rangle}}{\kappa_m: \|OrderNonDet/Det| & \kappa_k \sqsubseteq \kappa_m & \kappa_v \sqsubseteq \kappa_m & \beta_m \sqsubseteq \<Map>}$
+    $\infer[\textsc{ordernondet map}]{\wellformed{\kappa_m \  \beta_m \angles{\kappa_k \ \beta_k, \kappa_v \ \beta_v}}}{\kappa_m: \|OrderNonDet/Det| & \kappa_k \sqsubseteq \kappa_m & \kappa_v \sqsubseteq \kappa_m & \beta_m \sqsubseteq \<Map>}$
 
@@
 
-    $\infer[\textsc{nondet map}]{\wellformed{\|NonDet| \ \beta_m\langle\|NonDet| \ \beta_k, \|NonDet| \ \beta_v\rangle}}{\beta_m \sqsubseteq \<Map>}$
+    $\infer[\textsc{nondet map}]{\wellformed{\|NonDet| \ \beta_m\angles{\|NonDet| \ \beta_k, \|NonDet| \ \beta_v}}}{\beta_m \sqsubseteq \<Map>}$
 
@@
   is well-formed.  Likewise, antecedents are needed for the map rules.
-  Otherwise, these rules permit construction of $\<Map>\langle\tau'\rangle$
+  Otherwise, these rules permit construction of $\<Map>\angles{\tau'}$
   where $\tau'$ is itself not well-defined.}
@@
 \begin{figure}
-%     $\infer[\textsc{list sort}]{\Gamma \vdash x : \|Det \ List|\langle\|Det| \ \beta\rangle}{\Gamma \vdash x : \|OrderNonDet \ List|\langle\|Det| \ \beta\rangle & \|x.sort()|}$
+%     $\infer[\textsc{list sort}]{\Gamma \vdash x : \|Det \ List|\angles{\|Det| \ \beta}}{\Gamma \vdash x : \|OrderNonDet \ List|\angles{\|Det| \ \beta} & \|x.sort()|}$
 %     
@@
 
-    $\infer[\textsc{sort}]{\Gamma \vdash x : \|Det \ List|\langle\|Det| \ \beta\rangle}{\Gamma \vdash x : \|OrderNonDet \ List|\langle\|Det| \ \beta\rangle & \|Collections.sort(x)|}$
+    $\infer[\textsc{sort}]{\Gamma \vdash x : \|Det \ List|\angles{\|Det| \ \beta}}{\Gamma \vdash x : \|OrderNonDet \ List|\angles{\|Det| \ \beta} & \|Collections.sort(x)|}$
 
@@
     \todo{add line break}
-    $\infer[\textsc{set precision}]{\Gamma \vdash \|z : Det|}{\Gamma \vdash \|x| : \|OrderNonDet| \ \|Set|\langle \kappa_{x} \ \beta_{x}\rangle, \|y| : \|OrderNonDet| \ \|Set|\langle \kappa_{y} \ \beta_{y}\rangle & \|z = x.equals(y)| & \langle \kappa_{x} \ \beta_{x}\rangle \nsqsubseteq \langle \|OrderNonDet \ List| \rangle & \langle \kappa_{y} \ \beta_{y}\rangle \nsqsubseteq \langle \|OrderNonDet \ List| \rangle }$
+    $\infer[\textsc{set precision}]{\Gamma \vdash \|z : Det|}{\Gamma \vdash \|x| : \|OrderNonDet| \ \|Set|\angles{ \kappa_{x} \ \beta_{x}}, \|y| : \|OrderNonDet| \ \|Set|\angles{ \kappa_{y} \ \beta_{y}} & \|z = x.equals(y)| & \angles{ \kappa_{x} \ \beta_{x}} \nsqsubseteq \angles{ \|OrderNonDet \ List| } & \angles{ \kappa_{y} \ \beta_{y}} \nsqsubseteq \angles{ \|OrderNonDet \ List| } }$
 
@@
 
-    $\infer[\textsc{map precision}]{\Gamma \vdash \|z : Det|}{\Gamma \vdash \|x : OrderNonDet \ Map\langle Det \ \beta_k, \kappa_v\ \beta_v \rangle, y : Det \ \beta| & \|z = x.get(y)|}$
+    $\infer[\textsc{map precision}]{\Gamma \vdash \|z : Det|}{\Gamma \vdash \|x : OrderNonDet \ Map\angles{ Det \ \beta_k, \kappa_v\ \beta_v }, y : Det \ \beta| & \|z = x.get(y)|}$
 

### 2019-03-28 design.tex
Fix LaTeX error, complete figure
Edits in section 2.2

@@
 
-    $\infer[\textsc{invariant type param}]{\|method|(\overline{a_i}) : \tau_r}{\tau_r \  \|method|(\overline{\tau_i \  p_i}) & \overline{a_i : \tau_i}}$
+    $\infer[\textsc{invariant type param}]{C_1\angles{\overline{\tau_i}}
+      \sqsubseteq C_2\angles{\overline{\tau_i}}}{C_1 \sqsubseteq C_2}$
 
@@
     for brevity.  $\sqsubseteq$ is overloaded for classes and types.}
-    \label{typecheck-rules}
+    \label{typecheck-rules-standard}
 \end{figure}
@@
     \caption{Type checking rules.  $\sqsubseteq$ is overloaded for classes,
-    basetypes, type qualifiers, and types.  Most standard rules are omitted
-    for brevity.}
+    basetypes, type qualifiers, and types.}
     \label{typecheck-rules}
@@
 A map is a dictionary or associative array, such as a hash table.
-Our type system largely treats a map as a collection of key--value pairs.  Like any
-collection, a map may be \<Det>, \<OrderNonDet>, or \<NonDet>.
+Our type system largely treats a map as a collection of key--value pairs.  Like a
+collection, a map may be \<Det>, \<OrderNonDet>, or \<NonDet>.  (In Java,
+\<HashMap> cannot be deterministic, but \<TreeMap> and \<LinkedHashMap>
+can.\todo{forward reference?})
 
@@
 
-\subsection{Additional typing rules for collection elements}\label{collection-rules}
+\subsection{Collection types}\label{collection-rules}
 
@@
 rules of our type system. A type $\tau$ is represented as $\kappa \ \beta$
-where $\kappa$ is a type qualifier from the determinism type hierarchy and $\beta$ is a Java type.
-As stated in the rule $\textsc{any type}$, the types \<Det> and \<NonDet> can be written on any Java type.
-The (determinism) type of a collection must be a supertype or equal to
-its element type (Rule $\textsc{ordernondet collection}$) (see Figure~\ref{fig-determinism-collections}).
-If the Collection is \<NonDet>, then the type parameter may not be
+where $\kappa$ is a type qualifier (that is, \<NonDet>, \<OrderNonDet>, or
+\<Det>) and $\beta$ is a Java basetype.
+As stated in the rule $\textsc{any type}$, the types \<Det> and \<NonDet>
+can be written on any Java basetype.\todo{As noted in figure caption, that's
+  not true, in some contexts such as a type argument.}
+
+The determinism type qualifier on a collection must be a supertype or equal to
+its element type qualifier (Rule $\textsc{ordernondet collection}$ in Figure~\ref{fig-determinism-collections}).
+If the Collection is \<NonDet>, then the type argument may not be
 \<Det> or \<OrderNonDet> (Rule $\textsc{nondet collection}$ in Figure~\ref{type-validity}). Although such types could exist, they are
@@
 
-%\todo{Fewer examples, less variation.  Use \<List> and \<String>.}
+\todo{Some notes:}
+Typical type for \<get> is \\
+$\forall \tau . \<List>\angles{\tau} \times \<int> \rightarrow \tau$ \\
+but we need to change that.  If either argument is \<OrderNonDet> or \<NonDet>, then the
+result is \<NonDet>.
+If both arguments are \<Det>, the result is $\tau$ as usual.
+\todo{In that case, we know that $\tau = \<Det> \beta$.}
 
-Some examples of valid types are:\todo{double-check}
-\begin{itemize}
-    \item \codeid{NonDet\ \ \ \ \ \ List<NonDet\ \ \ \ \ \ Integer>}
-    \item \codeid{OrderNonDet List<OrderNonDet Set<...>\relax >}
-    \item \codeid{OrderNonDet List<Det Integer>}
-    \item \codeid{Det\ \ \ \ \ \ \ \ \ List<Det\ \ \ \ \ \ \ \ \ Integer>}
-\end{itemize}
+\smallskip
+\noindent
+\begin{minipage}{.48\textwidth}
+Some examples of valid types are:
+% Using "~" instead of "\ " does not work within \codeid :-(
+\begin{smaller}
+\begin{Verbatim}
+@NonDet      List<@NonDet      Integer>
+@OrderNonDet List<@OrderNonDet Set<...>>
+@OrderNonDet List<@Det         Integer>
+@Det         List<@Det         Integer>
 
+\end{Verbatim}
+\end{smaller}
+\end{minipage}
+\hfill
+\begin{minipage}{.46\textwidth}
 These types are invalid:
-\begin{itemize}
-    \item \codeid{NonDet\ \ \ \ \ \ List<OrderNonDet Integer>}
-    \item \codeid{NonDet\ \ \ \ \ \ List<Det\ \ \ \ \ \ \ \ \ Integer>}
-    \item \codeid{OrderNonDet List<NonDet\ \ \ \ \ \ Integer>}
-    \item \codeid{Det\ \ \ \ \ \ \ \ \ List<NonDet\ \ \ \ \ \ Set<...>\relax >}
-    \item \codeid{Det\ \ \ \ \ \ \ \ \ List<OrderNonDet Set<...>\relax >}
-\end{itemize}
+\begin{smaller}
+\begin{Verbatim}
+@NonDet      List<@OrderNonDet Set<...>>
+@NonDet      List<@Det         Integer>
+@OrderNonDet List<@NonDet      Integer>
+@Det         List<@NonDet      Integer>
+@Det         List<@OrderNonDet Set<...>>
+\end{Verbatim}
+\end{smaller}
+\end{minipage}
+\smallskip
 

### 2019-03-28 design.tex
Initial edits in section 2.3

@@
 
-\TheDeterminismChecker supports three types of polymorphism:  type
+Our type system supports three types of polymorphism:  type
 polymorphism, basetype polymorphism, and qualifier polymorphism.
+\todo{Give an example for each.}
+\begin{itemize}
+\item
+In \theDeterminismCheckerimplementation,
 Type polymorphism is handled by Java's generics mechanism, which
@@
 inference, etc.
+\item
 Basetype polymorphism is enabled by writing a type qualifier on a use of a
 type variable.
+\item
 Java does not provide a syntax that can be used for qualifier polymorphism,
 so we use a special type qualifier name, \<@PolyDet>.
-A method \<m> with signature $\forall \kappa . \kappa \<int> \times \<Det boolean> \rightarrow
+A qualifier-polymorphic method \<m> with signature $\forall \kappa . \kappa \<int> \times \<Det boolean> \rightarrow
 \kappa \<String>$ would be written in our dialect of Java as
@@
 \end{Verbatim}
-
 Each use of \<@PolyDet> stands for a use of the qualifier variable
 $\kappa$, and there is no need to declare the qualifier variable $\kappa$.
+\TheDeterminismChecker currently supports
+qualifier
+polymorphism on methods but not on classes.
+Therefore, \<@PolyDet> may be written in methods (signatures and bodies)
+but not on fields.
+\end{itemize}
 

### 2019-03-28 design.tex
Edits in section 2.3

@@
 \item
-In \theDeterminismCheckerimplementation,
-Type polymorphism is handled by Java's generics mechanism, which
+In \theDeterminismCheckerImplementation,
+type polymorphism is handled by Java's generics mechanism, which
 \theDeterminismChecker fully supports, including class and method generics,
 inference, etc.
+Given the Java declaration
+\begin{Verbatim}
+<T> T identity(T arg) { return arg; }
+\end{Verbatim}
+the type of \<identity(arg)> is the same as the type of
+\<arg>, and a programmer can write a type such as \codeid{List<@NonDet
+  Integer>} (but see below for details about collections).
 \item
 Basetype polymorphism is enabled by writing a type qualifier on a use of a
-type variable.
+type variable, which overrides the type qualifier at the instantiation
+site.
+For example, a \<choose> operation on sets could be defined in Java as
+\begin{Verbatim}
+class ChooseableSet<T> implements Set<T> {
+  // return an arbitrary element from the set
+  @NonDet T choose();
+}
+\end{Verbatim}
 \item
@@
 so we use a special type qualifier name, \<@PolyDet>.
-A qualifier-polymorphic method \<m> with signature $\forall \kappa . \kappa \<int> \times \<Det boolean> \rightarrow
-\kappa \<String>$ would be written in our dialect of Java as
+A qualifier-polymorphic method \<m> with signature $\forall \kappa . \kappa\ \<int> \times \<Det boolean> \rightarrow
+\kappa\ \<String>$ would be written in our dialect of Java as
 \begin{Verbatim}
@@
 
-The underlying checker-framework on which we build \theDeterminismChecker has support for
-polymorphic qualifiers. When a user writes a polymorphic qualifier on a method signature or a type parameter,
-it indicates that it could resolve to any type in the type system depending on how it is used.
+When a user writes a polymorphic qualifier on a method signature or a type parameter,
+it indicates that it could resolve\todo{Should all occurrences of ``resolve'' be ``can be instantiated''} to any type in the type system depending on how it is used.
 In \theDeterminismChecker, we define a polymorphic qualifier \<PolyDet>.
 One of the most common locations for a polymorphic qualifier is at a method signature.
-Consider the following method declaration annotated with polymorphic qualifiers:
+Consider the following declaration of addition:
 \begin{verbatim}
-PolyDet int foo(PolyDet String param1, PolyDet boolean param2) { }
+@PolyDet int plus(@PolyDet int a, @PolyDet int b) { return a+b; }
 \end{verbatim}
@@
 \begin{enumerate}
-    \item @NonDet int foo(@NonDet String param1, @NonDet boolean param2) { }
-    \item @Det int foo(@Det String param1, @Det boolean param2) { }
+    \item @NonDet int plus(@NonDet int a, @NonDet int b) { return a+b; }
+    \item @Det int plus(@Det int a, @Det int b) { return a+b; }
 \end{enumerate}
@@
  \<PolyDet> resolves to the least upper bound of
-the actual types on arguments. For instance, if this method is called as \codeid{foo(@Det String arg1, @NonDet boolean arg2)}, the 
-checker framework resolves the method declaration as \codeid{@NonDet int foo(@NonDet String param1, @NonDet boolean param2)}
+the actual types on arguments.
+For instance, if this method is called as \codeid{foo(@Det String arg1, @NonDet boolean arg2)}, \theDeterminismChecker resolves the method declaration as \codeid{@NonDet int foo(@NonDet String param1, @NonDet boolean param2)}
 causing the return type to have the type qualifier \<NonDet>.
@@
 with an argument of type \codeid{OrderNonDet List<Det String>} will result in \<PolyDet> getting resolved to \<OrderNonDet> which is invalid
-on the return type \<int>. We circumvent this scenario by introducing two variants to the \<PolyDet> qualifier, \<PolyDet$\uparrow$>
-and \<PolyDet$\downarrow$>.
+on the return type \<int>. We circumvent this scenario by introducing two variants to the \<PolyDet> qualifier, \PolyDetUp
+and \PolyDetDown.
 
@@
 \begin{itemize}
-    \item \<PolyDet$\uparrow$>: Same as \<PolyDet> except when it resolves to \<OrderNonDet>, it is replaced with \<NonDet>.
-    \item \<PolyDet$\downarrow$>: Same as \<PolyDet> except when it resolves to \<OrderNonDet>, it is replaced with \<Det>.
+    \item \PolyDetUp: Same as \<PolyDet> except when it resolves to \<OrderNonDet>, it is replaced with \<NonDet>.
+    \item \PolyDetDown: Same as \<PolyDet> except when it resolves to \<OrderNonDet>, it is replaced with \<Det>.
 \end{itemize} 
@@
     \caption{Polymorphic resolution rule}
+    \todo{I'm not sure what this is doing.  What code does it reject?}
+    \todo{This suggests that \PolyDetUp is a qualifier, rather than
+      $\uparrow$ being an operator on qualifiers.  We should clarify which
+      it is.}
     \label{fig-poly-resolution}
@@
         \hline
-        \textbf{\<PolyDet>} & \textbf{\<PolyDet$\uparrow$>} & \textbf{\<PolyDet$\downarrow$>} \\ \hline
+        \textbf{\<PolyDet>} & \textbf{\PolyDetUp} & \textbf{\PolyDetDown} \\ \hline
         \<NonDet> & \<NonDet> &  \<NonDet>\\ \hline
@@
 \subsection{Discussion on defaulting rules}\label{defaulting}
-Every type system in the Checker Framework must define a default type qualifier. This qualifier will be applied
-to every location that isn't explicitly annotated. In \theDeterminismChecker, we chose \<PolyDet> as the default for all 
+
+\TheDeterminismChecker applies a default qualifier at each unannotated type
+use (except uses of type variables, which already stand for a type that was
+defaulted at the instantiation site).
+This does not change the expressivity of the type system; it merely makes
+the system more convenient to use.
+
+In \theDeterminismChecker, we chose \<PolyDet> as the default for all 
 method parameters and return types including constructors.
@@
 \subsection{Dataflow analysis and type refinement}\label{dataflow}
-The checker framework performs automatic type inference for local types within method bodies.
+\TheDeterminismChecker performs automatic type inference for local types within method bodies.
 It also performs dataflow analysis and refines the types of these locals, thereby reducing the qualifier burden 
@@
 
-With the above rules, the checker framework will automatically annotate a local collection with type \<NonDet> and its
+With the above rules, \theDeterminismChecker will automatically annotate a local collection with type \<NonDet> and its
 type arguments will get the type qualifier \<Det>. For example, a locally declared \codeid{List} of \codeid{Strings} will be automatically annotated

### 2019-03-28 design.tex
Edits in section 2.0
Edits in section 2.1

@@
-\section{Design}\label{design}
+\section{A type system for determinism}\label{design}
+
+This section presents the key aspects of the design of our type system,
+called \ourTypeSystem, in the context of a core calculus for an
+object-oriented language.
+
+\subsection{Preliminaries}
 
@@
 % Throughout this paper, we use the notation \<@A> <: \<@B> to denote that type \<@A> is a subtype of type \<@B>.
-% As an illustration of the type hierarchy, consider the lattice of types shown in Figure~\ref{fig-example-lattice}.
+% As an illustration of the type hierarchy, consider the lattice of types shown in \cref{fig-example-lattice}.
 % \begin{figure}
@@
 % The type checker now performs additional type checking (similar to javac's type checking) with respect to this type hierarchy
-% and reports any violations. The snippet in Figure~\ref{code-invalid1} would result in an error at the assignment statements:
+% and reports any violations. The snippet in \cref{code-invalid1} would result in an error at the assignment statements:
 % \begin{figure}
@@
 % For method overriding, the usual rules apply for parameters and arguments i.e covariant subtyping for parameters
-% and contravariant subtyping for return types. Consider the example in Figure~\ref{code-invalid2} where method \<foo> of \<class X> is overridden
+% and contravariant subtyping for return types. Consider the example in \cref{code-invalid2} where method \<foo> of \<class X> is overridden
 % in \<class Y>.
@@
 
+\todo{Greek letters go after this paragraph.}
+
 In a type system,
@@
 A programming language provides \emph{basetypes}, such as \<int>.
-A \textit{type qualifier} on a \textit{type} adds additional constraints;
+A \textit{type qualifier} on a basetype adds additional constraints;
 that is, it reduces the size of the set of values.
-In Java, a type qualifier is written as a type annotation (which has a leading \<@>), as in
-\<@Positive int>.
-A type checker can verify the correctness of a \textit{type} via subtyping
-rules and dataflow analysis.
+An example type qualifier is \<Positive>,\todo{should all type qualifiers
+  be lowercase in this section??} and a type (which combines a qualifier
+anda basetype) is \<Positive int>.
+A polymorphic type such as \<List> can be instantiated by a type argument,
+as in \ptype{List}{Positive int}.
+
+A type checker verifies the types written in a program.
 
@@
     basetypes, type qualifiers, and types.}
-    \label{typecheck-rules}
+    \label{fig:typecheck-rules}
 \end{figure}
@@
     \caption{Determinism type qualifier hierarchy}
-    \label{fig-determinism-hierarchy}
+    \label{fig:determinism-hierarchy}
 \todo{Drop the at-signs (@).  The text is weirdly small for the box.
@@
 
-The core of the determinism type system is the following type qualifiers (see Figure~\ref{fig-determinism-hierarchy}):
+The core of the determinism type system is the following type qualifiers:
 %\todo{introduce just @Det and @NonDet first, and then @OrderNonDet after polymorphism.}
@@
 \end{itemize}
+\Cref{fig:typecheck-rules,fig:determinism-hierarchy} show their subtyping relationship.
 
@@
 
-In Figure~\ref{code-determinism}, we present some of the JDK methods that we have annotated with our determinism types and give examples of
+In \cref{code-determinism}, we present some of the JDK methods that we have annotated with our determinism types and give examples of
 client code that would produce errors at compile time.
@@
 
-Figure~\ref{type-validity} shows the type validity\todo{The caption says
+\Cref{type-validity} shows the type validity\todo{The caption says
   ``type well formedness''.  Choose one term and use it consistently
@@
 The determinism type qualifier on a collection must be a supertype or equal to
-its element type qualifier (Rule $\textsc{ordernondet collection}$ in Figure~\ref{fig-determinism-collections}).
+its element type qualifier (Rule $\textsc{ordernondet collection}$ in \cref{fig-determinism-collections}).
 If the Collection is \<NonDet>, then the type argument may not be
-\<Det> or \<OrderNonDet> (Rule $\textsc{nondet collection}$ in Figure~\ref{type-validity}). Although such types could exist, they are
+\<Det> or \<OrderNonDet> (Rule $\textsc{nondet collection}$ in \cref{type-validity}). Although such types could exist, they are
 disallowed to prevent the following bug:
@@
 
-Figure~\ref{fig-poly-resolution} presents the rule for type checking of methods annotated with \<PolyDet> or its variants.
+\Cref{fig-poly-resolution} presents the rule for type checking of methods annotated with \<PolyDet> or its variants.
 A polymorphic method type checks if every instantiation of \<PolyDet> type checks. If this method has \<PolyDet>$\uparrow$ or \<PolyDet>$\downarrow$
@@
 methods such as \<List.sort>, \<Collections.sort>, and \<Arrays.sort>:
- as shown in Figure~\ref{fig-sorting}. 
+ as shown in \cref{fig-sorting}.
 \begin{figure}
@@
 \TheDeterminismChecker type refines a receiver of type \codeid{OrderNonDet List} to a \codeid{Det List} if all the
-type argument of this \codeid{List} have the type qualifier \codeid{Det} (Rule LIST SORT in Figure~\ref{fig-sorting}).
+type argument of this \codeid{List} have the type qualifier \codeid{Det} (Rule LIST SORT in \cref{fig-sorting}).
 As per these rules, if \codeid{sort()} is called on a receiver of type \codeid{OrderNonDet List<Det Set<Det String>>}, it will
@@
 We prohibit the creation of a \codeid{Det HashMap} or an \codeid{OrderNonDet TreeMap} in our determinism checker.
-Figure~\ref{fig-creation-rules} formalizes these rules. Similar constraints apply to \<HashSet>s and \<TreeSet>s which are
-also expressed by the last two rules of Figure~\ref{fig-creation-rules}.
+\Cref{fig-creation-rules} formalizes these rules. Similar constraints apply to \<HashSet>s and \<TreeSet>s which are
+also expressed by the last two rules of \cref{fig-creation-rules}.
 
@@
 \end{verbatim}
-While this is sound, it is imprecise in the case of \codeid{OrderNonDet Sets}. We added the following rule shown in Figure~\ref{fig-precision-rules} to our checker to improve precision.
+While this is sound, it is imprecise in the case of \codeid{OrderNonDet Sets}. We added the following rule shown in \cref{fig-precision-rules} to our checker to improve precision.
 \begin{figure}
@@
 
-Similarly, the \codeid{get()} method in the \<Map> class is annotated with polymorphic qualifiers and the rule $\sc{\|MAP \ PRECISION|}$ in Figure~\ref{fig-precision-rules} improves precision for \<OrderNonDet> receivers.
+Similarly, the \codeid{get()} method in the \<Map> class is annotated with polymorphic qualifiers and the rule $\sc{\|MAP \ PRECISION|}$ in \cref{fig-precision-rules} improves precision for \<OrderNonDet> receivers.
 

### 2019-03-28 java-design.tex
Edits in section 2.1

\section{Determinism types for Java}\label{sec:java-types}

In Java, a type qualifier is written as a type annotation (which has a leading \<@>), as in
\<@Positive int>.

### 2019-03-28 design.tex
Change more \ref to \cref

@@
 A polymorphic method type checks if every instantiation of \<PolyDet> type checks. If this method has \<PolyDet>$\uparrow$ or \<PolyDet>$\downarrow$
-in its signature then the instantiation of \<PolyDet> at that location gets replaced by the appropriate type as shown in Table~\ref{tab-poly-resolutions}.
+in its signature then the instantiation of \<PolyDet> at that location gets replaced by the appropriate type as shown in \cref{tab-poly-resolutions}.
 
@@
 type arguments will get the type qualifier \<Det>. For example, a locally declared \codeid{List} of \codeid{Strings} will be automatically annotated
-as \codeid{NonDet List<Det String>}. Recall from~\ref{collection-rules} that \theDeterminismChecker does not allow this type on
+as \codeid{NonDet List<Det String>}. Recall from~\cref{collection-rules} that \theDeterminismChecker does not allow this type on
 any collection. While this is still sound, it will result in \theDeterminismChecker reporting a lot of errors because every locally declared 
@@
 %type arguments will get the type \<@Det>. For example, a locally declared \codeid{List} of \codeid{Strings} will be automatically annotated
-%as \codeid{@NonDet List<@Det String>}. Recall from~\ref{collection-rules} that \theDeterminismChecker does not allow this type on
+%as \codeid{@NonDet List<@Det String>}. Recall from~\cref{collection-rules} that \theDeterminismChecker does not allow this type on
 %any collection. While this is still sound, it will result in \theDeterminismChecker reporting a lot of errors because every locally declared 

### 2019-03-28 design.tex
More edits in section 2.1

@@
 
-    $\infer[\textsc{invariant type param}]{C_1\angles{\overline{\tau_i}}
+    $\infer[\textsc{invariant type arg}]{C_1\angles{\overline{\tau_i}}
       \sqsubseteq C_2\angles{\overline{\tau_i}}}{C_1 \sqsubseteq C_2}$
 
-    \caption{A sampling of standard type checking rules.  Most standard rules are omitted
+    \caption{A sampling of standard typing rules.  We omit other standard rules,
     for brevity.  $\sqsubseteq$ is overloaded for classes and types.}
@@
 
-    \caption{Type checking rules.  $\sqsubseteq$ is overloaded for classes,
+    \caption{Subtyping rules for \ourTypeSystem.  $\sqsubseteq$ is overloaded for classes,
     basetypes, type qualifiers, and types.}
@@
 
-\<OrderNonDet> may only be written on collections and maps.  Our implementation
-handles array types, and any class (including user-defined classes) that
-implements the \<Iterable> or \<Iterator> interfaces; this includes all
-Java collections such as \<List>s and \<Set>s.  For simplicity, and because
-the semantics and rules are similar even though the syntax varies, this
-paper uses the term ``collection'' and the type \<Collection> to represent
-all these types.
+\<OrderNonDet> may only be written on collections and maps.
 
 A map is a dictionary or associative array, such as a hash table.
-Our type system largely treats a map as a collection of key--value pairs.  Like a
-collection, a map may be \<Det>, \<OrderNonDet>, or \<NonDet>.  (In Java,
-\<HashMap> cannot be deterministic, but \<TreeMap> and \<LinkedHashMap>
-can.\todo{forward reference?})
+Our type system largely treats a map as a collection of key--value pairs.
 
-In \cref{code-determinism}, we present some of the JDK methods that we have annotated with our determinism types and give examples of
-client code that would produce errors at compile time.
-\begin{figure}
-    \begin{verbatim}
-    // Annotated JDK methods
-    public class Random implements java.io.Serializable {
-        public @NonDet Random() {}
-    }
-    public class PrintStream extends FilterOutputStream 
-        implements Appendable, Closeable {
-        public void println(@Det Object x) {}
-    }
-
-    // Client code
-    class Client {
-        void test() {
-            @Det double d = Math.random(); // Error - subtyping rules violated.
-            @NonDet double nd = Math.random(); // No error.
-            System.out.println(nd); // Error - println takes @Det arguments.
-        }
-    }
-    \end{verbatim}
-    \caption{Example: Errors detected by \theDeterminismChecker.}
-\todo{Use real examples of bugs from case studies, throughout.  Artificial
-  examples are less compelling (and lead the reader to wonder whether no
-  real example exist).  Put the first example of an error in a figure in
-  Section 1, to help with motivating the problem.}
-    \label{code-determinism}
-\end{figure}
+Both collections and map may be \<Det>, \<OrderNonDet>, or \<NonDet>, and
+the element types can be specified independently of the collection type,
+including qualifiers.
 

### 2019-03-28 java-design.tex
More edits in section 2.1

@@
 \<@Positive int>.
+
+Our implementation
+handles array types, and any class (including user-defined classes) that
+implements the \<Iterable> or \<Iterator> interfaces; this includes all
+Java collections such as \<List>s and \<Set>s.  For simplicity, and because
+the semantics and rules are similar even though the syntax varies, this
+paper uses the term ``collection'' and the type \<Collection> to represent
+all these types.
+
+(In Java,
+\<HashMap> cannot be deterministic, but \<TreeMap> and \<LinkedHashMap>
+can.\todo{forward reference?})
+
+\Cref{code-determinism} presents some of the JDK methods that we have annotated with our determinism types and give examples of
+client code that would produce errors at compile time.
+\begin{figure}
+    \begin{verbatim}
+    // Annotated JDK methods
+    public class Random implements java.io.Serializable {
+        public @NonDet Random() {}
+    }
+    public class PrintStream extends FilterOutputStream 
+        implements Appendable, Closeable {
+        public void println(@Det Object x) {}
+    }
+
+    // Client code
+    class Client {
+        void test() {
+            @Det double d = Math.random(); // Error - subtyping rules violated.
+            @NonDet double nd = Math.random(); // No error.
+            System.out.println(nd); // Error - println takes @Det arguments.
+        }
+    }
+    \end{verbatim}
+    \caption{Example: Errors detected by \theDeterminismChecker.}
+\todo{Use real examples of bugs from case studies, throughout.  Artificial
+  examples are less compelling (and lead the reader to wonder whether no
+  real example exist).  Put the first example of an error in a figure in
+  Section 1, to help with motivating the problem.}
+    \label{code-determinism}
+\end{figure}

### 2019-03-28 design.tex
Edits in section 2.2

@@
     \quad\quad
-    \infer[\textsc{subtyping}]{\Gamma \vdash x : \tau_2}{\Gamma \vdash x : \tau_1 & \tau_1 \sqsubseteq \tau_2}$
+    \infer[\rulename{subtyping}]{\Gamma \vdash x : \tau_2}{\Gamma \vdash x : \tau_1 & \tau_1 \sqsubseteq \tau_2}$
 
@@
 
-    $\infer[\textsc{method call}]{\|method|(\overline{a_i}) : \tau_r}{\tau_r \  \|method|(\overline{\tau_i \  p_i}) & \overline{a_i : \tau_i}}$
+    $\infer[\rulename{method call}]{\|method|(\overline{a_i}) : \tau_r}{\tau_r \  \|method|(\overline{\tau_i \  p_i}) & \overline{a_i : \tau_i}}$
 
@@
 
-    $\infer[\textsc{invariant type arg}]{C_1\angles{\overline{\tau_i}}
+    $\infer[\rulename{invariant type arg}]{C_1\angles{\overline{\tau_i}}
       \sqsubseteq C_2\angles{\overline{\tau_i}}}{C_1 \sqsubseteq C_2}$
@@
     \label{fig:determinism-hierarchy}
-\todo{Drop the at-signs (@).  The text is weirdly small for the box.
-  Probably just redraw this.}
+\todo{Drop the at-signs (@).}
+\todo{The text is weirdly small for the box.
+  Probably redraw this.}
 \end{wrapfigure}
@@
 A map is a dictionary or associative array, such as a hash table.
-Our type system largely treats a map as a collection of key--value pairs.
+\OurTypeSystem largely treats a map as a collection of key--value pairs.
 
@@
 
+\todo{A type $\tau$ is represented as $\kappa \ \beta$
+where $\kappa$ is a type qualifier (that is, \<NonDet>, \<OrderNonDet>, or
+\<Det>) and $\beta$ is a Java basetype.}
+
 \subsection{Collection types}\label{collection-rules}
 
-\todo{Note that the \textsc{ordernondet} versions also handle \<Det>,
+\todo{Note that the \rulename{ordernondet} versions also handle \<Det>,
   because $\<Det> \sqsubseteq \<OrderNonDet>$.}
@@
 \begin{figure}
-    $\infer[\textsc{any type}]{\wellformed{\kappa \  \beta}}{\kappa : \|Det / NonDet|}$
+    $\infer[\rulename{any type}]{\wellformed{\kappa \  \beta}}{\kappa : \|Det / NonDet|}$
 
@@
 
-    $\infer[\textsc{ordernondet collection}]{\wellformed{\kappa_c \  \beta_c \angles{\kappa_e \ \beta_e}}}{\kappa_c : \|OrderNonDet/Det| & \kappa_e \sqsubseteq \kappa_c & \beta_c \sqsubseteq \<Collection>}$
+    $\infer[\rulename{ordernondet collection}]{\wellformed{\kappa_c \  \beta_c \angles{\kappa_e \ \beta_e}}}{\kappa_c : \|OrderNonDet/Det| & \kappa_e \sqsubseteq \kappa_c & \beta_c \sqsubseteq \<Collection>}$
 
@@
 
-    $\infer[\textsc{nondet collection}]{\wellformed{\|NonDet| \ \beta_c\angles{\|NonDet| \ \beta_e}}}{\beta_c \sqsubseteq \<Collection>}$
+    $\infer[\rulename{nondet collection}]{\wellformed{\|NonDet| \ \beta_c\angles{\|NonDet| \ \beta_e}}}{\beta_c \sqsubseteq \<Collection>}$
 
@@
 
-%      $\infer[\textsc{ordernondet array}]{\wellformed{\kappa_e \ \beta_e \ \  \kappa_a []}}{\kappa_a : \|OrderNonDet/Det| & \kappa_e \sqsubseteq \kappa_a }$
+%      $\infer[\rulename{ordernondet array}]{\wellformed{\kappa_e \ \beta_e \ \  \kappa_a []}}{\kappa_a : \|OrderNonDet/Det| & \kappa_e \sqsubseteq \kappa_a }$
 %     
@@
 %     
-%     $\infer[\textsc{nondet array}]{\wellformed{\|NonDet| \ \beta_e \ \  \|NonDet| []}}{}$
+%     $\infer[\rulename{nondet array}]{\wellformed{\|NonDet| \ \beta_e \ \  \|NonDet| []}}{}$
 %     
@@
 
-    $\infer[\textsc{ordernondet map}]{\wellformed{\kappa_m \  \beta_m \angles{\kappa_k \ \beta_k, \kappa_v \ \beta_v}}}{\kappa_m: \|OrderNonDet/Det| & \kappa_k \sqsubseteq \kappa_m & \kappa_v \sqsubseteq \kappa_m & \beta_m \sqsubseteq \<Map>}$
+    $\infer[\rulename{ordernondet map}]{\wellformed{\kappa_m \  \beta_m \angles{\kappa_k \ \beta_k, \kappa_v \ \beta_v}}}{\kappa_m: \|OrderNonDet/Det| & \kappa_k \sqsubseteq \kappa_m & \kappa_v \sqsubseteq \kappa_m & \beta_m \sqsubseteq \<Map>}$
 
@@
 
-    $\infer[\textsc{nondet map}]{\wellformed{\|NonDet| \ \beta_m\angles{\|NonDet| \ \beta_k, \|NonDet| \ \beta_v}}}{\beta_m \sqsubseteq \<Map>}$
+    $\infer[\rulename{nondet map}]{\wellformed{\|NonDet| \ \beta_m\angles{\|NonDet| \ \beta_k, \|NonDet| \ \beta_v}}}{\beta_m \sqsubseteq \<Map>}$
 
     \caption{Type well formedness}
-\todo{The \textsc{any type} rule is not correct, because not every type is
+\todo{The \rulename{any type} rule is not correct, because not every type is
   well-formed.  It needs an antecedent stating that $\beta$ is not a
@@
   throughout.}
-rules of our type system. A type $\tau$ is represented as $\kappa \ \beta$
-where $\kappa$ is a type qualifier (that is, \<NonDet>, \<OrderNonDet>, or
-\<Det>) and $\beta$ is a Java basetype.
-As stated in the rule $\textsc{any type}$, the types \<Det> and \<NonDet>
+rules of \ourTypeSystem. 
+As stated in the rule \rulename{any type}, the types \<Det> and \<NonDet>
 can be written on any Java basetype.\todo{As noted in figure caption, that's
@@
 
-The determinism type qualifier on a collection must be a supertype or equal to
-its element type qualifier (Rule $\textsc{ordernondet collection}$ in \cref{fig-determinism-collections}).
-If the Collection is \<NonDet>, then the type argument may not be
-\<Det> or \<OrderNonDet> (Rule $\textsc{nondet collection}$ in \cref{type-validity}). Although such types could exist, they are
+\begin{figure}
+    \centering
+    \begin{tabular}{|l|l|l|l|l|}
+        \cline{3-5}
+        \multicolumn{2}{c|}{~}  &  \multicolumn{3}{c|}{element type} \\ \cline{3-5}
+        \multicolumn{2}{c|}{~}  & NonDet     & OrderNonDet & Det \\ \hline
+        & NonDet      &   valid    &  invalid    & invalid  \\ \cline{2-5}
+        Collection    & OrderNonDet &   invalid  &  valid      & valid  \\ \cline{2-5}
+        & Det         &   invalid  &  invalid    & valid      \\ \hline
+    \end{tabular}
+    \caption{Valid Collection declarations.  The Collection's type qualifier
+        must be a supertype or equal to the element type.}
+    \label{fig:determinism-collections}
+\end{figure}
+
+The element type of a \<NonDet> collection must be \<NonDet>.
+Otherwise, the type qualifier on a collection's element must be a subtype or equal to
+the collection's type qualifier (\rulename{collection}\ rules of
+\cref{type-validity}).
+\Cref{fig:determinism-collections} restates its restrictions, and here are examples.
+
+\smallskip
+\noindent
+\begin{minipage}{.48\textwidth}
+Some examples of valid types are:
+% Using "~" instead of "\ " does not work within \codeid :-(
+\begin{smaller}
+\begin{Verbatim}[commandchars=\\\{\}]
+NonDet      \lptype{List}{NonDet      Integer}
+OrderNonDet \lptype{List}{OrderNonDet \lptype{Set}{...}}
+OrderNonDet \lptype{List}{Det         Integer}
+Det         \lptype{List}{Det         Integer}
+
+\end{Verbatim}
+\end{smaller}
+\end{minipage}
+\hfill
+\begin{minipage}{.46\textwidth}
+These types are invalid:
+\begin{smaller}
+\begin{Verbatim}[commandchars=\\\{\}]
+NonDet      \lptype{List}{OrderNonDet \lptype{Set}{...}}
+NonDet      \lptype{List}{Det         Integer}
+OrderNonDet \lptype{List}{NonDet      Integer}
+Det         \lptype{List}{NonDet      Integer}
+Det         \lptype{List}{OrderNonDet \lptype{Set}{...}}
+\end{Verbatim}
+\end{smaller}
+\end{minipage}
+\smallskip
+
+ Although such types could exist, they are
 disallowed to prevent the following bug:
@@
 
-\smallskip
-\noindent
-\begin{minipage}{.48\textwidth}
-Some examples of valid types are:
-% Using "~" instead of "\ " does not work within \codeid :-(
-\begin{smaller}
-\begin{Verbatim}
-@NonDet      List<@NonDet      Integer>
-@OrderNonDet List<@OrderNonDet Set<...>>
-@OrderNonDet List<@Det         Integer>
-@Det         List<@Det         Integer>
-
-\end{Verbatim}
-\end{smaller}
-\end{minipage}
-\hfill
-\begin{minipage}{.46\textwidth}
-These types are invalid:
-\begin{smaller}
-\begin{Verbatim}
-@NonDet      List<@OrderNonDet Set<...>>
-@NonDet      List<@Det         Integer>
-@OrderNonDet List<@NonDet      Integer>
-@Det         List<@NonDet      Integer>
-@Det         List<@OrderNonDet Set<...>>
-\end{Verbatim}
-\end{smaller}
-\end{minipage}
-\smallskip
-
-\begin{figure}
-    \centering
-    \begin{tabular}{|l|l|l|l|l|}
-        \cline{3-5}
-        \multicolumn{2}{c|}{~}  &  \multicolumn{3}{c|}{element type} \\ \cline{3-5}
-        \multicolumn{2}{c|}{~}  & NonDet     & OrderNonDet & Det \\ \hline
-        & NonDet      &   valid    &  invalid    & invalid  \\ \cline{2-5}
-        Collection    & OrderNonDet &   invalid  &  valid      & valid  \\ \cline{2-5}
-        & Det         &   invalid  &  invalid    & valid      \\ \hline
-    \end{tabular}
-    \caption{Valid Collection declarations.  The Collection's type qualifier
-        must be a supertype or equal to the element type.}
-    \label{fig-determinism-collections}
-\end{figure}
-
 A collection or map annotated as \<OrderNonDet> has the following properties:
@@
 
-Our type system supports three types of polymorphism:  type
+\OurTypeSystem supports three types of polymorphism:  type
 polymorphism, basetype polymorphism, and qualifier polymorphism.
@@
 \begin{figure}
-    $\infer[\textsc{poly resolution}]{\pi_r \ \beta_r\ \<m>(\overline{\pi_i \ \beta_i})\{\ \}}{\forall \kappa \in \{\|NonDet, OrderNonDet, Det|\} : \kappa \ \beta_r\ \<m>(\overline{K \ \beta_i})\{\ \} & \pi_r, \pi_i \in \{\|PolyDet|, \|PolyDet|\uparrow, \|PolyDet|\downarrow    \}}$
+    $\infer[\rulename{poly resolution}]{\pi_r \ \beta_r\ \<m>(\overline{\pi_i \ \beta_i})\{\ \}}{\forall \kappa \in \{\|NonDet, OrderNonDet, Det|\} : \kappa \ \beta_r\ \<m>(\overline{K \ \beta_i})\{\ \} & \pi_r, \pi_i \in \{\|PolyDet|, \|PolyDet|\uparrow, \|PolyDet|\downarrow    \}}$
     \caption{Polymorphic resolution rule}
@@
 method parameters and return types including constructors.
-The rationale for this design choice was to not make our type checker too restrictive by making everything \<Det>.  This implies that
+The rationale for this design choice was to not make \theDeterminismChecker too restrictive by making everything \<Det>.  This implies that
 \theDeterminismChecker still allows some non-determinism as long as it doesn't break any determinism specifications explicitly written by the programmer.
@@
 \begin{figure}
-%     $\infer[\textsc{list sort}]{\Gamma \vdash x : \|Det \ List|\angles{\|Det| \ \beta}}{\Gamma \vdash x : \|OrderNonDet \ List|\angles{\|Det| \ \beta} & \|x.sort()|}$
+%     $\infer[\rulename{list sort}]{\Gamma \vdash x : \|Det \ List|\angles{\|Det| \ \beta}}{\Gamma \vdash x : \|OrderNonDet \ List|\angles{\|Det| \ \beta} & \|x.sort()|}$
 %     
@@
 
-    $\infer[\textsc{sort}]{\Gamma \vdash x : \|Det \ List|\angles{\|Det| \ \beta}}{\Gamma \vdash x : \|OrderNonDet \ List|\angles{\|Det| \ \beta} & \|Collections.sort(x)|}$
+    $\infer[\rulename{sort}]{\Gamma \vdash x : \|Det \ List|\angles{\|Det| \ \beta}}{\Gamma \vdash x : \|OrderNonDet \ List|\angles{\|Det| \ \beta} & \|Collections.sort(x)|}$
 
@@
 %     
-%     $\infer[\textsc{arrays sort}]{\Gamma \vdash x : \|Det \ \beta \ Det[]| }{\Gamma \vdash x : \|Det \ \beta \ OrderNonDet[]| & \|Arrays.sort( \ x)|}$
+%     $\infer[\rulename{arrays sort}]{\Gamma \vdash x : \|Det \ \beta \ Det[]| }{\Gamma \vdash x : \|Det \ \beta \ OrderNonDet[]| & \|Arrays.sort( \ x)|}$
 
@@
 \begin{figure}
-    $\infer[\textsc{hashmap creation}]{\|new \ \kappa\ HashMap()|}{\kappa = \|OrderDet/NonDet|}$
+    $\infer[\rulename{hashmap creation}]{\|new \ \kappa\ HashMap()|}{\kappa = \|OrderDet/NonDet|}$
 
@@
 
-    $\infer[\textsc{treemap creation}]{\|new \ \kappa\ TreeMap()|}{\kappa = \|Det/NonDet|}$
+    $\infer[\rulename{treemap creation}]{\|new \ \kappa\ TreeMap()|}{\kappa = \|Det/NonDet|}$
 
@@
 
-    $\infer[\textsc{hashset creation}]{\|new \ \kappa\ HashSet()|}{\kappa = \|OrderDet/NonDet|}$
+    $\infer[\rulename{hashset creation}]{\|new \ \kappa\ HashSet()|}{\kappa = \|OrderDet/NonDet|}$
 
@@
 
-    $\infer[\textsc{treeset creation}]{\|new \ \kappa\ TreeSet()|}{\kappa = \|Det/NonDet|}$
+    $\infer[\rulename{treeset creation}]{\|new \ \kappa\ TreeSet()|}{\kappa = \|Det/NonDet|}$
     \caption{\<Map> and \<Set> creation rules}
@@
     \todo{add line break}
-    $\infer[\textsc{set precision}]{\Gamma \vdash \|z : Det|}{\Gamma \vdash \|x| : \|OrderNonDet| \ \|Set|\angles{ \kappa_{x} \ \beta_{x}}, \|y| : \|OrderNonDet| \ \|Set|\angles{ \kappa_{y} \ \beta_{y}} & \|z = x.equals(y)| & \angles{ \kappa_{x} \ \beta_{x}} \nsqsubseteq \angles{ \|OrderNonDet \ List| } & \angles{ \kappa_{y} \ \beta_{y}} \nsqsubseteq \angles{ \|OrderNonDet \ List| } }$
+    $\infer[\rulename{set precision}]{\Gamma \vdash \|z : Det|}{\Gamma \vdash \|x| : \|OrderNonDet| \ \|Set|\angles{ \kappa_{x} \ \beta_{x}}, \|y| : \|OrderNonDet| \ \|Set|\angles{ \kappa_{y} \ \beta_{y}} & \|z = x.equals(y)| & \angles{ \kappa_{x} \ \beta_{x}} \nsqsubseteq \angles{ \|OrderNonDet \ List| } & \angles{ \kappa_{y} \ \beta_{y}} \nsqsubseteq \angles{ \|OrderNonDet \ List| } }$
 
@@
 
-    $\infer[\textsc{map precision}]{\Gamma \vdash \|z : Det|}{\Gamma \vdash \|x : OrderNonDet \ Map\angles{ Det \ \beta_k, \kappa_v\ \beta_v }, y : Det \ \beta| & \|z = x.get(y)|}$
+    $\infer[\rulename{map precision}]{\Gamma \vdash \|z : Det|}{\Gamma \vdash \|x : OrderNonDet \ Map\angles{ Det \ \beta_k, \kappa_v\ \beta_v }, y : Det \ \beta| & \|z = x.get(y)|}$
 

### 2019-03-28 introduction.tex
Edits in section 2.0
Edits in section 2.2

@@
 Programmers can write these to specify their program's behavior.
-Our type system contains a variety of other features that increase
+\OurTypeSystem contains a variety of other features that increase
 expressiveness and reduce false positive alarms.\todo{reference a section}
@@
     formal, machine-readable specification for the libraries, and it
-    demonstrates the richness of our type system.
+    demonstrates the richness of \ourTypeSystem.
 
@@
 
+\todo{It is essential that the introduction includes an example real-world
+  defect that \theDeterminismChecker found.}
+
 % LocalWords:  NONDEX NonDex DeFlaker Det OrderNonDet NonDet

### 2019-03-28 experiments.tex
nondex bullets

--- 2019-03-27 experiments.tex
@@
 \subsection{NONDEX case study}
+Methodology:
+\begin{itemize}
+    \item Objective: Measure recall.
+    \item Benchmarks from NONDEX paper.
+    \item That paper gives flaky test class and source of flakiness (nondeterministic method call).
+    \item Look at the source method under test.
+    \item annotate the local arrays and collections (This has been mostly sufficient for recall)
+\end{itemize}
+
+Discussion of results:
+
+\begin{itemize}
+    \item Refer to table and describe it.
+    \item P7: same source causing multiple tests to be flaky
+    \item Example for class.GetDeclaredMethod() and show annotations
+    \item Example for HashMap.entrySet() and show annotations.
+    \item qualitative comparison: orthogonal techniques (reporting tests vs reporting source locations).
+    \item From NonDex paper: "manually inspecting these failures was rather challenging"
+\end{itemize}
+
 \begin{table}[]
-    \begin{tabular}{|l|l|l|l|l|l|}
+    \begin{tabular}{|p{1.5cm}|p{1cm}|p{3cm}|p{4cm}|p{2cm}|p{1.5cm}|}
         \hline
-        Benchmark & Recall? & Error message source & # annotations & Type of annotation added & Comments \\ \hline
-        P1 &  &  &  &  &  \\
-        P2 &  &  &  &  &  \\
-        P3 &  &  &  &  &  \\ 
-        P4 &  &  &  &  &  \\ 
-        P5 &  &  &  &  &  \\ 
-        P5 &  &  &  &  &  \\ 
-        P6 &  &  &  &  &  \\ 
-        P6 &  &  &  &  &  \\
-        P7 &  &  &  &  &  \\ 
-        P7 &  &  &  &  &  \\
-        P7 &  &  &  &  &  \\
-        P7 &  &  &  &  &  \\
-        P7 &  &  &  &  &  \\
-        P8 &  &  &  &  &  \\
-        P8 &  &  &  &  &  \\
-        P8 &  &  &  &  &  \\
-        P9 &  &  &  &  &  \\
-        P10 &  &  &  &  &  \\
-        P10 &  &  &  &  &  \\
-        P11 &  &  &  &  &  \\
-        P12 &  &  &  &  &  \\
-        P13 &  &  &  &  &  \\
-        P14 &  &  &  &  &  \\
-        P15 &  &  &  &  &  \\
-        P15 &  &  &  &  &  \\
-        P16 &  &  &  &  &  \\
-        P17 &  &  &  &  &  \\
-        P18 &  &  &  &  &  \\
-        P19 &  &  &  &  &  \\
-        P19 &  &  &  &  &  \\
-        P19 &  &  &  &  &  \\
-        P20 &  &  &  &  &  \\
-        P20 &  &  &  &  &  \\
-        P21 &  &  &  &  &  \\
-        P21 &  &  &  &  &  \\ \hline
+        Benchmark & Recall? & Error message source & Error msg & \#annotations & Annotation type  \\ \hline
+        P1 & Yes & FieldAccess.java: [102,87] & assignment type incompatible &  &   \\
+        P2 & MI &  &  &  &  \\
+        P3 & Yes & DateTimeZone.java: [731,27] &  return type incompatible&  &  \\ 
+        P4 & MI &  &  &  &  \\ 
+        P5 & src &  &  &  &  \\ 
+        P5 & src &  &  &  &  \\ 
+        P6 & MI &  &  &  &  \\ 
+        P6 & MI &  &  &  &  \\
+        P7(test error) & Yes & checker/ tests/ determinism/P71.java: 5 & return type incompatible &  &  \\ 
+        P7(2) & Yes & FieldUtils.java: [219,84] & assignment type incompatible &  &  \\
+        P7(2) & Yes & Reflection ToStringBuilder.java: [543,65]  & assignment type incompatible &  &  \\
+        P7(2) & Yes & HashCodeBuilder.java: [193,69]  & assignment type incompatible &  &  \\
+        P7(2) & Yes & CompareToBuilder.java: [322,65] & assignment type incompatible &  &  \\
+        P7(2) & Yes & EqualsBuilder.java: [419,69]  & assignment type incompatible &  &  \\
+        P7 & Yes & FieldUtils.java: [200,69] & assignment type incompatible & &  \\
+        P7 & Yes & FieldUtils.java: [219,84] & assignment type incompatible &  &  \\
+        P7 & Yes & FieldUtils.java: [240,86] & assignment type incompatible &  &  \\
+        P8 & RM &  &  &  &  \\
+        P8 & RM &  &  &  &  \\
+        P8 & RM &  &  &  &  \\
+        P9 & MI &  &  &  &  \\
+        P10 & crash &  &  &  &  \\
+        P10 & crash &  &  &  &  \\
+        P11 & MI &  &  &  &  \\
+        P12 & MI &  &  &  &  \\
+        P13 & crash &  &  &  &  \\
+        P14 & MI &  &  &  &  \\
+        P15 & MI &  &  &  &  \\
+        P15 & MI &  &  &  &  \\
+        P16 & Yes & MethodSorter.java: [56,61]  & assignment type incompatible &  &  \\
+        P17 & Yes &  &  &  &  \\
+        P18 & src &  &  &  &  \\
+        P19 & MI &  &  &  &  \\
+        P19 & MI &  &  &  &  \\
+        P19 & MI &  &  &  &  \\
+        P20 & MI &  &  &  &  \\
+        P20 & MI &  &  &  &  \\
+        P21 & MI &  &  &  &  \\
+        P21 & MI &  &  &  &  \\ \hline
     \end{tabular}

### 2019-03-28 design.tex
Small changes

@@
 \begin{figure}
-    $\infer[\rulename{any type}]{\wellformed{\kappa \  \beta}}{\kappa : \|Det / NonDet|}$
+    $\infer[\rulename{noncollection}]
+    {\wellformed{\kappa \  \beta}}
+    {\kappa \in \{ \<Det>, \<NonDet> \} & \beta \not\sqsubseteq \<Collection> & \beta \not\sqsubseteq \<Map>}$
 
@@
 
-    $\infer[\rulename{ordernondet collection}]{\wellformed{\kappa_c \  \beta_c \angles{\kappa_e \ \beta_e}}}{\kappa_c : \|OrderNonDet/Det| & \kappa_e \sqsubseteq \kappa_c & \beta_c \sqsubseteq \<Collection>}$
+    $\infer[\rulename{ordernondet collection}]
+    {\wellformed{\kappa_c \  \beta_c \angles{\kappa_e \ \beta_e}}}
+    {\kappa_c \sqsubseteq \|OrderNonDet| & \kappa_e \sqsubseteq \kappa_c &
+      \beta_c \sqsubseteq \<Collection> & \wellformed{\kappa_e \ \beta_e}}$
 
@@
 
-%      $\infer[\rulename{ordernondet array}]{\wellformed{\kappa_e \ \beta_e \ \  \kappa_a []}}{\kappa_a : \|OrderNonDet/Det| & \kappa_e \sqsubseteq \kappa_a }$
+%      $\infer[\rulename{ordernondet array}]{\wellformed{\kappa_e \ \beta_e \ \  \kappa_a []}}{\kappa_a \sqsubseteq \|OrderNonDet| & \kappa_e \sqsubseteq \kappa_a }$
 %     
@@
 
-    $\infer[\rulename{ordernondet map}]{\wellformed{\kappa_m \  \beta_m \angles{\kappa_k \ \beta_k, \kappa_v \ \beta_v}}}{\kappa_m: \|OrderNonDet/Det| & \kappa_k \sqsubseteq \kappa_m & \kappa_v \sqsubseteq \kappa_m & \beta_m \sqsubseteq \<Map>}$
+    $\infer[\rulename{ordernondet map}]
+    {\wellformed{\kappa_m \  \beta_m \angles{\kappa_k \ \beta_k, \kappa_v \ \beta_v}}}
+    {\kappa_m \sqsubseteq \|OrderNonDet|
+      & \kappa_k \sqsubseteq \kappa_m
+      & \kappa_v \sqsubseteq \kappa_m
+      & \beta_m \sqsubseteq \<Map>
+      & \wellformed{\kappa_k \ \beta_k}
+      & \wellformed{\kappa_v \ \beta_v}
+    }$
 
@@
     \caption{Type well formedness}
-\todo{The \rulename{any type} rule is not correct, because not every type is
+\todo{The \rulename{noncollection} rule is not correct, because not every type is
   well-formed.  It needs an antecedent stating that $\beta$ is not a
@@
 
-\todo{The collection rules do not permit nested collections; similarly, the
+\todo{\todo{I think this will be OK, by redefining $\beta$.}
+  The collection rules do not permit nested collections; similarly, the
   (currently commented out) array rules did not permit any two-dimensional
@@
 
-\todo{``$\kappa : \|Det / NonDet|$'' should be ``$\kappa = \|Det /
-  NonDet|$'' or (better)
-  ``$\kappa \in \{\|Det|, \|NonDet|\}$''.
-  You could also express $\kappa_c : \|OrderNonDet/Det|$ as just
-  $\kappa_c \sqsubseteq \|OrderNonDet|$ since $\|Det| \sqsubseteq \|OrderNonDet|$.
-}
-
   \label{type-validity}
@@
 rules of \ourTypeSystem. 
-As stated in the rule \rulename{any type}, the types \<Det> and \<NonDet>
-can be written on any Java basetype.\todo{As noted in figure caption, that's
-  not true, in some contexts such as a type argument.}
+As stated in the rule \rulename{noncollection}, the types \<Det> and \<NonDet>
+can be written on any non-collection Java basetype.
 
@@
         \cline{3-5}
-        \multicolumn{2}{c|}{~}  &  \multicolumn{3}{c|}{element type} \\ \cline{3-5}
+        \multicolumn{2}{c|}{~}  &  \multicolumn{3}{c|}{Element type} \\ \cline{3-5}
         \multicolumn{2}{c|}{~}  & NonDet     & OrderNonDet & Det \\ \hline
                         & NonDet      &  valid     &  invalid    & invalid  \\ \cline{2-5}
-        Collection    & OrderNonDet &   invalid  &  valid      & valid  \\ \cline{2-5}
+        Collection type & OrderNonDet &  invalid   &  valid      & valid    \\ \cline{2-5}
                         & Det         &  invalid   &  invalid    & valid    \\ \hline
@@
 polymorphism, basetype polymorphism, and qualifier polymorphism.
-\todo{Give an example for each.}
+These apply to both classes and methods.
 \begin{itemize}
@@
 There are two valid instantiations for this declaration:
-\begin{enumerate}
-    \item @NonDet int plus(@NonDet int a, @NonDet int b) { return a+b; }
-    \item @Det int plus(@Det int a, @Det int b) { return a+b; }
-\end{enumerate}
+\begin{Verbatim}
+  @NonDet int plus(@NonDet int a, @NonDet int b) { return a+b; }
+  @Det int plus(@Det int a, @Det int b) { return a+b; }
+\end{Verbatim}
 This indicates that method \<foo> can be called with arguments having any of the \codeid{Det or NonDet} type qualifiers.
@@
 \Cref{fig-poly-resolution} presents the rule for type checking of methods annotated with \<PolyDet> or its variants.
-A polymorphic method type checks if every instantiation of \<PolyDet> type checks. If this method has \<PolyDet>$\uparrow$ or \<PolyDet>$\downarrow$
+A polymorphic method type checks if every instantiation of \<PolyDet> type checks. If this method has \PolyDetUp\ or \PolyDetDown\
 in its signature then the instantiation of \<PolyDet> at that location gets replaced by the appropriate type as shown in \cref{tab-poly-resolutions}.
@@
 \begin{figure}
-    $\infer[\rulename{poly resolution}]{\pi_r \ \beta_r\ \<m>(\overline{\pi_i \ \beta_i})\{\ \}}{\forall \kappa \in \{\|NonDet, OrderNonDet, Det|\} : \kappa \ \beta_r\ \<m>(\overline{K \ \beta_i})\{\ \} & \pi_r, \pi_i \in \{\|PolyDet|, \|PolyDet|\uparrow, \|PolyDet|\downarrow    \}}$
+    $\infer[\rulename{poly resolution}]{\pi_r \ \beta_r\ \<m>(\overline{\pi_i \ \beta_i})\{\ \}}{\forall \kappa \in \{\|NonDet, OrderNonDet, Det|\} : \kappa \ \beta_r\ \<m>(\overline{K \ \beta_i})\{\ \} & \pi_r, \pi_i \in \{\|PolyDet|, \PolyDetUp, \PolyDetDown    \}}$
     \caption{Polymorphic resolution rule}
@@
 irrespective of the type of \<index>, thereby preventing undesired side-effects.
-
-\subsection{Discussion on defaulting rules}\label{defaulting}
-
-\TheDeterminismChecker applies a default qualifier at each unannotated type
-use (except uses of type variables, which already stand for a type that was
-defaulted at the instantiation site).
-This does not change the expressivity of the type system; it merely makes
-the system more convenient to use.
-
-In \theDeterminismChecker, we chose \<PolyDet> as the default for all 
-method parameters and return types including constructors.
-The rationale for this design choice was to not make \theDeterminismChecker too restrictive by making everything \<Det>.  This implies that
-\theDeterminismChecker still allows some non-determinism as long as it doesn't break any determinism specifications explicitly written by the programmer.
-
-\begin{verbatim}
-@PolyDet int foo(@PolyDet int x) { ... }
-
-void bar(@NonDet int random, @Det int value) {
-    @NonDet int x = foo(random);      // No error.
-    @Det int y = foo(value);          // No error.
-    System.out.println(foo(random));  // Error.
-}
-\end{verbatim}
-In the example above, the method \codeid{foo} is annotated with the \<PolyDet> type qualifier which allows
-the caller method \codeid{bar} to invoke it with both \<Det> and \<NonDet> arguments. \TheDeterminismChecker flags an error 
-only when the returned value is used in a way that breaks any deterministic specification (Printing a \<NonDet> value
-breaks our determinism guarantees).
 
@@
 \begin{figure}
-    $\infer[\rulename{hashmap creation}]{\|new \ \kappa\ HashMap()|}{\kappa = \|OrderDet/NonDet|}$
+    $\infer[\rulename{hashmap creation}]{\|new \ \kappa\ HashMap()|}{\kappa \in \{ \<OrderNonDet>, \<NonDet> \}}$
 
@@
 
-    $\infer[\rulename{treemap creation}]{\|new \ \kappa\ TreeMap()|}{\kappa = \|Det/NonDet|}$
+    $\infer[\rulename{treemap creation}]{\|new \ \kappa\ TreeMap()|}{\kappa \in \{ \<Det>, \<NonDet> \}}$
 
@@
 
-    $\infer[\rulename{hashset creation}]{\|new \ \kappa\ HashSet()|}{\kappa = \|OrderDet/NonDet|}$
+    $\infer[\rulename{hashset creation}]{\|new \ \kappa\ HashSet()|}{\kappa \in \{ \<OrderNonDet>, \<NonDet> \}}$
 
@@
 
-    $\infer[\rulename{treeset creation}]{\|new \ \kappa\ TreeSet()|}{\kappa = \|Det/NonDet|}$
+    $\infer[\rulename{treeset creation}]{\|new \ \kappa\ TreeSet()|}{\kappa \in \{ \<Det>, \<NonDet> \}}$
     \caption{\<Map> and \<Set> creation rules}
@@
 \begin{verbatim}
-PolyDet$\uparrow$ boolean equals(PolyDet Set<E> this, PolyDet Object o)
+PolyDet("down") boolean equals(PolyDet Set<E> this, PolyDet Object o)
 \end{verbatim}
@@
 
-Similarly, the \codeid{get()} method in the \<Map> class is annotated with polymorphic qualifiers and the rule $\sc{\|MAP \ PRECISION|}$ in \cref{fig-precision-rules} improves precision for \<OrderNonDet> receivers.
+Similarly, the \codeid{get()} method in the \<Map> class is annotated with polymorphic qualifiers and the rule \rulename{map precision}\ in \cref{fig-precision-rules} improves precision for \<OrderNonDet> receivers.
 
@@
 
-\subsection{Extending \theDeterminismChecker}\label{extending-det-checker}
+\subsection{Extending \ourTypeSystem}\label{extending-det-checker}
 We have primarily focused on collections and annotated the corresponding classes in the JDK.

### 2019-03-28 java-design.tex
Small changes

@@
 \end{figure}
+
+\begin{itemize}
+\item
+In \theDeterminismCheckerImplementation,
+type polymorphism is handled by Java's generics mechanism, which
+\theDeterminismChecker fully supports, including class and method generics,
+inference, etc.
+Given the Java declaration
+\begin{Verbatim}
+<T> T identity(T arg) { return arg; }
+\end{Verbatim}
+the type of \<identity(arg)> is the same as the type of
+\<arg>, and a programmer can write a type such as \codeid{List<@NonDet
+  Integer>} (but see below for details about collections).
+\item
+Basetype polymorphism is enabled by writing a type qualifier on a use of a
+type variable, which overrides the type qualifier at the instantiation
+site.
+For example, a \<choose> operation on sets could be defined in Java as
+\begin{Verbatim}
+class ChooseableSet<T> implements Set<T> {
+  // return an arbitrary element from the set
+  @NonDet T choose();
+}
+\end{Verbatim}
+\item
+Java does not provide a syntax that can be used for qualifier polymorphism,
+so we use a special type qualifier name, \<@PolyDet>.
+A qualifier-polymorphic method \<m> with signature $\forall \kappa . \kappa\ \<int> \times \<Det boolean> \rightarrow
+\kappa\ \<String>$ would be written in our dialect of Java as
+\begin{Verbatim}
+@PolyDet String m(@PolyDet int, @Det boolean)
+\end{Verbatim}
+Each use of \<@PolyDet> stands for a use of the qualifier variable
+$\kappa$, and there is no need to declare the qualifier variable $\kappa$.
+\TheDeterminismChecker currently supports
+qualifier
+polymorphism on methods but not on classes.
+Therefore, \<@PolyDet> may be written in methods (signatures and bodies)
+but not on fields.
+\end{itemize}
+
+\subsection{Defaulting rules}\label{defaulting}
+
+\TheDeterminismChecker applies a default qualifier at each unannotated type
+use (except uses of type variables, which already stand for a type that was
+defaulted at the instantiation site).
+This does not change the expressivity of the type system; it merely makes
+the system more convenient to use.
+
+In \theDeterminismChecker, we chose \<PolyDet> as the default for all 
+method parameters and return types including constructors.
+The rationale for this design choice was to not make \theDeterminismChecker too restrictive by making everything \<Det>.  This implies that
+\theDeterminismChecker still allows some non-determinism as long as it doesn't break any determinism specifications explicitly written by the programmer.
+
+\begin{verbatim}
+@PolyDet int foo(@PolyDet int x) { ... }
+
+void bar(@NonDet int random, @Det int value) {
+    @NonDet int x = foo(random);      // No error.
+    @Det int y = foo(value);          // No error.
+    System.out.println(foo(random));  // Error.
+}
+\end{verbatim}
+In the example above, the method \codeid{foo} is annotated with the \<PolyDet> type qualifier which allows
+the caller method \codeid{bar} to invoke it with both \<Det> and \<NonDet> arguments. \TheDeterminismChecker flags an error 
+only when the returned value is used in a way that breaks any deterministic specification (Printing a \<NonDet> value
+breaks our determinism guarantees).

### 2019-03-28 design.tex
Edits in sections 2.3 and 2.4

@@
 
+\todo{Define the Greek.  For example:
+
+  A type $\tau = \kappa\ \beta$ consists of a qualifier and a basetype.
+  Neither may be omitted.
+
+  $\beta$ is a type that lacks its type qualifier, but has type qualifiers
+  on any type arguments:
+  $\beta = C | C\angles{\overline{\tau}}$.
+  }
+
 \subsection{Determinism types}\label{type-hierarchy}
@@
 
-    $\infer[\rulename{nondet collection}]{\wellformed{\|NonDet| \ \beta_c\angles{\|NonDet| \ \beta_e}}}{\beta_c \sqsubseteq \<Collection>}$
+    $\infer[\rulename{nondet collection}]
+    {\wellformed{\|NonDet| \ C\angles{\|NonDet| \ \beta_e}}}
+    {C \sqsubseteq \<Collection> & \wellformed{\|NonDet| \ \beta_e}}$
 
@@
 
-    $\infer[\rulename{nondet map}]{\wellformed{\|NonDet| \ \beta_m\angles{\|NonDet| \ \beta_k, \|NonDet| \ \beta_v}}}{\beta_m \sqsubseteq \<Map>}$
+    $\infer[\rulename{nondet map}]
+    {\wellformed{\|NonDet| \ \beta_m\angles{\|NonDet| \ \beta_k, \|NonDet| \ \beta_v}}}
+    {\beta_m \sqsubseteq \<Map> & \wellformed{\|NonDet| \ \beta_k} & \wellformed{\|NonDet| \ \beta_v}}$
 
     \caption{Type well formedness}
-\todo{The \rulename{noncollection} rule is not correct, because not every type is
-  well-formed.  It needs an antecedent stating that $\beta$ is not a
-  collection or map.}
 
@@
 These apply to both classes and methods.
+\todo{Greekify the following}
 \begin{itemize}

### 2019-03-28 experiments.tex
todo: redraw figure

@@
     \item From NonDex paper: "manually inspecting these failures was rather challenging"
+    \item Note: P7\_1 and P18: error in test case, so we added the same test to our microbenchmarks
 \end{itemize}
@@
         P4 & MI &  &  &  &  \\ 
-        P5 & src &  &  &  &  \\ 
+        P5 & visualee &  &  &  &  \\ 
         P5 & src &  &  &  &  \\ 
@@
         P17 & Yes &  &  &  &  \\
-        P18 & src &  &  &  &  \\
+        P18 & slf4j &  &  &  &  \\
         P19 & MI &  &  &  &  \\

### 2019-03-28 design.tex
todo: redraw figure
todo: type welformedness consistency
todo: resolve -> instantiate

@@
     \begin{center}
-        \includegraphics[scale=0.5]{determinism}
+        \includegraphics[scale=0.37]{detHierarchy}
     \end{center}
@@
     \label{fig:determinism-hierarchy}
-\todo{Drop the at-signs (@).}
-\todo{The text is weirdly small for the box.
-  Probably redraw this.}
+%\todo{Drop the at-signs (@).}
+%\todo{The text is weirdly small for the box.
+%  Probably redraw this.}
 \end{wrapfigure}
@@
 
-\Cref{type-validity} shows the type validity\todo{The caption says
-  ``type well formedness''.  Choose one term and use it consistently
-  throughout.}
+\Cref{type-validity} shows the type well formedness
+%\todo{The caption says
+%  ``type well formedness''.  Choose one term and use it consistently
+%  throughout.}
 rules of \ourTypeSystem. 
@@
 When a user writes a polymorphic qualifier on a method signature or a type parameter,
-it indicates that it could resolve\todo{Should all occurrences of ``resolve'' be ``can be instantiated''} to any type in the type system depending on how it is used.
+it indicates that it could be instantiated
+%\todo{Should all occurrences of ``resolve'' be ``can be instantiated''} 
+to any type in the type system depending on how it is used.
 In \theDeterminismChecker, we define a polymorphic qualifier \<PolyDet>.
@@
 %   (\<@OrderNonDet> is not allowed because it is invalid on primitive types).
- \<PolyDet> resolves to the least upper bound of
+ \<PolyDet> gets instantiated to the least upper bound of
 the actual types on arguments.
-For instance, if this method is called as \codeid{foo(@Det String arg1, @NonDet boolean arg2)}, \theDeterminismChecker resolves the method declaration as \codeid{@NonDet int foo(@NonDet String param1, @NonDet boolean param2)}
+For instance, if this method is called as \codeid{foo(@Det String arg1, @NonDet boolean arg2)}, \theDeterminismChecker instantiates the method declaration as \codeid{@NonDet int foo(@NonDet String param1, @NonDet boolean param2)}
 causing the return type to have the type qualifier \<NonDet>.
@@
 Suppose a method declaration is annotated as \codeid{PolyDet int size(PolyDet List<T> list)}. Invoking this method
-with an argument of type \codeid{OrderNonDet List<Det String>} will result in \<PolyDet> getting resolved to \<OrderNonDet> which is invalid
+with an argument of type \codeid{OrderNonDet List<Det String>} will result in \<PolyDet> getting instantiated to \<OrderNonDet> which is invalid
 on the return type \<int>. We circumvent this scenario by introducing two variants to the \<PolyDet> qualifier, \PolyDetUp
@@
 \begin{itemize}
-    \item \PolyDetUp: Same as \<PolyDet> except when it resolves to \<OrderNonDet>, it is replaced with \<NonDet>.
-    \item \PolyDetDown: Same as \<PolyDet> except when it resolves to \<OrderNonDet>, it is replaced with \<Det>.
+    \item \PolyDetUp: Same as \<PolyDet> except when it is instantiated as \<OrderNonDet>, it is replaced with \<NonDet>.
+    \item \PolyDetDown: Same as \<PolyDet> except when it is instantiated as \<OrderNonDet>, it is replaced with \<Det>.
 \end{itemize} 
@@
 Since \<PolyDet("use")> has no effect on type instantiations, calling this method on a \<Det List>
-will result in the method declaration being resolved as \codeid{public void add(@Det ArrayList<E> this, @Det int index, E element)}
+will result in the method declaration being instantiated as \codeid{public void add(@Det ArrayList<E> this, @Det int index, E element)}
 irrespective of the type of \<index>, thereby preventing undesired side-effects.
@@
 According to this rule, if the \<Set> receiver has type \codeid{@OrderNonDet Set<@Det List} and the argument to the \codeid{equals()} method is
-also of the same type, the return type gets the qualifier \<Det>. Without this precision rule, the return type would have resolved to \<NonDet> which
+also of the same type, the return type gets the qualifier \<Det>. Without this precision rule, the return type would have been instantiated as \<NonDet> which
 is sound but imprecise. This rule is careful enough to not annotate the return type of \<equals()> as \<Det> if a type argument to either the

### 2019-03-28 design.tex
typo

@@
   be lowercase in this section??} and a type (which combines a qualifier
-anda basetype) is \<Positive int>.
+and a basetype) is \<Positive int>.
 A polymorphic type such as \<List> can be instantiated by a type argument,

### 2019-03-28 experiments.tex
typo

@@
         P5 & visualee &  &  &  &  \\ 
-        P5 & src &  &  &  &  \\ 
+        P5 & visualee &  &  &  &  \\ 
         P6 & MI &  &  &  &  \\ 
@@
         P17 & Yes &  &  &  &  \\
-        P18 & slf4j &  &  &  &  \\
+        P18 & Arr Iter &  &  &  &  \\
         P19 & MI &  &  &  &  \\

### 2019-03-28 design.tex
iteration rules
todo: type rules

@@
 
- Although such types could exist, they are
+ Although \<NonDet> collections containing \<Det> or \<OrderNonDet> elements could exist, they are
 disallowed to prevent the following bug:
@@
 A collection or map annotated as \<OrderNonDet> has the following properties:
-\todo{Write these as type rules.}
+%\todo{Write these as type rules.}
 \begin{enumerate}
-    \item The individual elements extracted by iterating over the collection get the type \<NonDet>.
+    \item The individual elements extracted by iterating over the collection get the type \<NonDet> (Rules \rulename{iteration} and \rulename{next element} in Figure~\ref{fig-ordernondet-rules}). 
+    Rule \rulename{hasnext element} states that querying if the iterator of an \<OrderNonDet Collection> \<hasNext()> will result in a \<Det boolean>.
     \item Accessing properties such as \<size()> or querying if the collection \<isEmpty()> will return types
@@
 \end{enumerate}
+
+\begin{figure}
+    $\infer[\rulename{iteration}]
+    {\Gamma \vdash \|y| : \|OrderNonDet\ Iterator|\angles{\tau}}
+    {\Gamma \vdash \|x| : \|OrderNonDet \ C\angles{\tau}|
+    & C \sqsubseteq \|Collection|
+    & \|y = C.iterator()|
+    & \wellformed{\tau}
+    }$
+
+    \bigskip
+
+    $\infer[\rulename{next element}]
+    {\Gamma \vdash \|z| : \|NonDet \ \beta|}
+    {\Gamma \vdash \|y| : \|OrderNonDet \ Iterator\angles{\kappa \ \beta}|
+        & \|z = y.next()|
+        & \wellformed{\kappa \ \beta}
+    }$
+
+    \bigskip
+
+    $\infer[\rulename{hasnext element}]
+    {\Gamma \vdash \|z| : \|Det \ boolean|}
+    {\Gamma \vdash \|y| : \|OrderNonDet \ Iterator\angles{\tau}|
+        & \|z = y.hasNext()|
+        & \wellformed{\tau}
+    }$
+
+    \bigskip
+
+    $\infer[\rulename{size}]
+    {\Gamma \vdash \|y| : \|Det \ int|}
+    {\Gamma \vdash \|x| : \|OrderNonDet \ C\angles{\tau}|
+        & C \sqsubseteq \|Collection|
+        & \|y = x.size()|
+        & \wellformed{\tau}
+    }$
+\caption{\<OrderNonDet> Collection rules}
+\label{fig-ordernondet-rules}
+\end{figure}
 

### 2019-03-29 design.tex
todo: replace table with rules
todo: define greek

@@
 
-\todo{Greek letters go after this paragraph.}
+%\todo{Greek letters go after this paragraph.}
 
@@
 
-\todo{Define the Greek.  For example:
+%\todo{Define the Greek.  For example:
+%
+%  A type $\tau = \kappa\ \beta$ consists of a qualifier and a basetype.
+%  Neither may be omitted.
+%
+%  $\beta$ is a type that lacks its type qualifier, but has type qualifiers
+%  on any type arguments:
+%  $\beta = C | C\angles{\overline{\tau}}$.
+%  }
 
-  A type $\tau = \kappa\ \beta$ consists of a qualifier and a basetype.
-  Neither may be omitted.
-
-  $\beta$ is a type that lacks its type qualifier, but has type qualifiers
-  on any type arguments:
-  $\beta = C | C\angles{\overline{\tau}}$.
-  }
+Figure~\ref{typecheck-rules-standard} shows a sample of the standard typing rules applied to 
+any typed language. In the generic case, the type $\tau$ represents a valid basetype in a programming language.
+\ourTypeSystem extends a basetype with a type qualifier. Throughout this paper, $\tau$ represents a fully qualified type.
+In other words, type $\tau = \kappa \ \beta$ consists of both a qualifier $\kappa$ and a basetype $\beta$, neither of which may be omitted.
+Further, $\beta$ is a type without a type qualifier, but which has type qualifiers on all its type parameters.
+Formally, $\beta ::= C|C \angles{\tau}$ where $C$ represents a Java basetype.
 
@@
 
-\begin{table}[]
-    \begin{tabular}{|l|l|l|}
-        \hline
-        \textbf{\<PolyDet>} & \textbf{\PolyDetUp} & \textbf{\PolyDetDown} \\ \hline
-        \<NonDet> & \<NonDet> &  \<NonDet>\\ \hline
-        \<OrderNonDet> & \<NonDet> &  \<Det>\\ \hline
-        \<Det> & \<Det> &  \<Det>\\ \hline
-    \end{tabular}
+%\begin{table}[]
+%    \begin{tabular}{|l|l|l|}
+%        \hline
+%        \textbf{\<PolyDet>} & \textbf{\PolyDetUp} & \textbf{\PolyDetDown} \\ \hline
+%        \<NonDet> & \<NonDet> &  \<NonDet>\\ \hline
+%        \<OrderNonDet> & \<NonDet> &  \<Det>\\ \hline
+%        \<Det> & \<Det> &  \<Det>\\ \hline
+%    \end{tabular}
+%\caption{Polymorphic resolution}
+%\todo{Write this as a rule rather than a table.}
+%\label{tab-poly-resolutions}
+%\end{table}
+
+\begin{figure}
+    $\infer[\rulename{ordernondet polyup}]
+    {\tau \uparrow \ = \|NonDet \ \beta|}
+    {\tau \ = \|OrderNonNonDet \ \beta|
+        & \wellformed{\tau}
+    }$
+
+    \bigskip
+
+    $\infer[\rulename{ordernondet polydown}]
+    {\tau \downarrow \ = \|Det \ \beta|}
+    {\tau \ = \|OrderNonNonDet \ \beta|
+        & \wellformed{\tau}
+    }$
+
+    \bigskip
+
+    $\infer[\rulename{polyup}]
+    {\tau \uparrow \ = \kappa \ \beta}
+    {\tau \ =\kappa \ \beta
+        & \kappa \in \{\|Det, NonDet|\}
+        & \wellformed{\tau}
+    }$
+
+    \bigskip
+
+    $\infer[\rulename{polydown}]
+    {\tau \downarrow \ = \kappa \ \beta}
+    {\tau \ =\kappa \ \beta
+        & \kappa \in \{\|Det, NonDet|\}
+        & \wellformed{\tau}
+    }$
 \caption{Polymorphic resolution}
-\todo{Write this as a rule rather than a table.}
 \label{tab-poly-resolutions}
-\end{table}
+\end{figure}
 

### 2019-03-29 design.tex
todo: collection types

@@
 
-Figure~\ref{typecheck-rules-standard} shows a sample of the standard typing rules applied to 
+Figure~\cref{typecheck-rules-standard} shows a sample of the standard typing rules applied to 
 any typed language. In the generic case, the type $\tau$ represents a valid basetype in a programming language.
@@
 Further, $\beta$ is a type without a type qualifier, but which has type qualifiers on all its type parameters.
-Formally, $\beta ::= C|C \angles{\tau}$ where $C$ represents a Java basetype.
+Formally, $\beta ::= C|C \angles{\tau}$ where $C$ represents a valid Java basetype.
+\ourTypeSystem checks all the standard typing rules including the ones shown in Figure~\cref{typecheck-rules-standard} with
+the only difference of having an additional type qualifier on each of the basetypes. The type qualifiers can only 
+constrain the set of valid types. As a result, \ourTypeSystem doesn't allow any behavior that the original type system doesn't.
 
@@
 
-\todo{A type $\tau$ is represented as $\kappa \ \beta$
-where $\kappa$ is a type qualifier (that is, \<NonDet>, \<OrderNonDet>, or
-\<Det>) and $\beta$ is a Java basetype.}
+%\todo{A type $\tau$ is represented as $\kappa \ \beta$
+%where $\kappa$ is a type qualifier (that is, \<NonDet>, \<OrderNonDet>, or
+%\<Det>) and $\beta$ is a Java basetype.}
 
@@
 
-\todo{Note that the \rulename{ordernondet} versions also handle \<Det>,
-  because $\<Det> \sqsubseteq \<OrderNonDet>$.}
+%\todo{Note that the \rulename{ordernondet} versions also handle \<Det>,
+%  because $\<Det> \sqsubseteq \<OrderNonDet>$.}
 
@@
 
-\todo{\todo{I think this will be OK, by redefining $\beta$.}
-  The collection rules do not permit nested collections; similarly, the
-  (currently commented out) array rules did not permit any two-dimensional
-  array to be well-formed.  There is a bigger issue here.
-  Consider the ``nondet collection'' rule.  $\beta_c$ is a class (we need
-  to use a different symbol for it, probably $c$), but $\beta_e$ should be
-  a full type, which might contain type qualifiers within it.  (Actually,
-  $\|NonDet|\ \beta_e$ is a type, not $\beta_e$ itself.)  So, there should
-  be an antecedent stating that $\|NonDet| \beta_e$ is well-formed, and
-  for the ``ordernondet collection'' rule stating that $\kappa_e \beta_e$
-  is well-formed.  Likewise, antecedents are needed for the map rules.
-  Otherwise, these rules permit construction of $\<Map>\angles{\tau'}$
-  where $\tau'$ is itself not well-defined.}
+%\todo{\todo{I think this will be OK, by redefining $\beta$.}
+%  The collection rules do not permit nested collections; similarly, the
+%  (currently commented out) array rules did not permit any two-dimensional
+%  array to be well-formed.  There is a bigger issue here.
+%  Consider the ``nondet collection'' rule.  $\beta_c$ is a class (we need
+%  to use a different symbol for it, probably $c$), but $\beta_e$ should be
+%  a full type, which might contain type qualifiers within it.  (Actually,
+%  $\|NonDet|\ \beta_e$ is a type, not $\beta_e$ itself.)  So, there should
+%  be an antecedent stating that $\|NonDet| \beta_e$ is well-formed, and
+%  for the ``ordernondet collection'' rule stating that $\kappa_e \beta_e$
+%  is well-formed.  Likewise, antecedents are needed for the map rules.
+%  Otherwise, these rules permit construction of $\<Map>\angles{\tau'}$
+%  where $\tau'$ is itself not well-defined.}
 
@@
 the collection's type qualifier (\rulename{collection}\ rules of
-\cref{type-validity}).
+\cref{type-validity}). Note that the \rulename{ordernondet} versions also handle \<Det>,
+because $\<Det> \sqsubseteq \<OrderNonDet>$.
 \Cref{fig:determinism-collections} restates its restrictions, and here are examples.

### 2019-03-29 design.tex
ordernondet collection rules
todo: poly resolution rules

@@
 \begin{figure}
-    $\infer[\rulename{iteration}]
-    {\Gamma \vdash \|y| : \|OrderNonDet\ Iterator|\angles{\tau}}
-    {\Gamma \vdash \|x| : \|OrderNonDet \ C\angles{\tau}|
-    & C \sqsubseteq \|Collection|
-    & \|y = C.iterator()|
-    & \wellformed{\tau}
-    }$
+%    $\infer[\rulename{iteration}]
+%    {\Gamma \vdash \|y| : \|OrderNonDet\ Iterator|\angles{\tau}}
+%    {\Gamma \vdash \|x| : \|OrderNonDet \ C\angles{\tau}|
+%    & C \sqsubseteq \|Collection|
+%    & \|y = C.iterator()|
+%    & \wellformed{\tau}
+%    }$
+
+    \rulename{iteration}: 
+    $
+    \forall \tau . \|OrderNonDet \ C| \angles{\tau} \rightarrow 
+    \|OrderNonDet \ Iterator| \angles{\tau}
+    $
 
@@
 
-    $\infer[\rulename{next element}]
-    {\Gamma \vdash \|z| : \|NonDet \ \beta|}
-    {\Gamma \vdash \|y| : \|OrderNonDet \ Iterator\angles{\kappa \ \beta}|
-        & \|z = y.next()|
-        & \wellformed{\kappa \ \beta}
-    }$
+%    $\infer[\rulename{next element}]
+%    {\Gamma \vdash \|z| : \|NonDet \ \beta|}
+%    {\Gamma \vdash \|y| : \|OrderNonDet \ Iterator\angles{\kappa \ \beta}|
+%        & \|z = y.next()|
+%        & \wellformed{\kappa \ \beta}
+%    }$
+
+    \rulename{next element}:
+    $
+    \forall \kappa \beta . \|OrderNonDet \ Iterator|\angles{\kappa \ \beta} \rightarrow
+    \|NonDet|\ \beta
+    $
 
@@
 
-    $\infer[\rulename{hasnext element}]
-    {\Gamma \vdash \|z| : \|Det \ boolean|}
-    {\Gamma \vdash \|y| : \|OrderNonDet \ Iterator\angles{\tau}|
-        & \|z = y.hasNext()|
-        & \wellformed{\tau}
-    }$
+%    $\infer[\rulename{hasnext element}]
+%    {\Gamma \vdash \|z| : \|Det \ boolean|}
+%    {\Gamma \vdash \|y| : \|OrderNonDet \ Iterator\angles{\tau}|
+%        & \|z = y.hasNext()|
+%        & \wellformed{\tau}
+%    }$
 
+    \rulename{hasnext element}:
+    $
+    \forall \tau . \|OrderNonDet \ Iterator|\angles{\tau} \rightarrow
+    \|Det \ boolean|
+    $
     \bigskip
 
-    $\infer[\rulename{size}]
-    {\Gamma \vdash \|y| : \|Det \ int|}
-    {\Gamma \vdash \|x| : \|OrderNonDet \ C\angles{\tau}|
-        & C \sqsubseteq \|Collection|
-        & \|y = x.size()|
-        & \wellformed{\tau}
-    }$
+%    $\infer[\rulename{size}]
+%    {\Gamma \vdash \|y| : \|Det \ int|}
+%    {\Gamma \vdash \|x| : \|OrderNonDet \ C\angles{\tau}|
+%        & C \sqsubseteq \|Collection|
+%        & \|y = x.size()|
+%        & \wellformed{\tau}
+%    }$
+
+    \rulename{size}:
+    $
+    \forall \tau . \|OrderNonDet \ C|\angles{\tau} \rightarrow
+    \|Det \ int|
+    $
+
 \caption{\<OrderNonDet> Collection rules}
@@
 
-\Cref{fig-poly-resolution} presents the rule for type checking of methods annotated with \<PolyDet> or its variants.
+%\Cref{fig-poly-resolution} presents the rule for type checking of methods annotated with \<PolyDet> or its variants.
 A polymorphic method type checks if every instantiation of \<PolyDet> type checks. If this method has \PolyDetUp\ or \PolyDetDown\
-in its signature then the instantiation of \<PolyDet> at that location gets replaced by the appropriate type as shown in \cref{tab-poly-resolutions}.
+in its signature then the instantiation of \<PolyDet> at that location gets replaced by the appropriate type as shown in \cref{fig:poly-resolutions}.
 
-Annotating \<size()> as \codeid{PolyDet$\downarrow$ int size(PolyDet List<T> list)} will only allow the following type resolutions:
+Annotating \<size()> as \codeid{PolyDet$\downarrow$ int size(PolyDet List<T> list)} will only allow the following type instantiations:
 \begin{itemize}
@@
 
-\begin{figure}
-    $\infer[\rulename{poly resolution}]{\pi_r \ \beta_r\ \<m>(\overline{\pi_i \ \beta_i})\{\ \}}{\forall \kappa \in \{\|NonDet, OrderNonDet, Det|\} : \kappa \ \beta_r\ \<m>(\overline{K \ \beta_i})\{\ \} & \pi_r, \pi_i \in \{\|PolyDet|, \PolyDetUp, \PolyDetDown    \}}$
-    \caption{Polymorphic resolution rule}
-    \todo{I'm not sure what this is doing.  What code does it reject?}
-    \todo{This suggests that \PolyDetUp is a qualifier, rather than
-      $\uparrow$ being an operator on qualifiers.  We should clarify which
-      it is.}
-    \label{fig-poly-resolution}
-\end{figure}
+%\begin{figure}
+%    $\infer[\rulename{poly resolution}]{\pi_r \ \beta_r\ \<m>(\overline{\pi_i \ \beta_i})\{\ \}}{\forall \kappa \in \{\|NonDet, OrderNonDet, Det|\} : \kappa \ \beta_r\ \<m>(\overline{K \ \beta_i})\{\ \} & \pi_r, \pi_i \in \{\|PolyDet|, \PolyDetUp, \PolyDetDown    \}}$
+%    \caption{Polymorphic resolution rule}
+%    \todo{I'm not sure what this is doing.  What code does it reject?}
+%    \todo{This suggests that \PolyDetUp is a qualifier, rather than
+%      $\uparrow$ being an operator on qualifiers.  We should clarify which
+%      it is.}
+%    \label{fig-poly-resolution}
+%\end{figure}
 
@@
     {\tau \uparrow \ = \|NonDet \ \beta|}
-    {\tau \ = \|OrderNonNonDet \ \beta|
-        & \wellformed{\tau}
-    }$
-
-    \bigskip
-
-    $\infer[\rulename{ordernondet polydown}]
-    {\tau \downarrow \ = \|Det \ \beta|}
-    {\tau \ = \|OrderNonNonDet \ \beta|
+    {\tau \ = \|OrderNonDet \ \beta|
         & \wellformed{\tau}
@@
 
+    $\infer[\rulename{ordernondet polydown}]
+    {\tau \downarrow \ = \|Det \ \beta|}
+    {\tau \ = \|OrderNonDet \ \beta|
+        & \wellformed{\tau}
+    }$
+
+    \bigskip
+
     $\infer[\rulename{polydown}]
@@
     }$
-\caption{Polymorphic resolution}
-\label{tab-poly-resolutions}
+\caption{Polymorphic resolution ($\uparrow$ is overloaded for types and qualifiers)}
+\label{fig:poly-resolutions}
 \end{figure}

### 2019-03-30 experiments.tex
bib refs for tools

--- 2019-03-28 experiments.tex
@@
 \section{Evaluation}
+\begin{itemize}
+    \item goals of evaluation: Compare with NonDex and Deflaker
+    \item Quantitative comparison: \theDeterminismChecker finds all bugs reported by NonDex and Deflaker?
+    \item Qualitative study of the issues discovered.
+    \item System info: 2.3 GHz Intel Core i5, 8 GB 2133 MHz
+\end{itemize}
+
+\subsection{Running NONDEX on Randoop}
+To evaluate how existing tools perform on large programs, we ran NONDEX~\cite{nondex-tool} on Randoop~\cite{randoop-tool}.
+\begin{itemize} 
+    \item Ran randoop with NonDex, instructions from cite.
+    \item didn't find any flakiness issues in the FULL mode.
+    \item \todo{check if the bugs discovered by \theDeterminismChecker have test coverage}
+\end{itemize}
+
 \subsection{NONDEX case study}
@@
     \item annotate the local arrays and collections (This has been mostly sufficient for recall)
+    \item exceptions: unbuildable handlebars
+    \item the reported source of nondeterminism couldn't be found in the source code.
+    \item \todo{P4, P12 and P21}
 \end{itemize}
@@
     \item From NonDex paper: "manually inspecting these failures was rather challenging"
+    \item P18 sl4j: nondeterminism in the test case. we provide stub for junit.
+    \item Specification for Assert: parameters are annotated as Det
 \end{itemize}
@@
 \begin{table}[]
-    \begin{tabular}{|p{1.5cm}|p{1cm}|p{3cm}|p{4cm}|p{2cm}|p{1.5cm}|}
+    \begin{tabular}{|p{1.5cm}|p{1cm}|p{4cm}|p{5cm}|}
         \hline
-        Benchmark & Recall? & Error message source & Error msg & \#annotations & Annotation type  \\ \hline
-        P1 & Yes & FieldAccess.java: [102,87] & assignment type incompatible &  &   \\
-        P2 & MI &  &  &  &  \\
-        P3 & Yes & DateTimeZone.java: [731,27] &  return type incompatible&  &  \\ 
-        P4 & MI &  &  &  &  \\ 
-        P5 & src &  &  &  &  \\ 
-        P5 & src &  &  &  &  \\ 
-        P6 & MI &  &  &  &  \\ 
-        P6 & MI &  &  &  &  \\
-        P7(test error) & Yes & checker/ tests/ determinism/P71.java: 5 & return type incompatible &  &  \\ 
-        P7(2) & Yes & FieldUtils.java: [219,84] & assignment type incompatible &  &  \\
-        P7(2) & Yes & Reflection ToStringBuilder.java: [543,65]  & assignment type incompatible &  &  \\
-        P7(2) & Yes & HashCodeBuilder.java: [193,69]  & assignment type incompatible &  &  \\
-        P7(2) & Yes & CompareToBuilder.java: [322,65] & assignment type incompatible &  &  \\
-        P7(2) & Yes & EqualsBuilder.java: [419,69]  & assignment type incompatible &  &  \\
-        P7 & Yes & FieldUtils.java: [200,69] & assignment type incompatible & &  \\
-        P7 & Yes & FieldUtils.java: [219,84] & assignment type incompatible &  &  \\
-        P7 & Yes & FieldUtils.java: [240,86] & assignment type incompatible &  &  \\
-        P8 & RM &  &  &  &  \\
-        P8 & RM &  &  &  &  \\
-        P8 & RM &  &  &  &  \\
-        P9 & MI &  &  &  &  \\
-        P10 & crash &  &  &  &  \\
-        P10 & crash &  &  &  &  \\
-        P11 & MI &  &  &  &  \\
-        P12 & MI &  &  &  &  \\
-        P13 & crash &  &  &  &  \\
-        P14 & MI &  &  &  &  \\
-        P15 & MI &  &  &  &  \\
-        P15 & MI &  &  &  &  \\
-        P16 & Yes & MethodSorter.java: [56,61]  & assignment type incompatible &  &  \\
-        P17 & Yes &  &  &  &  \\
-        P18 & src &  &  &  &  \\
-        P19 & MI &  &  &  &  \\
-        P19 & MI &  &  &  &  \\
-        P19 & MI &  &  &  &  \\
-        P20 & MI &  &  &  &  \\
-        P20 & MI &  &  &  &  \\
-        P21 & MI &  &  &  &  \\
-        P21 & MI &  &  &  &  \\ \hline
+        Benchmark & Recall? & Error message source & Error msg  \\ \hline
+        P1 & Yes & FieldAccess.java: [102,87] & assignment type incompatible \\
+        P2 & Yes & YamlWriter.java: [224,28] & cast unsafe  \\
+        P3 & Yes & DateTimeZone.java: [731,27] &  return type incompatible \\ 
+        P4 & * & Det checker not found  &   \\ 
+        P5 & ** & Class.getDeclaredMethods not found &    \\ 
+        P6 & Yes & OptionGroup.java: [55,40]  & argument type incompatible \\ 
+        P6 & Yes & Option.java: [157,28]  &  assignment type incompatible\\
+        P7(test error) & Yes* & checker/ tests/ determinism/P71.java: 5 & return type incompatible \\ 
+        P7(2) & Yes & FieldUtils.java: [219,84] & assignment type incompatible \\
+        P7(2) & Yes & Reflection ToStringBuilder.java: [543,65]  & assignment type incompatible  \\
+        P7(2) & Yes & HashCodeBuilder.java: [193,69]  & assignment type incompatible   \\
+        P7(2) & Yes & CompareToBuilder.java: [322,65] & assignment type incompatible   \\
+        P7(2) & Yes & EqualsBuilder.java: [419,69]  & assignment type incompatible  \\
+        P7 & Yes & FieldUtils.java: [200,69] & assignment type incompatible \\
+        P7 & Yes & FieldUtils.java: [219,84] & assignment type incompatible  \\
+        P7 & Yes & FieldUtils.java: [240,86] & assignment type incompatible  \\
+        P8 & ** & getDeclaredFields not found &   \\
+        P9 & Yes & Printers.java: [269,97]  &  enhancedfor type incompatible \\
+        P10 & ** & getDeclaredFields not found  &  \\
+        P11 & Yes & MapUtils.java: [20,50]   & enhancedfor type incompatible  \\
+        P12 & *  & 0 errors &  \\
+        P13 & ** & unbuildable handlebars &  \\
+        P14 & Yes  & Maps.java: [91,69] & invalid element type   \\
+        P15 & Yes & PkiMessageDecoder.java: [144,29]  &  enhancedfor type incompatible  \\
+        P16 & Yes & MethodSorter.java: [56,61]  & assignment type incompatible  \\
+        P17 & Yes & JSONObject.java: [674,51]  & assignment type incompatible  \\
+        P18 & Yes & SubstitutableLoggerTest.java: [67,61]   & enhancedfor type incompatible  \\
+        P19 & Yes & DataDictionaryEventSource.java: [73,78]  &  enhancedfor type incompatible  \\
+        P20 & Yes & SaverImpl.java: [43,69]  &  return type incompatible  \\
+        P21 & * & 0 errors &   \\ \hline
     \end{tabular}

### 2019-03-30 experiments.tex
experiments: NONDEX on Randoop

@@
 \subsection{Running NONDEX on Randoop}
-To evaluate how existing tools perform on large programs, we ran NONDEX~\cite{nondex-tool} on Randoop~\cite{randoop-tool}.
-\begin{itemize} 
-    \item Ran randoop with NonDex, instructions from cite.
-    \item didn't find any flakiness issues in the FULL mode.
-    \item \todo{check if the bugs discovered by \theDeterminismChecker have test coverage}
-\end{itemize}
+To evaluate how existing tools perform on large programs, we ran the tool \href{https://github.com/TestingResearchIllinois/NonDex}{NONDEX~\cite{nondex}} on 
+\href{https://github.com/randoop/randoop}{Randoop~\cite{randoop}}.
+We followed the instructions to build the project with the gradle plugin. 
+NONDEX terminated in approximately 20 seconds and didn't report any flaky tests.
+\todo{check if the bugs discovered by \theDeterminismChecker have test coverage}
+\todo{We investigated the tests covering code that \theDeterminismChecker reported as nondeterministic.
+If code is covered by tests, NONDEX couldn't discover it. Else, its a limitation of NONDEX}.
+This indicates that the additional programmer effort required by \theDeterminismChecker is justified
+as our tool discovers non-trivial issues and doesn't depend on code coverage or runtime heuristics.
 

### 2019-03-30 experiments.tex
small fix

@@
 This indicates that the additional programmer effort required by \theDeterminismChecker is justified
-as our tool discovers non-trivial issues and doesn't depend on code coverage or runtime heuristics.
+as our tool discovers non-trivial issues and doesn't depend on code coverage or other runtime heuristics.
 

### 2019-03-30 experiments.tex
nondex case stidy

--- 2019-03-30 experiments.tex
@@
     \item goals of evaluation: Compare with NonDex and Deflaker
-    \item Quantitative comparison: \theDeterminismChecker finds all bugs reported by NonDex and Deflaker?
+    \item Quantitative comparison: \TheDeterminismChecker finds all bugs reported by NonDex and Deflaker?
     \item Qualitative study of the issues discovered.
@@
 We followed the instructions to build the project with the gradle plugin. 
-NONDEX terminated in approximately 20 seconds and didn't report any flaky tests.
-\todo{check if the bugs discovered by \theDeterminismChecker have test coverage}
-\todo{We investigated the tests covering code that \theDeterminismChecker reported as nondeterministic.
+NONDEX terminated without reporting any flaky tests.
+\todo{check if the bugs discovered by \TheDeterminismChecker have test coverage}
+\todo{We investigated the tests covering code that \TheDeterminismChecker reported as nondeterministic.
 If code is covered by tests, NONDEX couldn't discover it. Else, its a limitation of NONDEX}.
-This indicates that the additional programmer effort required by \theDeterminismChecker is justified
-as our tool discovers non-trivial issues and doesn't depend on code coverage or runtime heuristics.
+This indicates that the additional programmer effort required by \TheDeterminismChecker is justified
+as our tool discovers non-trivial issues and doesn't depend on code coverage or other runtime heuristics.
 
 \subsection{NONDEX case study}
-Methodology:
-\begin{itemize}
-    \item Objective: Measure recall.
-    \item Benchmarks from NONDEX paper.
-    \item That paper gives flaky test class and source of flakiness (nondeterministic method call).
-    \item Look at the source method under test.
-    \item annotate the local arrays and collections (This has been mostly sufficient for recall)
-    \item exceptions: unbuildable handlebars
-    \item the reported source of nondeterminism couldn't be found in the source code.
-    \item \todo{P4, P12 and P21}
-\end{itemize}
+\cite{nondex} reported flaky tests on 21 open source projects. The authors also reported the sources of flakiness
+after manually inspecting these tests. We ran \TheDeterminismChecker on all these benchmarks with the objective
+of measuring recall of our tool. We cloned all 21 repos and switched to the commit that had a flakiness issue as reported 
+in~\cite{nondex}.
+Of these projects \<handlebars> has a compile time failure which prevents us from running \TheDeterminismChecker on it.
+For benchmarks \<visualee>, \<easy-batch>, and \<vraptor>, we couldn't find
+the sources of flakiness as reported in~\cite{nondex} anywhere in the respective repositories.
+\TheDeterminismChecker reported the nondeterministic locations correctly in the rest of the projects.
+All these benchmarks and instructions to reproduce our results can be found at \href{https://github.com/t-rasmud/determinismCaseStudies}{https://github.com/t-rasmud/determinismCaseStudies}.
+\todo{P4, P12 and P21 det checker isn't recognized?}
 
@@
 \begin{itemize}
+    \item Look at the source method under test.
+    \item annotate the local arrays and collections (This has been mostly sufficient for recall)
     \item Refer to table and describe it.
@@
     \item Specification for Assert: parameters are annotated as Det
+    \item \todo{Link to det checker repo?}
 \end{itemize}

### 2019-03-31 experiments.tex
nondex case study

@@
 in~\cite{nondex}.
-Of these projects \<handlebars> has a compile time failure which prevents us from running \TheDeterminismChecker on it.
+Of these projects, \<handlebars> has a compile time failure which prevents us from running \TheDeterminismChecker on it.
 For benchmarks \<visualee>, \<easy-batch>, and \<vraptor>, we couldn't find
 the sources of flakiness as reported in~\cite{nondex} anywhere in the respective repositories.
-\TheDeterminismChecker reported the nondeterministic locations correctly in the rest of the projects.
+\TheDeterminismChecker reported the nondeterministic source locations correctly in the rest of the projects.
 All these benchmarks and instructions to reproduce our results can be found at \href{https://github.com/t-rasmud/determinismCaseStudies}{https://github.com/t-rasmud/determinismCaseStudies}.
@@
 
-Discussion of results:
+All the flakiness that NONDEX found was due to 7 methods in 3 classes:
+\<Class.getDeclaredFields>, \<Class.getDeclaredMethods>, \<Class.getFields>, \<DateFormatSymbols.getZoneStrings>,
+\<HashMap.entrySet>, \<HashMap.keySet>, and \<HashMap.values>.  
+We provide samples of the annotated source code form these benchmarks and the errors as reported by \TheDeterminismChecker 
+in Figures~\ref{code-reflectasm} and~\ref{code-actiongenerator}.
+\begin{figure}
+    \begin{verbatim}
+    static public FieldAccess get (Class type) {
+        @Det ArrayList<@Det Field> fields = new ArrayList<Field>();
+        ...
+        while (nextClass != Object.class) {
+            @Det Field @Det[] declaredFields = nextClass.getDeclaredFields();
+            ...
+        }
+    }
+    [ERROR] FieldAccess.java:[102,87] [assignment.type.incompatible] 
+    incompatible types in assignment.
+    found   : @Det Field @OrderNonDet []
+    required: @Det Field @Det []
+    \end{verbatim}
+    \caption{Annotated source code from reflectasm.}
+    \label{code-reflectasm}
+\end{figure}
+\begin{figure}
+    \begin{verbatim}
+    protected SimpleDataEvent createNextEvent() {
+      for (@Det Entry<@Det String, @Det FieldType> entry : fields.entrySet()) {
+          ...
+      }
+      ...
+    }
+    [ERROR] DataDictionaryEventSource.java:[73,78] [enhancedfor.type.incompatible] 
+    incompatible types in enhanced for loop.
+    found   : @NonDet Entry<@Det String, @Det FieldType>
+    required: @Det Entry<@Det String, @Det FieldType>
+    \end{verbatim}
+    \caption{Annotated source code from ActionGenerator.}
+    \label{code-actiongenerator}
+\end{figure}
+As can be seen from these figures, annotating local arrays and collections
+has been sufficient to discover the sources of nondeterminism in most of the projects.
+The rest of the projects required annotating the return types of methods.
 
-\begin{itemize}
-    \item Look at the source method under test.
-    \item annotate the local arrays and collections (This has been mostly sufficient for recall)
-    \item Refer to table and describe it.
-    \item P7: same source causing multiple tests to be flaky
-    \item Example for class.GetDeclaredMethod() and show annotations
-    \item Example for HashMap.entrySet() and show annotations.
-    \item qualitative comparison: orthogonal techniques (reporting tests vs reporting source locations).
-    \item From NonDex paper: "manually inspecting these failures was rather challenging"
-    \item P18 sl4j: nondeterminism in the test case. we provide stub for junit.
-    \item Specification for Assert: parameters are annotated as Det
-    \item \todo{Link to det checker repo?}
-\end{itemize}
+For the benchmark \<commons-lang>, \cite{nondex} report multiple failing tests (14 flaky tests) due to calls
+to \<Class.getDeclaredFields>. Upon inspecting the source code, we find that all of invocations to
+\<Class.getDeclaredFields> are made from 5 locations. This means that annotating 5 lines of source code
+would have been sufficient to identify the nondeterminism. Having said that, we admit that there could be
+significant programmer effort involved in annotating the whole program. On the other hand,~\cite{nondex}
+state "we found that manually inspecting these failures was
+rather challenging, and we leave it as future work to automate
+debugging test failures due to ADINS code". \TheDeterminismChecker reports source locations
+which makes it easier for the programmer to fix issues and is orthogonal to the approach taken by NONDEX which
+report failing tests.
+
+\TheDeterminismChecker is also capable of detecting nondetermism in test cases. We provide a stub file
+for \<junit> in which we annotated all the \<assert> methods as accepting only \<Det> parameters.
+This will report any \<junit> test that passes nondeterministic arguments to \<assert>s and the test doesn't
+even have to be executed in our case.
+\todo{It took one of our authors about an hour to create this stub file - proof of easy estensibility of tool?} 
+\todo{Link to det checker repo?}
 

### 2019-03-31 conclusion.tex
Randoop case study

\section{Conclusion}\label{sec:conclusion}

\todo{TODO}

\todo{\TheDeterminismChecker found errors that the
  Randoop developers had missed, despite spending over 100 hours on the
  problem of nondeterminism.}

### 2019-03-31 randoop-case-study.tex
Randoop case study

\section{Case study}\label{sec:randoop-case-study}

As a case study, we applied \theDeterminismChecker to the Randoop test
generation tool~\ifanonymous\citepalias{randoop-tool}\else\cite{PachecoLEB2007}\fi.
We chose Randoop because 
it is frequently used in software engineering experiments,
it is actively maintained,
and its developers have struggled with nondeterminism (personal
communication, 2019).

Randoop is intended to be deterministic, when invoked on a deterministic
program.\footnote{Users of Randoop can pass in a different seed in order to
  obtain a different deterministic output.  Randoop has command-line
  options that enable concurrency and timeouts, both of which can lead to
  nondeterministic behavior.}

However, Randoop was not deterministic.  This caused the developers
problems in 
reproducing bugs reported by users, 
reproducing test failures during development, and
understanding the effect of changes to Randoop (by comparing executions of two
similar variants of Randoop).

The developers took extensive action to detect and mitigate the effect of, nondeterminism.
They used Docker images to run tests, to avoid system dependencies such as
a different JDK having a different number of classes or methods.
They wrote tests with relaxed oracles (assertions) that permit multiple
possible answers --- for example, in code coverage of generated tests.
They used linters such as Error Prone, which warns if toString is used on
array objects, which don't override Object.toString and therefore print a
hash code which may vary from run to run.
They used libraries that make hash codes deterministic, by giving each
object of a type a unique ID that counts up from 1, rather than using a
memory address as \<Object.hashCode> does.
They wrote preprocess output and logs to make nondeterministic ones easier
to compare, such as by removing or canonicalizing hash codes, dates, and
other nondeterministing output.
Each of these efforts wase helpful, but they were not enough to solve the problem.

In July 2017, the Randoop developers spent two weeks of full-time work to
eliminate unintentional nondeterministic behavior in Randoop.
Their methodology was to repeatedly
 run Randoop with verbose logging enabled,
 look for differences in logging output,
 find the root cause of nondeterminism,
 and eliminate it.
The most common causes were toString() routines and iteration order of sets and maps.
Some of the nondeterminism was in libraries, such as the JDK\@.

The developers chose not to make every Map a LinkedHashMap, because that
would have increased memory and CPU costs.  They chose not to make every
order-nondeterminism List a Set, for similar reasons:  deduplication was
not always needed and would have increased costs.

\todo{When we discuss annotation effort, compare to this.
  The annotation effort is
  high, but it's no more than the hundreds of hours that the Randoop
  developers spent, with less success.}

That coding sprint did not find all the problems.
The developers debugged and fixed 5 additional determinism defects over the
next 12 months, using a similar methodology.
% [2413 changed lines in the diffs.
% (This number slightly overstates the number of code changes made, because sometimes changing code would result in code formatting changes, and some changes in those commits are unrelated to determinism, such as documentation improvements.)]
We analyzed a version of Randoop from February 2019 (commit 0a8e63fb), over 6 months after the most recent determinism bug fix.

We annotated the core of Randoop (the \<src/main/java> directory), which
contains 24K non-comment, non-blank lines of code.
% (41 KLOC including blanks and comments).
% [Maybe we will add 1600 LOC of agent code.]
We did not annotate Randoop's test suite.

\todo{Report the number of annotations required, and say a few words about
  the process and the effort required.}

\TheDeterminismChecker found three previously-unknown nondeterminism bugs in Randoop.
The Randoop developers accepted our bug reports and committed fixes to the repository.

\begin{itemize}
\item
  A use of \<HashSet> that the developers changed to \<LinkedHashSet>\commit{c975a9f7}.
The old code could cause a problem if a type variable's lower or upper
bound has a type parameter that the type variable itself does not have.
This situation does occur, even in Randoop's test suite.

\item 
 Diagnostic output that printed a hash code for brevity, but should have had deterministic output\commit{661a4970}.  By contrast to the other two bugs, this one is not serious.

\item
Dependency on the CLASSPATH environment variable in preference to the
classpath passed on the command line\commit{330e3c56}.
This could have led to incorrect behavior if a user set the environment variable.
The developers changed Randoop to not read the environment variable.
\end{itemize}

We reported another suspicous case of order-nondeterminism, in the
\<findOverridden> method.  The Randoop developers explained it was
acceptable, after tracing the flow through the program.  (One said, ``This
looks OK to me \ldots\ it took me a little while to decide that.'')  The fact
depended on subtle, undocumented invariants about Randoop that we had not
been able to reverse-engineer.

### 2019-03-31 related.tex
Randoop case study

\section{Related work}\label{sec:related}

\todo{TODO}

### 2019-03-31 experiments.tex
Edits in experiments.tex

--- 2019-03-30 experiments.tex
@@
-\section{Evaluation}
+\section{Comparison to previous work}
+
+The state of the art in flaky test detection is NonDex~\cite{nondex}.
+\Cref{sec:related} explains how NonDex works.
+
 \begin{itemize}
@@
 
-\subsection{Running NONDEX on Randoop}
-To evaluate how existing tools perform on large programs, we ran the tool \href{https://github.com/TestingResearchIllinois/NonDex}{NONDEX~\cite{nondex}} on 
-\href{https://github.com/randoop/randoop}{Randoop~\cite{randoop}}.
-We followed the instructions to build the project with the gradle plugin. 
-NONDEX terminated in approximately 20 seconds and didn't report any flaky tests.
+\subsection{Case study with NonDex}\label{sec:nondex-randoop}
+
+We ran NonDex on the same commit of Randoop as we used for the case study
+of \cref{sec:case-study}.  This version contains 3 nondeterminism bugs that
+\ourTool found.
+
+We modified Randoop by deleting tests that were skipped by its buildfile,
+because the NonDex gradle plugin does not respect those settings.
+
+After that, NonDex ran without problems on Randoop.
+It terminated in approximately 20 seconds and reported no flaky tests.
+
 \todo{check if the bugs discovered by \theDeterminismChecker have test coverage}
@@
 If code is covered by tests, NONDEX couldn't discover it. Else, its a limitation of NONDEX}.
-This indicates that the additional programmer effort required by \theDeterminismChecker is justified
-as our tool discovers non-trivial issues and doesn't depend on code coverage or other runtime heuristics.
 
-\subsection{NONDEX case study}
+\subsection{\OurTool on NonDex benchmarks}\label{sec:nondex-benchmarks}
+
+\Cref{sec:nondex-randoop} shows that \ourTool finds errors that NonDex does not.
+
 Methodology:

### 2019-03-31 design.tex

@@
 \TheDeterminismChecker has additional type refinement rules for the sorting
-methods such as \<List.sort>, \<Collections.sort>, and \<Arrays.sort>:
+methods of \<List>, \<Collections>, and \<Arrays> classes
  as shown in \cref{fig-sorting}.

### 2019-03-31 experiments.tex

--- 2019-03-31 experiments.tex
@@
 
-\subsection{Running NONDEX on Randoop}
-To evaluate how existing tools perform on large programs, we ran the tool \href{https://github.com/TestingResearchIllinois/NonDex}{NONDEX~\cite{nondex}} on 
-\href{https://github.com/randoop/randoop}{Randoop~\cite{randoop}}.
-We followed the instructions to build the project with the gradle plugin. 
-NONDEX terminated without reporting any flaky tests.
-\todo{check if the bugs discovered by \TheDeterminismChecker have test coverage}
-\todo{We investigated the tests covering code that \TheDeterminismChecker reported as nondeterministic.
+\subsection{Case study with NonDex}\label{sec:nondex-randoop}
+
+We ran NonDex on the same commit of Randoop as we used for the case study
+of \cref{sec:case-study}.  This version contains 3 nondeterminism bugs that
+\ourTool found.
+
+We modified Randoop by deleting tests that were skipped by its buildfile,
+because the NonDex gradle plugin does not respect those settings.
+
+After that, NonDex ran without problems on Randoop.
+It completed successfully without reporting any flaky tests.
+
+\todo{check if the bugs discovered by \theDeterminismChecker have test coverage}
+\todo{We investigated the tests covering code that \theDeterminismChecker reported as nondeterministic.
 If code is covered by tests, NONDEX couldn't discover it. Else, its a limitation of NONDEX}.

### 2019-03-31 experiments.tex
Tweak in NonDex experiments

@@
 
-\subsection{NONDEX case study}
-\cite{nondex} reported flaky tests on 21 open source projects. The authors also reported the sources of flakiness
+\subsection{\OurTool on NonDex benchmarks}\label{sec:nondex-benchmarks}
+
+\Cref{sec:nondex-randoop} shows that \ourTool finds errors that NonDex does not.
+This section determines whether NonDex finds errors that \ourTool does not.
+Its authors ran NonDex on 195 open-source projects, and NonDex found flaky tests in
+21 of them.
+
+The authors also reported the sources of flakiness
 after manually inspecting these tests. We ran \TheDeterminismChecker on all these benchmarks with the objective

### 2019-03-31 experiments.tex
Fix a macro

--- 2019-03-31 experiments.tex
@@
 of \cref{sec:case-study}.  This version contains 3 nondeterminism bugs that
-\ourTool found.
+\theDeterminismChecker found.
 
@@
 
-\subsection{NONDEX case study}
-\cite{nondex} reported flaky tests on 21 open source projects. The authors also reported the sources of flakiness
+\subsection{\TheDeterminismChecker on NonDex benchmarks}\label{sec:nondex-benchmarks}
+
+\Cref{sec:nondex-randoop} shows that \theDeterminismChecker finds errors that NonDex does not.
+This section determines whether NonDex finds errors that \theDeterminismChecker does not.
+Its authors ran NonDex on 195 open-source projects, and NonDex found flaky tests in
+21 of them.
+
+The authors also reported the sources of flakiness
 after manually inspecting these tests. We ran \TheDeterminismChecker on all these benchmarks with the objective

### 2019-03-31 related.tex
Notes about related work
Fix typo
Todo notes

@@
 \todo{TODO}
+
+The state of the art is NonDex.  It is a highly effective tool.
+
+% https://github.com/TestingResearchIllinois/NonDex/wiki/Supported-APIs
+NonDex uses a hand-crafted list of 47 methods (25 unique method names) in
+in 13 classes.
+This is analogous to our annotations on the JDK\@.
+\todo{Say how much code our annotations cover?}
+\todo{Check whether our JDK is missing anything they support.}
+\todo{Did the NonDex authors contact the maintainers of the tools in which
+  they found nondeterminism?}

### 2019-03-31 experiments.tex
Todo notes
typo

@@
 even have to be executed in our case.
-\todo{It took one of our authors about an hour to create this stub file - proof of easy estensibility of tool?} 
+\todo{It took one of our authors about an hour to create this stub file - proof of easy extensibility of tool?} 
 \todo{Link to det checker repo?}
@@
 \end{table}
+
+\todo{If there is time, we could do a second case study, on a program that
+  NonDex did not find nondeterminism in.  Our hope would be to find
+  nondeterminism that NonDex did not.  Choose a small project that is
+  maintained (it has recent commits), so we can communicate with the authors.

### 2019-03-31 experiments.tex
Fix cross-reference

@@
 We ran NonDex on the same commit of Randoop as we used for the case study
-of \cref{sec:case-study}.  This version contains 3 nondeterminism bugs that
+of \cref{sec:randoop-case-study}.  This version contains 3 nondeterminism bugs that
 \theDeterminismChecker found.
@@
 even have to be executed in our case.
-\todo{It took one of our authors about an hour to create this stub file - proof of easy extensibility of tool?} 
+\todo{It took one of our authors about an hour to create this stub file - proof of easy estensibility of tool?} 
 \todo{Link to det checker repo?}

### 2019-03-31 experiments.tex
Fix unclosed group
typo

--- 2019-03-31 experiments.tex
@@
   nondeterminism that NonDex did not.  Choose a small project that is
-  maintained (it has recent commits), so we can communicate with the authors.
+  maintained (it has recent commits), so we can communicate with the authors.}
+
+\section{deflaker case study}
+\begin{table}[]
+    \begin{tabular}{|l|l|}
+        \hline
+        Benchmark, issue & found by det checker? \\ \hline
+        achilles, [2] issue \#309 &  Race condition\\ \hline
+        achilles, [3] issue \#310 & Same fix for \#309 and \#310 \\ \hline
+        checkstyle [4] issue \#4664& det checker not found \\ \hline
+        jackrabbit-oak [6] \#6512&  Unresolved on issue tracker \\ \hline
+        jackrabbit-oak [7] issue \#6524&  Dead link, didn't find 6524 in issue tracker \\ \hline
+        nutz [10] issue \#1283 &  \\ \hline
+        &  \\ \hline
+        &  \\ \hline
+    \end{tabular}
+\end{table}

### 2019-03-31 experiments.tex
more passing benchmarks

--- 2019-03-31 experiments.tex
@@
         P11 & Yes & MapUtils.java: [20,50]   & enhancedfor type incompatible  \\
-        P12 & *  & 0 errors &  \\
+        P12 & Yes  & GSLayerEncoder21.java: [113,91]  & enhancedfor type incompatible \\
         P13 & ** & unbuildable handlebars &  \\
@@
         P20 & Yes & SaverImpl.java: [43,69]  &  return type incompatible  \\
-        P21 & * & 0 errors &   \\ \hline
+        P21 & Yes & RestDocumentation.java: [419,50] &  argument type incompatible   \\ \hline
     \end{tabular}
@@
   nondeterminism that NonDex did not.  Choose a small project that is
-  maintained (it has recent commits), so we can communicate with the authors.
+  maintained (it has recent commits), so we can communicate with the authors.}
+
+\section{deflaker case study}
+\begin{table}[]
+    \begin{tabular}{|l|l|}
+        \hline
+        Benchmark, issue & found by det checker? \\ \hline
+        achilles, [2] issue \#309 &  Race condition\\ \hline
+        achilles, [3] issue \#310 & Same fix for \#309 and \#310 \\ \hline
+        checkstyle [4] issue \#4664& det checker not found \\ \hline
+        jackrabbit-oak [6] \#6512&  Unresolved on issue tracker \\ \hline
+        jackrabbit-oak [7] issue \#6524&  Dead link, didn't find 6524 in issue tracker \\ \hline
+        nutz [10] issue \#1283 &  Removed tests (Printing (HTTP's) Response.getstatus())\\ \hline
+        togglz [19] issue \#223 & (use NONDEX for debugging) \\ \hline
+        togglz [20] issue \#240 &  \\ \hline
+    \end{tabular}
+\end{table}

### 2019-04-01 design.tex
Edits in sections 2-2.3

@@
 
-\subsection{Preliminaries}
+\subsection{Preliminaries and notation}
 
@@
 
+\begin{figure}
+  \newcommand{\bnfalt}{\ \ |\ \ }
+  \begin{tabular}{lcll}
+    $C$ & ::= & \<Int>\bnfalt \<List>\bnfalt \ldots & class name \\
+    $\kappa$ & ::= & \<NonDet>\bnfalt \<OrderNonDet>\bnfalt\<Det> & type qualifier \\
+    $\tau$ & ::= & $\kappa\ \beta$ & \mbox{type} \\
+    $\beta$ & ::= & $C$\bnfalt $C\angles{\overline{\tau}}$ & \mbox{unqualified type}
+  \end{tabular}
+  \caption{Grammar of types}
+  \label{fig:grammar}
+\end{figure}
+
 In a type system,
@@
 set of possible run-time values that an expression may evaluate to.
-A programming language provides \emph{basetypes}, such as \<int>.
+A programming language provides \emph{basetypes}, such as \<Int>.
 A \textit{type qualifier} on a basetype adds additional constraints;
 that is, it reduces the size of the set of values.
-An example type qualifier is \<Positive>,\todo{should all type qualifiers
-  be lowercase in this section??} and a type (which combines a qualifier
-and a basetype) is \<Positive int>.
-A polymorphic type such as \<List> can be instantiated by a type argument,
-as in \ptype{List}{Positive int}.
+An example type qualifier is \<Positive>, and a type (which combines a qualifier
+and a basetype) is \<Positive Int>.
+A polymorphic type abstraction such as \<List> can be instantiated by a type argument,
+as in \ptype{List}{Positive Int}.
+
+\Cref{fig:grammar} formalizes these notions.
+\Cref{fig:grammar}'s $\beta$ is not a basetype as described above:  $\beta$
+lacks a top-level type, but any type arguments are full types
+$\tau$.  This paper does not formally define basetype, and uses $\beta$,
+for simplicity of exposition.
+In our core calculus, a type $\tau$ always has a qualifier.
+Our implementation, \theDeterminismChecker, uses inference and defaults to
+permit users to omit type qualifiers\todo{Forward reference}.
 
 A type checker verifies the types written in a program.
+A type qualifier constrain the set of possible run-time values, that is,
+$\kappa \ /beta \sqsubseteq \beta$.
+As a result, a qualifier type system does not allow any behavior that the
+original type system does not.
+However, the qualifier type system may reject more programs, and thus
+affords stronger guarantees.
+(Our implementation \theDeterminismChecker runs the underlying base type
+system, then issues additional errors and warnings.)
+\todo{Some qualifier type systems can be defined independently of the underlying
+type system.  We don't do that because there are interactions between the
+basetypes and the type qualifiers.  This allows greater precision, which
+is important in practice.}
+
+\begin{figure}
+    \bigskip
+
+    $\infer{\Det \sqsubseteq \OrderNonDet}{}
+    \quad\quad
+    \infer{\OrderNonDet \sqsubseteq \NonDet}{}$
+
+    \bigskip
+
+    $\infer{\kappa_1 \beta \sqsubseteq \kappa_2 \beta}{\kappa_1 \sqsubseteq \kappa_2}
+    \quad\quad
+    \infer{\kappa \beta_1 \sqsubseteq \kappa \beta_2}{\beta_1 \sqsubseteq \beta_2}$
+
+    \caption{Subtyping rules for \ourTypeSystem.  $\sqsubseteq$ is overloaded for classes,
+    type qualifiers, types, and unqualified types $\beta$.}
+    \label{fig:typecheck-rules}
+\end{figure}
 
@@
     \caption{A sampling of standard typing rules.  We omit other standard rules,
-    for brevity.  $\sqsubseteq$ is overloaded for classes and types.}
+    for brevity.}
     \label{typecheck-rules-standard}
-\end{figure}
-
-\begin{figure}
-    \bigskip
-
-    $\infer{\Det \sqsubseteq \OrderNonDet}{}
-    \quad\quad
-    \infer{\OrderNonDet \sqsubseteq \NonDet}{}$
-
-    \bigskip
-
-    $\infer{\kappa_1 \beta \sqsubseteq \kappa_2 \beta}{\kappa_1 \sqsubseteq \kappa_2}
-    \quad\quad
-    \infer{\kappa \beta_1 \sqsubseteq \kappa \beta_2}{\beta_1 \sqsubseteq \beta_2}$
-
-    \caption{Subtyping rules for \ourTypeSystem.  $\sqsubseteq$ is overloaded for classes,
-    basetypes, type qualifiers, and types.}
-    \label{fig:typecheck-rules}
 \end{figure}
@@
 
-Figure~\cref{typecheck-rules-standard} shows a sample of the standard typing rules applied to 
-any typed language. In the generic case, the type $\tau$ represents a valid basetype in a programming language.
-\ourTypeSystem extends a basetype with a type qualifier. Throughout this paper, $\tau$ represents a fully qualified type.
-In other words, type $\tau = \kappa \ \beta$ consists of both a qualifier $\kappa$ and a basetype $\beta$, neither of which may be omitted.
-Further, $\beta$ is a type without a type qualifier, but which has type qualifiers on all its type parameters.
-Formally, $\beta ::= C|C \angles{\tau}$ where $C$ represents a valid Java basetype.
-\ourTypeSystem checks all the standard typing rules including the ones shown in Figure~\cref{typecheck-rules-standard} with
-the only difference of having an additional type qualifier on each of the basetypes. The type qualifiers can only 
-constrain the set of valid types. As a result, \ourTypeSystem doesn't allow any behavior that the original type system doesn't.
+\Cref{typecheck-rules-standard} shows a sample of standard typing rules for
+an object-oriented programming language.
+
+\OurTypeSystem checks all the standard typing rules including the ones shown in \cref{typecheck-rules-standard} with
+the only difference of having an additional type qualifier on each of the basetypes.
 
@@
       in a different order.
-    \item \<Det> indicates that
-    the expression evaluates to the same value (with respect to \<.equals()>) in all
-    executions; for a collection, iteration also yields the values in the same
-    order.
+    \item \<Det> indicates that the expression evaluates to equal values in
+      all executions; for an unordered collection such as a set, iteration
+      also yields the values in the same order.
 \end{itemize}
@@
 \subsection{Collection types}\label{collection-rules}
+
+\subsubsection{Type well-formedness}
 
@@
 Otherwise, the type qualifier on a collection's element must be a subtype or equal to
-the collection's type qualifier (\rulename{collection}\ rules of
+the collection's type qualifier (\rulename{collection} rules of
 \cref{type-validity}). Note that the \rulename{ordernondet} versions also handle \<Det>,
 because $\<Det> \sqsubseteq \<OrderNonDet>$.
-\Cref{fig:determinism-collections} restates its restrictions, and here are examples.
+\Cref{fig:determinism-collections} restates the \rulename{collection} rules
+of
+\cref{type-validity}.
 
@@
 % Using "~" instead of "\ " does not work within \codeid :-(
-\begin{smaller}
 \begin{Verbatim}[commandchars=\\\{\}]
-NonDet      \lptype{List}{NonDet      Integer}
+NonDet      \lptype{List}{NonDet      Int}
 OrderNonDet \lptype{List}{OrderNonDet \lptype{Set}{...}}
-OrderNonDet \lptype{List}{Det         Integer}
-Det         \lptype{List}{Det         Integer}
+OrderNonDet \lptype{List}{Det         Int}
+Det         \lptype{List}{Det         Int}
 
 \end{Verbatim}
-\end{smaller}
 \end{minipage}
@@
 NonDet      \lptype{List}{OrderNonDet \lptype{Set}{...}}
-NonDet      \lptype{List}{Det         Integer}
-OrderNonDet \lptype{List}{NonDet      Integer}
-Det         \lptype{List}{NonDet      Integer}
+NonDet      \lptype{List}{Det         Int}
+OrderNonDet \lptype{List}{NonDet      Int}
+Det         \lptype{List}{NonDet      Int}
 Det         \lptype{List}{OrderNonDet \lptype{Set}{...}}
@@
 
- Although \<NonDet> collections containing \<Det> or \<OrderNonDet> elements could exist, they are
-disallowed to prevent the following bug:
+It is possible to store an expression of type \<Det> in a \<NonDet>
+collection, but all elements retrieved from a \<NonDet> collection have
+type \<NonDet>.  If the type \ptype{NonDet List}{Det String} were
+permitted, the following type hole would exist:
 
-\begin{verbatim}
-public static void add(@NonDet List<@Det String> list, @NonDet int index, @Det String s) {
-    list.add(index, s);
-}
+\begin{Verbatim}[commandchars=\\\{\}]
+\lptype{Det    List}{Det String} ddlist
+\lptype{NonDet List}{Det String} ndlist = ddlist   // permitted by subtyping rules
+NonDet Int ni = ...
+ndlist[ni] = "anywhere"
+\end{Verbatim}
 
-public static void f(@Det List<@Det String> list, @NonDet int index, @Det String s) {
-    add(list, index, s);
-}
-\end{verbatim}
+\noindent
+The deterministic string \<"anywhere"> is placed in the list at an
+arbitrary (that is, nondeterministic) index.  This must be prohibited,
+because it unsoundly permits the deterministic list \<ddlist> to differ
+from execution to execution.
 
-The possibility of mutation allows us to add to the \<Det List> at a
-\<NonDet> index, which is unsound.
+\subsubsection{Behavior of order-nondeterministic collections}
 
-\todo{Some notes:}
-Typical type for \<get> is \\
-$\forall \tau . \<List>\angles{\tau} \times \<int> \rightarrow \tau$ \\
+A collection of type \<OrderNonDet> has special properties, including the following
+
+\begin{enumerate}
+\item
+The individual elements retrieved from it have type \<NonDet>.  This
+affects access, iteration, searching, and more.
+(Rules \rulename{iteration} and \rulename{next element} in Figure~\ref{fig-ordernondet-rules})
+\item
+Size-related operations return a deterministic result.  This also affects
+queries of whether an iterator has more elements.
+\item
+If the collection is sorted, or its elements are placed in a collection
+that does sorting, the result is deterministic.
+\end{enumerate}
+
+To state the first point more formally, the 
+typical type for list access, such as \<get>, is \\
+$$\forall \tau . \<List>\angles{\tau} \times \<int> \rightarrow \tau$$
+but this type is incorrect for \ourTypeSystem.  Each of the following is a 
+$$
+\forall \kappa\ \beta . \<List>\angles{\tau} \times \<int> \rightarrow \tau \\
+\forall \tau . \<List>\angles{\tau} \times \<int> \rightarrow \tau \\
+\forall \tau . \<List>\angles{\tau} \times \<int> \rightarrow \tau
+$$
+
 but we need to change that.  If either argument is \<OrderNonDet> or \<NonDet>, then the
@@
 \todo{In that case, we know that $\tau = \<Det> \beta$.}
-
-A collection or map annotated as \<OrderNonDet> has the following properties:
-%\todo{Write these as type rules.}
-\begin{enumerate}
-    \item The individual elements extracted by iterating over the collection get the type \<NonDet> (Rules \rulename{iteration} and \rulename{next element} in Figure~\ref{fig-ordernondet-rules}). 
-    Rule \rulename{hasnext element} states that querying if the iterator of an \<OrderNonDet Collection> \<hasNext()> will result in a \<Det boolean>.
-    \item Accessing properties such as \<size()> or querying if the collection \<isEmpty()> will return types
-    that are annotated as \<Det> as they do not depend on the iteration order. 
-\end{enumerate}
 

### 2019-04-01 experiments.tex
formatting

@@
 All these benchmarks and instructions to reproduce our results can be found at \href{https://github.com/t-rasmud/determinismCaseStudies}{https://github.com/t-rasmud/determinismCaseStudies}.
-\todo{P4, P12 and P21 det checker isn't recognized?}
+\todo{P4 compile error: dependencies}
 
@@
         P3 & Yes & DateTimeZone.java: [731,27] &  return type incompatible \\ 
-        P4 & * & Det checker not found  &   \\ 
+        P4 & * & compile dependency failure  &   \\ 
         P5 & ** & Class.getDeclaredMethods not found &    \\ 

### 2019-04-01 design.tex
formatting
Edits in section 2.3

@@
 
-\begin{wrapfigure}{R}{0.3\textwidth}
+%% Reinstate wrapfigure later.  But, if it spans pages, every following
+%% page to the end of the document is narrowed to accommodate it.
+% \begin{wrapfigure}{R}{0.3\textwidth}
+\begin{figure}
     \begin{center}
@@
     \label{fig:determinism-hierarchy}
-%\todo{Drop the at-signs (@).}
-%\todo{The text is weirdly small for the box.
-%  Probably redraw this.}
-\end{wrapfigure}
+\end{figure}
+% \end{wrapfigure}
 
@@
 
-A collection of type \<OrderNonDet> has special properties, including the following
+A collection of type \<OrderNonDet> has special properties, including the following.
 
@@
 The individual elements retrieved from it have type \<NonDet>.  This
-affects access, iteration, searching, and more.
+affects access, iteration, searching, etc.
 (Rules \rulename{iteration} and \rulename{next element} in Figure~\ref{fig-ordernondet-rules})
@@
 To state the first point more formally, the 
-typical type for list access, such as \<get>, is \\
-$$\forall \tau . \<List>\angles{\tau} \times \<int> \rightarrow \tau$$
+typical type for a list access operation, such as \<get>, is
+$$\forall \tau . \<List>\angles{\tau} \times \<Det Int> \rightarrow \tau$$
 but this type is incorrect for \ourTypeSystem.  Each of the following is a
-$$
-\forall \kappa\ \beta . \<List>\angles{\tau} \times \<int> \rightarrow \tau \\
-\forall \tau . \<List>\angles{\tau} \times \<int> \rightarrow \tau \\
-\forall \tau . \<List>\angles{\tau} \times \<int> \rightarrow \tau
-$$
+correct (but overly restrictive) type for \<get>:
+\begin{eqnarray*}
+\forall \kappa\ \beta . \<NonDet>\ \<List>\angles{\kappa\ \beta} \times \<Det Int>
+& \strut\hspace{-1em} \rightarrow \<NonDet>\ \beta
+& \mbox{\quad in this case $\kappa = \<NonDet>$} \\
+\forall \kappa\ \beta . \<OrderNonDet>\ \<List>\angles{\kappa\ \beta} \times \<Det Int>
+& \strut\hspace{-1em} \rightarrow \<NonDet>\ \beta
+& \mbox{\quad in this case $\kappa \in \{ \<OrderNonDet>, \<Det> \} $} \\
+\forall \kappa\ \beta . \<Det>\ \<List>\angles{\kappa\ \beta} \times \<Det Int>
+& \strut\hspace{-1em} \rightarrow \<Det>\ \beta\phantom{\<Non>}
+& \mbox{\quad in this case $\kappa = \<Det>$}
+\end{eqnarray*}
 
-but we need to change that.  If either argument is \<OrderNonDet> or \<NonDet>, then the
-result is \<NonDet>.
-If both arguments are \<Det>, the result is $\tau$ as usual.
-\todo{In that case, we know that $\tau = \<Det> \beta$.}
+\noindent
+and no type introduced so far can express this polymorphism.
+(The typing is actually even more complex, because if \emph{either}
+argument to \<get> is \<OrderNonDet> or \<NonDet>, then the
+result is \<NonDet>.  The above examples only show the case where the index
+is deterministic.)
+The typical type fro the list access operation only applies if both
+arguments are \<Det>,
+and our typing must have that as a special case, which is the first line above.
 
@@
 
-\OurTypeSystem supports three types of polymorphism:  type
-polymorphism, basetype polymorphism, and qualifier polymorphism.
-These apply to both classes and methods.
+\todo{Cut parts of the below; show the rest in Greek.}
+
+\OurTypeSystem supports both basetype polymorphism and qualifier polymorphism.
+When both are applied, it gives the effect of type polymorphism.
+Polymorphism applies to both classes and methods.
+
 \todo{Greekify the following}
@@
 \item
+Type polymorphism has the usual semantics.  For example, the type of the
+identity function is $\forall \tau . \tau \rightarrow \tau$, which can be
+equivalently written as
+$\forall \kappa \ \beta . \kappa \ \beta \rightarrow \kappa \ \beta$.
+\item
+Basetype polymorphism lets the basetype vary without applies a
+
+An example of a type-polymorphic method is identity
+
 In \theDeterminismCheckerImplementation,

### 2019-04-01 experiments.tex
expts

@@
 in~\cite{nondex}.
-Of these projects, \<handlebars> has a compile time failure which prevents us from running \TheDeterminismChecker on it.
+Of these projects, \<handlebars> and \<oryx> have compile time failures which prevents us from running \TheDeterminismChecker on them.
 For benchmarks \<visualee>, \<easy-batch>, and \<vraptor>, we couldn't find
@@
 All these benchmarks and instructions to reproduce our results can be found at \href{https://github.com/t-rasmud/determinismCaseStudies}{https://github.com/t-rasmud/determinismCaseStudies}.
-\todo{P4 compile error: dependencies}
 
@@
 \begin{table}[]
-    \begin{tabular}{|p{1.5cm}|p{1cm}|p{4cm}|p{5cm}|}
+    \begin{tabular}{|p{1.5cm}|p{1cm}|p{5cm}|p{5cm}|}
         \hline
@@
         P6 & Yes & Option.java: [157,28]  &  assignment type incompatible\\
-        P7(test error) & Yes* & checker/ tests/ determinism/P71.java: 5 & return type incompatible \\ 
+        P7(test error & Yes & MultilineRecursiveToString..: [95,35] argument type incompatible & MultilineRecursiveToString...: [208,50]  return type incompatible \\ 
         P7(2) & Yes & FieldUtils.java: [219,84] & assignment type incompatible \\

### 2019-04-01 related.tex
related todo

@@
 \todo{Check whether our JDK is missing anything they support.}
-\todo{Did the NonDex authors contact the maintainers of the tools in which
-  they found nondeterminism?}
+%\todo{Did the NonDex authors contact the maintainers of the tools in which
+%  they found nondeterminism?}
+\todo{The NONDEX authors didn't contact the maintainers of tools.
+From the paper: "NONDEX detected 57 previously unknown
+flaky tests in open-source projects and three flaky tests
+that have been already fixed by the open-source software
+developers."}

### 2019-04-01 experiments.tex
Fix capitalization of NonDex

@@
 \todo{We investigated the tests covering code that \theDeterminismChecker reported as nondeterministic.
-If code is covered by tests, NONDEX couldn't discover it. Else, its a limitation of NONDEX}.
+If code is covered by tests, NonDex couldn't discover it. Else, its a limitation of NonDex.}
 This indicates that the additional programmer effort required by \TheDeterminismChecker is justified
@@
 
-All the flakiness that NONDEX found was due to 7 methods in 3 classes:
+All the flakiness that NonDex found was due to 7 methods in 3 classes:
 \<Class.getDeclaredFields>, \<Class.getDeclaredMethods>, \<Class.getFields>, \<DateFormatSymbols.getZoneStrings>,
@@
 debugging test failures due to ADINS code". \TheDeterminismChecker reports source locations
-which makes it easier for the programmer to fix issues and is orthogonal to the approach taken by NONDEX which
+which makes it easier for the programmer to fix issues and is orthogonal to the approach taken by NonDex which
 report failing tests.
@@
         nutz [10] issue \#1283 &  Removed tests (Printing (HTTP's) Response.getstatus())\\ \hline
-        togglz [19] issue \#223 & (use NONDEX for debugging) \\ \hline
+        togglz [19] issue \#223 & (use NonDex for debugging) \\ \hline
         togglz [20] issue \#240 &  \\ \hline

### 2019-04-01 related.tex
Fix capitalization of NonDex

--- 2019-03-31 related.tex
@@
 \todo{Check whether our JDK is missing anything they support.}
-\todo{Did the NonDex authors contact the maintainers of the tools in which
-  they found nondeterminism?}
+%\todo{Did the NonDex authors contact the maintainers of the tools in which
+%  they found nondeterminism?}
+\todo{The NonDex authors didn't contact the maintainers of tools.
+From the paper: "NonDex detected 57 previously unknown
+flaky tests in open-source projects and three flaky tests
+that have been already fixed by the open-source software
+developers."}

### 2019-04-02 experiments.tex
minor

--- 2019-04-01 experiments.tex
@@
         achilles, [3] issue \#310 & Same fix for \#309 and \#310 \\ \hline
-        checkstyle [4] issue \#4664& det checker not found \\ \hline
+        checkstyle [4] issue \#4664& det checker on test files \\ \hline
         jackrabbit-oak [6] \#6512&  Unresolved on issue tracker \\ \hline
@@
         nutz [10] issue \#1283 &  Removed tests (Printing (HTTP's) Response.getstatus())\\ \hline
-        togglz [19] issue \#223 & (use NONDEX for debugging) \\ \hline
+        togglz [19] issue \#223 & build issue \\ \hline
         togglz [20] issue \#240 &  \\ \hline

### 2019-04-02 experiments.tex

--- 2019-04-01 experiments.tex
@@
         nutz [10] issue \#1283 &  Removed tests (Printing (HTTP's) Response.getstatus())\\ \hline
-        togglz [19] issue \#223 & (use NonDex for debugging) \\ \hline
+        togglz [19] issue \#223 & build issue \\ \hline
         togglz [20] issue \#240 &  \\ \hline

### 2019-04-02 experiments.tex
edits to randoop case study

@@
 \begin{table}[]
-    \begin{tabular}{|l|l|}
+    \begin{tabular}{|l|p{5cm}|}
         \hline
@@
         achilles, [3] issue \#310 & Same fix for \#309 and \#310 \\ \hline
-        checkstyle [4] issue \#4664& det checker on test files \\ \hline
-        jackrabbit-oak [6] \#6512&  Unresolved on issue tracker \\ \hline
+        checkstyle [4] issue \#4664&test (added to test cases in det checker) \\ \hline
+        jackrabbit-oak [6] \#6512&  Unresolved on issue tracker (Race condition issue) \\ \hline
         jackrabbit-oak [7] issue \#6524&  Dead link, didn't find 6524 in issue tracker \\ \hline
-        nutz [10] issue \#1283 &  Removed tests (Printing (HTTP's) Response.getstatus())\\ \hline
+        nutz [10] issue \#1283 &  Removed tests (Printing (HTTP's) Response.getstatus()) has Chinese characters,
+        compilation with det checker stuck!\\ \hline
         togglz [19] issue \#223 & build issue \\ \hline
-        togglz [20] issue \#240 &  \\ \hline
+        togglz [20] issue \#240 &  build issue\\ \hline
     \end{tabular}

### 2019-04-02 randoop-case-study.tex
edits to randoop case study

@@
 
-The developers took extensive action to detect and mitigate the effect of, nondeterminism.
+The developers took extensive action to detect and mitigate the effect of nondeterminism.
 They used Docker images to run tests, to avoid system dependencies such as
@@
 possible answers --- for example, in code coverage of generated tests.
-They used linters such as Error Prone, which warns if toString is used on
-array objects, which don't override Object.toString and therefore print a
+They used linters such as Error Prone, which warns if \<toString> is used on
+array objects, which don't override \<Object.toString> and therefore print a
 hash code which may vary from run to run.
@@
 other nondeterministing output.
-Each of these efforts wase helpful, but they were not enough to solve the problem.
+Each of these efforts was helpful, but they were not enough to solve the problem.
 
@@
  and eliminate it.
-The most common causes were toString() routines and iteration order of sets and maps.
+The most common causes were \<toString()> routines and iteration order of sets and maps.
 Some of the nondeterminism was in libraries, such as the JDK\@.
 
-The developers chose not to make every Map a LinkedHashMap, because that
+The developers chose not to make every \<Map> a \<LinkedHashMap>, because that
 would have increased memory and CPU costs.  They chose not to make every
-order-nondeterminism List a Set, for similar reasons:  deduplication was
+order-nondeterministic \<List> a \<Set>, for similar reasons:  deduplication was
 not always needed and would have increased costs.
@@
 \TheDeterminismChecker found three previously-unknown nondeterminism bugs in Randoop.
-The Randoop developers accepted our bug reports and committed fixes to the repository.
+The Randoop developers accepted our bug reports and committed fixes to the repository. A summary
+of these bugs follows:
 
@@
 \item
- Diagnostic output that printed a hash code for brevity, but should have had deterministic output\commit{661a4970}.  By contrast to the other two bugs, this one is not serious.
-
-\item
 Dependency on the CLASSPATH environment variable in preference to the
@@
 The developers changed Randoop to not read the environment variable.
+
+\item 
+Diagnostic output that printed a hash code for brevity, but should have had deterministic output\commit{661a4970}.  By contrast to the other two bugs, this one is not serious.
 \end{itemize}
 
-We reported another suspicous case of order-nondeterminism, in the
+We reported another suspicious case of order-nondeterminism, in the
 \<findOverridden> method.  The Randoop developers explained it was

### 2019-04-03 design.tex
Edits in section 2.4

@@
 %     \end{verbatim}
-% \caption{Example: Invalid assingment}
+% \caption{Example: Invalid assignment}
 % \label{code-invalid1}
@@
     $\tau$ & ::= & $\kappa\ \beta$ & \mbox{type} \\
-    $\beta$ & ::= & $C$\bnfalt $C\angles{\overline{\tau}}$ & \mbox{unqualified type}
+    $\beta$ & ::= & $C$\bnfalt $C\angles{\overline{\tau}}$ 
+    & \mbox{unqualified type}
   \end{tabular}
@@
 rules of \ourTypeSystem. 
+(For brevity, we use ``valid'' synonymously with ``well-formed'' and
+``invalid'' synonymously with ``non-well-formed''.)
 As stated in the rule \rulename{noncollection}, the types \<Det> and \<NonDet>
@@
 
-\subsubsection{Behavior of order-nondeterministic collections}
+\subsubsection{Behavior of order-nondeterministic collections}\label{sec:ond-behavior}
 
@@
 typical type for a list access operation, such as \<get>, is
-$$\forall \tau . \<List>\angles{\tau} \times \<Det Int> \rightarrow \tau$$
+$$\forallt{\tau} \<List>\angles{\tau} \times \<Det Int> \rightarrow \tau$$
 but this type is incorrect for \ourTypeSystem.  Each of the following is a
@@
 \begin{eqnarray*}
-\forall \kappa\ \beta . \<NonDet>\ \<List>\angles{\kappa\ \beta} \times \<Det Int>
+\forallt{\kappa\ \beta} \<NonDet>\ \<List>\angles{\kappa\ \beta} \times \<Det Int>
 & \strut\hspace{-1em} \rightarrow \<NonDet>\ \beta
 & \mbox{\quad in this case $\kappa = \<NonDet>$} \\
-\forall \kappa\ \beta . \<OrderNonDet>\ \<List>\angles{\kappa\ \beta} \times \<Det Int>
+\forallt{\kappa\ \beta} \<OrderNonDet>\ \<List>\angles{\kappa\ \beta} \times \<Det Int>
 & \strut\hspace{-1em} \rightarrow \<NonDet>\ \beta
 & \mbox{\quad in this case $\kappa \in \{ \<OrderNonDet>, \<Det> \} $} \\
-\forall \kappa\ \beta . \<Det>\ \<List>\angles{\kappa\ \beta} \times \<Det Int>
+\forallt{\kappa\ \beta} \<Det>\ \<List>\angles{\kappa\ \beta} \times \<Det Int>
 & \strut\hspace{-1em} \rightarrow \<Det>\ \beta\phantom{\<Non>}
@@
     $
-    \forall \tau . \|OrderNonDet \ C| \angles{\tau} \rightarrow 
+    \forallt{\tau} \|OrderNonDet \ C| \angles{\tau} \rightarrow 
     \|OrderNonDet \ Iterator| \angles{\tau}
@@
     $
-    \forall \kappa \beta . \|OrderNonDet \ Iterator|\angles{\kappa \ \beta} \rightarrow
+    \forallt{kappa \beta} \|OrderNonDet \ Iterator|\angles{\kappa \ \beta} \rightarrow
     \|NonDet|\ \beta
@@
     $
-    \forall \tau . \|OrderNonDet \ Iterator|\angles{\tau} \rightarrow
+    \forallt{\tau} \|OrderNonDet \ Iterator|\angles{\tau} \rightarrow
     \|Det \ boolean|
@@
     $
-    \forall \tau . \|OrderNonDet \ C|\angles{\tau} \rightarrow
+    \forallt{\tau} \|OrderNonDet \ C|\angles{\tau} \rightarrow
     \|Det \ int|
@@
 
-\todo{Cut parts of the below; show the rest in Greek.}
+\todo{Can we come up with more examples of interesting functions to
+  describe here?  Or put them in section 3.}
 
@@
 Type polymorphism has the usual semantics.  For example, the type of the
-identity function is $\forall \tau . \tau \rightarrow \tau$, which can be
+identity function is $\forallt{\tau} \tau \rightarrow \tau$, which can be
 equivalently written as
-$\forall \kappa \ \beta . \kappa \ \beta \rightarrow \kappa \ \beta$.
+$\forallt{\kappa \ \beta} \kappa \ \beta \rightarrow \kappa \ \beta$.
+(Compare this to 
+$\forallt{\kappa} \forallt{\beta} \kappa \ \beta \rightarrow \kappa \ \beta$
+which differs because the qualifier and basetype can vary independently;
+see below.)\todo{Does the foregoing make sense?}
+
 \item
-Basetype polymorphism lets the basetype vary without applies a
+  Basetype polymorphism lets the basetype vary indepentently of the
+  qualifier, which might be fixed or might be polymorphic.
+  An example is a nondeterministic \<choose> function, which might have type
 
-An example of a type-polymorphic method is identity
+  $\forallt{\beta} \<Set>\angles{\<Det>\ \beta} \rightarrow \<NonDet>\ \beta$
 
-In \theDeterminismCheckerImplementation,
-type polymorphism is handled by Java's generics mechanism, which
-\theDeterminismChecker fully supports, including class and method generics,
-inference, etc.
-Given the Java declaration
-\begin{Verbatim}
-<T> T identity(T arg) { return arg; }
-\end{Verbatim}
-the type of \<identity(arg)> is the same as the type of
-\<arg>, and a programmer can write a type such as \codeid{List<@NonDet
-  Integer>} (but see below for details about collections).
+  Basetype polymorphism is rarely used.  Full type polymorphism is more
+  common, even if the type is decomposed and the parts used independently,
+  as in this more general signature for the \<choose> function:
+
+  $\forallt{\kappa\ \beta} \<Set>\angles{\kappa\ \beta} \rightarrow \<NonDet> \beta$
+
 \item
-Basetype polymorphism is enabled by writing a type qualifier on a use of a
-type variable, which overrides the type qualifier at the instantiation
-site.
-For example, a \<choose> operation on sets could be defined in Java as
-\begin{Verbatim}
-class ChooseableSet<T> implements Set<T> {
-  // return an arbitrary element from the set
-  @NonDet T choose();
-}
-\end{Verbatim}
-\item
-Java does not provide a syntax that can be used for qualifier polymorphism,
-so we use a special type qualifier name, \<@PolyDet>.
-A qualifier-polymorphic method \<m> with signature $\forall \kappa . \kappa\ \<int> \times \<Det boolean> \rightarrow
-\kappa\ \<String>$ would be written in our dialect of Java as
-\begin{Verbatim}
-@PolyDet String m(@PolyDet int, @Det boolean)
-\end{Verbatim}
-Each use of \<@PolyDet> stands for a use of the qualifier variable
-$\kappa$, and there is no need to declare the qualifier variable $\kappa$.
-\TheDeterminismChecker currently supports
-qualifier
-polymorphism on methods but not on classes.
-Therefore, \<@PolyDet> may be written in methods (signatures and bodies)
-but not on fields.
+Qualifier polymorphism is common.  For example, here is the signature for
+the \<length> method on strings:
+
+$\<length>\ :\ \forallt{\kappa} . \kappa\ \<String> \rightarrow \kappa \<Int>$
+
+\noindent
+and here is that for the addition operation:
+
+$\<plus>\ :\ \forallt{\kappa} . \kappa\ \<Int> \times \kappa\ \<Int> \rightarrow \kappa \<Int>$
+
 \end{itemize}
 
-When a user writes a polymorphic qualifier on a method signature or a type parameter,
-it indicates that it could be instantiated
-%\todo{Should all occurrences of ``resolve'' be ``can be instantiated''} 
-to any type in the type system depending on how it is used.
-In \theDeterminismChecker, we define a polymorphic qualifier \<PolyDet>.
-One of the most common locations for a polymorphic qualifier is at a method signature.
-Consider the following declaration of addition:
-\begin{verbatim}
-@PolyDet int plus(@PolyDet int a, @PolyDet int b) { return a+b; }
-\end{verbatim}
-\todo{It has the standard meaning.  There are two ways to interpret this.}
-There are two valid instantiations for this declaration:
+\OurTypeSystem uses the standard meaning of parametric
+polymorphism~\cite{Abadi:1989:FIM:77350.77373,Plotkin:1993:LPP:645891.671433}.
+A polymorphic abstraction (such as a class or function) is written and
+typechecked once, but can be instantiated differently at different uses,
+each of which behaves as if it was written with a different non-polymorphic
+type.
+
+Informally, the \<length> above acts as if it had multiple
+(overloaded) definitions
+
+$\<NonDet>\ \<String> \rightarrow \<NonDet> \<Int>$
+
+$\<Det>\ \<String> \rightarrow \<Det> \<Int>$
+
+The body must typecheck at every possible instantiation, and at a use site,
+the most specific applicable instantiation is chosen.
+Here is an example:
+(put it in a table)
+
 \begin{Verbatim}
-  @NonDet int plus(@NonDet int a, @NonDet int b) { return a+b; }
-  @Det int plus(@Det int a, @Det int b) { return a+b; }
+NonDet Int ni;
+   Det Int di;
+
+plus(ni, ni)  // has type NonDet Int; most specific applicable instantiation is NonDet Int -> NonDet Int
+plus(di, di)  // has type Det Int; most specific applicable instantiation is Det Int -> Det Int
+plus(ni, di)  // has type NonDet Int; most specific applicable instantiation is NonDet Int -> NonDet Int
 \end{Verbatim}
-This indicates that method \<foo> can be called with arguments having any of the \codeid{Det or NonDet} type qualifiers.
-%   (\<@OrderNonDet> is not allowed because it is invalid on primitive types).
- \<PolyDet> gets instantiated to the least upper bound of
-the actual types on arguments.
-For instance, if this method is called as \codeid{foo(@Det String arg1, @NonDet boolean arg2)}, \theDeterminismChecker instantiates the method declaration as \codeid{@NonDet int foo(@NonDet String param1, @NonDet boolean param2)}
-causing the return type to have the type qualifier \<NonDet>.
 
-\subsection{Polymorphism rules for collections}\label{polymorphism-up-down}
+We adopt the convention that  polymorphism is not instantiated in ways
+that would create invalid types.  For example, the \<plus> polymorphic
+function would not be instantiated at $\kappa = \<OrderNonDet>.$
 
-Some types are invalid and must not be permitted (during polymorphic instantiation).
+\subsubsection{Polymorphism rules for collections}\label{polymorphism-up-down}
 
-Recall that \theDeterminismChecker does not permit \<OrderNonDet> on a type
-that isn't a collection.
-As a consequence, resolution of polymorphic type qualifiers at method signatures could have undesirable effects.
-Suppose a method declaration is annotated as \codeid{PolyDet int size(PolyDet List<T> list)}. Invoking this method
-with an argument of type \codeid{OrderNonDet List<Det String>} will result in \<PolyDet> getting instantiated to \<OrderNonDet> which is invalid
-on the return type \<int>. We circumvent this scenario by introducing two variants to the \<PolyDet> qualifier, \PolyDetUp
-and \PolyDetDown.
+As described so far, polymorphism cannot express the collection behaviors
+of \cref{sec:ond-behavior}.
+Consider this typing for the \<size> function:
+
+$\<size>\ :\ \forallt{\kappa} \kappa\ \<Collection> \rightarrow \kappa\ \<Int>$
+
+It cannot be instantiated as
+
+$\<size>\ :\ \<OrderNonDet>\ \<Collection> \rightarrow \<OrderNonDet> \<Int>$
+
+because such an instantiation would include the invalid type \<OrderNonDet Int>.
+A similar problem occurs for the \<first> operation that returns the first
+element of a list.
+
+\OurTypeSystem resolves this problem by introducing two variants to the
+\<PolyDet> qualifier, \PolyDetUp\ and \PolyDetDown.
+The $\uparrow$ operator converts \<OrderNonDet> to \<NonDet> and leaves the
+other qualifiers unchanged.  The upward-pointing arrow is mnemonic of how
+\<OrderNonDet> is changed:  it replaced by something higher in the type hierarchy.
+The $\downarrow$ operator is analogous, but converts \<OrderNonDet> to
+\<Det>, which is lower in the type hierarchy.
+\Cref{fig:poly-resolutions} formalizes their behavior.
 
@@
 
-The semantics of these variants are as follows:
-\begin{itemize}
-    \item \PolyDetUp: Same as \<PolyDet> except when it is instantiated as \<OrderNonDet>, it is replaced with \<NonDet>.
-    \item \PolyDetDown: Same as \<PolyDet> except when it is instantiated as \<OrderNonDet>, it is replaced with \<Det>.
-\end{itemize} 
-The modifiers $\uparrow$ and $\downarrow$ are indicative of the subtyping relationship of \<OrderNonDet> with \<NonDet> and \<Det> respectively
-in the determinism type hierarchy.
+%\Cref{fig-poly-resolution} presents the rule for type checking of methods annotated with \<PolyDet> or its variants.
 
-%\Cref{fig-poly-resolution} presents the rule for type checking of methods annotated with \<PolyDet> or its variants.
-A polymorphic method type checks if every instantiation of \<PolyDet> type checks. If this method has \PolyDetUp\ or \PolyDetDown\
-in its signature then the instantiation of \<PolyDet> at that location gets replaced by the appropriate type as shown in \cref{fig:poly-resolutions}.
+The correct type for \<size> is
 
-Annotating \<size()> as \codeid{PolyDet$\downarrow$ int size(PolyDet List<T> list)} will only allow the following type instantiations:
-\begin{itemize}
-    \item \codeid{@Det int size(@Det List<T> list)}
-    \item \codeid{@Det int size(@OrderNonDet List<T> list)}
-    \item \codeid{@NonDet int size(@NonDet List<T> list)}
-\end{itemize}
+$\<size> : \forallt{\kappa} \kappa\ \<Collection> \rightarrow \down{\kappa} \<Int>$
+
+This can be instantiated at all three type qualifiers without creating any
+invalid types, and the instantiations implement the semantics of \cref{sec:ond-behavior}:
+
+$\<size> : \<NonDet>\ \<Collection> \rightarrow \<NonDet> \<Int>$
+
+$\<size> : \<OrderNonDet>\ \<Collection> \rightarrow \<Det> \<Int>$
+
+$\<size> : \<Det>\ \<Collection> \rightarrow \<Det> \<Int>$
+
+An example use of $\uparrow$ is the \<first> operation.  A simplified
+version of its type is
+
+$\<first> : \forallt{\kappa\ \beta} \kappa\ \<List>\angles{\kappa\ \beta} \rightarrow \up{\kappa}\ \beta$
+
+whose type qualifier can be instantiated as
+
+$\<first> : \forallt{\beta} \<NonDet>\ \<List>\angles{\<NonDet>\ \beta} \rightarrow \<NonDet>\ \beta$
+
+$\<first> : \forallt{\beta} \<OrderNonDet>\ \<List>\angles{\<OrderNonDet>\ \beta} \rightarrow \<NonDet>\ \beta$
+
+$\<first> : \forallt{\beta} \<Det>\ \<List>\angles{\<Det>\ \beta} \rightarrow \<Det>\ \beta$
+
+The actual type of \<first> accounts for more possible types for the type arguments.
 
 %\begin{figure}
-%    $\infer[\rulename{poly resolution}]{\pi_r \ \beta_r\ \<m>(\overline{\pi_i \ \beta_i})\{\ \}}{\forall \kappa \in \{\|NonDet, OrderNonDet, Det|\} : \kappa \ \beta_r\ \<m>(\overline{K \ \beta_i})\{\ \} & \pi_r, \pi_i \in \{\|PolyDet|, \PolyDetUp, \PolyDetDown    \}}$
+%    $\infer[\rulename{poly resolution}]{\pi_r \ \beta_r\ \<m>(\overline{\pi_i \ \beta_i})\{\ \}}{\forallt{\kappa \in \{\|NonDet, OrderNonDet, Det|\}} \kappa \ \beta_r\ \<m>(\overline{K \ \beta_i})\{\ \} & \pi_r, \pi_i \in \{\|PolyDet|, \PolyDetUp, \PolyDetDown    \}}$
 %    \caption{Polymorphic resolution rule}
@@
     $\infer[\rulename{ordernondet polyup}]
-    {\tau \uparrow \ = \|NonDet \ \beta|}
-    {\tau \ = \|OrderNonDet \ \beta|
+    {\up{\tau} = \|NonDet \ \beta|}
+    {\tau = \|OrderNonDet \ \beta|
         & \wellformed{\tau}
@@
     $\infer[\rulename{polyup}]
-    {\tau \uparrow \ = \kappa \ \beta}
-    {\tau \ =\kappa \ \beta
+    {\up{\tau} = \kappa \ \beta}
+    {\tau =\kappa \ \beta
         & \kappa \in \{\|Det, NonDet|\}
@@
     $\infer[\rulename{ordernondet polydown}]
-    {\tau \downarrow \ = \|Det \ \beta|}
-    {\tau \ = \|OrderNonDet \ \beta|
+    {\down{\tau} = \|Det \ \beta|}
+    {\tau = \|OrderNonDet \ \beta|
         & \wellformed{\tau}
@@
     $\infer[\rulename{polydown}]
-    {\tau \downarrow \ = \kappa \ \beta}
-    {\tau \ =\kappa \ \beta
+    {\down{\tau} = \kappa \ \beta}
+    {\tau =\kappa \ \beta
         & \kappa \in \{\|Det, NonDet|\}
@@
 
-In order to provide soundness guarantees, we must ensure that a collection that is
-declared as \<Det> is not side-effected in non-deterministic ways. Consider the following qualifier 
+A simplified type for the list \<get> method for lists of strings is
+
+$\<get> : \forallt{\kappa} \kappa\ \<List>\angles{\kappa\ \<String>} \times \kappa\ \<Int> \rightarrow \up{\kappa}\ \<String>$
+
+(The actual typing permits the qualifier on the type argument to differ
+from the type qualifier on the list, and it handles arbitrary basetypes
+including collections that may themselves be order-nondeterministic.)
+
+Instantiations are:
+
+$\<get> : \forallt{\kappa\ \beta} \<List>\angles{\kappa\ \beta} \times \kappa\ \<Int> \rightarrow \up{\kappa}\ \beta$
+
+$\<get> : \forallt{\kappa\ \beta} \<List>\angles{\kappa\ \beta} \times \kappa\ \<Int> \rightarrow \up{\kappa}\ \beta$
+
+$\<get> : \forallt{\kappa\ \beta} \<List>\angles{\kappa\ \beta} \times \kappa\ \<Int> \rightarrow \up{\kappa}\ \beta$
+
+\todo{We don't need the OrderNonDet instantiation.  The other two
+  instantiations are enough.
+  We don't need the following instantiation:}
+\begin{Verbatim}
+   // instantiation: OrderNonDet List<Det String> x _ Int -> NonDet String 
+\end{Verbatim}
+
+If either the list or the index is non-deterministic, then the result is
+nondeterministic:
+
+\begin{Verbatim}
+NonDet Int ni;
+   Det Int di;
+NonDet List<NonDet String> nl;
+OrderNonDet List<Det String> ol;
+Det List<Det String> dl;
+
+get(nl, ni) : NonDet String // instantiation: NonDet List<NonDet String> x NonDet Int -> NonDet String
+get(nl, di) : NonDet String
+get(ol, ni) : NonDet String
+get(ol, di) : NonDet String
+get(dl, ni) : NonDet String
+get(dl, di) : Det String    // instantiation: NonDet List<Det String> x Det Int -> Det String
+\end{Verbatim}
+
+A similar typing does \emph{not} work for the \<set> function that sets an
+element of a list, because it would permit a deterministic collection must
+not be side-effected in non-deterministic ways.  Consider the following
+type for
+
+$\<set> : \forall \forallt{\tau}$
+
+qualifier 
 on the List add method: \codeid{public void add(PolyDet ArrayList<E> this, PolyDet int index, E element)}.

### 2019-04-03 java-design.tex
Edits in sections 2-2.3
Edits in section 2.3
Fix LaTeX error
Edits in section 2.4

@@
 so we use a special type qualifier name, \<@PolyDet>.
-A qualifier-polymorphic method \<m> with signature $\forall \kappa . \kappa\ \<int> \times \<Det boolean> \rightarrow
+A qualifier-polymorphic method \<m> with signature $\forallt{\kappa} \kappa\ \<int> \times \<Det boolean> \rightarrow
 \kappa\ \<String>$ would be written in our dialect of Java as
@@
 breaks our determinism guarantees).
+
+\<@Det> equality is with respect to \<.equals()>.
+
+\subsection{Polymorphism}\label{java-polymorphism}
+
+\OurTypeSystem supports three types of polymorphism:  type
+polymorphism, basetype polymorphism, and qualifier polymorphism.
+These apply to both classes and methods.
+
+\begin{itemize}
+\item
+In \theDeterminismCheckerImplementation,
+type polymorphism is handled by Java's generics mechanism, which
+\theDeterminismChecker fully supports, including class and method generics,
+inference, etc.
+Given the Java declaration
+\begin{Verbatim}
+<T> T identity(T arg) { return arg; }
+\end{Verbatim}
+the type of \<identity(arg)> is the same as the type of
+\<arg>, and a programmer can write a type such as \codeid{List<@NonDet
+  Integer>} (but see below for details about collections).
+\item
+Basetype polymorphism is enabled by writing a type qualifier on a use of a
+type variable, which overrides the type qualifier at the instantiation
+site.
+For example, a \<choose> operation on sets could be defined in Java as
+\begin{Verbatim}
+class ChooseableSet<T> implements Set<T> {
+  // return an arbitrary element from the set
+  @NonDet T choose();
+}
+\end{Verbatim}
+\item
+Java does not provide a syntax that can be used for qualifier polymorphism,
+so we use a special type qualifier name, \<@PolyDet>.
+A qualifier-polymorphic method \<m> with signature $\forallt{\kappa} \kappa\ \<int> \times \<Det boolean> \rightarrow
+\kappa\ \<String>$ would be written in our dialect of Java as
+\begin{Verbatim}
+@PolyDet String m(@PolyDet int, @Det boolean)
+\end{Verbatim}
+Each use of \<@PolyDet> stands for a use of the qualifier variable
+$\kappa$, and there is no need to declare the qualifier variable $\kappa$.
+\TheDeterminismChecker currently supports
+qualifier
+polymorphism on methods but not on classes.
+Therefore, \<@PolyDet> may be written in methods (signatures and bodies)
+but not on fields.
+\end{itemize}
+
+When a user writes a polymorphic qualifier on a method signature or a type parameter,
+it indicates that it could be instantiated
+%\todo{Should all occurrences of ``resolve'' be ``can be instantiated''} 
+to any type in the type system depending on how it is used.
+In \theDeterminismChecker, we define a polymorphic qualifier \<PolyDet>.
+One of the most common locations for a polymorphic qualifier is at a method signature.
+Consider the following declaration of addition:
+\begin{verbatim}
+@PolyDet int plus(@PolyDet int a, @PolyDet int b) { return a+b; }
+\end{verbatim}
+\todo{It has the standard meaning.  There are two ways to interpret this.}
+There are two valid instantiations for this declaration:
+\begin{Verbatim}
+  @NonDet int plus(@NonDet int a, @NonDet int b) { return a+b; }
+  @Det int plus(@Det int a, @Det int b) { return a+b; }
+\end{Verbatim}
+This indicates that method \<foo> can be called with arguments having any of the \codeid{Det or NonDet} type qualifiers.
+%   (\<@OrderNonDet> is not allowed because it is invalid on primitive types).
+ \<PolyDet> gets instantiated to the least upper bound of
+the actual types on arguments.
+For instance, if this method is called as \codeid{foo(@Det String arg1, @NonDet boolean arg2)}, \theDeterminismChecker instantiates the method declaration as \codeid{@NonDet int foo(@NonDet String param1, @NonDet boolean param2)}
+causing the return type to have the type qualifier \<NonDet>.

### 2019-04-03 experiment_tables.tex
edits to nondex expts

--- 2019-04-02 experiments.tex
@@
-\section{Comparison to previous work}
-
-The state of the art in flaky test detection is NonDex~\cite{nondex}.
-\Cref{sec:related} explains how NonDex works.
-
-\begin{itemize}
-    \item goals of evaluation: Compare with NonDex and Deflaker
-    \item Quantitative comparison: \TheDeterminismChecker finds all bugs reported by NonDex and Deflaker?
-    \item Qualitative study of the issues discovered.
-    \item System info: 2.3 GHz Intel Core i5, 8 GB 2133 MHz
-\end{itemize}
-
-\subsection{Case study with NonDex}\label{sec:nondex-randoop}
-
-We ran NonDex on the same commit of Randoop as we used for the case study
-of \cref{sec:randoop-case-study}.  This version contains 3 nondeterminism bugs that
-\theDeterminismChecker found.
-
-We modified Randoop by deleting tests that were skipped by its buildfile,
-because the NonDex gradle plugin does not respect those settings.
-
-After that, NonDex ran without problems on Randoop.
-It completed successfully without reporting any flaky tests.
-
-\todo{check if the bugs discovered by \theDeterminismChecker have test coverage}
-\todo{We investigated the tests covering code that \theDeterminismChecker reported as nondeterministic.
-If code is covered by tests, NONDEX couldn't discover it. Else, its a limitation of NONDEX}.
-This indicates that the additional programmer effort required by \TheDeterminismChecker is justified
-as our tool discovers non-trivial issues and doesn't depend on code coverage or other runtime heuristics.
-
-\subsection{\TheDeterminismChecker on NonDex benchmarks}\label{sec:nondex-benchmarks}
-
-\Cref{sec:nondex-randoop} shows that \theDeterminismChecker finds errors that NonDex does not.
-This section determines whether NonDex finds errors that \theDeterminismChecker does not.
-Its authors ran NonDex on 195 open-source projects, and NonDex found flaky tests in
-21 of them.
-
-The authors also reported the sources of flakiness
-after manually inspecting these tests. We ran \TheDeterminismChecker on all these benchmarks with the objective
-of measuring recall of our tool. We cloned all 21 repos and switched to the commit that had a flakiness issue as reported 
-in~\cite{nondex}.
-Of these projects, \<handlebars> and \<oryx> have compile time failures which prevents us from running \TheDeterminismChecker on them.
-For benchmarks \<visualee>, \<easy-batch>, and \<vraptor>, we couldn't find
-the sources of flakiness as reported in~\cite{nondex} anywhere in the respective repositories.
-\TheDeterminismChecker reported the nondeterministic source locations correctly in the rest of the projects.
-All these benchmarks and instructions to reproduce our results can be found at \href{https://github.com/t-rasmud/determinismCaseStudies}{https://github.com/t-rasmud/determinismCaseStudies}.
-
-All the flakiness that NONDEX found was due to 7 methods in 3 classes:
-\<Class.getDeclaredFields>, \<Class.getDeclaredMethods>, \<Class.getFields>, \<DateFormatSymbols.getZoneStrings>,
-\<HashMap.entrySet>, \<HashMap.keySet>, and \<HashMap.values>.  
-We provide samples of the annotated source code form these benchmarks and the errors as reported by \TheDeterminismChecker 
-in Figures~\ref{code-reflectasm} and~\ref{code-actiongenerator}.
-\begin{figure}
-    \begin{verbatim}
-    static public FieldAccess get (Class type) {
-        @Det ArrayList<@Det Field> fields = new ArrayList<Field>();
-        ...
-        while (nextClass != Object.class) {
-            @Det Field @Det[] declaredFields = nextClass.getDeclaredFields();
-            ...
-        }
-    }
-    [ERROR] FieldAccess.java:[102,87] [assignment.type.incompatible] 
-    incompatible types in assignment.
-    found   : @Det Field @OrderNonDet []
-    required: @Det Field @Det []
-    \end{verbatim}
-    \caption{Annotated source code from reflectasm.}
-    \label{code-reflectasm}
-\end{figure}
-\begin{figure}
-    \begin{verbatim}
-    protected SimpleDataEvent createNextEvent() {
-      for (@Det Entry<@Det String, @Det FieldType> entry : fields.entrySet()) {
-          ...
-      }
-      ...
-    }
-    [ERROR] DataDictionaryEventSource.java:[73,78] [enhancedfor.type.incompatible] 
-    incompatible types in enhanced for loop.
-    found   : @NonDet Entry<@Det String, @Det FieldType>
-    required: @Det Entry<@Det String, @Det FieldType>
-    \end{verbatim}
-    \caption{Annotated source code from ActionGenerator.}
-    \label{code-actiongenerator}
-\end{figure}
-As can be seen from these figures, annotating local arrays and collections
-has been sufficient to discover the sources of nondeterminism in most of the projects.
-The rest of the projects required annotating the return types of methods.
-
-For the benchmark \<commons-lang>, \cite{nondex} report multiple failing tests (14 flaky tests) due to calls
-to \<Class.getDeclaredFields>. Upon inspecting the source code, we find that all of invocations to
-\<Class.getDeclaredFields> are made from 5 locations. This means that annotating 5 lines of source code
-would have been sufficient to identify the nondeterminism. Having said that, we admit that there could be
-significant programmer effort involved in annotating the whole program. On the other hand,~\cite{nondex}
-state "we found that manually inspecting these failures was
-rather challenging, and we leave it as future work to automate
-debugging test failures due to ADINS code". \TheDeterminismChecker reports source locations
-which makes it easier for the programmer to fix issues and is orthogonal to the approach taken by NONDEX which
-report failing tests.
-
-\TheDeterminismChecker is also capable of detecting nondetermism in test cases. We provide a stub file
-for \<junit> in which we annotated all the \<assert> methods as accepting only \<Det> parameters.
-This will report any \<junit> test that passes nondeterministic arguments to \<assert>s and the test doesn't
-even have to be executed in our case.
-\todo{It took one of our authors about an hour to create this stub file - proof of easy extensibility of tool?} 
-\todo{Link to det checker repo?}
-
 \begin{table}[]
@@
 \end{table}
-
-\todo{If there is time, we could do a second case study, on a program that
-  NonDex did not find nondeterminism in.  Our hope would be to find
-  nondeterminism that NonDex did not.  Choose a small project that is
-  maintained (it has recent commits), so we can communicate with the authors.}
-
-\section{deflaker case study}
-\begin{table}[]
-    \begin{tabular}{|l|l|}
-        \hline
-        Benchmark, issue & found by det checker? \\ \hline
-        achilles, [2] issue \#309 &  Race condition\\ \hline
-        achilles, [3] issue \#310 & Same fix for \#309 and \#310 \\ \hline
-        checkstyle [4] issue \#4664& det checker on test files \\ \hline
-        jackrabbit-oak [6] \#6512&  Unresolved on issue tracker \\ \hline
-        jackrabbit-oak [7] issue \#6524&  Dead link, didn't find 6524 in issue tracker \\ \hline
-        nutz [10] issue \#1283 &  Removed tests (Printing (HTTP's) Response.getstatus())\\ \hline
-        togglz [19] issue \#223 & build issue \\ \hline
-        togglz [20] issue \#240 &  \\ \hline
-    \end{tabular}
-\end{table}

### 2019-04-03 experiments.tex
edits to nondex expts

@@
 The state of the art in flaky test detection is NonDex~\cite{nondex}.
-\Cref{sec:related} explains how NonDex works.
-
-\begin{itemize}
-    \item goals of evaluation: Compare with NonDex and Deflaker
-    \item Quantitative comparison: \TheDeterminismChecker finds all bugs reported by NonDex and Deflaker?
-    \item Qualitative study of the issues discovered.
-    \item System info: 2.3 GHz Intel Core i5, 8 GB 2133 MHz
-\end{itemize}
+\Cref{sec:related} explains how NonDex works. In this section, we perform a qualitative comparison of the
+errors reported by NonDex and those reported by \TheDeterminismChecker. We ran all our
+experiments on a 2.3 GHz Intel Core i5 processor with a RAM configuration of 8 GB 2133 MHz.
+\TheDeterminismChecker doesn't show any noticeable performance change at compile time.
+\TheDeterminismChecker runs on any project that builds with maven or gradle and requires
+Java version 8 or higher (support for annotations was added in Java 8).
 
@@
 because the NonDex gradle plugin does not respect those settings. 
-
 After that, NonDex ran without problems on Randoop.
 It completed successfully without reporting any flaky tests.
-
-\todo{check if the bugs discovered by \theDeterminismChecker have test coverage}
-\todo{We investigated the tests covering code that \theDeterminismChecker reported as nondeterministic.
-If code is covered by tests, NonDex couldn't discover it. Else, its a limitation of NonDex.}
+We investigated the tests covering code that \theDeterminismChecker reported as nondeterministic.
+Bug 1 and Bug 2 have corresponding test cases whereas Bug 3 does not.
+As NonDex is a \textit{flaky test} detection tool, Bug 3 is outside of its scope.
+The reason for nondeterminism in Bug 2 was a call to the \<System.getProperty()> method which
+is not modeled by NonDex.
 This indicates that the additional programmer effort required by \TheDeterminismChecker is justified
@@
 Its authors ran NonDex on 195 open-source projects, and NonDex found flaky tests in
-21 of them.
+21 of them. We defer the reader to~\cite{nondex} for a complete list of these projects
+and the details of commits hashes in which NonDex detected flaky tests.
+The authors also reported the sources of flakiness
+after manually inspecting these tests.
 
-The authors also reported the sources of flakiness
-after manually inspecting these tests. We ran \TheDeterminismChecker on all these benchmarks with the objective
-of measuring recall of our tool. We cloned all 21 repos and switched to the commit that had a flakiness issue as reported 
+We ran \TheDeterminismChecker on all these benchmarks with the objective
+of measuring the recall of our tool. We cloned all 21 repos and switched to the commit that had a flakiness issue as reported 
 in~\cite{nondex}.
-Of these projects, \<handlebars> and \<oryx> have compile time failures which prevents us from running \TheDeterminismChecker on them.
+Of these projects, \<handlebars> and \<oryx> have compile time failures which prevent us from running \TheDeterminismChecker on them.
 For benchmarks \<visualee>, \<easy-batch>, and \<vraptor>, we couldn't find
@@
 
-All the flakiness that NonDex found was due to 7 methods in 3 classes:
-\<Class.getDeclaredFields>, \<Class.getDeclaredMethods>, \<Class.getFields>, \<DateFormatSymbols.getZoneStrings>,
-\<HashMap.entrySet>, \<HashMap.keySet>, and \<HashMap.values>.  
-We provide samples of the annotated source code form these benchmarks and the errors as reported by \TheDeterminismChecker 
-in Figures~\ref{code-reflectasm} and~\ref{code-actiongenerator}.
+All the flakiness that NonDex found was due to 7 methods in 3 classes. We enumerate all the methods corresponding to
+each class in Figure~\ref{fig:flaky-sources}.
 \begin{figure}
+    \begin{tabular}{lll}
+        \<Class> & : \<getDeclaredFields>, \<getDeclaredMethods>, \<getFields> \\
+        \<DateFormatSymbols> & : \<getZoneStrings> \\
+        \<HashMap> & : \<entrySet>, \<keySet>, \<values>
+    \end{tabular}
+\caption{Souces of flakiness in~\cite{nondex}.}
+\label{fig:flaky-sources}
+\end{figure}
+
+We provide samples of the annotated source code from these benchmarks and the errors as reported by \TheDeterminismChecker 
+in Figure~\ref{fig:nondex-source}.
+\begin{figure}
+    \centering
+    \begin{subfigure}[b]{0.95\textwidth}
         \begin{verbatim}
@@
     }
-    [ERROR] FieldAccess.java:[102,87] [assignment.type.incompatible] 
+    [ERROR] FieldAccess.java:[102,87] 
     incompatible types in assignment.
@@
         \end{verbatim}
-    \caption{Annotated source code from reflectasm.}
+        \caption{Annotated source code from reflectasm. \vspace{0.5cm}}
         \label{code-reflectasm}
-\end{figure}
-\begin{figure}
+    \end{subfigure}
+
+    \begin{subfigure}[b]{0.95\textwidth}
         \begin{verbatim}
@@
     }
-    [ERROR] DataDictionaryEventSource.java:[73,78] [enhancedfor.type.incompatible] 
+    [ERROR] DataDictionaryEventSource.java:[73,78] 
     incompatible types in enhanced for loop.
@@
         \label{code-actiongenerator}
+    \end{subfigure}    
+    \caption{Errors detected by \TheDeterminismChecker.}
+    \label{fig:nondex-source}
 \end{figure}
@@
 which makes it easier for the programmer to fix issues and is orthogonal to the approach taken by NonDex which
-report failing tests.
+reports failing tests.
 
 \TheDeterminismChecker is also capable of detecting nondetermism in test cases. We provide a stub file
-for \<junit> in which we annotated all the \<assert> methods as accepting only \<Det> parameters.
-This will report any \<junit> test that passes nondeterministic arguments to \<assert>s and the test doesn't
-even have to be executed in our case.
-\todo{It took one of our authors about an hour to create this stub file - proof of easy extensibility of tool?} 
+for \<junit> in which we annotated the parameters of all the \<assert> methods as \<Det>.
+This will cause \TheDeterminismChecker to report any \<junit> test that passes nondeterministic arguments to 
+\<assert>s. The test case doesn't even have to be executed for out our tool to detect errors.
+It took one of our authors under an hour to create the stub file for junit. This gives us
+empirical evidence of the ease of extensibility of our tool.
+
 \todo{Link to det checker repo?}
-
-\begin{table}[]
-    \begin{tabular}{|p{1.5cm}|p{1cm}|p{5cm}|p{5cm}|}
-        \hline
-        Benchmark & Recall? & Error message source & Error msg  \\ \hline
-        P1 & Yes & FieldAccess.java: [102,87] & assignment type incompatible \\
-        P2 & Yes & YamlWriter.java: [224,28] & cast unsafe  \\
-        P3 & Yes & DateTimeZone.java: [731,27] &  return type incompatible \\ 
-        P4 & * & compile dependency failure  &   \\ 
-        P5 & ** & Class.getDeclaredMethods not found &    \\ 
-        P6 & Yes & OptionGroup.java: [55,40]  & argument type incompatible \\ 
-        P6 & Yes & Option.java: [157,28]  &  assignment type incompatible\\
-        P7(test error & Yes & MultilineRecursiveToString..: [95,35] argument type incompatible & MultilineRecursiveToString...: [208,50]  return type incompatible \\ 
-        P7(2) & Yes & FieldUtils.java: [219,84] & assignment type incompatible \\
-        P7(2) & Yes & Reflection ToStringBuilder.java: [543,65]  & assignment type incompatible  \\
-        P7(2) & Yes & HashCodeBuilder.java: [193,69]  & assignment type incompatible   \\
-        P7(2) & Yes & CompareToBuilder.java: [322,65] & assignment type incompatible   \\
-        P7(2) & Yes & EqualsBuilder.java: [419,69]  & assignment type incompatible  \\
-        P7 & Yes & FieldUtils.java: [200,69] & assignment type incompatible \\
-        P7 & Yes & FieldUtils.java: [219,84] & assignment type incompatible  \\
-        P7 & Yes & FieldUtils.java: [240,86] & assignment type incompatible  \\
-        P8 & ** & getDeclaredFields not found &   \\
-        P9 & Yes & Printers.java: [269,97]  &  enhancedfor type incompatible \\
-        P10 & ** & getDeclaredFields not found  &  \\
-        P11 & Yes & MapUtils.java: [20,50]   & enhancedfor type incompatible  \\
-        P12 & Yes  & GSLayerEncoder21.java: [113,91]  & enhancedfor type incompatible \\
-        P13 & ** & unbuildable handlebars &  \\
-        P14 & Yes  & Maps.java: [91,69] & invalid element type   \\
-        P15 & Yes & PkiMessageDecoder.java: [144,29]  &  enhancedfor type incompatible  \\
-        P16 & Yes & MethodSorter.java: [56,61]  & assignment type incompatible  \\
-        P17 & Yes & JSONObject.java: [674,51]  & assignment type incompatible  \\
-        P18 & Yes & SubstitutableLoggerTest.java: [67,61]   & enhancedfor type incompatible  \\
-        P19 & Yes & DataDictionaryEventSource.java: [73,78]  &  enhancedfor type incompatible  \\
-        P20 & Yes & SaverImpl.java: [43,69]  &  return type incompatible  \\
-        P21 & Yes & RestDocumentation.java: [419,50] &  argument type incompatible   \\ \hline
-    \end{tabular}
-\end{table}
-
 \todo{If there is time, we could do a second case study, on a program that

### 2019-04-03 randoop-case-study.tex
edits to nondex expts

@@
 \begin{itemize}
-\item
+\item \textbf{Bug 1:}
   A use of \<HashSet> that the developers changed to \<LinkedHashSet>\commit{c975a9f7}.
@@
 
-\item
+\item \textbf{Bug 2:}
 Dependency on the CLASSPATH environment variable in preference to the
@@
 
-\item 
+\item  \textbf{Bug 3:}
 Diagnostic output that printed a hash code for brevity, but should have had deterministic output\commit{661a4970}.  By contrast to the other two bugs, this one is not serious.

### 2019-04-03 experiment_tables.tex
deflaker expts

@@
 \end{table}
+
+\begin{table}[]
+    \begin{tabular}{|l|p{5cm}|}
+        \hline
+        Benchmark, issue & found by det checker? \\ \hline
+        achilles, [2] issue \#309 &  Race condition\\ \hline
+        achilles, [3] issue \#310 & Same fix for \#309 and \#310 \\ \hline
+        checkstyle [4] issue \#4664&test (added to test cases in det checker) \\ \hline
+        jackrabbit-oak [6] \#6512&  Unresolved on issue tracker (Race condition issue) \\ \hline
+        jackrabbit-oak [7] issue \#6524&  Dead link, didn't find 6524 in issue tracker \\ \hline
+        nutz [10] issue \#1283 &  Removed tests (Printing (HTTP's) Response.getstatus()) has Chinese characters,
+        compilation with det checker stuck!\\ \hline
+        togglz [19] issue \#223 & build issue \\ \hline
+        togglz [20] issue \#240 &  build issue\\ \hline
+    \end{tabular}
+\end{table}

### 2019-04-03 experiments.tex
deflaker expts

@@
-\section{Comparison to previous work}
+\section{Comparison to NonDex}
 
@@
 
-\section{deflaker case study}
-\begin{table}[]
-    \begin{tabular}{|l|p{5cm}|}
-        \hline
-        Benchmark, issue & found by det checker? \\ \hline
-        achilles, [2] issue \#309 &  Race condition\\ \hline
-        achilles, [3] issue \#310 & Same fix for \#309 and \#310 \\ \hline
-        checkstyle [4] issue \#4664&test (added to test cases in det checker) \\ \hline
-        jackrabbit-oak [6] \#6512&  Unresolved on issue tracker (Race condition issue) \\ \hline
-        jackrabbit-oak [7] issue \#6524&  Dead link, didn't find 6524 in issue tracker \\ \hline
-        nutz [10] issue \#1283 &  Removed tests (Printing (HTTP's) Response.getstatus()) has Chinese characters,
-        compilation with det checker stuck!\\ \hline
-        togglz [19] issue \#223 & build issue \\ \hline
-        togglz [20] issue \#240 &  build issue\\ \hline
-    \end{tabular}
-\end{table}
+\section{Comparison to Deflaker}
+Deflaker~\cite{deflaker}, like NonDex, reports tests that could be flaky. Ideally, we would have 
+liked to run all three tools (NonDex, DeFlaker, and \TheDeterminismChecker) on an open source project
+that wasn't explored by any of the tool for a fair comparison. Unfortunately, there are issues with the
+DeFlaker plugin and we weren't able to run it on any project. We have created a Github issue for the same
+in its repository.
+
+\subsection{\TheDeterminismChecker on DeFlaker benchmarks}\label{sec:deflaker-benchmarks}
+DeFlaker found 87 previously unknown flaky tests in 93 projects that were being actively developed at the the time
+the paper was written. The authors reported 19 of these tests, out of which 7 were addressed by the maintainers of
+those projects. Since we can only verify if \TheDeterminismChecker finds the same bugs by looking at the source code
+locations, we ran our tool on the benchmarks where the reported bug was fixed. We again defer the reader to~\cite{deflaker}
+for a complete list of the open source projects and their details.
+
+Two of the seven flaky tests (in \<achilles> and \<jackrabbit-oak>) that DeFlaker reported were caused by a race condition. \TheDeterminismChecker
+is designed for sequential programs and hence cannot detect this category of bugs.  
+We couldn't find the link to one of the issues in the issue tracker of \<jackrabbit-oak>
+We ran into build issues
+for \<tigglz> and \<nutz> which had two and one flakiness issues respectively. 
+\TheDeterminismChecker was able to find the source of flakiness in \<checkstyle>. This bug was in a test
+case that deterministically accessed an \<OrderNonDet> array returned by \<Class.getDeclaredConstructors()>.

### 2019-04-03 design.tex
Edits in section 2

@@
 
+This section first describes basic polymorphism over type qualifiers and
+over basetypes (\cref{sec:basic-polymorphism}).  Then, it describes two
+extensions.
+One extension resolves \<OrderNonDet> to \<NonDet> or \<Det> where
+needed (\cref{polymorphism-up-down}).
+The other extension permits use of polymorphic (type) variables without
+affecting binding of those variables --- that is, it affects which
+instantiation of an abstraction is chosen at a use site 
+(\cref{bindings-uses}).
+
 \todo{Can we come up with more examples of interesting functions to
   describe here?  Or put them in section 3.}
+
+\subsubsection{Qualifier and basetype polymorphism}\label{sec:basic-polymorphism}
+
+\OurTypeSystem supports parametric
+polymorphism~\cite{Abadi:1989:FIM:77350.77373,Plotkin:1993:LPP:645891.671433}
+over both classes and methods.
+A polymorphic abstraction (a class or method) is written and
+typechecked once.
+It acts as if it has multiple different types, and each use site is
+typechecked using the most specific applicable (instantiated,
+non-polymorphic) type.
 
@@
 $\forallt{\kappa \ \beta} \kappa \ \beta \rightarrow \kappa \ \beta$.
-(Compare this to 
-$\forallt{\kappa} \forallt{\beta} \kappa \ \beta \rightarrow \kappa \ \beta$
-which differs because the qualifier and basetype can vary independently;
-see below.)\todo{Does the foregoing make sense?}
 
@@
 
-  Basetype polymorphism is rarely used.  Full type polymorphism is more
+  Basetype polymorphism is rarely used on its own.  Full type polymorphism is more
   common, even if the type is decomposed and the parts used independently,
@@
 
-  $\forallt{\kappa\ \beta} \<Set>\angles{\kappa\ \beta} \rightarrow \<NonDet> \beta$
+  $\forallt{\kappa\ \beta} \<Set>\angles{\kappa\ \beta} \rightarrow \<NonDet>\ \beta$
 
@@
 
-$\<length>\ :\ \forallt{\kappa} . \kappa\ \<String> \rightarrow \kappa \<Int>$
+$\<length>\ :\ \forallt{\kappa} \kappa\ \<String> \rightarrow \kappa\ \<Int>$
 
@@
 
-$\<plus>\ :\ \forallt{\kappa} . \kappa\ \<Int> \times \kappa\ \<Int> \rightarrow \kappa \<Int>$
+$\<plus>\ :\ \forallt{\kappa} \kappa\ \<Int> \times \kappa\ \<Int> \rightarrow \kappa\ \<Int>$
 
@@
 
-\OurTypeSystem uses the standard meaning of parametric
-polymorphism~\cite{Abadi:1989:FIM:77350.77373,Plotkin:1993:LPP:645891.671433}.
-A polymorphic abstraction (such as a class or function) is written and
-typechecked once, but can be instantiated differently at different uses,
-each of which behaves as if it was written with a different non-polymorphic
-type.
-
-Informally, the \<length> above acts as if it had multiple
+Informally, the \<length> method above acts as if it had multiple
 (overloaded) definitions
@@
 
-The body must typecheck at every possible instantiation, and at a use site,
+The body of \<length> must typecheck at every possible instantiation of its
+polymorphic function type.  At a use site of \<length>,
 the most specific applicable instantiation is chosen.
 Here is an example:
-(put it in a table)
+\todo{(put it in a table)}
 
@@
 of \cref{sec:ond-behavior}.
-Consider this typing for the \<size> function:
+Consider this potential typing for the \<size> method in class \CollectionE:
 
-$\<size>\ :\ \forallt{\kappa} \kappa\ \<Collection> \rightarrow \kappa\ \<Int>$
+$\<size>\ :\ \forallt{\kappa} \kappa\ \CollectionE \rightarrow \kappa\ \<Int>$
 
-It cannot be instantiated as
+\noindent
+It cannot be instantiated at $\kappa = \<OrderNonDet>$ as
 
-$\<size>\ :\ \<OrderNonDet>\ \<Collection> \rightarrow \<OrderNonDet> \<Int>$
+$\<size>\ :\ \<OrderNonDet>\ \CollectionE \rightarrow \<OrderNonDet>\ \<Int>$
 
-because such an instantiation would include the invalid type \<OrderNonDet Int>.
+\noindent
+because such an instantiation would include the invalid return type \<OrderNonDet Int>.
+This means that the only two instantiations are at 
+$\kappa = \<NonDet>$ and $\kappa = \<Det>$.  An invocation on a
+\<OrderNonDet> collection would choose the \<NonDet> instantiation and the
+size would be \<NonDet Int>.
+%
 A similar problem occurs for the \<first> operation that returns the first
@@
 
-\OurTypeSystem resolves this problem by introducing two variants to the
-\<PolyDet> qualifier, \PolyDetUp\ and \PolyDetDown.
+\OurTypeSystem resolves this problem by introducing two operators over
+polymorhic (type) variables, $\uparrow$ and $\downarrow$.
 The $\uparrow$ operator converts \<OrderNonDet> to \<NonDet> and leaves the
-other qualifiers unchanged.  The upward-pointing arrow is mnemonic of how
-\<OrderNonDet> is changed:  it replaced by something higher in the type hierarchy.
+other qualifiers unchanged.  The upward-pointing arrow is mnemonic for replacing
+\<OrderNonDet> by something higher in the type hierarchy.
 The $\downarrow$ operator is analogous, but converts \<OrderNonDet> to
@@
 
-$\<size> : \forallt{\kappa} \kappa\ \<Collection> \rightarrow \down{\kappa} \<Int>$
+$\<size> : \forallt{\kappa} \kappa\ \CollectionE \rightarrow \down{\kappa}\ \<Int>$
 
+\noindent
 This can be instantiated at all three type qualifiers without creating any
-invalid types, and the instantiations implement the semantics of \cref{sec:ond-behavior}:
+invalid types:
 
-$\<size> : \<NonDet>\ \<Collection> \rightarrow \<NonDet> \<Int>$
+$\<size> : \<NonDet>\ \CollectionE \rightarrow \<NonDet>\ \<Int>$
 
-$\<size> : \<OrderNonDet>\ \<Collection> \rightarrow \<Det> \<Int>$
+$\<size> : \<OrderNonDet>\ \CollectionE \rightarrow \<Det>\ \<Int>$
 
-$\<size> : \<Det>\ \<Collection> \rightarrow \<Det> \<Int>$
+$\<size> : \<Det>\ \CollectionE \rightarrow \<Det>\ \<Int>$
 
-An example use of $\uparrow$ is the \<first> operation.  A simplified
-version of its type is
+\noindent
+The above instantiations implement the semantics of \cref{sec:ond-behavior}.
 
-$\<first> : \forallt{\kappa\ \beta} \kappa\ \<List>\angles{\kappa\ \beta} \rightarrow \up{\kappa}\ \beta$
+An example use of $\uparrow$ is in the \<first> method in class \ListKB\ that
+returns the first element of a list.  Its type is
 
-whose type qualifier can be instantiated as
+$\<first> : \forallt{\kappa} \kappa\ \ListKB \rightarrow \up{\kappa}\ \beta_e$
 
-$\<first> : \forallt{\beta} \<NonDet>\ \<List>\angles{\<NonDet>\ \beta} \rightarrow \<NonDet>\ \beta$
+\noindent
+which can be instantiated as
 
-$\<first> : \forallt{\beta} \<OrderNonDet>\ \<List>\angles{\<OrderNonDet>\ \beta} \rightarrow \<NonDet>\ \beta$
+$\<first> : \<NonDet>\ \ListKB \rightarrow \<NonDet>\ \beta_e$
 
-$\<first> : \forallt{\beta} \<Det>\ \<List>\angles{\<Det>\ \beta} \rightarrow \<Det>\ \beta$
+$\<first> : \<OrderNonDet>\ \ListKB \rightarrow \<NonDet>\ \beta_e$
 
-The actual type of \<first> accounts for more possible types for the type arguments.
+$\<first> : \<Det>\ \ListKB \rightarrow \<Det>\ \beta_e$
+
+\todo{As I mentioned in email, $\<NonDet>\ \beta_e$ looks wrong to me,
+  because $\beta_e$ may be a collection whose type arguments are not
+  compatible with \<NonDet>.  Maybe just remark that the instantiation does
+  not happen in that case.  If we try to adjust the type arguments to fit
+  \<NonDet>, then unsoundness occurs.}
 
@@
     }$
-\caption{Polymorphic resolution ($\uparrow$ is overloaded for types and qualifiers)}
+\caption{Polymorphic resolution of $\uparrow$ and $\downarrow$ operators ($\uparrow$ is overloaded for types and qualifiers)}
+\todo{I think this needs to do something to nested collections.  Or maybe
+  just clarify that if there is a nested collection such that this would
+  yield an invalid type, then the instantiation does not apply.  That might
+  be cleanest.}
 \label{fig:poly-resolutions}
@@
 
-$\<get> : \forallt{\kappa} \kappa\ \<List>\angles{\kappa\ \<String>} \times \kappa\ \<Int> \rightarrow \up{\kappa}\ \<String>$
+$\<get> : \forallt{\kappa} \kappa\ \<List>\angles{\kappa\ \<String>} \times \kappa\ \<Int> \rightarrow \kappa\ \<String>$
+\todo{Used to be $\up{\kappa}$ for return type, but I don't think that is needed.}
+\todo{Oh, maybe it's because we need to be able to instantiate in the tye
+  parameter differently?  How do we enable that?}
 
@@
 
-$\<get> : \forallt{\kappa\ \beta} \<List>\angles{\kappa\ \beta} \times \kappa\ \<Int> \rightarrow \up{\kappa}\ \beta$
+$\<get> : \<NonDet List>\angles{\<NonDet String>} \times \<NonDet Int> \rightarrow \<NonDet String>$
 
-$\<get> : \forallt{\kappa\ \beta} \<List>\angles{\kappa\ \beta} \times \kappa\ \<Int> \rightarrow \up{\kappa}\ \beta$
-
-$\<get> : \forallt{\kappa\ \beta} \<List>\angles{\kappa\ \beta} \times \kappa\ \<Int> \rightarrow \up{\kappa}\ \beta$
+$\<get> : \<Det List>\angles{\<Det String>} \times \<Det Int> \rightarrow \<Det String>$
 
@@
 
-If either the list or the index is non-deterministic, then the result is
+If either the list or the index is (order) non-deterministic, then the result is
 nondeterministic:
@@
 A similar typing does \emph{not} work for the \<set> function that sets an
-element of a list, because it would permit a deterministic collection must
-not be side-effected in non-deterministic ways.  Consider the following
-type for
+element of a list, because it would permit a deterministic collection to
+be side-effected in non-deterministic ways.  Consider the following
+possible type for
 
@@
 \subsection{Dataflow analysis and type refinement}\label{dataflow}
-\TheDeterminismChecker performs automatic type inference for local types within method bodies.
-It also performs dataflow analysis and refines the types of these locals, thereby reducing the qualifier burden 
-on the programmer. Consider the example below:
-\begin{verbatim}
-void foo(@Det int x) {
-    @NonDet int y = x;    // Type of y gets refined to @Det.
-    @Det int z = y;       //  No error here.
-}
-\end{verbatim}
-In the example above, even though the local variable \codeid{y} was declared to be \codeid{NonDet}, it gets
-type refined to \codeid{Det} after the first assignment statement. As the result, the assignment of \codeid{y}
-to a \codeid{Det} variable \codeid{z} is sound. In the following example,
-\begin{verbatim}
-void foo(@NonDet Object x, @Det Object y) {
-    x = y;                // No type refinement for x.
-    @Det Object z = x;    // Error.
-}
-\end{verbatim}
-\theDeterminismChecker reports a warning at the second assignment statement of \codeid{foo}.
-Type refinement only applies to locals.
-Refining non-locals can result in side effects making the analysis unsound.
+\TheDeterminismChecker performs local type inference within method bodies.
+It does not perform whole-program type inference.
+This makes separate compilation possible.
+It forces programmers to write specifications (type qualifiers) on methods,
+which is good style and valuable documentation.
 
-With the above rules, \theDeterminismChecker will automatically annotate a local collection with type \<NonDet> and its
-type arguments will get the type qualifier \<Det>. For example, a locally declared \codeid{List} of \codeid{Strings} will be automatically annotated
-as \codeid{NonDet List<Det String>}. Recall from~\cref{collection-rules} that \theDeterminismChecker does not allow this type on
-any collection. While this is still sound, it will result in \theDeterminismChecker reporting a lot of errors because every locally declared 
-collection gets annotated with this invalid type. To avoid this situation, we make an exception to the defaulting rules for local collections
-and annotate their type parameters also with \<NonDet>.
-So a local list of Strings gets the default qualifier of \codeid{NonDet List<NonDet String>}.
-While this defaulting eliminates the possibility of automatically annotating locals collections with an invalid
-type, it could result in a high number of false positives. In the following code snippet,
-\begin{verbatim}
-void test(@Det List<@Det String> argList) {
-    List<String> localList = argList;
-}    
-\end{verbatim}
-\theDeterminismChecker would report an error at the assignment statement since the inferred type for \codeid{localList} is 
-\codeid{NonDet List<NonDet> String} and since collections types are invariant.
-We therefore recommend that programmers explicitly annotate
-all local variables that are collections to reduce false positives.
+The local type inference is flow-sensitive.  That is, each variable may
+have a different type on every line of the program, depending on
+assignments to the variable.  The inferred type always respects any
+declared type:  that is, it is always a refinement (a subtype) of the
+declared type, if any.
+
+In addition to assignments, some method calls refine the types of their
+arguments.  One example is the \<sort> routine, as shown in 
+as shown in \cref{fig-sorting}.  One notable fact about the \rulename{sort}
+rule is that it requires the type argument to be deterministic (type
+qualifier \<Det>).  This is necessary because of two constraints.
+First, refinement must not change the type argument, because of invariant
+type argument subtyping.
+Second, the only possible type argument for a \<Det> collection is \<Det>.
+
+It is not permitted to
 
@@
 methods of \<List>, \<Collections>, and \<Arrays> classes
- as shown in \cref{fig-sorting}.
 \begin{figure}
@@
 
-    $\infer[\rulename{sort}]{\Gamma \vdash x : \|Det \ List|\angles{\|Det| \ \beta}}{\Gamma \vdash x : \|OrderNonDet \ List|\angles{\|Det| \ \beta} & \|Collections.sort(x)|}$
+  $\infer[\rulename{sort}]
+  {\Gamma \vdash x : \<Det List>\angles{\<Det>\ \beta}}
+  {\Gamma \vdash x : \<OrderNonDet List>\angles{\<Det>\ \beta} & \<sort(x)>}$
 
@@
 
-    \caption{Sorting type refinement rules}
+    \caption{Sorting type refinement rules.  \<sort> is a method that works
+    by side effect and changes the value---and the type---of its receiver.}
+\todo{These don't capture the notion of ``the type \emph{after} the sort call''.}
     \label{fig-sorting}
@@
 \TheDeterminismChecker type refines a receiver of type \codeid{OrderNonDet List} to a \codeid{Det List} if all the
-type argument of this \codeid{List} have the type qualifier \codeid{Det} (Rule LIST SORT in \cref{fig-sorting}).
+type argument of this \codeid{List} have the type qualifier \codeid{Det} (Rule \rulename{sort} in \cref{fig-sorting}).
 As per these rules, if \codeid{sort()} is called on a receiver of type \codeid{OrderNonDet List<Det Set<Det String>>}, it will
@@
 
-\subsection{Maps discussion}\label{maps}
-A \<Map> is a set of mappings from keys to values. A \<Map> is deterministic if its \<entrySet> is deterministic.
-In other words, iterating over this \<entrySet> is guaranteed to produce the same values in the exact same order across executions.
-The \<Map> interface in Java doesn't define an \<iterator()> interface. Therefore it's sound to equate the determinism of the \<Set>
-returned by \<entrySet()> to the determinism of the \<Map>. Interestingly, we found that every implementation of the \<Map> 
-interface had the following property: if the \<Set> returned by \<entrySet()> has a determinism type $\kappa$, then
-the \<Set>s returned by \<keySet()> and \<values()> also have the exact same determinism type $\kappa$.
+\subsection{Maps}\label{maps}
+A map is a collection of keys--value pairs.
+Like other collections, a map can be nondeterministic,
+order-nondeterministic, or deterministic.
+Analogously to set membership, lookups are deterministic in
+order-nondeterministic and deterministic maps; in addition, the iteration
+order is fixed in a deterministic map.
 
-We noticed the following properties with respect to determinism regarding the most widely used \<Map> implementations:
-\begin{itemize}
-    \item \<HashMap> can never guarantee fully deterministic iteration over its entries. A \codeid{Det HashMap} does not exist.
-    \item \<LinkedHashMap> can have any of \<Det>, \<OrderNonDet>, or \<NonDet> types. Iterating over a \<LinkedHashMap> returns
-    its entries in the order of their insertion. An \<OrderNonDet LinkedHashMap> can be created by calling its constructor with a \<HashMap>.
-    \item \<TreeMap> can either be \<Det> or \<NonDet>. An \<OrderNonDet TreeMap> doesn't exist because the entries always ordered according
-    to the natural ordering of the keys. 
-\end{itemize}
+We explain why each of these is possible by considering three common
+implementation strategies.
 
-We prohibit the creation of a \codeid{Det HashMap} or an \codeid{OrderNonDet TreeMap} in our determinism checker.
-\Cref{fig-creation-rules} formalizes these rules. Similar constraints apply to \<HashSet>s and \<TreeSet>s which are
-also expressed by the last two rules of \cref{fig-creation-rules}.
+A hash table is order-nondeterministic, because hash codes are
+nondeterministic in general.
 
-\begin{figure}
-    $\infer[\rulename{hashmap creation}]{\|new \ \kappa\ HashMap()|}{\kappa \in \{ \<OrderNonDet>, \<NonDet> \}}$
+A hash table can be augmented by a linked list recording insertion order,
+so that iteration order is deterministic.
 
-    \bigskip
-
-    $\infer[\rulename{treemap creation}]{\|new \ \kappa\ TreeMap()|}{\kappa \in \{ \<Det>, \<NonDet> \}}$
-
-    \bigskip
-
-    $\infer[\rulename{hashset creation}]{\|new \ \kappa\ HashSet()|}{\kappa \in \{ \<OrderNonDet>, \<NonDet> \}}$
-
-    \bigskip
-
-    $\infer[\rulename{treeset creation}]{\|new \ \kappa\ TreeSet()|}{\kappa \in \{ \<Det>, \<NonDet> \}}$
-    \caption{\<Map> and \<Set> creation rules}
-    \label{fig-creation-rules}
-\end{figure}
-
-In the JDK, \<LinkedHashSet> is implemented as a subclass of \<HashMap>.
-\todo{Talk about this design choice. Josh Bloch's book: "If the code surrounding the first declaration depended on LinkedHashSet's ordering policy,
-    then it would be incorrect to substitute HashSet for LinkedHashSet in the declaration, because HashSet makes no guarantee concerning iteration order."}
+A map can be backed by a sorted collection, such as a tree.  Such a map is
+either \<NonDet> (if \<NonDet> keys or values are inserted into it) or
+\<Det>, but never \<OrderNonDet>.
 
@@
 \end{verbatim}
-While this is sound, it is imprecise in the case of \codeid{OrderNonDet Sets}. We added the following rule shown in \cref{fig-precision-rules} to our checker to improve precision.
+While this is sound, it is imprecise in the case of \codeid{OrderNonDet
+  Set}s.\todo{Explain why.}
+ We added the following rule shown in \cref{fig-precision-rules} to our checker to improve precision.
 \begin{figure}
     \todo{add line break}
-    $\infer[\rulename{set precision}]{\Gamma \vdash \|z : Det|}{\Gamma \vdash \|x| : \|OrderNonDet| \ \|Set|\angles{ \kappa_{x} \ \beta_{x}}, \|y| : \|OrderNonDet| \ \|Set|\angles{ \kappa_{y} \ \beta_{y}} & \|z = x.equals(y)| & \angles{ \kappa_{x} \ \beta_{x}} \nsqsubseteq \angles{ \|OrderNonDet \ List| } & \angles{ \kappa_{y} \ \beta_{y}} \nsqsubseteq \angles{ \|OrderNonDet \ List| } }$
+    $\infer[\rulename{set precision}]{\Gamma \vdash \|x.equals(y) : Det|}{\Gamma \vdash \|x| : \|OrderNonDet| \ \|Set|\angles{ \kappa_{x} \ \beta_{x}}, \|y| : \|OrderNonDet| \ \|Set|\angles{ \kappa_{y} \ \beta_{y}} & \kappa_{x} \ \beta_{x} \nsqsubseteq \|OrderNonDet \ List| & \kappa_{y} \ \beta_{y} \nsqsubseteq \|OrderNonDet \ List| }$
 
@@
 
-    $\infer[\rulename{map precision}]{\Gamma \vdash \|z : Det|}{\Gamma \vdash \|x : OrderNonDet \ Map\angles{ Det \ \beta_k, \kappa_v\ \beta_v }, y : Det \ \beta| & \|z = x.get(y)|}$
+    $\infer[\rulename{map precision}]{\Gamma \vdash \|x.get(y) : Det|}{\Gamma \vdash \|x : OrderNonDet \ Map\angles{ Det \ \beta_k, \kappa_v\ \beta_v }, y : Det \ \beta|}$
 
     \caption{Improving precision}
+\todo{In the \rulename{map precision} rule, should the \<Det> in the
+  conclusion be $\beta_v$?}
     \label{fig-precision-rules}
@@
 is sound but imprecise. This rule is careful enough to not annotate the return type of \<equals()> as \<Det> if a type argument to either the
-receiver or the argument happens to be an \codeid{OrderNonDet List} (An order nondeterministic list does not provide any guarantee on the
+receiver or the argument happens to be an \codeid{OrderNonDet List}.
+\todo{The following is the key!  Put in earlier.}(An order nondeterministic list does not provide any guarantee on the
 equality as it is dependent on the iteration order).
@@
 Similarly, the \codeid{get()} method in the \<Map> class is annotated with polymorphic qualifiers and the rule \rulename{map precision}\ in \cref{fig-precision-rules} improves precision for \<OrderNonDet> receivers.
-
-\codeid{System.getProperty()} returns a \<NonDet> type since we can't provide any determinism guarantees for
-executions across machines. However, we make exceptions when the argument to this method is one of 
-\codeid{"line.separator", "path.separator", or "file.separator"} as they return OS dependent values which are guaranteed to be deterministic on that OS.
 

### 2019-04-03 java-design.tex
Edits in section 2

@@
 causing the return type to have the type qualifier \<NonDet>.
+
+\subsection{Dataflow analysis and type refinement}\label{sec:dataflow-java}
+
+\todo{Relate this to defaulting.}
+
+It also performs dataflow analysis and refines the types of these locals, thereby reducing the qualifier burden 
+on the programmer. Consider the example below:
+\begin{verbatim}
+void foo(@Det int x) {
+    @NonDet int y = x;    // Type of y gets refined to @Det.
+    @Det int z = y;       // No error here.
+}
+\end{verbatim}
+In the example above, even though the local variable \codeid{y} was declared to be \codeid{NonDet}, it gets
+type refined to \codeid{Det} after the first assignment statement. As the result, the assignment of \codeid{y}
+to a \codeid{Det} variable \codeid{z} is sound. In the following example,
+\begin{verbatim}
+void foo(@NonDet Object x, @Det Object y) {
+    x = y;                // No type refinement for x.
+    @Det Object z = x;    // Error.
+}
+\end{verbatim}
+\theDeterminismChecker reports a warning at the second assignment statement of \codeid{foo}.
+Type refinement only applies to locals.
+Refining non-locals can result in side effects making the analysis unsound.
+
+With the above rules, \theDeterminismChecker will automatically annotate a local collection with type \<NonDet> and its
+type arguments will get the type qualifier \<Det>. For example, a locally declared \codeid{List} of \codeid{Strings} will be automatically annotated
+as \codeid{NonDet List<Det String>}. Recall from~\cref{collection-rules} that \theDeterminismChecker does not allow this type on
+any collection. While this is still sound, it will result in \theDeterminismChecker reporting a lot of errors because every locally declared 
+collection gets annotated with this invalid type. To avoid this situation, we make an exception to the defaulting rules for local collections
+and annotate their type parameters also with \<NonDet>.
+So a local list of Strings gets the default qualifier of \codeid{NonDet List<NonDet String>}.
+While this defaulting eliminates the possibility of automatically annotating locals collections with an invalid
+type, it could result in a high number of false positives. In the following code snippet,
+\begin{verbatim}
+void test(@Det List<@Det String> argList) {
+    List<String> localList = argList;
+}    
+\end{verbatim}
+\theDeterminismChecker would report an error at the assignment statement since the inferred type for \codeid{localList} is 
+\codeid{NonDet List<NonDet> String} and since collections types are invariant.
+We therefore recommend that programmers explicitly annotate
+all local variables that are collections to reduce false positives.
+
+\subsection{Maps}\label{sec:maps-java}
+
+ A \<Map> is deterministic if its \<entrySet> is deterministic.
+In other words, iterating over this \<entrySet> is guaranteed to produce the same values in the exact same order across executions.
+The \<Map> interface in Java doesn't define an \<iterator()> interface. Therefore it's sound to equate the determinism of the \<Set>
+returned by \<entrySet()> to the determinism of the \<Map>. Interestingly, we found that every implementation of the \<Map> 
+interface had the following property: if the \<Set> returned by \<entrySet()> has a determinism type $\kappa$, then
+the \<Set>s returned by \<keySet()> and \<values()> also have the exact same determinism type $\kappa$.
+
+We noticed the following properties with respect to determinism regarding the most widely used \<Map> implementations:
+\begin{itemize}
+    \item \<HashMap> can never guarantee fully deterministic iteration over its entries. A \codeid{Det HashMap} does not exist.
+    \item \<LinkedHashMap> can have any of \<Det>, \<OrderNonDet>, or \<NonDet> types. Iterating over a \<LinkedHashMap> returns
+    its entries in the order of their insertion. An \<OrderNonDet LinkedHashMap> can be created by calling its constructor with a \<HashMap>.
+    \item \<TreeMap> can either be \<Det> or \<NonDet>. An \<OrderNonDet TreeMap> doesn't exist because the entries always ordered.
+\end{itemize}
+
+We prohibit the creation of a \codeid{Det HashMap} or an \codeid{OrderNonDet TreeMap} in our determinism checker.
+\Cref{fig-creation-rules} formalizes these rules. Similar constraints apply to \<HashSet>s and \<TreeSet>s which are
+also expressed by the last two rules of \cref{fig-creation-rules}.
+
+\begin{figure}
+    $\infer[\rulename{hashmap creation}]{\|new \ \kappa\ HashMap()|}{\kappa \in \{ \<OrderNonDet>, \<NonDet> \}}$
+
+    \bigskip
+
+    $\infer[\rulename{treemap creation}]{\|new \ \kappa\ TreeMap()|}{\kappa \in \{ \<Det>, \<NonDet> \}}$
+
+    \bigskip
+
+    $\infer[\rulename{hashset creation}]{\|new \ \kappa\ HashSet()|}{\kappa \in \{ \<OrderNonDet>, \<NonDet> \}}$
+
+    \bigskip
+
+    $\infer[\rulename{treeset creation}]{\|new \ \kappa\ TreeSet()|}{\kappa \in \{ \<Det>, \<NonDet> \}}$
+    \caption{\<Map> and \<Set> creation rules}
+    \todo{Are these type well-formedness rules?}
+\todo{What about \<new TreeSet(myOrderedCollection<Det ...>)> which yields a
+  deterministic set?}
+    \label{fig-creation-rules}
+\end{figure}
+
+In the JDK, \<LinkedHashSet> is implemented as a subclass of \<HashMap>.
+\todo{Talk about this design choice. Josh Bloch's book: "If the code surrounding the first declaration depended on LinkedHashSet's ordering policy,
+    then it would be incorrect to substitute HashSet for LinkedHashSet in the declaration, because HashSet makes no guarantee concerning iteration order."}
+
+\subsection{Precision improvements}\label{sec:precision-java}
+
+\codeid{System.getProperty()} returns a \<NonDet> type since we can't provide any determinism guarantees for
+executions across machines. However, we make exceptions when the argument to this method is one of 
+\codeid{"line.separator", "path.separator", or "file.separator"} as they return OS dependent values which are guaranteed to be deterministic on that OS.

### 2019-04-03 related.tex
Edits in section 2
related nondex

--- 2019-03-31 related.tex
@@
 
-The state of the art is NonDex.  It is a highly effective tool.
+The state of the art is NonDex~\cite{nondex}.  It is a highly effective tool.
+% https://github.com/TestingResearchIllinois/NonDex/wiki/Supported-APIs
+NonDex uses a hand-crafted list of 47 methods (25 unique method names)
+in 13 classes that were identified as potential sources of flakiness. 
+This is analogous to our annotations on the JDK\@. So far, we have annotated
+928 methods across 41 classes.
+For each of the identified methods, the authors built models that
+return different results when called consecutively. A modified JVM then
+runs a given test multiple times and reports the test as being flaky if it observes
+diverging test output. While this approach produces precise results, it requires manual inspection
+and considerable debugging effort to locate and fix the source of flakiness. \TheDeterminismChecker, in
+contrast, reports the cause of nondeterminism at compile requiring little to no debugging effort with the
+caveat of producing higher false positives compared to NonDex.
+%\todo{Check whether our JDK is missing anything they support - No}
+%\todo{Did the NonDex authors contact the maintainers of the tools in which
+%  they found nondeterminism - No!}
+%\todo{The NonDex authors didn't contact the maintainers of tools.
+%From the paper: "NonDex detected 57 previously unknown
+%flaky tests in open-source projects and three flaky tests
+%that have been already fixed by the open-source software
+%developers."}
 
-% https://github.com/TestingResearchIllinois/NonDex/wiki/Supported-APIs
-NonDex uses a hand-crafted list of 47 methods (25 unique method names) in
-in 13 classes.
-This is analogous to our annotations on the JDK\@.
-\todo{Say how much code our annotations cover?}
-\todo{Check whether our JDK is missing anything they support.}
-\todo{Did the NonDex authors contact the maintainers of the tools in which
-  they found nondeterminism?}
+\todo{A NonDet annotation:
+  \url{http://www.swi-prolog.org/pldoc/man?section=modes}}
+
+%% Added this info in the experiments section
+%\todo{All nondeterminism that NonDex found was due to 7 methods in 3 classes:
+%Class\#getDeclaredFields
+%Class\#getDeclaredMethods
+%Class\#getFields
+%DateFormatSymbols\#getZoneStrings
+%HashMap\#entrySet
+%HashMap\#keySet
+%HashMap\#values
+%}

### 2019-04-03 design.tex
Added closing bracket

@@
 \end{figure}
-According to this rule, if the \<Set> receiver has type \codeid{@OrderNonDet Set<@Det List} and the argument to the \codeid{equals()} method is
+According to this rule, if the \<Set> receiver has type \codeid{@OrderNonDet Set<@Det List>} and the argument to the \codeid{equals()} method is
 also of the same type, the return type gets the qualifier \<Det>. Without this precision rule, the return type would have been instantiated as \<NonDet> which

### 2019-04-03 randoop-case-study.tex
Explain the large number of `@Det` annotations

@@
 been able to reverse-engineer.
+
+The large number of \<@Det> annotations has two main reasons.
+
+One is that we annotated some core components of Randoop to require all
+boolean expressions used in conditionals to be deterministic.  (This is an
+option in our implementation, which increases soundness.)  This forced
+their clients, and much of the Randoop codebase, to be deterministic.
+These annotations are not incorrect, but they are stricter than necessary
+because some types could otherwise be @PolyDet, which is the default for
+formal parameter and return types.  We plan to re-do the annotation effort
+from scratch, and the number of annotations will be much lower.
+
+The second reason is a limitation in \theDeterminismCheckerImplementation.
+\TheDeterminismChecker does not support qualifier polymorphism over
+classes.  This means that fields must be deterministic.  Our annotations in
+Randoop use qualifier polymorphism extensively on methods but cannot use it
+on classes, and this leads to many more Det annotations than would be
+needed otherwise.

### 2019-04-03 related.tex
flake related

@@
 
+DeFlaker~\cite{deflaker} is another approach for flake test detection. It relies on
+version control history to detect tests that behave differently. It computes a diff of the code covered
+in the current version and the previous one. If there exists a tests case whose code coverage does not include
+this diff but still produces different results on the two versions being compared, it is flagged as being flaky.
+This approach doesn't require JVM modifications
+and integrates easily with a production software. DeFlaker reported 19 previously unknown bugs
+in open source projects, 7 of which were addressed by the developers of these projects.
+
+~\cite{LuoHEM2014} study 51 open source projects and analyze the most common root causes of
+test flakiness. The findings in the paper suggest that most of the flakiness in tests is caused by
+the use of async away, concurrency, or due to test order dependency.~\cite{Plotkin:1993:LPP:645891.671433}
+performs an empirical study on open source benchmarks that concur with the results of~\cite{LuoHEM2014}.
+Additionally, the authors of this paper study the effects of removing these test smells. Their results suggest that
+refactoring the test code smells fixed more than half of the identified flaky tests.
 \todo{A NonDet annotation:

### 2019-04-03 java-design.tex
No qualifier polymorphism on classes
small changes

@@
 \item
+\todo{We do not support qualifier polymorphism on classes, only on methods.}
 Java does not provide a syntax that can be used for qualifier polymorphism,
@@
     its entries in the order of their insertion. An \<OrderNonDet LinkedHashMap> can be created by calling its constructor with a \<HashMap>.
-    \item \<TreeMap> can either be \<Det> or \<NonDet>. An \<OrderNonDet TreeMap> doesn't exist because the entries always ordered.
+    \item \<TreeMap> can either be \<Det> or \<NonDet>. An \<OrderNonDet TreeMap> doesn't exist because the entries are always ordered.
 \end{itemize}

### 2019-04-03 randoop-case-study.tex
Add another bug found, and a description of the process

@@
 \item \textbf{Bug 3:}
-Diagnostic output that printed a hash code for brevity, but should have had deterministic output\commit{661a4970}.  By contrast to the other two bugs, this one is not serious.
+Diagnostic output that printed a hash code for brevity, but should have had
+deterministic output\commit{661a4970}.  By contrast to the first two bugs,
+this one is not serious.
+
+\item \textbf{Bug 4:}
+Diagnostic output that printed a timestamp, but should have obeyed an
+option about whether to do so\commit{a460df97}.  By contrast to the first
+two bugs, this one is not serious.
+
 \end{itemize}
@@
 We reported another suspicious case of order-nondeterminism, in the
-\<findOverridden> method.  The Randoop developers explained it was
+\<SpecificationCollection>\-\<.findOverridden> method.  The Randoop developers explained it was
 acceptable, after tracing the flow through the program.  (One said, ``This
@@
 been able to reverse-engineer.
+
+We annotated one package at a time, starting with the most central packges
+that are most depended upon.  Within a package, we followed a similar
+strategy, annotating supertypes first.  If the classes and methods had been
+documented with respect to their determinism, then our annotation effort
+would have been easy, just converting English into type qualifiers.
+Unfortunately, this was not the case: we reverse-engineered each
+specification, largely from the methods it calls.  Sometimes after we
+annotated a subclass, we had to go back and change existing annotation on
+the supertype or elsewhere.
+After resolving most type-checking warnings by adding or changing
+annotations, we manually examined each remaining one to determine whether
+it was a bug or a false positive warning.
 

### 2019-04-04 design.tex
Threats to validity, and changes suggested by Jason

@@
 
-Both collections and map may be \<Det>, \<OrderNonDet>, or \<NonDet>, and
+Both collections and maps may be \<Det>, \<OrderNonDet>, or \<NonDet>, and
 the element types can be specified independently of the collection type,

### 2019-04-04 introduction.tex
Improve a sentence based on Jason's suggestion
Threats to validity, and changes suggested by Jason

--- 2019-03-24 introduction.tex
@@
 programs that do not flip coins.
-For example, any program that iterates over a set or map, and that uses the
-results to affect the output, may produce different output on different runs.
+For example, a program that iterates over a hash table
+may produce different output on different runs.
 So may any program that uses default formatting, such as Java's
@@
 %
-Our analysis is sound:  if it issues no warnings, then the program does not
-suffer nondeterminism of the given types.  (If the program uses concurrency,
+Our analysis is largely sound:  if it issues no warnings, then the program does not
+suffer nondeterminism of the given types, up to the limitations of the
+analysis (see \cref{sec:threats}).  If the program uses concurrency,
 it might still be nondeterministic.)
@@
 
-% LocalWords:  NONDEX NonDex DeFlaker Det OrderNonDet NonDet
+\todo{It is essential that the introduction includes an example real-world
+  defect that \theDeterminismChecker found.}
+
+% LocalWords:  NonDex DeFlaker Det OrderNonDet NonDet

### 2019-04-04 java-design.tex
Threats to validity, and changes suggested by Jason

@@
 
-\codeid{System.getProperty()} returns a \<NonDet> type since we can't provide any determinism guarantees for
-executions across machines. However, we make exceptions when the argument to this method is one of 
-\codeid{"line.separator", "path.separator", or "file.separator"} as they return OS dependent values which are guaranteed to be deterministic on that OS.
+The return type of \<System.getenv}, which reads an operating system
+environment variable, is always \<NonDet>.
+The return type of \codeid{System.getProperty} is usually \<NonDet>. However,
+the result is \<Det> if the argument is \codeid{"line.separator"}; although
+it varies by operating system, many programs including \<diff> hide those
+differences from the user.
+The result is also deterministic for \codeid{"path.separator"} and
+\codeid{"file.separator"}, because these lead to the same behavior given
+corresponding environments.
+The user can specify input properties that must be passed on the \<java>
+command line and thus act like inputs to the program; these are also
+treated as deterministic.
+The inputs to a program are also treated as \<Det>.

### 2019-04-04 threats.tex
Threats to validity, and changes suggested by Jason

\section{Threats to validity}\label{sec:threats}

Our type system and implementation suffer some unsoundness.  The most
important is that they do not capture all sources of nondeterminism.  One
important such source is concurrency.
%
In addition, our annotations of the JDK (the Java standard library) are incomplete.
Some of the unannotated parts might introduce nondeterminism without
the verification tool being aware of it.

The type system presented in this paper does not account for the fact that
a nondeterministic conditional expression (in an \<if> or \<switch>
statement or a \<for> or \<while> loop, for example) may behave differently
from execution to execution.
These ``implicit flows'' are a well-known challenge for taint analysis.
Suppose that an expression used in a conditional is tainted.
A standard approach\todo{Add citations} for a dynamic taint analysis is to taint any value that may
be assigned in either branch of the conditional statement.  This tends to rapidly lead to most of
the program state being tainted.
In a static analysis, every possibly-assigned lvalue would be considered
possibly tainted; again, this tends to snowball through the program.

Our implementation, \theDeterminismChecker, can enforce that conditional
expressions are deterministic, but we found this to be too restrictive:  it
causes \theDeterminismChecker to issue large numbers of false positive
warnings, or it forces types to be \<Det> unnecessarily.
We disabled this check in most of our case study.

It our case study, we also disabled a check that required all caught
exceptions to have \<NonDet> type.  This type is a safe approximation to an
exception that can be thrown from arbitrary code, including libraries.
Enabling this check led to many unhelpful false positive warnings.

\TheDeterminismChecker only examines the code it is run on.  Unchecked
libraries and native code with incorrect specifications might introduce
nondeterminism that it does not check.

We have performed only one case study.  This case study found important
previously-unknown errors, but its results might not carry over to other
programs or to programs written in other programming lanugages.  We
mitigated this problem by showing that \theDeterminismChecker finds most or
all of the sources of nondeterminism identified by other tools.

### 2019-04-04 related.tex
small changes

@@
 
-~\cite{LuoHEM2014} study 51 open source projects and analyze the most common root causes of
+\cite{LuoHEM2014} study 51 open source projects and analyze the most common root causes of
 test flakiness. The findings in the paper suggest that most of the flakiness in tests is caused by
@@
 refactoring the test code smells fixed more than half of the identified flaky tests.
+
+\cite{Jin:2012:UDR:2254064.2254075} study the effect of inefficient code sequences on the performance of software.
+Their findings suggest that incorrect understanding of performance specifications of APIs is a very common root cause for
+performance bugs. ~\cite{Rui:2013:180377} analyze the APIs related to security such as authorization and authentication.
+These APIs which come as a part of the SDKs provided by online providers are used by application developers to 
+build secure apps. The paper discovers that the security guarantees provided by these SDKs are violated due to
+the implicit assumptions made by the APIs that the users are unaware of. \cite{Xiao:2014:NMC:2591062.2591177}
+study the effect of nondeterminism in MapReduce programs with a specific focus on nondeterminism caused by
+non-commutative reducers. While they found bugs that violated correctness due to this bug pattern,
+the authors reported that several of these were harmless as they relied on an implicit assumption on data
+which ensured correctness.
+
+\cite{Petrenko1996,Petrenko:1993:NSM:648128.761244,Savor:1997:639710,Hierons:2004:TCD:1040993.1040998} propose several techniques to test whether a
+deterministic implementation conforms to its non-deterministic finite state machine.
+\cite{Cook:2013:RNP:2491956.2491969} present an approach that can automatically verify properties
+in branching time temporal logic systems that are inherently non-deterministic.
+
 \todo{A NonDet annotation:

### 2019-04-04 introduction.tex
more citations

@@
   Nondeterminism may make outputs incompatible with previous outputs relied
-  on by users or by other systems.\todo{Citations about bugs related to this?}
+  on by users or by other systems~\cite{Herzig:2015:EDF:2819009.2819018,Huo:2014:IOQ:2635868.2635917,Zhang:2014:ERT:2610384.2610404,BellKMD2015,Fowler,Gyori:2015:RTD:2771783.2771793,Dan:2013:10.1007/978-3-642-39038-8_25,Vahabzadeh:2015:7332456,Sudarshan}.
+%  \todo{Citations about bugs related to this?}
 \item

### 2019-04-04 paper.tex
moving text around
Edits in section 2.1
expts
Edits in section 2.2
references
experiments: NONDEX on Randoop
Randoop case study
Edits in sections 2-2.3
Threats to validity, and changes suggested by Jason
Draft of abstract

@@
 %% Some recommended packages.
+\usepackage{proof}
+\usepackage{amsmath,amssymb,amsthm,textcomp}
+
 \usepackage{booktabs}   %% For formal tables:
                         %% http://ctan.org/pkg/booktabs
+\usepackage{wrapfig}
 \usepackage{subcaption} %% For complex figures with subfigures/subcaptions
@@
 \usepackage{url}
+\usepackage{xspace}
+
+\usepackage{fancyvrb}
+\fvset{fontsize=\smaller}
 
 \usepackage{color}
-%% Comment out one of these two definitions.
-% \newcommand{\todo}[1]{\relax}
-\newcommand{\todo}[1]{{\color{red}\bfseries [[#1]]}}
 
-% \|name| or \mathid{name} denotes identifiers and slots in formulas
-\def\|#1|{\mathid{#1}}
-\newcommand{\mathid}[1]{\ensuremath{\mathit{#1}}}
-% \<name> or \codeid{name} denotes computer code identifiers
-\def\<#1>{\codeid{#1}}
-\protected\def\codeid#1{\ifmmode{\mbox{\sf{#1}}}\else{\sf #1}\fi}
-\protected\def\codeid#1{\ifmmode{\mbox{\ttfamily{#1}}}\else{\ttfamily #1}\fi}
-\protected\def\codeid#1{\ifmmode{\mbox{\smaller\ttfamily{#1}}}\else{\smaller\ttfamily #1}\fi}
+\usepackage{cleveref}
+\usepackage{hyperref}
+
+\input{macros.tex}
 
@@
 \begin{abstract}
-\todo{Text of abstract \ldots.}
+\todo{The title ``Making Sequential Programs Deterministic'' is misleading
+  because our tool doesn't transform programs, though it does help
+  programmers transform programs.}
+
+When a program is nondeterministic, it is difficult to test and debug.
+Nondeterminism can occur even in sequential programs, for example as a
+result of iterating over the elements of a hash table.
+
+We have created a type system for determinism.  A type checker warns if a
+program is nondeterministic or verifies that the program is deterministic.
+The key ideas in the type system are type qualifiers for nondeterminism,
+order-nondeterminism, and determinism; type well-formedness rules to
+restrict the typings for collections; and enhancements to polymorphism that
+improve precision when analyzing collection operations.
+
+We implemented our type system for Java.  In a case study, it found
+previously-unknown nondeterminism errors in a program that had been heavily
+vetted by its developers.  In an experiment, it found all of the
+non-concurrency-related nondeterminism that was found by other tools.
+
 \end{abstract}
@@
 \input{design.tex}
+\input{java-design.tex}
+\input{randoop-case-study}
+\input{experiments.tex}
+\input{threats}
+\input{related}
+\input{conclusion}
 
@@
 %\bibliographystyle{ACM-Reference-Format.bst}
+\clearpage
 \bibliography{paper,bibstring-unabbrev,testing,crossrefs}

### 2019-04-04 design.tex
Edits in section 1

@@
   \begin{tabular}{lcll}
-    $C$ & ::= & \<Int>\bnfalt \<List>\bnfalt \ldots & class name \\
+    $C$ & ::= & \<Int>\bnfalt \<String>\bnfalt \<Collection>\bnfalt \<List>\bnfalt \<Set>\bnfalt \<Map> & class name \\
     $\kappa$ & ::= & \<NonDet>\bnfalt \<OrderNonDet>\bnfalt\<Det> & type qualifier \\

### 2019-04-04 introduction.tex
Edits in section 1

--- 2019-04-04 introduction.tex
@@
   oracles must account for all possible behaviors while still enforcing
-  correct behaviors.  Test oracles that are too strict lead to flaky tests
+  correct behaviors.  Test oracles that are too strict lead to flaky
+% ,Gyori:2015:RTD:2771783.2771793
+  tests~\cite{LuoHEM2014,nondex,deflaker,Rahman:2018:IFF:3236024.3275529}
   that sometimes pass and sometimes fail.  Flaky tests must be re-run, or
@@
   Nondeterminism may make outputs incompatible with previous outputs relied
-  on by users or by other systems.\todo{Citations about bugs related to this?}
+  on by users or by other systems~\todo{These do not seem to be supporting
+    the claim about bugs related to incompatibilities between outputs.
+    These seem to be general citations about test flakiness or dependence.
+    Readers will be unhappy if we make a claim but then do not support it.}\cite{Herzig:2015:EDF:2819009.2819018,Huo:2014:IOQ:2635868.2635917,ZhangJWMLEN2014,BellKMD2015,Fowler,Gyori:2015:RTD:2771783.2771793,Dan:2013:10.1007/978-3-642-39038-8_25,Vahabzadeh:2015:7332456,Sudarshan}.
+%  \todo{Citations about bugs related to this?}
 \item
@@
 So may any program that uses default formatting, such as Java's
-\<Object.toString()> which includes an address or hash code.
+\<Object.toString()> which includes a memory address.
 Other nondeterministic APIs include date-and-time functions and
@@
 absence.
-Our analysis permits programmers to specify which parts of their program
+Our analysis permits a programmer to specify which parts of their program
 are intentionally nondeterministic, and it verifies that the remainder is deterministic.
 %
-Our analysis is largely sound:  if it issues no warnings, then the program does not
-suffer nondeterminism of the given types, up to the limitations of the
-analysis (see \cref{sec:threats}).  If the program uses concurrency,
-it might still be nondeterministic.)
+If our analysis issues no warnings, then the program produces the same
+output when executed twice on the same inputs.  This guarantee is modulo
+the limitations of the
+analysis, notably concurrency, implicit control flow, and unanalyzed libraries (see \cref{sec:threats}).
 %
@@
 specifying deterministic properties in a program and verifying them
-at compile time.
-
-Other researchers have also recognized the importance of this problem.
+statically.
+%
+Other researchers have also recognized the importance of the problem of nondeterminism.
 Previous work in program analysis for nondeterminism has focused on unsound dynamic
 approaches that identify flaky test cases.
-The authors of~\cite{nondex} manually identify methods in the JDK that have nondeterministic specifications and build models for
-these methods. They  modified that JVM so that it runs the test suite multiple times.
+NonDex uses a modified JVM that returns different results on different
+executions, for a few key JDK methods with loose specifications.  Running a
+test suite multiple times can reveal unwarranted dependence on those
+methods.
 %NonDex~\cite{nondex} is a tool that manually\todo{This is confusing:  the
@@
 %identifies 
-Diverging test results in this setting indicate a flaky test. DeFlaker~\cite{deflaker} looks at a range of commit versions
+DeFlaker~\cite{deflaker} looks at a range of commit versions
 of a code, and marks a test as flaky if it doesn't execute any modified code but still fails in the newer version. These techniques
@@
 \begin{description}
+\item[\<NonDet>] represents values that might differ from run to run.
+\item[\<OrderNonDet>] represents collections that are guaranteed to contain
+  the same elements, albeit in possibly different iteration order.
 \item[\<Det>] represents values that will be the same across executions.
-\item[\<OrderNonDet>] represents collections that are guaranteed to contain the same elements albeit in possibly
-different iteration order.
-\item[\<NonDet>] represents values that
-might differ from run to run.
 \end{description}
@@
 Programmers can write these to specify their program's behavior.
-Our type system contains a variety of other features that increase
-expressiveness and reduce false positive alarms.\todo{reference a section}
+Our full analysis contains other features that increase
+expressiveness.
 
-We have implemented this analysis as a pluggable type system for Java.
+\begin{figure}
 
-We have run experiments\todo{describe them}.
-We found this type hierarchy
-is expressive enough to identify nondeterministic behaviors in production software.
+\noindent
+In \<TypeVariable.java>:
+
+\begin{Verbatim}
+160:   public List<TypeVariable> getTypeParameters() {
+161:-    Set<TypeVariable> parameters = new HashSet<>(super.getTypeParameters());
+161:+    Set<TypeVariable> parameters = new LinkedHashSet<>(super.getTypeParameters());
+162:     parameters.add(this);
+163:     return new ArrayList<>(parameters);
+164:   }
+\end{Verbatim}
+
+\caption{Fixes made by the Randoop developers in response to our bug report
+  about improper use of a HashSet.  Lines starting with ``\<->'' were
+ removed and those starting with ``\<+>'' were added.
+ Our tool, \theDeterminismChecker, confirmed that 
+25 other uses of \<new HashSet> were acceptable, as were 15 uses of \<new HashMap>.}
+\label{fig:randoop-bug-hashset}
+\end{figure}
 
@@
 \begin{enumerate}
-  \item We designed a type type system for expressing determinism properties.
+  \item We designed a type system, named \ourTypeSystem, for expressing determinism properties.
 
-  \item We implemented the type system for Java.
+  \item We implemented analysis, as a pluggable type system for Java, in a
+    tool called \theDeterminismChecker.
 
-  \item In a case study, we annotated several libraries, including the
-    relevant parts of the JDK (Java's standard library).  This provides a
-    formal, machine-readable specification for the libraries, and it
-    demonstrates the richness of our type system.
+  \item
+  \todo{It's a bit weird that we call this a contribution but it is never
+    explicitly discussed in the paper.  I think it's probably OK.}
+  In a case study, we annotated several libraries, including the collection
+    classes and other parts of the JDK (Java's standard library).  This
+    provides a formal, machine-readable specification for the libraries, and
+    it demonstrates the expressiveness of our type system.
 
   \item In another case study, we ran our analysis on a 24 KLOC project that
-    developers had spent weeks of effort to make deterministic.  Our system
-    discovered nondeterminism that the developers had overlooked, and they
-    accepted our patches.
+    developers had already spent weeks of testing and inspection effort to
+    make deterministic.  \TheDeterminismChecker
+    discovered 5 instances of nondeterminism that the developers had
+    overlooked.
+    The developers fixed all these issues when we reported them.
+    \Cref{fig:randoop-bug-hashset,fig:randoop-bug-getenv} show two examples.
 
-  \item We compared our tool against state-of-the-art flaky test
-    detectors.\todo{Say something about the results.}
+  \item We compared \theDeterminismChecker our tool against state-of-the-art flaky test
+    detectors, on their benchmarks.
+    \TheDeterminismChecker found all the non-concurrency nondeterminism
+    found by the other tools.
+
 \end{enumerate}

### 2019-04-04 paper.tex
Edits in section 1

@@
 
+% \usepackage[svgnames]{xcolor}
+%   \definecolor{diffstart}{named}{Grey}
+%   \definecolor{diffincl}{named}{Green}
+%   \definecolor{diffrem}{named}{OrangeRed}
+
+% \usepackage{listings}
+%   \lstdefinelanguage{diff}{
+%     basicstyle=\ttfamily\small,
+%     morecomment=[f][\color{Dark gray}]{@@},
+%     morecomment=[f][\color{Forest green}]{+\ },
+%     morecomment=[f][\color{Firebrick}]{-\ },
+%   }
+% % \lstset{numbers=left,numberblanklines=false,escapeinside=||}
+% \lstset{escapeinside=||}
+% \let\origthelstnumber\thelstnumber
+% \makeatletter
+% \newcommand*\Suppressnumber{%
+%   \lst@AddToHook{OnNewLine}{%
+%     \let\thelstnumber\relax%
+%      \advance\c@lstnumber-\@ne\relax%
+%     }%
+% }
+% 
+% % This causes line numbers to increment by 2 every time!
+% \newcommand*\Reactivatenumber[1]{%
+%   \setcounter{lstnumber}{\numexpr#1-1\relax}
+%   \lst@AddToHook{OnNewLine}{%
+%    \let\thelstnumber\origthelstnumber%
+%    \refstepcounter{lstnumber}%
+%   }%
+% }
+
 \input{macros.tex}
@@
 % \title{Compile Time Verification of Deterministic properties for Sequential Programs}
+% \todo{The title ``Making Sequential Programs Deterministic'' is misleading
+%   because our tool doesn't transform programs, though it does help
+%   programmers transform programs.}
 \title{Making Sequential Programs Deterministic}
@@
 \begin{abstract}
-\todo{The title ``Making Sequential Programs Deterministic'' is misleading
-  because our tool doesn't transform programs, though it does help
-  programmers transform programs.}
-
 When a program is nondeterministic, it is difficult to test and debug.
@@
 \clearpage
-\bibliography{paper,bibstring-unabbrev,testing,crossrefs}
+\bibliography{paper,bibstring-unabbrev,ernst,testing,crossrefs}
 

### 2019-04-04 randoop-case-study.tex
Edits in section 1

@@
 
+\begin{figure}
+
+\noindent
+In \<Minimize.java>:
+
+\begin{Verbatim}
+151:-  private static final String SYSTEM_CLASS_PATH = System.getProperty("java.class.path");
+
+913:-    String command = "javac -classpath " + SYSTEM_CLASS_PATH + PATH_SEPARATOR + ".";
+913:+    String command = "javac -classpath .";
+914:     if (classpath != null) {
+915:       // Add specified classpath to command.
+916:       command += PATH_SEPARATOR + classpath;
+917:     }
+
+948:-    String classpath = SYSTEM_CLASS_PATH + PATH_SEPARATOR + dirPath;
+948:+    String classpath = dirPath;
+949:     if (userClassPath != null) {
+950:       classpath += PATH_SEPARATOR + userClassPath;
+951:     }
+\end{Verbatim}
+
+\noindent
+In \<MinimizerTests.java>:
+
+\begin{Verbatim}
+55:-    String classPath = "";
+55:+    String classPath = JUNIT_JAR;
+56:     if (dependencies != null) {
+57:       for (String s : dependencies) {
+58:         Path file = Paths.get(s);
+59:         classPath += (pathSeparator + file.toAbsolutePath().toString());
+60:       }
+61:     }
+\end{Verbatim}
+
+\caption{Fixes made by the Randoop developers in response to our bug report
+  about use of environment variables.  Lines starting with ``\<->'' were
+  removed and those starting with ``\<+>'' were added.}
+\label{fig:randoop-bug-getenv}
+\end{figure}
+
 We reported another suspicious case of order-nondeterminism, in the

### 2019-04-04 design.tex
Edits in sections 2.1 and 2.2

@@
 
-This section presents the key aspects of the design of our type system,
+This section presents the key aspects of our type system,
 called \ourTypeSystem, in the context of a core calculus for an
@@
 \Cref{fig:grammar} formalizes these notions.
-\Cref{fig:grammar}'s $\beta$ is not a basetype as described above:  $\beta$
+For simplicity of exposition,
+\cref{fig:grammar}'s $\beta$ is not a basetype as described above:  $\beta$
 lacks a top-level type, but any type arguments are full types
-$\tau$.  This paper does not formally define basetype, and uses $\beta$,
-for simplicity of exposition.
+$\tau$.
 In our core calculus, a type $\tau$ always has a qualifier.
-Our implementation, \theDeterminismChecker, uses inference and defaults to
-permit users to omit type qualifiers\todo{Forward reference}.
+Our implementation, \theDeterminismChecker, uses inference (\cref{sec:dataflow-java}) and
+defaults (\cref{defaulting}) to
+permit users to omit type qualifiers.
 
 A type checker verifies the types written in a program.
-A type qualifier constrain the set of possible run-time values, that is,
-$\kappa \ /beta \sqsubseteq \beta$.
-As a result, a qualifier type system does not allow any behavior that the
+\OurTypeSystem checks all the standard typing rules including the ones
+shown in \cref{typecheck-rules-standard}.  For \ourTypeSystem, $\tau$
+represents a qualified type rather than a basetype.
+
+A type qualifier constrains the set of possible run-time values, that is,
+$\kappa \ \beta \sqsubseteq \beta$.
+As a result, a qualifier type system does not allow any values that the
 original type system does not.
@@
 affords stronger guarantees.
-(Our implementation \theDeterminismChecker runs the underlying base type
-system, then issues additional errors and warnings.)
-\todo{Some qualifier type systems can be defined independently of the underlying
-type system.  We don't do that because there are interactions between the
-basetypes and the type qualifiers.  This allows greater precision, which
-is important in practice.}
+% (Our implementation \theDeterminismChecker runs the underlying base type
+% system, then issues additional errors and warnings.)
+
+Type qualifier systems are sometimes defined independently of the
+underlying type system.  Because there are interactions between the
+basetypes and the type qualifiers, defining them together improves
+precision, which is important in practice.
+
+\begin{figure}
+    \bigskip
+
+    $\infer{\tau_1 \sqsubseteq \tau_3}{\tau_1 \sqsubseteq \tau_2 \quad \tau_2 \sqsubseteq \tau_3}
+    \quad\quad
+    \infer[\rulename{subtyping}]{\Gamma \vdash x : \tau_2}{\Gamma \vdash x : \tau_1 & \tau_1 \sqsubseteq \tau_2}$
+
+    \bigskip
+
+    $\infer[\rulename{call}]{\Gamma \vdash f(\overline{a_i}) : \tau_r}
+    {\Gamma \vdash f : \overline{\tau_i} \rightarrow \tau_r & \Gamma \vdash \overline{a_i : \tau_i}}$
+
+    \bigskip
+
+    $\infer[\rulename{invariant type arg}]{C_1\angles{\overline{\tau_i}}
+      \sqsubseteq C_2\angles{\overline{\tau_i}}}{C_1 \sqsubseteq C_2}$
+
+    \caption{A sampling of standard typing rules.  We omit other standard rules,
+    for brevity.}
+    \label{typecheck-rules-standard}
+\end{figure}
 
@@
     \quad\quad
-    \infer{\kappa \beta_1 \sqsubseteq \kappa \beta_2}{\beta_1 \sqsubseteq \beta_2}$
+    \infer{\kappa\ \beta_1 \sqsubseteq \kappa\ \beta_2}{\beta_1 \sqsubseteq \beta_2}$
 
-    \caption{Subtyping rules for \ourTypeSystem.  $\sqsubseteq$ is overloaded for classes,
+    \caption{Subtyping rules for \ourTypeSystem's type qualifiers.  $\sqsubseteq$ is overloaded for classes,
     type qualifiers, types, and unqualified types $\beta$.}
     \label{fig:typecheck-rules}
-\end{figure}
-
-\begin{figure}
-    \bigskip
-
-    $\infer{\tau_1 \sqsubseteq \tau_3}{\tau_1 \sqsubseteq \tau_2 \quad \tau_2 \sqsubseteq \tau_3}
-    \quad\quad
-    \infer[\rulename{subtyping}]{\Gamma \vdash x : \tau_2}{\Gamma \vdash x : \tau_1 & \tau_1 \sqsubseteq \tau_2}$
-
-    \bigskip
-
-    $\infer[\rulename{method call}]{\|method|(\overline{a_i}) : \tau_r}{\tau_r \  \|method|(\overline{\tau_i \  p_i}) & \overline{a_i : \tau_i}}$
-
-    \bigskip
-
-    $\infer[\rulename{invariant type arg}]{C_1\angles{\overline{\tau_i}}
-      \sqsubseteq C_2\angles{\overline{\tau_i}}}{C_1 \sqsubseteq C_2}$
-
-    \caption{A sampling of standard typing rules.  We omit other standard rules,
-    for brevity.}
-    \label{typecheck-rules-standard}
 \end{figure}
@@
 an object-oriented programming language.
-
-\OurTypeSystem checks all the standard typing rules including the ones shown in \cref{typecheck-rules-standard} with
-the only difference of having an additional type qualifier on each of the basetypes.
 
@@
 The core of the determinism type system is the following type qualifiers:
-%\todo{introduce just @Det and @NonDet first, and then @OrderNonDet after polymorphism.}
 \begin{itemize}
@@
     \item \<Det> indicates that the expression evaluates to equal values in
-      all executions; for an unordered collection such as a set, iteration
+      all executions; for a collection, iteration
       also yields the values in the same order.
 \end{itemize}
-\Cref{fig:typecheck-rules,fig:determinism-hierarchy} show their subtyping relationship.
+\Cref{fig:typecheck-rules,fig:determinism-hierarchy} show the subtyping
+relationship among the qualifiers.
 
 \<OrderNonDet> may only be written on collections and maps.
-
 A map is a dictionary or associative array, such as a hash table.
 \OurTypeSystem largely treats a map as a collection of key--value pairs.
-
 Both collections and maps may be \<Det>, \<OrderNonDet>, or \<NonDet>, and
@@
     $\infer[\rulename{ordernondet collection}]
-    {\wellformed{\kappa_c \  \beta_c \angles{\kappa_e \ \beta_e}}}
-    {\kappa_c \sqsubseteq \|OrderNonDet| & \kappa_e \sqsubseteq \kappa_c &
-      \beta_c \sqsubseteq \<Collection> & \wellformed{\kappa_e \ \beta_e}}$
+    {\wellformed{\kappa_c \  C \angles{\kappa_e \ \beta_e}}}
+    {\kappa_c \sqsubseteq \<OrderNonDet> &
+      C \sqsubseteq \<Collection> &
+      \kappa_e \sqsubseteq \kappa_c &
+      \wellformed{\kappa_e \ \beta_e}}$
 
@@
     $\infer[\rulename{nondet collection}]
-    {\wellformed{\|NonDet| \ C\angles{\|NonDet| \ \beta_e}}}
-    {C \sqsubseteq \<Collection> & \wellformed{\|NonDet| \ \beta_e}}$
+    {\wellformed{\<NonDet> \ C\angles{\<NonDet> \ \beta_e}}}
+    {C \sqsubseteq \<Collection> & \wellformed{\<NonDet> \ \beta_e}}$
 
@@
 
-%      $\infer[\rulename{ordernondet array}]{\wellformed{\kappa_e \ \beta_e \ \  \kappa_a []}}{\kappa_a \sqsubseteq \|OrderNonDet| & \kappa_e \sqsubseteq \kappa_a }$
+%      $\infer[\rulename{ordernondet array}]{\wellformed{\kappa_e \ \beta_e \ \  \kappa_a []}}{\kappa_a \sqsubseteq \<OrderNonDet> & \kappa_e \sqsubseteq \kappa_a }$
 %     
@@
 %     
-%     $\infer[\rulename{nondet array}]{\wellformed{\|NonDet| \ \beta_e \ \  \|NonDet| []}}{}$
+%     $\infer[\rulename{nondet array}]{\wellformed{\<NonDet> \ \beta_e \ \  \<NonDet> []}}{}$
 %     
@@
     $\infer[\rulename{ordernondet map}]
-    {\wellformed{\kappa_m \  \beta_m \angles{\kappa_k \ \beta_k, \kappa_v \ \beta_v}}}
-    {\kappa_m \sqsubseteq \|OrderNonDet|
+    {\wellformed{\kappa_m \  C \angles{\kappa_k \ \beta_k, \kappa_v \ \beta_v}}}
+    {\kappa_m \sqsubseteq \<OrderNonDet>
+      & C \sqsubseteq \<Map>
       & \kappa_k \sqsubseteq \kappa_m
       & \kappa_v \sqsubseteq \kappa_m
-      & \beta_m \sqsubseteq \<Map>
       & \wellformed{\kappa_k \ \beta_k}
@@
     $\infer[\rulename{nondet map}]
-    {\wellformed{\|NonDet| \ \beta_m\angles{\|NonDet| \ \beta_k, \|NonDet| \ \beta_v}}}
-    {\beta_m \sqsubseteq \<Map> & \wellformed{\|NonDet| \ \beta_k} & \wellformed{\|NonDet| \ \beta_v}}$
+    {\wellformed{\<NonDet> \ C\angles{\<NonDet> \ \beta_k, \<NonDet> \ \beta_v}}}
+    {C \sqsubseteq \<Map> & \wellformed{\<NonDet> \ \beta_k} & \wellformed{\<NonDet> \ \beta_v}}$
 
     \caption{Type well formedness}
+\todo{What if $\beta$ has \<List> within?  Be sure everything works in that case.}
 
@@
 %  array to be well-formed.  There is a bigger issue here.
-%  Consider the ``nondet collection'' rule.  $\beta_c$ is a class (we need
+%  Consider the ``nondet collection'' rule.  $C_c$ is a class (we need
 %  to use a different symbol for it, probably $c$), but $\beta_e$ should be
 %  a full type, which might contain type qualifiers within it.  (Actually,
-%  $\|NonDet|\ \beta_e$ is a type, not $\beta_e$ itself.)  So, there should
-%  be an antecedent stating that $\|NonDet| \beta_e$ is well-formed, and
+%  $\<NonDet>\ \beta_e$ is a type, not $\beta_e$ itself.)  So, there should
+%  be an antecedent stating that $\<NonDet> \beta_e$ is well-formed, and
 %  for the ``ordernondet collection'' rule stating that $\kappa_e \beta_e$
@@
         \multicolumn{2}{c|}{~}  &  \multicolumn{3}{c|}{Element type} \\ \cline{3-5}
-        \multicolumn{2}{c|}{~}  & NonDet     & OrderNonDet & Det \\ \hline
-                        & NonDet      &  valid     &  invalid    & invalid  \\ \cline{2-5}
-        Collection type & OrderNonDet &  invalid   &  valid      & valid    \\ \cline{2-5}
-                        & Det         &  invalid   &  invalid    & valid    \\ \hline
+        \multicolumn{2}{c|}{~}  & \<NonDet>     & \<OrderNonDet> & \<Det> \\ \hline
+                        & \<NonDet>      &  valid     &  invalid    & invalid  \\ \cline{2-5}
+        Collection type & \<OrderNonDet> &  invalid   &  valid      & valid    \\ \cline{2-5}
+                        & \<Det>         &  invalid   &  invalid    & valid    \\ \hline
     \end{tabular}
@@
     $
-    \forallt{\tau} \|OrderNonDet \ C| \angles{\tau} \rightarrow 
-    \|OrderNonDet \ Iterator| \angles{\tau}
+    \forallt{\tau} \<OrderNonDet> \ C \angles{\tau} \rightarrow 
+    \<OrderNonDet> \ \<Iterator> \angles{\tau}
     $
@@
     $
-    \forallt{kappa \beta} \|OrderNonDet \ Iterator|\angles{\kappa \ \beta} \rightarrow
-    \|NonDet|\ \beta
+    \forallt{kappa\ \beta} \<OrderNonDet>\ \<Iterator>\angles{\kappa \ \beta} \rightarrow
+    \<NonDet>\ \beta
     $
@@
     $
-    \forallt{\tau} \|OrderNonDet \ Iterator|\angles{\tau} \rightarrow
-    \|Det \ boolean|
+    \forallt{\tau} \<OrderNonDet>\ \<Iterator>\angles{\tau} \rightarrow
+    \<Det> \ \<boolean>
     $
@@
     $
-    \forallt{\tau} \|OrderNonDet \ C|\angles{\tau} \rightarrow
-    \|Det \ int|
+    \forallt{\tau} \<OrderNonDet>\ C\angles{\tau} \rightarrow
+    \<Det> \ \<int>
     $
@@
     $\infer[\rulename{ordernondet polyup}]
-    {\up{\tau} = \|NonDet \ \beta|}
-    {\tau = \|OrderNonDet \ \beta|
+    {\up{\tau} = \<NonDet> \ \beta}
+    {\tau = \<OrderNonDet> \ \beta
         & \wellformed{\tau}
@@
     {\tau =\kappa \ \beta
-        & \kappa \in \{\|Det, NonDet|\}
+        & \kappa \in \{\<Det>, \<NonDet>\}
         & \wellformed{\tau}
@@
     $\infer[\rulename{ordernondet polydown}]
-    {\down{\tau} = \|Det \ \beta|}
-    {\tau = \|OrderNonDet \ \beta|
+    {\down{\tau} = \<Det> \ \beta}
+    {\tau = \<OrderNonDet> \ \beta}
         & \wellformed{\tau}
@@
 
-\todo{We don't need the OrderNonDet instantiation.  The other two
+\todo{We don't need the \<OrderNonDet> instantiation.  The other two
   instantiations are enough.
@@
 \begin{figure}
-%     $\infer[\rulename{list sort}]{\Gamma \vdash x : \|Det \ List|\angles{\|Det| \ \beta}}{\Gamma \vdash x : \|OrderNonDet \ List|\angles{\|Det| \ \beta} & \|x.sort()|}$
+%     $\infer[\rulename{list sort}]{\Gamma \vdash x : \|Det \ List|\angles{\<Det> \ \beta}}{\Gamma \vdash x : \|OrderNonDet \ List|\angles{\<Det> \ \beta} & \|x.sort()|}$
 %     
@@
     \todo{add line break}
-    $\infer[\rulename{set precision}]{\Gamma \vdash \|x.equals(y) : Det|}{\Gamma \vdash \|x| : \|OrderNonDet| \ \|Set|\angles{ \kappa_{x} \ \beta_{x}}, \|y| : \|OrderNonDet| \ \|Set|\angles{ \kappa_{y} \ \beta_{y}} & \kappa_{x} \ \beta_{x} \nsqsubseteq \|OrderNonDet \ List| & \kappa_{y} \ \beta_{y} \nsqsubseteq \|OrderNonDet \ List| }$
+    $\infer[\rulename{set precision}]{\Gamma \vdash \|x.equals(y)| : \<Det>}{\Gamma \vdash \|x| : \<OrderNonDet> \ \|Set|\angles{ \kappa_{x} \ \beta_{x}}, \|y| : \<OrderNonDet> \ \|Set|\angles{ \kappa_{y} \ \beta_{y}} & \kappa_{x} \ \beta_{x} \nsqsubseteq \<OrderNonDet> \ \<List> & \kappa_{y} \ \beta_{y} \nsqsubseteq \<OrderNonDet> \ \<List> }$
 
@@
 
-    $\infer[\rulename{map precision}]{\Gamma \vdash \|x.get(y) : Det|}{\Gamma \vdash \|x : OrderNonDet \ Map\angles{ Det \ \beta_k, \kappa_v\ \beta_v }, y : Det \ \beta|}$
+    $\infer[\rulename{map precision}]{\Gamma \vdash \|x.get(y)| : \<Det>}{\Gamma \vdash \|x : \<OrderNonDet> \ \<Map>\angles{ \<Det> \ \beta_k, \kappa_v\ \beta_v }, y : \<Det> \ \beta|}$
 

### 2019-04-04 java-design.tex
Fix LaTeX error
Edits in sections 2.1 and 2.2

@@
 \end{Verbatim}
-This indicates that method \<foo> can be called with arguments having any of the \codeid{Det or NonDet} type qualifiers.
+This indicates that method \<foo> can be called with arguments having any of the \codeid{@Det} or \codeid{@NonDet} type qualifiers.
 %   (\<@OrderNonDet> is not allowed because it is invalid on primitive types).
@@
 
-We prohibit the creation of a \codeid{Det HashMap} or an \codeid{OrderNonDet TreeMap} in our determinism checker.
+We prohibit the creation of a \codeid{@Det HashMap} or an \codeid{@OrderNonDet TreeMap} in our determinism checker.
 \Cref{fig-creation-rules} formalizes these rules. Similar constraints apply to \<HashSet>s and \<TreeSet>s which are
@@
 \begin{figure}
-    $\infer[\rulename{hashmap creation}]{\|new \ \kappa\ HashMap()|}{\kappa \in \{ \<OrderNonDet>, \<NonDet> \}}$
+    $\infer[\rulename{valid hashmap}]{\vdash: \|\kappa\ HashMap()|}{\kappa \in \{ \<OrderNonDet>, \<NonDet> \}}$
 
@@
 
-    $\infer[\rulename{treemap creation}]{\|new \ \kappa\ TreeMap()|}{\kappa \in \{ \<Det>, \<NonDet> \}}$
+    $\infer[\rulename{valid treemap}]{\vdash: \|\kappa\ TreeMap()|}{\kappa \in \{ \<Det>, \<NonDet> \}}$
 
@@
 
-    $\infer[\rulename{hashset creation}]{\|new \ \kappa\ HashSet()|}{\kappa \in \{ \<OrderNonDet>, \<NonDet> \}}$
+    $\infer[\rulename{valid hashset}]{\vdash: \|\kappa\ HashSet()|}{\kappa \in \{ \<OrderNonDet>, \<NonDet> \}}$
 
@@
 
-    $\infer[\rulename{treeset creation}]{\|new \ \kappa\ TreeSet()|}{\kappa \in \{ \<Det>, \<NonDet> \}}$
+    $\infer[\rulename{valid treeset}]{\vdash: \|\kappa\ TreeSet()|}{\kappa \in \{ \<Det>, \<NonDet> \}}$
     \caption{\<Map> and \<Set> creation rules}
@@
 
-The return type of \<System.getenv}, which reads an operating system
+The return type of \<System.getenv>, which reads an operating system
 environment variable, is always \<NonDet>.

### 2019-04-04 design.tex
Edits in 2.3.1

@@
 
-    \caption{Type well formedness}
+    \caption{Type well-formedness}
 \todo{What if $\beta$ has \<List> within?  Be sure everything works in that case.}
@@
 
-\Cref{type-validity} shows the type well formedness
-%\todo{The caption says
-%  ``type well formedness''.  Choose one term and use it consistently
-%  throughout.}
+% "invalid" is easier to read than "non-well-formed".
+\Cref{type-validity} shows the type well-formedness, or type validity,
 rules of \ourTypeSystem.
-(For brevity, we use ``valid'' synonymously with ``well-formed'' and
-``invalid'' synonymously with ``non-well-formed''.)
-As stated in the rule \rulename{noncollection}, the types \<Det> and \<NonDet>
-can be written on any non-collection Java basetype.
 
@@
 the collection's type qualifier (\rulename{collection} rules of
-\cref{type-validity}). Note that the \rulename{ordernondet} versions also handle \<Det>,
+\cref{type-validity}). Note that the \rulename{ordernondet} rules also
+permit \<Det> collections,
 because $\<Det> \sqsubseteq \<OrderNonDet>$.
@@
 It is possible to store an expression of type \<Det> in a \<NonDet>
-collection, but all elements retrieved from a \<NonDet> collection have
-type \<NonDet>.  If the type \ptype{NonDet List}{Det String} were
+collection, but any value retrieved from a \<NonDet> collection has
+type \<NonDet>.
+
+If the type \ptype{NonDet List}{Det String} were
 permitted, the following type hole would exist:
@@
 \lptype{Det    List}{Det String} ddlist
-\lptype{NonDet List}{Det String} ndlist = ddlist   // permitted by subtyping rules
+\lptype{NonDet List}{Det String} ndlist = ddlist   // assignment permitted by subtyping rules
 NonDet Int ni = ...
@@
 \noindent
-The deterministic string \<"anywhere"> is placed in the list at an
+In this code snippet,
+the deterministic string \<"anywhere"> is placed in the list at an
 arbitrary (that is, nondeterministic) index.  This must be prohibited,
 because it unsoundly permits the deterministic list \<ddlist> to differ
-from execution to execution.
+from execution to execution.  \OurTypeSystem does so by forbidding the type
+\ptype{NonDet List}{Det String}.
 
@@
     {\down{\tau} = \<Det> \ \beta}
-    {\tau = \<OrderNonDet> \ \beta}
+    {\tau = \<OrderNonDet> \ \beta
         & \wellformed{\tau}

### 2019-04-04 experiments.tex
deflaker expts
Tweaks

@@
 All the flakiness that NonDex found was due to 7 methods in 3 classes. We enumerate all the methods corresponding to
-each class in Figure~\ref{fig:flaky-sources}.
+each class in \cref{fig:flaky-sources}.
 \begin{figure}
@@
 We provide samples of the annotated source code from these benchmarks and the errors as reported by \TheDeterminismChecker 
-in Figure~\ref{fig:nondex-source}.
+in \cref{fig:nondex-source}.
 \begin{figure}
@@
 in its repository.
+\todo{Deflaker works now. Tried on retrofit: only works where there is sufficient change in code coverage between head and its parent.
+Looked throught= about 10 commits, didn't find differential code coverage and therefore no flakes}
+\todo{Deflaker requires maven build. Randoop is on gradle.}
 

### 2019-04-04 experiments.tex
deflaker todo

@@
 \section{Comparison to Deflaker}
-Deflaker~\cite{deflaker}, like NonDex, reports tests that could be flaky. Ideally, we would have 
-liked to run all three tools (NonDex, DeFlaker, and \TheDeterminismChecker) on an open source project
-that wasn't explored by any of the tool for a fair comparison. Unfortunately, there are issues with the
-DeFlaker plugin and we weren't able to run it on any project. We have created a Github issue for the same
-in its repository.
-\todo{Deflaker works now. Tried on retrofit: only works where there is sufficient change in code coverage between head and its parent.
-Looked throught= about 10 commits, didn't find differential code coverage and therefore no flakes}
-\todo{Deflaker requires maven build. Randoop is on gradle.}
+Deflaker~\cite{deflaker}, like NonDex, reports tests that could be flaky. 
+While we would have like to run DeFlaker on Randoop for a fair comparison of the tools,
+we were unable to do so because Deflaker works with projects built with maven and Randoop
+uses gradle for its build system. We plan to convert Randoop's build system to maven and 
+perform this case study.
+%Ideally, we would have 
+%liked to run all three tools (NonDex, DeFlaker, and \TheDeterminismChecker) on an open source project
+%that wasn't explored by any of the tool for a fair comparison. Unfortunately, there are issues with the
+%DeFlaker plugin and we weren't able to run it on any project. We have created a Github issue for the same
+%in its repository.
+%\todo{Deflaker works now. Tried on retrofit: only works where there is sufficient change in code coverage between head and its parent.
+%Looked throught= about 10 commits, didn't find differential code coverage and therefore no flakes}
+%\todo{Deflaker requires maven build. Randoop is on gradle.}
 

### 2019-04-04 design.tex
Tweaks

@@
 affects access, iteration, searching, etc.
-(Rules \rulename{iteration} and \rulename{next element} in Figure~\ref{fig-ordernondet-rules})
+(Rules \rulename{iteration} and \rulename{next element} in \cref{fig-ordernondet-rules}.)
 \item
@@
 \begin{figure}
+\todo{Give the types for get and set, which are more basic list/array operations,
+  at the beginning before iteration.}
+
 %    $\infer[\rulename{iteration}]
@@
     $
-    \forallt{kappa\ \beta} \<OrderNonDet>\ \<Iterator>\angles{\kappa \ \beta} \rightarrow
+    \forallt{\kappa\ \beta} \<OrderNonDet>\ \<Iterator>\angles{\kappa \ \beta} \rightarrow
     \<NonDet>\ \beta
@@
 
-\caption{\<OrderNonDet> Collection rules}
+\caption{\<OrderNonDet> Collection rules.  $C$ is a collection class.
+    These are partial types indicating the methods' behavior on
+    \<OrderNonDet> arguments.  Full types are given in \cref{polymorphism}.}
+\todo{These should be method names, not rule names, right?  The figure is
+  just giving the types of methods.}
 \label{fig-ordernondet-rules}
@@
 affecting binding of those variables --- that is, it affects which
-instantiation of an abstraction is chosen at a use site 
+instantiation of a (method) abstraction is chosen at a use site
 (\cref{bindings-uses}).
@@
 It acts as if it has multiple different types, and each use site is
-typechecked using the most specific applicable (instantiated,
-non-polymorphic) type.
+typechecked using the most specific applicable instantiated
+non-polymorphic type.
 
@@
 
-\todo{Greekify the following}
 \begin{itemize}

### 2019-04-04 design.tex
Edits in section 2.4.2

@@
 
-  $\forallt{\beta} \<Set>\angles{\<Det>\ \beta} \rightarrow \<NonDet>\ \beta$
+  $\<choose> : \forallt{\beta} \<Set>\angles{\<Det>\ \beta} \rightarrow \<NonDet>\ \beta$
 
@@
   common, even if the type is decomposed and the parts used independently,
-  as in this more general signature for the \<choose> function:
+  as in this more general type for the \<choose> function:
 
-  $\forallt{\kappa\ \beta} \<Set>\angles{\kappa\ \beta} \rightarrow \<NonDet>\ \beta$
+  $\<choose> : \forallt{\kappa\ \beta} \<Set>\angles{\kappa\ \beta} \rightarrow \<NonDet>\ \beta$
 
 \item
-Qualifier polymorphism is common.  For example, here is the signature for
-the \<length> method on strings:
+Qualifier polymorphism is common.  For example, here are types for
+the \<length> method on strings and the addition operation:
 
 $\<length>\ :\ \forallt{\kappa} \kappa\ \<String> \rightarrow \kappa\ \<Int>$
-
-\noindent
-and here is that for the addition operation:
 
@@
 
-$\<NonDet>\ \<String> \rightarrow \<NonDet> \<Int>$
+$\<length>\ : \<NonDet>\ \<String> \rightarrow \<NonDet> \<Int>$
 
-$\<Det>\ \<String> \rightarrow \<Det> \<Int>$
+$\<length>\ :\<Det>\ \<String> \rightarrow \<Det> \<Int>$
 
@@
 Here is an example:
-\todo{(put it in a table)}
 
-\begin{Verbatim}
-NonDet Int ni;
-   Det Int di;
+Suppose we have two variables $\<ni> : \<NonDet Int>$ and $\<di> : \<Det Int>$.
+This table shows, for several method invocations, the 
+most specific applicable instantiation for that invocation and the type of
+the call expression:
 
-plus(ni, ni)  // has type NonDet Int; most specific applicable instantiation is NonDet Int -> NonDet Int
-plus(di, di)  // has type Det Int; most specific applicable instantiation is Det Int -> Det Int
-plus(ni, di)  // has type NonDet Int; most specific applicable instantiation is NonDet Int -> NonDet Int
-\end{Verbatim}
+\begin{tabular}{lll}
+  Expression & instantiation & type \\
+\<plus(ni, ni)> & $\<NonDet>\ \<String> \rightarrow \<NonDet> \<Int>$ & \<NonDet> \\
+\<plus(di, di)> & $\<Det>\ \<String> \rightarrow \<Det> \<Int>$ & \<Det> \\
+\<plus(ni, di)> & $\<NonDet>\ \<String> \rightarrow \<NonDet> \<Int>$ & \<NonDet> \\
+\<plus(di, ni)> & $\<NonDet>\ \<String> \rightarrow \<NonDet> \<Int>$ & \<NonDet> \\
+\end{tabular}
+
+\smallskip
 
 We adopt the convention that  polymorphism is not instantiated in ways
-that would create invalid types.  For example, the \<plus> polymorphic
-function would not be instantiated at $\kappa = \<OrderNonDet>.$
+that would create invalid types.  For example, the \<length> and \<plus> polymorphic
+functions would not be instantiated at $\kappa = \<OrderNonDet>.$
+This makes no difference for these particular functions:  such an
+instantiation would never be the most specific applicable one.
+This is important for some of the type rules given in this paper, which
+would otherwise need complex preconditions that recapitulate the type
+well-formedness rules.
 
@@
 \<OrderNonDet> collection would choose the \<NonDet> instantiation and the
-size would be \<NonDet Int>.
-%
-A similar problem occurs for the \<first> operation that returns the first
-element of a list.
+return type at that call site would be \<NonDet Int> rather than the
+desired \<Det Int>.
 
@@
 \<OrderNonDet> by something higher in the type hierarchy.
-The $\downarrow$ operator is analogous, but converts \<OrderNonDet> to
+The $\downarrow$ operator is analogous, but it converts \<OrderNonDet> to
 \<Det>, which is lower in the type hierarchy.
@@
 
-\todo{As I mentioned in email, $\<NonDet>\ \beta_e$ looks wrong to me,
-  because $\beta_e$ may be a collection whose type arguments are not
-  compatible with \<NonDet>.  Maybe just remark that the instantiation does
-  not happen in that case.  If we try to adjust the type arguments to fit
-  \<NonDet>, then unsoundness occurs.}
+\paragraph{Discussion of the type of \<length>}
+
+In the type of \<first>, the $\uparrow$ is not strictly necessary.
+
+Suppose that $\<myStrings> : \ptype{OrderNonDet List}{Det String}$.
+The type of \<first(myStrings)> should be \<NonDet String>.
+
+If the type of \<first> were
+
+$\<first> : \forallt{\kappa} \kappa\ \ListKB \rightarrow \kappa\ \beta_e$
+
+\noindent
+then the instantiation
+
+$\<first> : \<OrderNonDet>\ \ptype{List}{$\kappa$ String} \rightarrow \<OrderNonDet>\ \<String>$
+
+\noindent
+would be invalid.
+At the call \<first(myStrings)>, the most specific instantiation would be
+
+$\<first> : \<NonDet>\ \ptype{List}{$\kappa$ String} \rightarrow \<NonDet>\ \<String>$
+
+\noindent making the type of the expression \<NonDet String> as desired.
+We prefer the type
+
+$\<first> : \forallt{\kappa} \kappa\ \ListKB \rightarrow \up{\kappa}\ \beta_e$
+
+\noindent
+because it is more explicit about the method's effect and more
+instantiations, rather than requiring
+a reader to reason about valid types.
+
+The rules prevent some invocations.  Suppose we have the following variable:
+
+$\<listOfLists> : \ptype{OND List}{\ptype{Det List}{Det String}}$
+
+\noindent
+Then the invocation \<first(listOfLists)> is illegal.
+Its type would be \ptype{NonDet List}{Det String}, which is an invalid
+type.
+
+This restriction prevents programmers from writing some code, but we never
+found this to be a problem in our case studies.
 

### 2019-04-04 design.tex
Edits through section 2.5

@@
 
+\begin{figure}
+    $\infer{\up{\<NonDet>} = \<NonDet>}{}
+    \quad
+    \infer{\up{\<OrderNonDet>} = \<NonDet>}{}
+    \quad
+    \infer[\rulename{polyup}]{\up{\<Det>} = \<Det>}{}
+    $
+
+    \bigskip
+
+    $\infer{\down{\<NonDet>} = \<NonDet>}{}
+    \quad
+    \infer{\down{\<OrderNonDet>} = \<Det>}{}
+    \quad
+    \infer[\rulename{polydown}]{\down{\<Det>} = \<Det>}{}
+    $
+
+\caption{Polymorphic resolution of $\uparrow$ and $\downarrow$ operators.  $\uparrow$ is overloaded for types and qualifiers.}
+\label{fig:poly-resolutions}
+\end{figure}
+
 \OurTypeSystem resolves this problem by introducing two operators over
@@
 
-\begin{figure}
-    $\infer[\rulename{ordernondet polyup}]
-    {\up{\tau} = \<NonDet> \ \beta}
-    {\tau = \<OrderNonDet> \ \beta
-        & \wellformed{\tau}
-    }$
-
-    \bigskip
-
-    $\infer[\rulename{polyup}]
-    {\up{\tau} = \kappa \ \beta}
-    {\tau =\kappa \ \beta
-        & \kappa \in \{\<Det>, \<NonDet>\}
-        & \wellformed{\tau}
-    }$
-
-    \bigskip
-
-    $\infer[\rulename{ordernondet polydown}]
-    {\down{\tau} = \<Det> \ \beta}
-    {\tau = \<OrderNonDet> \ \beta
-        & \wellformed{\tau}
-    }$
-
-    \bigskip
-
-    $\infer[\rulename{polydown}]
-    {\down{\tau} = \kappa \ \beta}
-    {\tau =\kappa \ \beta
-        & \kappa \in \{\<Det>, \<NonDet>\}
-        & \wellformed{\tau}
-    }$
-\caption{Polymorphic resolution of $\uparrow$ and $\downarrow$ operators ($\uparrow$ is overloaded for types and qualifiers)}
-\todo{I think this needs to do something to nested collections.  Or maybe
-  just clarify that if there is a nested collection such that this would
-  yield an invalid type, then the instantiation does not apply.  That might
-  be cleanest.}
-\label{fig:poly-resolutions}
-\end{figure}
-
 \subsubsection{Differentiating binding and use}\label{bindings-uses}
+
+\todo{TODO: Need to work on this section.}
+
+In the annotated JDK, the type of get is:
+
+\begin{Verbatim}
+    @PolyDet("up") E get(@PolyDet List<E> this, @PolyDet int index);
+\end{Verbatim}
 
@@
 
-\subsection{Dataflow analysis and type refinement}\label{dataflow}
+\subsection{Type inference, dataflow analysis, and type refinement}\label{dataflow}
+
+\todo{Move all of this to Section 3 since the core calculus has no type
+  inference?
+  But we do want to discuss \<sort>.  So maybe just move the discussion of type inference?}
+
+The core calculus requires a qualifier on every type.}
+
 \TheDeterminismChecker performs local type inference within method bodies.
@@
 type argument subtyping.
-Second, the only possible type argument for a \<Det> collection is \<Det>.
-
-It is not permitted to
+Second, the only possible element type for a \<Det> collection is \<Det>.
 
@@
   $\infer[\rulename{sort}]
-  {\Gamma \vdash x : \<Det List>\angles{\<Det>\ \beta}}
-  {\Gamma \vdash x : \<OrderNonDet List>\angles{\<Det>\ \beta} & \<sort(x)>}$
+  {\Gamma \vdash x_{\mathrm{post}} : \<Det List>\angles{\<Det>\ \beta}}
+  {\Gamma \vdash x_{\mathrm{pre}} : \<OrderNonDet List>\angles{\<Det>\ \beta} & \<sort(>x_{\mathrm{pre}}\<)>}$
 
@@
     \caption{Sorting type refinement rules.  \<sort> is a method that works
-    by side effect and changes the value---and the type---of its receiver.}
-\todo{These don't capture the notion of ``the type \emph{after} the sort call''.}
+      by side effect and changes the value---and the type---of its receiver.
+    ``$x_{\mathrm{pre}}$ stands for a value before the call to \<sort>, and
+      ``$x_{\mathrm{post}}$ stands for the value after the call.
+    }
     \label{fig-sorting}
 \end{figure}
-\TheDeterminismChecker type refines a receiver of type \codeid{OrderNonDet List} to a \codeid{Det List} if all the
-type argument of this \codeid{List} have the type qualifier \codeid{Det} (Rule \rulename{sort} in \cref{fig-sorting}).
-As per these rules, if \codeid{sort()} is called on a receiver of type \codeid{OrderNonDet List<Det Set<Det String>>}, it will
-be type refined to \codeid{Det List<Det Set<Det String>>}. However a receiver of type \codeid{OrderNonDet List<OrderNonDet Set<Det String>>} will not be type refined.
+As an example, if \codeid{sort()} is called on a receiver of type
+\codeid{OrderNonDet List<Det Set<Det String>>}, it will be type refined to
+\codeid{Det List<Det Set<Det String>>}. However, it is not permitted to
+call \<sort> on a receiver of type \codeid{OrderNonDet List<OrderNonDet
+  Set<Det String>>}.  The rule is not applicable, because doing so would
+instantiate invalid types.
+
+As another example, it is only sometimes permitted to call \<shuffle> on a receiver of
+type \ptype{Det List}{Det String}.  It is illegal to call \<shuffle> on a
+receiver of \emph{declared} type \ptype{Det List}{Det String}, because doing so would change
+the receiver's type to be inconsistent with its declaration.  However, it
+is permitted to call \<shuffle> on a receiver of type \ptype{Det List}{Det
+  String} that was declared as \ptype{OrderNonDet List}{Det
+  String} but then refined to \ptype{Det List}{Det String}, for example by
+calling \<sort> or via assignment.
 

### 2019-04-04 design.tex
Finished with partial pass through section 2

@@
 \subsection{Maps}\label{maps}
-A map is a collection of keys--value pairs.
-Like other collections, a map can be nondeterministic,
+A map is a collection of key--value pairs.
+Like other collections, a map or set can be nondeterministic,
 order-nondeterministic, or deterministic.
+
+It might seem that the notion of an order-nondeterministic set or map is
+nonsensical, since the set and map specifications make no promises about
+iteration order.  However, some subtypes do.
+We explain why each of these is possible by considering three common
+implementation strategies.
+
+\begin{itemize}
+\item
+A hash table is order-nondeterministic, because hash codes are
+nondeterministic in general.
+\item
+A hash table can be augmented by a linked list recording insertion order,
+so that iteration order is deterministic.
+\item
+A map or set can be backed by a sorted collection, such as a tree.  Such a map is
+either \<NonDet> (if \<NonDet> keys or values are inserted into it) or
+\<Det>, but never \<OrderNonDet>.
+\end{itemize}
+
 Analogously to set membership, lookups are deterministic in
@@
 
-We explain why each of these is possible by considering three common
-implementation strategies.
-
-A hash table is order-nondeterministic, because hash codes are
-nondeterministic in general.
-
-A hash table can be augmented by a linked list recording insertion order,
-so that iteration order is deterministic.
-
-A map can be backed by a sorted collection, such as a tree.  Such a map is
-either \<NonDet> (if \<NonDet> keys or values are inserted into it) or
-\<Det>, but never \<OrderNonDet>.
-
 \subsection{Determinism checker specific rules for precision}\label{precision}
+
+\todo{Sets vs lists.  Set equality is independent of order.}
+An order nondeterministic list does not provide any guarantee on the
+equality as it is dependent on the iteration order
+
 We annotated the \codeid{equals()} method of the \<Set> class in the JDK as
@@
 
-    \caption{Improving precision}
+    \caption{Special rules for set equality and map lookup.}
 \todo{In the \rulename{map precision} rule, should the \<Det> in the
@@
 receiver or the argument happens to be an \codeid{OrderNonDet List}.
-\todo{The following is the key!  Put in earlier.}(An order nondeterministic list does not provide any guarantee on the
-equality as it is dependent on the iteration order).
-
-Similarly, the \codeid{get()} method in the \<Map> class is annotated with polymorphic qualifiers and the rule \rulename{map precision}\ in \cref{fig-precision-rules} improves precision for \<OrderNonDet> receivers.
-
-\subsection{Extending \ourTypeSystem}\label{extending-det-checker}
-We have primarily focused on collections and annotated the corresponding classes in the JDK.
-These qualifiers provide machine verifiable specifications of determinism properties.
-Another source of nondeterminism that we haven't analyzed closely is the File system methods.
-A program analysis developer interested in verifying determinism guarantees with respect to 
-the File system can easily do so by annotating the parameter and return types of those methods in the JDK.
-Alternatively, a user can write these qualifiers in a \textit{stub} file and supply it to the type checker. A \textit{stub}
-only needs to contain method declarations along with their qualifiers.
+
+Similarly, the rule \rulename{map precision}\ in \cref{fig-precision-rules}
+improves precision for \<OrderNonDet> receivers.
 

### 2019-04-04 java-design.tex
structuring section 3
Finished with partial pass through section 2
benchmark example

@@
 client code that would produce errors at compile time.
+%\begin{figure}
+%    \begin{verbatim}
+%    // Annotated JDK methods
+%    public class Random implements java.io.Serializable {
+%        public @NonDet Random() {}
+%    }
+%    public class PrintStream extends FilterOutputStream 
+%        implements Appendable, Closeable {
+%        public void println(@Det Object x) {}
+%    }
+%    
+%    // Client code
+%    class Client {
+%        void test() {
+%            @Det double d = Math.random(); // Error - subtyping rules violated.
+%            @NonDet double nd = Math.random(); // No error.
+%            System.out.println(nd); // Error - println takes @Det arguments.
+%        }
+%    }
+%    \end{verbatim}
+%    \caption{Example: Errors detected by \theDeterminismChecker.}
+%\todo{Use real examples of bugs from case studies, throughout.  Artificial
+%  examples are less compelling (and lead the reader to wonder whether no
+%  real example exist).  Put the first example of an error in a figure in
+%  Section 1, to help with motivating the problem.}
+%    \label{code-determinism}
+%\end{figure}
+
 \begin{figure}
@@
     // Annotated JDK methods
-    public class Random implements java.io.Serializable {
-        public @NonDet Random() {}
+    public interface Map<K,V> {
+        @PolyDet Set<Map.Entry<K, V>> entrySet(@PolyDet Map<K,V> this);
     }
-    public class PrintStream extends FilterOutputStream 
-        implements Appendable, Closeable {
-        public void println(@Det Object x) {}
+    public interface Iterator<E> {
+        @PolyDet("up") E next(@PolyDet Iterator<E> this);
     }
@@
     // Client code
-    class Client {
-        void test() {
-            @Det double d = Math.random(); // Error - subtyping rules violated.
-            @NonDet double nd = Math.random(); // No error.
-            System.out.println(nd); // Error - println takes @Det arguments.
+    public class MapUtils {
+        public static <K extends @PolyDet Object, V extends @PolyDet Object> 
+                                @Det String toString(@PolyDet Map<K,V> map) {
+            ...
+            for(Map. @Det Entry<K,V> entry : map.entrySet()) { ... }
         }
     }
+    [ERROR] MapUtils.java:[20,50] [enhancedfor.type.incompatible] 
+    incompatible types in enhanced for loop.
+    found : @PolyDet Entry<K extends @PolyDet Object,V extends @PolyDet Object>
+    required: @Det Entry<K extends @PolyDet Object,V extends @PolyDet Object>
     \end{verbatim}
-    \caption{Example: Errors detected by \theDeterminismChecker.}
-\todo{Use real examples of bugs from case studies, throughout.  Artificial
-  examples are less compelling (and lead the reader to wonder whether no
-  real example exist).  Put the first example of an error in a figure in
-  Section 1, to help with motivating the problem.}
+    \caption{Example: Error detected by \theDeterminismChecker in \<scribe-java>~\cite{nondex}.}
+%    \todo{Use real examples of bugs from case studies, throughout.  Artificial
+%        examples are less compelling (and lead the reader to wonder whether no
+%        real example exist).  Put the first example of an error in a figure in
+%        Section 1, to help with motivating the problem.}
     \label{code-determinism}
@@
 \end{itemize}
+
+\subsection{Java collection types}\label{sec:maps-java}
+
+A \<Map> is deterministic if its \<entrySet> is deterministic.
+In other words, iterating over this \<entrySet> is guaranteed to produce the same values in the exact same order across executions.
+The \<Map> interface in Java doesn't define an \<iterator()> interface. Therefore it's sound to equate the determinism of the \<Set>
+returned by \<entrySet()> to the determinism of the \<Map>. Interestingly, we found that every implementation of the \<Map> 
+interface had the following property: if the \<Set> returned by \<entrySet()> has a determinism type $\kappa$, then
+the \<Set>s returned by \<keySet()> and \<values()> also have the exact same determinism type $\kappa$.
+
+We noticed the following properties with respect to determinism regarding the most widely used \<Map> implementations:
+\begin{itemize}
+    \item \<HashMap> can never guarantee fully deterministic iteration over its entries. A \codeid{Det HashMap} does not exist.
+    \item \<LinkedHashMap> can have any of \<Det>, \<OrderNonDet>, or \<NonDet> types. Iterating over a \<LinkedHashMap> returns
+    its entries in the order of their insertion. An \<OrderNonDet LinkedHashMap> can be created by calling its constructor with a \<HashMap>.
+    \item \<TreeMap> can either be \<Det> or \<NonDet>. An \<OrderNonDet TreeMap> doesn't exist because the entries are always ordered.
+\end{itemize}
+
+We prohibit the creation of a \codeid{@Det HashMap} or an \codeid{@OrderNonDet TreeMap} in our determinism checker.
+\Cref{fig-creation-rules} formalizes the type wellformedness rules for maps and sets. Similar constraints apply to \<HashSet>s and \<TreeSet>s which are
+also expressed by the last two rules of \cref{fig-creation-rules}.
+
+\begin{figure}
+    $\infer[\rulename{valid hashmap}]{\vdash: \|\kappa\ HashMap()|}{\kappa \in \{ \<OrderNonDet>, \<NonDet> \}}$
+
+    \bigskip
+
+    $\infer[\rulename{valid treemap}]{\vdash: \|\kappa\ TreeMap()|}{\kappa \in \{ \<Det>, \<NonDet> \}}$
+
+    \bigskip
+
+    $\infer[\rulename{valid hashset}]{\vdash: \|\kappa\ HashSet()|}{\kappa \in \{ \<OrderNonDet>, \<NonDet> \}}$
+
+    \bigskip
+
+    $\infer[\rulename{valid treeset}]{\vdash: \|\kappa\ TreeSet()|}{\kappa \in \{ \<Det>, \<NonDet> \}}$
+    \caption{\<Map> and \<Set> creation rules}
+    \todo{Are these type well-formedness rules?}
+    \todo{What about \<new TreeSet(myOrderedCollection<Det ...>)> which yields a
+        deterministic set?}
+    \label{fig-creation-rules}
+\end{figure}
+
+In the JDK, \<LinkedHashSet> is implemented as a subclass of \<HashMap>.
+\todo{Talk about this design choice. Josh Bloch's book: "If the code surrounding the first declaration depended on LinkedHashSet's ordering policy,
+    then it would be incorrect to substitute HashSet for LinkedHashSet in the declaration, because HashSet makes no guarantee concerning iteration order."}
 
@@
 
-\subsection{Maps}\label{sec:maps-java}
-
- A \<Map> is deterministic if its \<entrySet> is deterministic.
-In other words, iterating over this \<entrySet> is guaranteed to produce the same values in the exact same order across executions.
-The \<Map> interface in Java doesn't define an \<iterator()> interface. Therefore it's sound to equate the determinism of the \<Set>
-returned by \<entrySet()> to the determinism of the \<Map>. Interestingly, we found that every implementation of the \<Map> 
-interface had the following property: if the \<Set> returned by \<entrySet()> has a determinism type $\kappa$, then
-the \<Set>s returned by \<keySet()> and \<values()> also have the exact same determinism type $\kappa$.
-
-We noticed the following properties with respect to determinism regarding the most widely used \<Map> implementations:
-\begin{itemize}
-    \item \<HashMap> can never guarantee fully deterministic iteration over its entries. A \codeid{Det HashMap} does not exist.
-    \item \<LinkedHashMap> can have any of \<Det>, \<OrderNonDet>, or \<NonDet> types. Iterating over a \<LinkedHashMap> returns
-    its entries in the order of their insertion. An \<OrderNonDet LinkedHashMap> can be created by calling its constructor with a \<HashMap>.
-    \item \<TreeMap> can either be \<Det> or \<NonDet>. An \<OrderNonDet TreeMap> doesn't exist because the entries are always ordered.
-\end{itemize}
-
-We prohibit the creation of a \codeid{@Det HashMap} or an \codeid{@OrderNonDet TreeMap} in our determinism checker.
-\Cref{fig-creation-rules} formalizes these rules. Similar constraints apply to \<HashSet>s and \<TreeSet>s which are
-also expressed by the last two rules of \cref{fig-creation-rules}.
-
-\begin{figure}
-    $\infer[\rulename{valid hashmap}]{\vdash: \|\kappa\ HashMap()|}{\kappa \in \{ \<OrderNonDet>, \<NonDet> \}}$
-
-    \bigskip
-
-    $\infer[\rulename{valid treemap}]{\vdash: \|\kappa\ TreeMap()|}{\kappa \in \{ \<Det>, \<NonDet> \}}$
-
-    \bigskip
-
-    $\infer[\rulename{valid hashset}]{\vdash: \|\kappa\ HashSet()|}{\kappa \in \{ \<OrderNonDet>, \<NonDet> \}}$
-
-    \bigskip
-
-    $\infer[\rulename{valid treeset}]{\vdash: \|\kappa\ TreeSet()|}{\kappa \in \{ \<Det>, \<NonDet> \}}$
-    \caption{\<Map> and \<Set> creation rules}
-    \todo{Are these type well-formedness rules?}
-\todo{What about \<new TreeSet(myOrderedCollection<Det ...>)> which yields a
-  deterministic set?}
-    \label{fig-creation-rules}
-\end{figure}
-
-In the JDK, \<LinkedHashSet> is implemented as a subclass of \<HashMap>.
-\todo{Talk about this design choice. Josh Bloch's book: "If the code surrounding the first declaration depended on LinkedHashSet's ordering policy,
-    then it would be incorrect to substitute HashSet for LinkedHashSet in the declaration, because HashSet makes no guarantee concerning iteration order."}
-
 \subsection{Precision improvements}\label{sec:precision-java}

### 2019-04-05 introduction.tex
Edits in sections 2.1 and 2.2
intro edits

@@
 We have created an analysis that detects nondeterminism or verifies its
-absence.
+absence in sequential programs.
 Our analysis permits a programmer to specify which parts of their program
@@
 approaches that identify flaky test cases.
-NonDex uses a modified JVM that returns different results on different
+NonDex~\cite{nondex} uses a modified JVM that returns different results on different
 executions, for a few key JDK methods with loose specifications.  Running a
@@
 
-  \item We implemented analysis, as a pluggable type system for Java, in a
+  \item We implemented the analysis, as a pluggable type system for Java, in a
     tool called \theDeterminismChecker.
@@
 
-  \item We compared \theDeterminismChecker our tool against state-of-the-art flaky test
+  \item We compared our tool, \theDeterminismChecker, against state-of-the-art flaky test
     detectors, on their benchmarks.
@@
 
-\todo{It is essential that the introduction includes an example real-world
-  defect that \theDeterminismChecker found.}
+\todo{1211 LOC of Java in the determinism checker}
+
+%\todo{It is essential that the introduction includes an example real-world
+%  defect that \theDeterminismChecker found.}
 

### 2019-04-05 introduction.tex
into edits

@@
 
+We implemented \TheDeterminismChecker as a pluggable type checker on top of the checker-framework available at \href{https://checkerframework.org/}{https://checkerframework.org/}.
+\TheDeterminismChecker is written in Java and
+spans approximately 1.2K lines of code.
+
 This paper makes the following contributions:
@@
 
-\todo{1211 LOC of Java in the determinism checker}
-
 %\todo{It is essential that the introduction includes an example real-world

### 2019-04-05 experiments.tex
Beginnings of edits to section 5

--- 2019-04-03 experiments.tex
@@
 The state of the art in flaky test detection is NonDex~\cite{nondex}.
-\Cref{sec:related} explains how NonDex works. In this section, we perform a qualitative comparison of the
-errors reported by NonDex and those reported by \TheDeterminismChecker. We ran all our
+\Cref{sec:related} explains how NonDex works. This section compares the
+errors reported by NonDex and \theDeterminismChecker. We ran all our
 experiments on a 2.3 GHz Intel Core i5 processor with a RAM configuration of 8 GB 2133 MHz.

### 2019-04-05 conclusion.tex

@@
 
-\todo{\TheDeterminismChecker found errors that the
+We designed a type system that helps programmers express determinism specifications
+in sequential programs. To the best of our knowledge, ours is the
+first compile time verification approach addressing the problem of nondeterminism.
+We implemented our type system in Java, applied it to real world 
+software and found it to be effective in finding previously unknown determinism bugs. 
+\TheDeterminismChecker found errors that the
 Randoop developers had missed, despite spending over 100 hours on the
-  problem of nondeterminism.}
+problem of nondeterminism. We compared our tool with existing state of the art flake test detectors
+and found that \TheDeterminismChecker found all the bugs arising due to nondeterminism
+in sequential programs that were used for evaluation in NonDex~\cite{nondex} and DeFlaker~\cite{deflaker}.

### 2019-04-05 experiments.tex
Capitalization

--- 2019-04-04 experiments.tex
@@
 The state of the art in flaky test detection is NonDex~\cite{nondex}.
-\Cref{sec:related} explains how NonDex works. In this section, we perform a qualitative comparison of the
-errors reported by NonDex and those reported by \TheDeterminismChecker. We ran all our
+\Cref{sec:related} explains how NonDex works. This section compares the
+errors reported by NonDex and \theDeterminismChecker. We ran all our
 experiments on a 2.3 GHz Intel Core i5 processor with a RAM configuration of 8 GB 2133 MHz.
 \TheDeterminismChecker doesn't show any noticeable performance change at compile time.
-\TheDeterminismChecker runs on any project that builds with maven or gradle and requires
+\TheDeterminismChecker runs on any project that builds with Maven or Gradle and requires
 Java version 8 or higher (support for annotations was added in Java 8).
@@
 We modified Randoop by deleting tests that were skipped by its buildfile,
-because the NonDex gradle plugin does not respect those settings. 
+because the NonDex Gradle plugin does not respect those settings.
 After that, NonDex ran without problems on Randoop.
@@
 While we would have like to run DeFlaker on Randoop for a fair comparison of the tools,
-we were unable to do so because Deflaker works with projects built with maven and Randoop
-uses gradle for its build system. We plan to convert Randoop's build system to maven and 
+we were unable to do so because Deflaker works with projects built with Maven and Randoop
+uses Gradle for its build system. We plan to convert Randoop's build system to Maven and
 perform this case study.
@@
 %Looked throught= about 10 commits, didn't find differential code coverage and therefore no flakes}
-%\todo{Deflaker requires maven build. Randoop is on gradle.}
+%\todo{Deflaker requires Maven build. Randoop is on Gradle.}
 

### 2019-04-05 design.tex
design minor edits

@@
 
-It is possible to store an expression of type \<Det> in a \<NonDet>
-collection, but any value retrieved from a \<NonDet> collection has
-type \<NonDet>.
+%It is possible to store an expression of type \<Det> in a \<NonDet>
+%collection, but any value retrieved from a \<NonDet> collection has
+%type \<NonDet>.
 
@@
 is deterministic.)
-The typical type fro the list access operation only applies if both
+The typical type for the list access operation only applies if both
 arguments are \<Det>,
@@
 This means that the only two instantiations are at 
-$\kappa = \<NonDet>$ and $\kappa = \<Det>$.  An invocation on a
+$\kappa = \<NonDet>$ and $\kappa = \<Det>$.  An invocation on an
 \<OrderNonDet> collection would choose the \<NonDet> instantiation and the
@@
 
-$\<listOfLists> : \ptype{OND List}{\ptype{Det List}{Det String}}$
+$\<listOfLists> : \ptype{OrderNonDet List}{\ptype{Det List}{Det String}}$
 
@@
 
-The core calculus requires a qualifier on every type.}
+\todo{The core calculus requires a qualifier on every type.}
 
@@
 \TheDeterminismChecker has additional type refinement rules for the sorting
-methods of \<List>, \<Collections>, and \<Arrays> classes
+methods of \<List>, \<Collections>, and \<Arrays> classes.
 \begin{figure}
@@
   Set}s.\todo{Explain why.}
- We added the following rule shown in \cref{fig-precision-rules} to our checker to improve precision.
+ We added the rule shown in \cref{fig-precision-rules} to our checker to improve precision.
 \begin{figure}

### 2019-04-05 experiments.tex
Continued work on case study and experiments

@@
 \Cref{sec:related} explains how NonDex works. This section compares the
-errors reported by NonDex and \theDeterminismChecker. We ran all our
-experiments on a 2.3 GHz Intel Core i5 processor with a RAM configuration of 8 GB 2133 MHz.
-\TheDeterminismChecker doesn't show any noticeable performance change at compile time.
-\TheDeterminismChecker runs on any project that builds with Maven or Gradle and requires
-Java version 8 or higher (support for annotations was added in Java 8).
+errors reported by NonDex and \theDeterminismChecker.
 
@@
 We ran NonDex on the same commit of Randoop as we used for the case study
-of \cref{sec:randoop-case-study}.  This version contains 3 nondeterminism bugs that
+of \cref{sec:randoop-case-study}.  This version contains all the nondeterminism bugs that
 \theDeterminismChecker found.
@@
 After that, NonDex ran without problems on Randoop.
-It completed successfully without reporting any flaky tests.
+NonDex reported no flaky tests.
+\todo{Revise the following text.
+  I renumbered the bugs without realizing that this section refers to them
+  by number, and I don't know which one these refer to any
+  more.
+  Please use cross-references or macros.  Also, this needs to be expanded
+  to account for the additional bugs.}
 We investigated the tests covering code that \theDeterminismChecker reported as nondeterministic.
@@
 is not modeled by NonDex.
-This indicates that the additional programmer effort required by \TheDeterminismChecker is justified
-as our tool discovers non-trivial issues and doesn't depend on code coverage or other runtime heuristics.
 
@@
 
-\Cref{sec:nondex-randoop} shows that \theDeterminismChecker finds errors that NonDex does not.
-This section determines whether NonDex finds errors that \theDeterminismChecker does not.
-Its authors ran NonDex on 195 open-source projects, and NonDex found flaky tests in
-21 of them. We defer the reader to~\cite{nondex} for a complete list of these projects
-and the details of commits hashes in which NonDex detected flaky tests.
-The authors also reported the sources of flakiness
-after manually inspecting these tests.
-
-We ran \TheDeterminismChecker on all these benchmarks with the objective
-of measuring the recall of our tool. We cloned all 21 repos and switched to the commit that had a flakiness issue as reported 
-in~\cite{nondex}.
-Of these projects, \<handlebars> and \<oryx> have compile time failures which prevent us from running \TheDeterminismChecker on them.
-For benchmarks \<visualee>, \<easy-batch>, and \<vraptor>, we couldn't find
-the sources of flakiness as reported in~\cite{nondex} anywhere in the respective repositories.
-\TheDeterminismChecker reported the nondeterministic source locations correctly in the rest of the projects.
-All these benchmarks and instructions to reproduce our results can be found at \href{https://github.com/t-rasmud/determinismCaseStudies}{https://github.com/t-rasmud/determinismCaseStudies}.
-
-All the flakiness that NonDex found was due to 7 methods in 3 classes. We enumerate all the methods corresponding to
-each class in \cref{fig:flaky-sources}.
 \begin{figure}
@@
     \end{tabular}
-\caption{Souces of flakiness in~\cite{nondex}.}
+\caption{Souces of flakiness known to NonDex~\cite{nondex}.}
 \label{fig:flaky-sources}
@@
 
-We provide samples of the annotated source code from these benchmarks and the errors as reported by \TheDeterminismChecker 
-in \cref{fig:nondex-source}.
+\Cref{sec:nondex-randoop} shows that \theDeterminismChecker finds errors that NonDex does not.
+This section determines whether NonDex finds errors that \theDeterminismChecker does not.
+Its authors ran NonDex on 195 open-source projects, and NonDex found flaky tests in
+21 of them~\cite{nondex}.
+The authors also reported the sources of flakiness
+after manually inspecting these tests. 
+The flakiness that NonDex found was due to 7 methods in 3 classes (\cref{fig:flaky-sources}).
+
+We tried to run \theDeterminismChecker on all these benchmarks, at the
+commit given in the NonDex paper.
+Some of the projects had moved, or did not build because their dependencies
+had moved or were no longer available.  We repaired all these issues and
+were able to compile all but two
+projects, handlebars and oryx.  Some of the remaining projects did not pass
+their tests, but that did not hinder us since \theDeterminismChecker works
+at compile time.
+
+For three of the projects, we could not find the flakiness reported in the
+NonDex paper.
+\todo{Important: give a sentence description of each one.}
+\todo{visualee}
+\todo{easy-batch}
+\todo{vraptor}.
+
+This left 16 projects.  We ran \theDeterminismChecker on the part of the
+project that the NonDex authors determined as flaky.  In every case,
+\theDeterminismChecker issued a warning on the nondeterministic code.  In
+other words, \theDeterminismChecker's recall was 100\%.
+
+All these benchmarks and instructions to reproduce our results can be found at \myurl{https://github.com/t-rasmud/determinismCaseStudies}.
+
 \begin{figure}
@@
         \begin{verbatim}
-    static public FieldAccess get (Class type) {
-        @Det ArrayList<@Det Field> fields = new ArrayList<Field>();
+    static public FieldAccess get(Class type) {
         ...
         while (nextClass != Object.class) {
-            @Det Field @Det[] declaredFields = nextClass.getDeclaredFields();
+            Field[] declaredFields = nextClass.getDeclaredFields();
             ...
@@
     }
-    [ERROR] FieldAccess.java:[102,87] 
-    incompatible types in assignment.
-    found   : @Det Field @OrderNonDet []
-    required: @Det Field @Det []
         \end{verbatim}
-        \caption{Annotated source code from reflectasm. \vspace{0.5cm}}
+        \caption{Nondeterministic source code from reflectasm.}
         \label{code-reflectasm}
@@
     protected SimpleDataEvent createNextEvent() {
-        for (@Det Entry<@Det String, @Det FieldType> entry : fields.entrySet()) {
+        for (Entry<String, FieldType> entry : fields.entrySet()) {
             ...
@@
     }
-    [ERROR] DataDictionaryEventSource.java:[73,78] 
-    incompatible types in enhanced for loop.
-    found   : @NonDet Entry<@Det String, @Det FieldType>
-    required: @Det Entry<@Det String, @Det FieldType>
         \end{verbatim}
@@
 \end{figure}
-As can be seen from these figures, annotating local arrays and collections
-has been sufficient to discover the sources of nondeterminism in most of the projects.
-The rest of the projects required annotating the return types of methods.
 
-For the benchmark \<commons-lang>, \cite{nondex} report multiple failing tests (14 flaky tests) due to calls
-to \<Class.getDeclaredFields>. Upon inspecting the source code, we find that all of invocations to
-\<Class.getDeclaredFields> are made from 5 locations. This means that annotating 5 lines of source code
-would have been sufficient to identify the nondeterminism. Having said that, we admit that there could be
-significant programmer effort involved in annotating the whole program. On the other hand,~\cite{nondex}
-state "we found that manually inspecting these failures was
+\Cref{fig:nondex-source} show samples of nondeterministic code from these
+benchmarks.
+We annotated the benchmarks based on 
+assumptions made downstream of the shown code.
+
+In reflectasm (\cref{code-reflectasm}), we changed the type of field
+\<declaredFields> from \<Field[]> to \<@Det Field @Det []>.  That type
+means a deterministic array of deterministic \<Field>, analogously to
+\codeid{@Det List<@Det Field>}.  Then, \theDeterminismChecker issued a
+warning at the assignment, because \<getDeclaredMethods> returns \<@Det
+Field @Det []>, which is an order-nondeterministic array of deterministic
+\<Field>s.
+
+The ActionGenerator code (\cref{code-actiongenerator}) is similar.  Other
+code assumes that \<entry> is deterministic, but annotating it as \<@Det>
+leads to a warning from \theDeterminismChecker that iterating over
+\<fields.entrySet()> (which is itself \<@OrderNonDet>) yields \<@NonDet>
+\<Entry> values..
+
+Nondex found 14 flaky tests in Apache Commans Lang due to calls
+to \<Class.getDeclaredFields>. 
+There are only 5 invocations of \<Class.getDeclaredFields>, so annotating 5 lines of source code
+would have been sufficient to identify all that nondeterminism.
+Having said that, we admit that there could be
+significant programmer effort involved in annotating the whole program. On
+the other hand, the NonDex authors
+state ``we found that manually inspecting these failures was
 rather challenging, and we leave it as future work to automate
-debugging test failures due to ADINS code". \TheDeterminismChecker reports source locations
-which makes it easier for the programmer to fix issues and is orthogonal to the approach taken by NonDex which
-reports failing tests.
+debugging test failures.''
+\TheDeterminismChecker reports source locations
+which makes it easier for the programmer to fix issues, and the annotation
+effort serves as valuable documentation and prevents regressions.
 
-\TheDeterminismChecker is also capable of detecting nondetermism in test cases. We provide a stub file
-for \<junit> in which we annotated the parameters of all the \<assert> methods as \<Det>.
-This will cause \TheDeterminismChecker to report any \<junit> test that passes nondeterministic arguments to 
-\<assert>s. The test case doesn't even have to be executed for out our tool to detect errors.
-It took one of our authors under an hour to create the stub file for junit. This gives us
-empirical evidence of the ease of extensibility of our tool.
+\TheDeterminismChecker is capable of detecting nondetermism in test
+cases, as some of these examples are.
+We wrote specifications for the
+JUnit library (in addition to the JDK and some other libraries).
+The specifications are the \<@Det> type qualifier on the formal parameters
+to the \<assert>* methods.
+\TheDeterminismChecker reports any \<junit> test that passes nondeterministic arguments to
+\<assert>*, even if the test case is not executed.
 
@@
 Deflaker~\cite{deflaker}, like NonDex, reports tests that could be flaky. 
-While we would have like to run DeFlaker on Randoop for a fair comparison of the tools,
-we were unable to do so because Deflaker works with projects built with Maven and Randoop
+We were unable to run DeFlaker on Randoop, because Deflaker works with projects built with Maven and Randoop
 uses Gradle for its build system. We plan to convert Randoop's build system to Maven and

### 2019-04-05 randoop-case-study.tex
Continued work on case study and experiments

@@
 \section{Case study}\label{sec:randoop-case-study}
+
+%% This is not relevant in this section of the paper, where no times are given.
+% We ran all our
+% experiments on a 2.3 GHz Intel Core i5 processor with 8 GB of RAM.
 
@@
 
+% run: ./gradlew clean compileJava;./gradlew clean compileJava;./gradlew clean compileJava;./gradlew clean compileJava;./gradlew clean compileJava
+% then read the times reported by Gradle.
+% javac 5 trials:   31, 19, 18, 18, 18.
+% \theDeterminismChecker 5 trials: 45s 39s 32s 31s 31s
+\<javac> takes 18 seconds to compile all files of Randoop.
+While also running \theDeterminismChecker as a compiler plugin, \<javac> takes 32 seconds to verify determinism and compile all files of Randoop.
+% godwit.  Info comes from `cat /proc/cpuinfo` and from `top`.
+These numbers are the median of 5 trials on an 8-core Intel i7-3770 CPU running at 3.40GHz with 
+32GB of memory.
+
 \todo{Report the number of annotations required, and say a few words about
@@
   about use of environment variables.  Lines starting with ``\<->'' were
-  removed and those starting with ``\<+>'' were added.}
+  removed and those starting with ``\<+>'' were added.
+  \TheDeterminismChecker verified all other uses of system and Java properties.}
 \label{fig:randoop-bug-getenv}

### 2019-04-05 experiments.tex
Edits in DeFlaker and threats to validity

@@
 Deflaker~\cite{deflaker}, like NonDex, reports tests that could be flaky. 
-We were unable to run DeFlaker on Randoop, because Deflaker works with projects built with Maven and Randoop
+We were unable to run DeFlaker on Randoop, because Deflaker works with
+projects built with Maven, but Randoop
 uses Gradle for its build system. We plan to convert Randoop's build system to Maven and
@@
 the paper was written. The authors reported 19 of these tests, out of which 7 were addressed by the maintainers of
-those projects. Since we can only verify if \TheDeterminismChecker finds the same bugs by looking at the source code
-locations, we ran our tool on the benchmarks where the reported bug was fixed. We again defer the reader to~\cite{deflaker}
-for a complete list of the open source projects and their details.
+those projects~\cite{deflaker}.
+We ran \theDeterminismChecker on the part of each of these
+codebases where the reported bug was fixed, as in \cref{sec:nondex-benchmarks}.
+% The DeFlaker authors graciously fixed problems we discovered while using
+% their tool.
 
-Two of the seven flaky tests (in \<achilles> and \<jackrabbit-oak>) that DeFlaker reported were caused by a race condition. \TheDeterminismChecker
-is designed for sequential programs and hence cannot detect this category of bugs.  
-We couldn't find the link to one of the issues in the issue tracker of \<jackrabbit-oak>
-We ran into build issues
-for \<tigglz> and \<nutz> which had two and one flakiness issues respectively. 
+Two of the seven flaky tests (in \<achilles> and \<jackrabbit-oak>)
+were caused by a race condition, which \theDeterminismChecker
+cannot detect.  (This is a strength of DeFlaker over \theDeterminismChecker.)
+We couldn't find the link to one of the issues in the issue tracker of \<jackrabbit-oak>.
+We were unable to build \<tigglz> and \<nutz> which had two and one flakiness issues respectively.
 \TheDeterminismChecker was able to find the source of flakiness in \<checkstyle>. This bug was in a test
-case that deterministically accessed an \<OrderNonDet> array returned by \<Class.getDeclaredConstructors()>.
+case that treated an array returned by \<Class.getDeclaredConstructors()>
+as deterministic.

### 2019-04-05 threats.tex
taint citations
Edits in DeFlaker and threats to validity

@@
 Suppose that an expression used in a conditional is tainted.
-A standard approach\todo{Add citations} for a dynamic taint analysis is to taint any value that may
+A standard approach
+\cite{Kang2011DTADT,Arzt:2014:FPC:2594291.2594299}
+%\todo{Add citations}
+ for a dynamic taint analysis is to taint any value that may
 be assigned in either branch of the conditional statement.  This tends to rapidly lead to most of
@@
 exception that can be thrown from arbitrary code, including libraries.
-Enabling this check led to many unhelpful false positive warnings.
+Enabling this check led to many unhelpful false positive warnings, which
+are only relevant if unverified libraries throw nondeterministic exceptions.
 
@@
 libraries and native code with incorrect specifications might introduce
-nondeterminism that it does not check.
+nondeterminism even if \theDeterminismChecker issues no warnings.
 
@@
 programs or to programs written in other programming lanugages.  We
-mitigated this problem by showing that \theDeterminismChecker finds most or
-all of the sources of nondeterminism identified by other tools.
+mitigated this problem by showing that \theDeterminismChecker finds many
+of the sources of nondeterminism identified by other tools.

### 2019-04-05 randoop-case-study.tex
macros

@@
 \begin{itemize}
-\item \textbf{Bug 1:}
+\item \bugOne:
   A use of \<HashSet> that the developers changed to \<LinkedHashSet>\commit{c975a9f7}.
@@
 
-\item \textbf{Bug 2:}
+\item \bugTwo:
 Dependency on the CLASSPATH environment variable in preference to the
@@
 
-\item \textbf{Bug 3:}
+\item \bugThree:
 Diagnostic output that printed a hash code for brevity, but should have had
@@
 
-\item \textbf{Bug 4:}
+\item \bugFour:
 Diagnostic output that printed a timestamp, but should have obeyed an

### 2019-04-05 randoop-case-study.tex
Expanding case study

@@
 
-%% This is not relevant in this section of the paper, where no times are given.
-% We ran all our
-% experiments on a 2.3 GHz Intel Core i5 processor with 8 GB of RAM.
+To evaluate the usability of \theDeterminismChecker,
+we applied it to the Randoop test
+generation tool~\ifanonymous\citepalias{randoop-tool}\else\cite{PachecoLEB2007}\fi.
+We also wrote specifications for libraries it uses, such as the JDK, JUnit,
+and others.
+Like \theDeterminismCheckerImplementation,
+all the case study materials are publicly available at \myurl{\todo{url for
+    case study}} for reproducibility.
 
-As a case study, we applied \theDeterminismChecker to the Randoop test
-generation tool~\ifanonymous\citepalias{randoop-tool}\else\cite{PachecoLEB2007}\fi.
+\subsection{Subject program}
+
 We chose Randoop because 
@@
 it is actively maintained,
-and its developers have struggled with nondeterminism (personal
-communication, 2019).
+and its developers have struggled with nondeterminism~\cite{randoop-issue-tracker,randoop-mailing-lists}.
 
 Randoop is intended to be deterministic, when invoked on a deterministic
-program.\footnote{Users of Randoop can pass in a different seed in order to
+program~\cite{randoop-manual}.\footnote{Users of Randoop can pass in a different seed in order to
   obtain a different deterministic output.  Randoop has command-line
@@
   nondeterministic behavior.}
-
 However, Randoop was not deterministic.  This caused the developers
@@
 reproducing test failures during development, and
-understanding the effect of changes to Randoop (by comparing executions of two
-similar variants of Randoop).
+understanding the effect of changes to Randoop by comparing executions of two
+similar variants of Randoop.
 
@@
 hash code which may vary from run to run.
-They used libraries that make hash codes deterministic, by giving each
-object of a type a unique ID that counts up from 1, rather than using a
+They used a library that makes hash codes deterministic, by giving each
+object of a type a unique ID that counts up from 1 rather than using a
 memory address as \<Object.hashCode> does.
-They wrote preprocess output and logs to make nondeterministic ones easier
+They wrote specialized tools to preprocess output and logs to make them easier
 to compare, such as by removing or canonicalizing hash codes, dates, and
-other nondeterministing output.
+other nondeterministing output (commits e15f9155--b473fd14).
 Each of these efforts was helpful, but they were not enough to solve the problem.
@@
  and eliminate it.
-The most common causes were \<toString()> routines and iteration order of sets and maps.
 Some of the nondeterminism was in libraries, such as the JDK\@.
-
-The developers chose not to make every \<Map> a \<LinkedHashMap>, because that
+The most common causes were \<toString> routines and iteration order of sets and maps.
+The most common fixes were to change the implementation of \<toString>
+and to use \<LinkedHashSet> and \<LinkedHashMap> or to sort collections
+before iterating over them.
+The developers did not make every \<Set> and \<Map> a \<LinkedHashSet> or \<LinkedHashMap>, because that
 would have increased memory and CPU costs.  They chose not to make every
@@
 
-\todo{When we discuss annotation effort, compare to this.
-  The annotation effort is
-  high, but it's no more than the hundreds of hours that the Randoop
-  developers spent, with less success.}
-
 That coding sprint did not find all the problems.
 The developers debugged and fixed 5 additional determinism defects over the
-next 12 months, using a similar methodology.
+next 12 months, using a similar methodology
+(commits c15ccbf2, 44bdeebd, 5ff5b4c4, 22eda87f, and b473fd14).
+
 % [2413 changed lines in the diffs.
 % (This number slightly overstates the number of code changes made, because sometimes changing code would result in code formatting changes, and some changes in those commits are unrelated to determinism, such as documentation improvements.)]
-We analyzed a version of Randoop from February 2019 (commit 0a8e63fb), over 6 months after the most recent determinism bug fix.
+We analyzed a version of Randoop from late February 2019 (commit 0a8e63fb), over 6 months after the most recent determinism bug fix.
+
+\subsection{Methodology}
+
+In brief, we wrote type qualifiers in the Randoop source code to express its
+determinism specification
+We then ran
+\theDeterminismChecker.  Each warning indicated a mismatch between the
+specification and the implementation, which we addressed by changing our
+specification or reporting a bug in the implementation.
 
@@
 We did not annotate Randoop's test suite.
+
+We annotated one package at a time, starting with the most central packges
+that are most depended upon.  Within a package, we followed a similar
+strategy, annotating supertypes first.  If the classes and methods had been
+documented with respect to their determinism, then our annotation effort
+would have been easy, just converting English into type qualifiers.
+Unfortunately, this was not the case: we reverse-engineered each
+specification, largely from the methods it calls.  Sometimes after we
+annotated a subclass, we had to go back and change existing annotation on
+the supertype or elsewhere.
+After resolving most type-checking warnings by adding or changing
+annotations, we manually examined each remaining one to determine whether
+it was a bug or a false positive warning.
 
@@
 
-\todo{Report the number of annotations required, and say a few words about
-  the process and the effort required.}
+\subsection{Results}
 
-\TheDeterminismChecker found three previously-unknown nondeterminism bugs in Randoop.
+\TheDeterminismChecker found \numRandoopBugs previously-unknown nondeterminism bugs in Randoop.
 The Randoop developers accepted our bug reports and committed fixes to the repository. A summary
-of these bugs follows:
+of these bugs follows, according to the Randoop developers' categorization:
 
-\begin{itemize}
-\item \bugOne:
-  A use of \<HashSet> that the developers changed to \<LinkedHashSet>\commit{c975a9f7}.
-The old code could cause a problem if a type variable's lower or upper
+\begin{description}
+\item[Severe issues]
+  \strut
+\begin{description}
+\item[\bugHashSet]
+One use of \<HashSet> could cause a problem if a type variable's lower or upper
 bound has a type parameter that the type variable itself does not have.
 This situation does occur, even in Randoop's test suite.
+The developers fixed this by changing a \<HashSet> to \<LinkedHashSet>
+(commit c975a9f7, shown in \cref{fig:randoop-bug-hashset}).
+\TheDeterminismChecker confirmed that 
+25 other uses of \<new HashSet> were acceptable, as were 15 uses of \<new HashMap>.
 
-\item \bugTwo:
-Dependency on the CLASSPATH environment variable in preference to the
-classpath passed on the command line\commit{330e3c56}.
-This could have led to incorrect behavior if a user set the environment variable.
-The developers changed Randoop to not read the environment variable.
+\item[\bugClasspath]
+Randoop used the CLASSPATH environment variable in preference to the
+classpath passed on the command line.
+This could have led to incorrect behavior, both in Randoop's test suite and in the field,
+if a user set the environment variable.
+The developers fixed it by changing Randoop to not read the environment variable
+(commit 330e3c56, shown in \cref{fig:randoop-bug-getenv}).
+\TheDeterminismChecker verified that all other uses of system and Java
+properties did not lead to nondeterministic behavior.
+\end{description}
 
-\item \bugThree:
-Diagnostic output that printed a hash code for brevity, but should have had
-deterministic output\commit{661a4970}.  By contrast to the first two bugs,
-this one is not serious.
+\item[Minor issues]
+\strut
+\begin{description}
 
-\item \bugFour:
-Diagnostic output that printed a timestamp, but should have obeyed an
-option about whether to do so\commit{a460df97}.  By contrast to the first
-two bugs, this one is not serious.
+\item \bugHashcodeOutput:
+  Diagnostic output printed a hash code for brevity.
+  The developers changed it to have deterministic output (commit 661a4970).
+  Although the Randoop developers classified this issue as minor, it is
+  very similar to ones they fixed during their two-week coding sprint.
 
-\end{itemize}
+\item \bugTimestampOutput:
+  Diagnostic output printed a timestamp.
+  The developers fixed it by making it obeyed an
+  option about whether to print timestamps (commit a460df97).
+  The developers may not have noticed this because their log-postprocessing
+  tools removed them from the log when doing comparisons.
+
+\end{description}
+
+\end{description}
 
@@
 \end{Verbatim}
-
 \caption{Fixes made by the Randoop developers in response to our bug report
@@
 
-We annotated one package at a time, starting with the most central packges
-that are most depended upon.  Within a package, we followed a similar
-strategy, annotating supertypes first.  If the classes and methods had been
-documented with respect to their determinism, then our annotation effort
-would have been easy, just converting English into type qualifiers.
-Unfortunately, this was not the case: we reverse-engineered each
-specification, largely from the methods it calls.  Sometimes after we
-annotated a subclass, we had to go back and change existing annotation on
-the supertype or elsewhere.
-After resolving most type-checking warnings by adding or changing
-annotations, we manually examined each remaining one to determine whether
-it was a bug or a false positive warning.
+\subsection{False positive warnings}
+
+Java 8 lambdas.
+
+\subsection{Annotation effort}
+
+\todo{Report the number of annotations required, and say a few words about
+  the process and the effort required.}
 
@@
 needed otherwise.
+
+\todo{When we discuss annotation effort, compare to the manual work.
+  The annotation effort is
+  high, but it's no more than the hundreds of hours that the Randoop
+  developers spent, with less success.}

### 2019-04-05 randoop-case-study.tex
More edits in case study section

@@
 
-\subsection{Subject program}
+\subsection{Subject program}\label{sec:subject-program}
 
@@
 that are most depended upon.  Within a package, we followed a similar
-strategy, annotating supertypes first.  If the classes and methods had been
-documented with respect to their determinism, then our annotation effort
+strategy, annotating supertypes first.  If the determinism of classes and methods had been
+documented, then our annotation effort
 would have been easy, just converting English into type qualifiers.
-Unfortunately, this was not the case: we reverse-engineered each
-specification, largely from the methods it calls.  Sometimes after we
+Unfortunately, this was not the case: we had to reverse-engineer each
+specification, largely from the methods it calls.  If a method $m$ calls a
+method that we had previously annotated as nondeterministic, we annotated
+$m$ as nondeterministic.
+Sometimes after we
 annotated a subclass, we had to go back and change existing annotation on
-the supertype or elsewhere.
+the supertype or elsewhere because the subclass implementation was
+inconsistent with the specification we had guessed for the supertype.
 After resolving most type-checking warnings by adding or changing
@@
 depended on subtle, undocumented invariants about Randoop that we had not
-been able to reverse-engineer.
+been able to reverse-engineer on our own.
 
@@
 
-Java 8 lambdas.
+\todo{Java 8 lambdas could help.}
 
@@
 
+\todo{Easier for someone familiar with the codebase.  Easier on new code.}
+
+It's higher than we would prefer.
+
 The large number of \<@Det> annotations has two main reasons.
@@
 boolean expressions used in conditionals to be deterministic.  (This is an
-option in our implementation, which increases soundness.)  This forced
-their clients, and much of the Randoop codebase, to be deterministic.
+option in \theDetminismCheckerImplementation, which increases soundness;
+see \cref{sec:threats}.)  This forced
+their clients --- transitively, much of the Randoop codebase --- to be
+annotated as deterministic.
 These annotations are not incorrect, but they are stricter than necessary
-because some types could otherwise be @PolyDet, which is the default for
+because some method types could otherwise be qualifier-polymorphic.
+This would make them usable in more contexts.  It would also reduce the
+number of necessary annotations, since qualifier-polymorphism (the
+\<@PolyDet> annotation) is the default for
 formal parameter and return types.  We plan to re-do the annotation effort
@@
 
-The second reason is a limitation in \theDeterminismCheckerImplementation.
-\TheDeterminismChecker does not support qualifier polymorphism over
-classes.  This means that fields must be deterministic.  Our annotations in
+The second reason is a limitation in \theDeterminismCheckerImplementation,
+which is inherited from the Checker Framework~\cite{checker-framework-tool} upon which
+\theDeterminismChecker is built.
+The Checker Framework does not support qualifier polymorphism over
+classes.  This means that fields must be annotated as deterministic.  Our annotations in
 Randoop use qualifier polymorphism extensively on methods but cannot use it
-on classes, and this leads to many more Det annotations than would be
+on classes, and this leads to many more \<@Det> annotations than would be
 needed otherwise.
 
-\todo{When we discuss annotation effort, compare to the manual work.
-  The annotation effort is
-  high, but it's no more than the hundreds of hours that the Randoop
-  developers spent, with less success.}
+Even though the number of annotations is higher for our prototype
+implementation than we would prefer, it still compares favorably to the
+extensive effort by the Randoop developers (\cref{sec:subject-program}),
+especially given that that \theDeterminismChecker found issues that the
+developers had not.

### 2019-04-05 randoop-case-study.tex
Fix LaTeX error: typo in macro name

@@
 boolean expressions used in conditionals to be deterministic.  (This is an
-option in \theDetminismCheckerImplementation, which increases soundness;
+option in \theDeterminismCheckerImplementation, which increases soundness;
 see \cref{sec:threats}.)  This forced

### 2019-04-05 experiments.tex
randoop bugs

@@
 NonDex reported no flaky tests.
-\todo{Revise the following text.
-  I renumbered the bugs without realizing that this section refers to them
-  by number, and I don't know which one these refer to any
-  more.
-  Please use cross-references or macros.  Also, this needs to be expanded
-  to account for the additional bugs.}
+%\todo{Revise the following text.
+%  I renumbered the bugs without realizing that this section refers to them
+%  by number, and I don't know which one these refer to any
+%  more.
+%  Please use cross-references or macros.  Also, this needs to be expanded
+%  to account for the additional bugs.}
 We investigated the tests covering code that \theDeterminismChecker reported as nondeterministic.
-Bug 1 and Bug 2 have corresponding test cases whereas Bug 3 does not.
-As NonDex is a \textit{flaky test} detection tool, Bug 3 is outside of its scope.
-The reason for nondeterminism in Bug 2 was a call to the \<System.getProperty()> method which
+\bugHashSet and \bugClasspath have corresponding test cases whereas \bugTimestampOutput and 
+\bugTimestampOutput do not.
+As NonDex is a \textit{flaky test} detection tool, detecting \bugTimestampOutput and \bugTimestampOutput 
+is outside of its scope.
+The reason for nondeterminism in \bugClasspath was a call to the \<System.getProperty()> method which
 is not modeled by NonDex.

### 2019-04-05 design.tex
minor edits
macros
randoop bugs
Copy text from manual

@@
     \todo{add line break}
-    $\infer[\rulename{set precision}]{\Gamma \vdash \|x.equals(y)| : \<Det>}{\Gamma \vdash \|x| : \<OrderNonDet> \ \|Set|\angles{ \kappa_{x} \ \beta_{x}}, \|y| : \<OrderNonDet> \ \|Set|\angles{ \kappa_{y} \ \beta_{y}} & \kappa_{x} \ \beta_{x} \nsqsubseteq \<OrderNonDet> \ \<List> & \kappa_{y} \ \beta_{y} \nsqsubseteq \<OrderNonDet> \ \<List> }$
-
-    \bigskip
-
-    $\infer[\rulename{map precision}]{\Gamma \vdash \|x.get(y)| : \<Det>}{\Gamma \vdash \|x : \<OrderNonDet> \ \<Map>\angles{ \<Det> \ \beta_k, \kappa_v\ \beta_v }, y : \<Det> \ \beta|}$
+    $\infer[\rulename{set precision}]{\Gamma \vdash \|x.equals(y)| : \<Det> \ \|boolean|}{\Gamma \vdash \|x| : \<OrderNonDet> \ \|Set|\angles{ \kappa_{x} \ \beta_{x}}, \|y| : \<OrderNonDet> \ \|Set|\angles{ \kappa_{y} \ \beta_{y}} & \kappa_{x} \ \beta_{x} \nsqsubseteq \<OrderNonDet> \ \<List> & \kappa_{y} \ \beta_{y} \nsqsubseteq \<OrderNonDet> \ \<List> }$
+
+    \bigskip
+
+    $\infer[\rulename{map precision}]{\Gamma \vdash \|x.get(y)| : \<Det> \ \beta_v}{\Gamma \vdash \|x : \<OrderNonDet> \ \<Map>\angles{ \<Det> \ \beta_k, \kappa_v\ \beta_v }, y : \<Det> \ \beta|}$
 
     \caption{Special rules for set equality and map lookup.}
-\todo{In the \rulename{map precision} rule, should the \<Det> in the
-  conclusion be $\beta_v$?}
+%\todo{In the \rulename{map precision} rule, should the \<Det> in the
+%  conclusion be $\beta_v$?}
     \label{fig-precision-rules}
 \end{figure}
+
+\todo{
+  In the JDK, the \<equals()> method of \<Set> is annotated as
+
+@PolyDet("up") boolean equals(@PolyDet Set<E> this, @PolyDet Object o);
+
+While this is sound, it is imprecise.
+For example, consider the code below:
+
+    @OrderNonDet Set<@Det Integer> a; \\
+    @OrderNonDet Set<@Det Integer> b; \\
+    a.equals(b);
+
+Here, the return type of \<equals()> gets the annotation \<@NonDet>.
+But it would be more precise to annotate it as \<@Det>.
+To achieve this precision, the checker annotates the return type
+of \<equals()> called on a receiver of type \<Set> as \<@Det> if the
+following conditions are true:
+\begin{itemize}
+    \item the receiver does not have {@code List} or its subtype as a type parameter,
+    \item the argument to equals() is also an {@code @OrderNonDet Set}, and
+    \item the argument to equals() also does not have {@code List} or its subtype as a
+    type parameter
+\end{itemize}
+}
+
 According to this rule, if the \<Set> receiver has type \codeid{@OrderNonDet Set<@Det List>} and the argument to the \codeid{equals()} method is

### 2019-04-05 java-design.tex
Continued work on case study and experiments
Tweak
Copy text from manual

@@
 \section{Determinism types for Java}\label{sec:java-types}
+
+We have implemented our type system for Java, in a tool named \theDeterminismChecker.
+
+\TheDeterminismChecker requires Java version 8.
 
@@
 
+\todo{
+The default annotation for most unannotated types is \<@Det>.
+
+\begin{itemize}
+\item
+  Formal parameter and return types default to \<@PolyDet>.
+\item
+  If a formal parameter/return type is an array, its component type defaults to
+  \<@PolyDet>.
+\item
+  As an exception, the return type of a method with no unannotated or
+  \<@PolyDet> formal parameters (including the receiver) defaults to
+  \<@Det>.
+\end{itemize}
+
+These choices permit all possible arguments to an unannotated method.
+}
+
 \subsection{Polymorphism}\label{java-polymorphism}
@@
 \end{itemize}
+
+\todo{Explain @PolyDet("up"), "down", "use".}
 
@@
 treated as deterministic.
-The inputs to a program are also treated as \<Det>.
+The inputs to a program are also treated as \<Det>.  That is, the parameter
+type of the \<main> method is implicitly annotated as
+\<@Det String @Det []>, a deterministic array of deterministic strings.

### 2019-04-05 experiments.tex
remove todo
expt todos

@@
 \todo{Important: give a sentence description of each one.}
-\todo{visualee}
-\todo{easy-batch}
-\todo{vraptor}.
+\todo{visualee : Class#getDeclaredMethods}
+\todo{easy-batch: Class#getDeclaredFields}
+\todo{vraptor : Class#getDeclaredFields}.
 
@@
         \end{verbatim}
-        \caption{Nondeterministic source code from reflectasm.}
+        \caption{Nondeterministic source code from reflectasm.\vspace{0.5cm}}
         \label{code-reflectasm}
@@
 
-\Cref{fig:nondex-source} show samples of nondeterministic code from these
+\Cref{fig:nondex-source} shows samples of nondeterministic code from these
 benchmarks.
@@
 
-Nondex found 14 flaky tests in Apache Commans Lang due to calls
+Nondex found 14 flaky tests in Apache Commons Lang due to calls
 to \<Class.getDeclaredFields>. 
@@
 
-\todo{Link to det checker repo?}
-\todo{If there is time, we could do a second case study, on a program that
-  NonDex did not find nondeterminism in.  Our hope would be to find
-  nondeterminism that NonDex did not.  Choose a small project that is
-  maintained (it has recent commits), so we can communicate with the authors.}
+%\todo{Link to det checker repo?}
+%\todo{If there is time, we could do a second case study, on a program that
+%  NonDex did not find nondeterminism in.  Our hope would be to find
+%  nondeterminism that NonDex did not.  Choose a small project that is
+%  maintained (it has recent commits), so we can communicate with the authors.}
 

### 2019-04-05 java-design.tex
Intro text about implementation

@@
 We have implemented our type system for Java, in a tool named \theDeterminismChecker.
+The implementation consists of 1200 lines of Java
+code, plus 1800 lines of tests (all line measurements are non-comment,
+non-blank lines), a manual, examples, etc.  It is publicly available at
+\myurl{URL for \theDeterminismChecker}.
 
@@
 
-In Java, a type qualifier is written as a type annotation (which has a leading \<@>), as in
-\<@Positive int>.
+\subsection{Determinism type qualifiers}
 
-Our implementation
-handles array types, and any class (including user-defined classes) that
+A type qualifier is written in Java source code as a type annotation.
+A type annotation has a leading ``\<@>'' and is written immediately before a
+Java basetype, as in \<@Positive int> or \codeid{@NonEmpty List<@NonNull String>}.
+
+\TheDeterminismChecker supports the type qualifiers \<@NonDet>,
+\<@OrderNonDet>, and \<@Det> (and others described below).
+
+Our implementation treats arrays analogously to collections:
+the semantics and rules are similar even though the syntax varies.
+For simplicity this
+paper uses the term ``collection'' and the type \<Collection> to represent
+collections and arrays.
+Collections include any class (including user-defined classes) that
 implements the \<Iterable> or \<Iterator> interfaces; this includes all
-Java collections such as \<List>s and \<Set>s.  For simplicity, and because
-the semantics and rules are similar even though the syntax varies, this
-paper uses the term ``collection'' and the type \<Collection> to represent
-all these types.
-
-(In Java,
-\<HashMap> cannot be deterministic, but \<TreeMap> and \<LinkedHashMap>
-can.\todo{forward reference?})
+Java collections such as \<List>s and \<Set>s.
 

### 2019-04-05 experiments.tex
Fix LaTeX error
deflaker togglz

@@
 For three of the projects, we could not find the flakiness reported in the
-NonDex paper.
-\todo{Important: give a sentence description of each one.}
-\todo{visualee : Class#getDeclaredMethods}
-\todo{easy-batch: Class#getDeclaredFields}
-\todo{vraptor : Class#getDeclaredFields}.
+NonDex paper. The reported root cause of flakiness in \<easy-batch> and
+\<vraptor> was a call to \<Class.getDeclaredFields>. We couldn't find a call
+to this method in any of the source files of these two repositories.
+Similarly, the flakiness in \<visualee> was attributed to an invocation of
+\<Class.getDeclaredMethods> which we didn't find in the source code.
+%\todo{Important: give a sentence description of each one.}
 
@@
 \subsection{\TheDeterminismChecker on DeFlaker benchmarks}\label{sec:deflaker-benchmarks}
-DeFlaker found 87 previously unknown flaky tests in 93 projects that were being actively developed at the the time
+DeFlaker found 87 previously unknown flaky tests in 93 projects that were being actively developed at the time
 the paper was written. The authors reported 19 of these tests, out of which 7 were addressed by the maintainers of
@@
 
-Two of the seven flaky tests (in \<achilles> and \<jackrabbit-oak>)
+Four of the seven flaky tests (two in \<achilles>, one each in \<jackrabbit-oak> and \<togglz>)
 were caused by a race condition, which \theDeterminismChecker
 cannot detect.  (This is a strength of DeFlaker over \theDeterminismChecker.)
-We couldn't find the link to one of the issues in the issue tracker of \<jackrabbit-oak>.
-We were unable to build \<tigglz> and \<nutz> which had two and one flakiness issues respectively.
+We were unable to build \<togglz> and \<nutz> which had one flakiness issue each.
+However, we extracted the source code causing the flakiness in \<togglz> into a test case 
+after looking at the issue on Github. This bug was caused by a copy method that iterated over
+an \<OrderNonDet Set> and expected it to be deterministic. \TheDeterminismChecker correctly flagged
+an error in the loop that iterated over the \<Set>.
 \TheDeterminismChecker was able to find the source of flakiness in \<checkstyle>. This bug was in a test

### 2019-04-05 java-design.tex

@@
 collections and arrays.
-Collections include any class (including user-defined classes) that
+Collections include any type (including user-defined classes) that
 implements the \<Iterable> or \<Iterator> interfaces; this includes all
 Java collections such as \<List>s and \<Set>s.
+Our implementation also supports any type that implements the \<Map>
+interface.
 
@@
 
-We noticed the following properties with respect to determinism regarding the most widely used \<Map> implementations:
+The most widely used \<Map> implementations have the following properties:
 \begin{itemize}
-    \item \<HashMap> can never guarantee fully deterministic iteration over its entries. A \codeid{Det HashMap} does not exist.
-    \item \<LinkedHashMap> can have any of \<Det>, \<OrderNonDet>, or \<NonDet> types. Iterating over a \<LinkedHashMap> returns
-    its entries in the order of their insertion. An \<OrderNonDet LinkedHashMap> can be created by calling its constructor with a \<HashMap>.
-    \item \<TreeMap> can either be \<Det> or \<NonDet>. An \<OrderNonDet TreeMap> doesn't exist because the entries are always ordered.
+    \item \<HashMap> is implemented in terms of a hash table, which never
+      guarantees deterministic iteration over its entries. A \codeid{@Det
+        HashMap} does not exist.
+    \item \<LinkedHashMap>, like \<List>, can have any of the \<@NonDet>,
+      \<@OrderNonDet>, or \<@Det> type qualifiers. Iterating over a
+      \<LinkedHashMap> returns
+    its entries in the order of their insertion. An \<OrderNonDet
+    LinkedHashMap> can be created by calling its constructor with a
+    \<@OrderNonDet HashMap>.
+    \item \<TreeMap> can either be \<@Det> or \<@NonDet>. An \<@OrderNonDet
+      TreeMap> doesn't exist because the entries are always ordered.
 \end{itemize}
 
-We prohibit the creation of a \codeid{@Det HashMap} or an \codeid{@OrderNonDet TreeMap} in our determinism checker.
-\Cref{fig-creation-rules} formalizes the type wellformedness rules for maps and sets. Similar constraints apply to \<HashSet>s and \<TreeSet>s which are
-also expressed by the last two rules of \cref{fig-creation-rules}.
+\TheDeterminismChecker prohibits the creation of a \codeid{@Det HashMap} or an \codeid{@OrderNonDet TreeMap}.
+\Cref{fig-creation-rules} formalizes the type well-formedness rules for
+Java \<Map>s and \<Set>s.
 
 \begin{figure}
-    $\infer[\rulename{valid hashmap}]{\vdash: \|\kappa\ HashMap()|}{\kappa \in \{ \<OrderNonDet>, \<NonDet> \}}$
+    $\infer[\rulename{valid hashmap}]{\vdash: \|\kappa\ \<HashMap>|}{\kappa \in \{ \<OrderNonDet>, \<NonDet> \}}$
 
@@
 
-    $\infer[\rulename{valid treemap}]{\vdash: \|\kappa\ TreeMap()|}{\kappa \in \{ \<Det>, \<NonDet> \}}$
+    $\infer[\rulename{valid treemap}]{\vdash: \|\kappa\ \<TreeMap>|}{\kappa \in \{ \<Det>, \<NonDet> \}}$
 
@@
 
-    $\infer[\rulename{valid hashset}]{\vdash: \|\kappa\ HashSet()|}{\kappa \in \{ \<OrderNonDet>, \<NonDet> \}}$
+    $\infer[\rulename{valid hashset}]{\vdash: \|\kappa\ \<HashSet>|}{\kappa \in \{ \<OrderNonDet>, \<NonDet> \}}$
 
@@
 
-    $\infer[\rulename{valid treeset}]{\vdash: \|\kappa\ TreeSet()|}{\kappa \in \{ \<Det>, \<NonDet> \}}$
-    \caption{\<Map> and \<Set> creation rules}
-    \todo{Are these type well-formedness rules?}
-    \todo{What about \<new TreeSet(myOrderedCollection<Det ...>)> which yields a
-        deterministic set?}
+    $\infer[\rulename{valid treeset}]{\vdash: \|\kappa\ \<TreeSet>|}{\kappa \in \{ \<Det>, \<NonDet> \}}$
+    \caption{Type well-formedness rules for subclasses of \<Map> and \<Set>.}
     \label{fig-creation-rules}
@@
 
-In the JDK, \<LinkedHashSet> is implemented as a subclass of \<HashMap>.
+The JDK implements \<LinkedHashSet> as a subclass of \<HashMap>\todo{should
+  this be HashSet?  It's not true as written.}
+in order to inherit parts of \<HashMap>'s implementation.
 \todo{Talk about this design choice. Josh Bloch's book: "If the code surrounding the first declaration depended on LinkedHashSet's ordering policy,
     then it would be incorrect to substitute HashSet for LinkedHashSet in the declaration, because HashSet makes no guarantee concerning iteration order."} 
+\todo{The above is not clear.  What are the consequences for our
+  implementation.  What goes wrong or is hard to annotate or is surprising?
+  I'm not sure what is the point of this paragraph.}
 

### 2019-04-05 experiments.tex
deflaker nutz

@@
 We were unable to build \<togglz> and \<nutz> which had one flakiness issue each.
-However, we extracted the source code causing the flakiness in \<togglz> into a test case 
-after looking at the issue on Github. This bug was caused by a copy method that iterated over
+However, we extracted the source code causing the flakiness in these repositories into test cases
+after looking at the corresponding issues on Github. The bug in \<togglz> was caused by a copy method that iterated over
 an \<OrderNonDet Set> and expected it to be deterministic. \TheDeterminismChecker correctly flagged
 an error in the loop that iterated over the \<Set>.
+The flakiness in \<nutz> was a result of printing a response received over http.
+Since network operations are nondeterministic, we annotated the method in \<nutz> that
+returns this response as \<NonDet> which lead \TheDeterminismChecker to report an error at the print statement. 
 \TheDeterminismChecker was able to find the source of flakiness in \<checkstyle>. This bug was in a test

### 2019-04-05 java-design.tex
Edits in defaulting section

@@
 code, plus 1800 lines of tests (all line measurements are non-comment,
-non-blank lines), a manual, examples, etc.  It is publicly available at
+non-blank lines), a manual, examples, etc.
+\TheDeterminismChecker requires Java version 8.
+It is publicly available at
 \myurl{URL for \theDeterminismChecker}.
-
-\TheDeterminismChecker requires Java version 8.
 
@@
 \<@OrderNonDet>, and \<@Det> (and others described below).
+The meaning of \<@Det> is with respect to value
+equality, not reference equality; that is,
+values on different executions are the same with respect to \<.equals()>,
+not \<==>.
 
@@
 This does not change the expressivity of the type system; it merely makes
-the system more convenient to use.
+the system more convenient to use by reducing programmer effort and code clutter.
 
-In \theDeterminismChecker, we chose \<PolyDet> as the default for all 
-method parameters and return types including constructors.
-The rationale for this design choice was to not make \theDeterminismChecker too restrictive by making everything \<Det>.  This implies that
-\theDeterminismChecker still allows some non-determinism as long as it doesn't break any determinism specifications explicitly written by the programmer.
+The default annotation for most unannotated types is \<@Det>, because
+programmers generally expect their programs to behave the same when re-run
+on the same inputs.
 
-\begin{verbatim}
-@PolyDet int foo(@PolyDet int x) { ... }
+Formal parameter and return types default to \<@PolyDet>.  That is, a
+programmer-written method
 
-void bar(@NonDet int random, @Det int value) {
-    @NonDet int x = foo(random);      // No error.
-    @Det int y = foo(value);          // No error.
-    System.out.println(foo(random));  // Error.
-}
-\end{verbatim}
-In the example above, the method \codeid{foo} is annotated with the \<PolyDet> type qualifier which allows
-the caller method \codeid{bar} to invoke it with both \<Det> and \<NonDet> arguments. \TheDeterminismChecker flags an error 
-only when the returned value is used in a way that breaks any deterministic specification (Printing a \<NonDet> value
-breaks our determinism guarantees).
+\begin{Verbatim}
+  int plus(int a, int b) { ... }
+\end{Verbatim}
 
-\<@Det> equality is with respect to \<.equals()>.
+\noindent
+is treated as if the programmer had written
 
-\todo{
-The default annotation for most unannotated types is \<@Det>.
+\begin{Verbatim}
+  @PolyDet int plus(@PolyDet int a, @PolyDet int b) { ... }
+\end{Verbatim}
+
+\noindent
+and its function type is
+
+$\forallt{\kappa}  \kappa\ \<int> \times \kappa\ \<int> \rightarrow \kappa\ \<int>$
+
+\noindent
+This choice type-checks if the method body does not make calls to any
+interfaces that require \<@Det> arguments or produce \<@NonDet> results.
 
 \begin{itemize}
-\item
-  Formal parameter and return types default to \<@PolyDet>.
 \item
@@
 These choices permit all possible arguments to an unannotated method.
+
+In \theDeterminismChecker, we chose \<PolyDet> as the default for all 
+method parameters and return types including constructors.
+The rationale for this design choice was to not make \theDeterminismChecker too restrictive by making everything \<Det>.
+
+\begin{verbatim}
+@PolyDet int foo(@PolyDet int x) { ... }
+
+void bar(@NonDet int random, @Det int value) {
+    @NonDet int x = foo(random);      // no error
+    @Det int y = foo(value);          // no error
+    System.out.println(foo(random));  // error
 }
+\end{verbatim}
+In the example above, the method \codeid{foo} is annotated with the \<PolyDet> type qualifier which allows
+the caller method \codeid{bar} to invoke it with both \<Det> and \<NonDet> arguments. \TheDeterminismChecker flags an error 
+only when the returned value is used in a way that breaks any deterministic specification (Printing a \<NonDet> value
+breaks our determinism guarantees).
 

### 2019-04-05 java-design.tex
Move and edit defaults section

@@
 
-\subsection{Defaulting rules}\label{defaulting}
-
-\TheDeterminismChecker applies a default qualifier at each unannotated type
-use (except uses of type variables, which already stand for a type that was
-defaulted at the instantiation site).
-This does not change the expressivity of the type system; it merely makes
-the system more convenient to use by reducing programmer effort and code clutter.
-
-The default annotation for most unannotated types is \<@Det>, because
-programmers generally expect their programs to behave the same when re-run
-on the same inputs.
-
-Formal parameter and return types default to \<@PolyDet>.  That is, a
-programmer-written method
-
-\begin{Verbatim}
-  int plus(int a, int b) { ... }
-\end{Verbatim}
-
-\noindent
-is treated as if the programmer had written
-
-\begin{Verbatim}
-  @PolyDet int plus(@PolyDet int a, @PolyDet int b) { ... }
-\end{Verbatim}
-
-\noindent
-and its function type is
-
-$\forallt{\kappa}  \kappa\ \<int> \times \kappa\ \<int> \rightarrow \kappa\ \<int>$
-
-\noindent
-This choice type-checks if the method body does not make calls to any
-interfaces that require \<@Det> arguments or produce \<@NonDet> results.
-
-\begin{itemize}
-\item
-  If a formal parameter/return type is an array, its component type defaults to
-  \<@PolyDet>.
-\item
-  As an exception, the return type of a method with no unannotated or
-  \<@PolyDet> formal parameters (including the receiver) defaults to
-  \<@Det>.
-\end{itemize}
-
-These choices permit all possible arguments to an unannotated method.
-
-In \theDeterminismChecker, we chose \<PolyDet> as the default for all 
-method parameters and return types including constructors.
-The rationale for this design choice was to not make \theDeterminismChecker too restrictive by making everything \<Det>.
-
-\begin{verbatim}
-@PolyDet int foo(@PolyDet int x) { ... }
-
-void bar(@NonDet int random, @Det int value) {
-    @NonDet int x = foo(random);      // no error
-    @Det int y = foo(value);          // no error
-    System.out.println(foo(random));  // error
-}
-\end{verbatim}
-In the example above, the method \codeid{foo} is annotated with the \<PolyDet> type qualifier which allows
-the caller method \codeid{bar} to invoke it with both \<Det> and \<NonDet> arguments. \TheDeterminismChecker flags an error 
-only when the returned value is used in a way that breaks any deterministic specification (Printing a \<NonDet> value
-breaks our determinism guarantees).
-
 \subsection{Polymorphism}\label{java-polymorphism}
@@
 
+\subsection{Defaulting}\label{defaulting}
+
+\TheDeterminismChecker applies a default qualifier at each unqualified Java
+basetype (except uses of type variables, which already stand for a type that was
+defaulted at the instantiation site).
+This does not change the expressivity of the type system; it merely makes
+the system more convenient to use by reducing programmer effort and code clutter.
+Defaulted type qualifiers are not trusted:  they are type-checked just as
+explicitly-written ones are.
+
+Formal parameter and return types default to \<@PolyDet>.  That is, a
+programmer-written method
+
+\begin{Verbatim}
+  int plus(int a, int b) { ... }
+\end{Verbatim}
+
+\noindent
+is treated as if the programmer had written
+
+\begin{Verbatim}
+  @PolyDet int plus(@PolyDet int a, @PolyDet int b) { ... }
+\end{Verbatim}
+
+\noindent
+and its function type is
+
+$\forallt{\kappa}  \kappa\ \<int> \times \kappa\ \<int> \rightarrow \kappa\ \<int>$
+
+\noindent
+This choice type-checks if the method body does not make calls to any
+interfaces that require \<@Det> arguments or produce \<@NonDet> results.
+Otherwise, the programmer must write an explicit type qualifier.
+
+If a formal parameter or return type is a collection, its element type
+defaults to \<@PolyDet>.  No other choice would result in a valid type
+after instantiation.
+
+As an exception to the above rules about return types, if a method's formal
+parameters (including the receiver) are all \<@Det> --- that is, there are
+no unannotated or \<@PolyDet> formal parameters --- then an unannotated
+return type defaults to \<@Det>.  This is particularly useful for methods
+that take no formal parameters.  A type like $\forallt{\kappa} ()
+\rightarrow \kappa int$ does not make sense, becauses there is no way to
+instantiate the type argument $\kappa$.  Treating the type as $()
+\rightarrow \<Det> int$ permits just as many uses.
+
+The default annotation for local variables is \<@NonDet>, which is the top
+qualifier in the hierarchy.  \Cref{sec:dataflow-java} explains why.
+
+The default annotation for other unannotated types is \<@Det>, because
+programmers generally expect their programs to behave the same when re-run
+on the same inputs.
+
 \subsection{Dataflow analysis and type refinement}\label{sec:dataflow-java}

### 2019-04-05 java-design.tex
Edits on page 15

@@
 The return type of \<System.getenv>, which reads an operating system
-environment variable, is always \<NonDet>.
-The return type of \codeid{System.getProperty} is usually \<NonDet>. However,
-the result is \<Det> if the argument is \codeid{"line.separator"}; although
-it varies by operating system, many programs including \<diff> hide those
+environment variable, is always \<@NonDet>.
+
+The return type of \codeid{System.getProperty} is usually \<@NonDet>. However,
+the result is \<@Det> if the argument is \codeid{"line.separator"}; although
+it varies by operating system, many programs including \<diff> and editors hide those
 differences from the user.
@@
 treated as deterministic.
-The inputs to a program are also treated as \<Det>.  That is, the parameter
-type of the \<main> method is implicitly annotated as
+
+The inputs to a program are also treated as \<Det>.  That is, the type of
+the formal parameter to \<main> is
 \<@Det String @Det []>, a deterministic array of deterministic strings.

### 2019-04-05 randoop-case-study.tex
randoop bugs
Edits on page 15

@@
 we applied it to the Randoop test
-generation tool~\ifanonymous\citepalias{randoop-tool}\else\cite{PachecoLEB2007}\fi.
-We also wrote specifications for libraries it uses, such as the JDK, JUnit,
+generator~\ifanonymous\citepalias{randoop-tool}\else\cite{PachecoLEB2007}\fi.
+We also wrote specifications for libraries Randoop uses, such as the JDK, JUnit,
 and others.
@@
   Diagnostic output printed a timestamp.
-  The developers fixed it by making it obeyed an
+  The developers fixed it by making it obey an
   option about whether to print timestamps (commit a460df97).

### 2019-04-05 design.tex
Edits in section 3.1
types for get and set
Polymorphism syntax in Java

@@
 
+    \rulename{get}: 
+    $
+    \forallt{\tau} \kappa \in \{\<Det>, \<NonDet>\}\ \<OrderNonDet> \ C \angles{\tau} \ \times \kappa \ \<Int> \rightarrow 
+    \<NonDet> \ \<Int>
+    $
+
+    \bigskip
+
+    \rulename{set}: 
+    $
+    \forallt{\kappa, \beta} \ \<OrderNonDet> \ C \angles{\kappa \ \beta} \ \times \<Det> \ \<Int> \times \kappa \ \beta_n \rightarrow 
+    \<NonDet> \ \beta
+    $
+
+    \bigskip
+
+    \rulename{set invalid}: 
+    $
+    \forallt{\kappa, \beta} \ \<OrderNonDet> \ C \angles{\kappa \ \beta} \ \times \<NonDet> \ \<Int> \times \kappa \ \beta_n \rightarrow 
+    \|INVALID|
+    $
+
+    \bigskip
+
     \rulename{iteration}: 
@@
 In addition to assignments, some method calls refine the types of their
-arguments.  One example is the \<sort> routine, as shown in 
+arguments.  One example is the \<sort> routine,
 as shown in \cref{fig-sorting}.  One notable fact about the \rulename{sort}
@@
 
-\subsection{Determinism checker specific rules for precision}\label{precision}
+\subsection{Improving precision}\label{precision}
 
@@
 \begin{verbatim}
-PolyDet("down") boolean equals(PolyDet Set<E> this, PolyDet Object o)
+PolyDet("up") boolean equals(PolyDet Set<E> this, PolyDet Object o)
 \end{verbatim}

### 2019-04-05 java-design.tex
Polymorphism syntax in Java

@@
 \end{Verbatim}
+the type of \<identity> is $\forallt{\tau} \tau \rightarrow \tau$, and
 the type of \<identity(arg)> is the same as the type of
-\<arg>, and a programmer can write a type such as \codeid{List<@NonDet
-  Integer>} (but see below for details about collections).
+\<arg>.  A programmer can also write a type qualifier on a type argument,
+as in \codeid{@NonDet List<@NonDet Integer>}.
 \item
@@
 \begin{Verbatim}
-class ChooseableSet<T> implements Set<T> {
+interface ChooseableSet<T> implements Set<T> {
   // return an arbitrary element from the set
@@
 Java does not provide a syntax that can be used for qualifier polymorphism,
-so we use a special type qualifier name, \<@PolyDet>.
+so \theDeterminismChecker uses a special type qualifier name, \<@PolyDet>.  (\<@PolyDet> stands
+for ``polymorphic determinism qualifier.)
 A qualifier-polymorphic method \<m> with signature $\forallt{\kappa} \kappa\ \<int> \times \<Det boolean> \rightarrow
@@
 
-\todo{Explain @PolyDet("up"), "down", "use".}
+Qualifier  polymorphism is common on methods that a programmer might think
+of as deterministic.  For example, an addition method could be defined as
 
-When a user writes a polymorphic qualifier on a method signature or a type parameter,
-it indicates that it could be instantiated
-%\todo{Should all occurrences of ``resolve'' be ``can be instantiated''} 
-to any type in the type system depending on how it is used.
-In \theDeterminismChecker, we define a polymorphic qualifier \<PolyDet>.
-One of the most common locations for a polymorphic qualifier is at a method signature.
-Consider the following declaration of addition:
 \begin{verbatim}
@@
 \end{verbatim}
-\todo{It has the standard meaning.  There are two ways to interpret this.}
-There are two valid instantiations for this declaration:
-\begin{Verbatim}
-  @NonDet int plus(@NonDet int a, @NonDet int b) { return a+b; }
+
+This can be used in more contexts than
+
+\begin{verbatim}
     @Det int plus(@Det int a, @Det int b) { return a+b; }
-\end{Verbatim}
-This indicates that method \<foo> can be called with arguments having any of the \codeid{@Det} or \codeid{@NonDet} type qualifiers.
-%   (\<@OrderNonDet> is not allowed because it is invalid on primitive types).
- \<PolyDet> gets instantiated to the least upper bound of
-the actual types on arguments.
-For instance, if this method is called as \codeid{foo(@Det String arg1, @NonDet boolean arg2)}, \theDeterminismChecker instantiates the method declaration as \codeid{@NonDet int foo(@NonDet String param1, @NonDet boolean param2)}
-causing the return type to have the type qualifier \<NonDet>.
+\end{verbatim}
+
+\noindent
+can be, as was shown in \cref{sec:basic-polymorphism}.
+
+Just as a qualifier variable $\kappa$ is written as \<@PolyDet> in Java
+source code, $\up{\kappa}$ is written as \<@PolyDet("up")>, and
+$\down{\kappa}$ is written as \<@PolyDet("down")>.  An occurrence of a
+qualifier variable that does not affect the binding of that variable
+(\cref{bindings-uses}) is written \<@PolyDet("use")>.
 

### 2019-04-05 java-design.tex
todo map subtyping relationship

@@
 
-The JDK implements \<LinkedHashSet> as a subclass of \<HashMap>\todo{should
-  this be HashSet?  It's not true as written.}
+The JDK implements \<LinkedHashMap> as a subclass of \<HashMap>
 in order to inherit parts of \<HashMap>'s implementation.
-\todo{Talk about this design choice. Josh Bloch's book: "If the code surrounding the first declaration depended on LinkedHashSet's ordering policy,
-    then it would be incorrect to substitute HashSet for LinkedHashSet in the declaration, because HashSet makes no guarantee concerning iteration order."} 
-\todo{The above is not clear.  What are the consequences for our
-  implementation.  What goes wrong or is hard to annotate or is surprising?
-  I'm not sure what is the point of this paragraph.}
+%\todo{Talk about this design choice.
+Josh Bloch's book: "If the code surrounding the first declaration depended on LinkedHashSet's ordering policy, then it would be incorrect to substitute HashSet for LinkedHashSet in the declaration, because HashSet makes no guarantee concerning iteration order."}
+%\todo{The above is not clear.  What are the consequences for our
+%  implementation.  What goes wrong or is hard to annotate or is surprising?
+%  I'm not sure what is the point of this paragraph.}
+\todo{From the discussion in this section, it is clear that \<LinkedHashMap> and \<HashMap> do not
+have a subtyping relationship with respect to their determinism types. We looked at the implementation of
+these two classes in the JDK and found that \<LinkedHashMap> doesn't reuse code that corresponds to 
+iterating over its keys or values with \<HashMap>.}
 

### 2019-04-05 introduction.tex
minor edits
intro remove todos

--- 2019-04-05 introduction.tex
@@
   and prevents use of techniques such as Delta Debugging~\cite{Zeller1999}.
-\item
-  Nondeterminism may make outputs incompatible with previous outputs relied
-  on by users or by other systems~\todo{These do not seem to be supporting
-    the claim about bugs related to incompatibilities between outputs.
-    These seem to be general citations about test flakiness or dependence.
-    Readers will be unhappy if we make a claim but then do not support it.}\cite{Herzig:2015:EDF:2819009.2819018,Huo:2014:IOQ:2635868.2635917,ZhangJWMLEN2014,BellKMD2015,Fowler,Gyori:2015:RTD:2771783.2771793,Dan:2013:10.1007/978-3-642-39038-8_25,Vahabzadeh:2015:7332456,Sudarshan}.
+%\item
+%  Nondeterminism may make outputs incompatible with previous outputs relied
+%  on by users or by other systems~\todo{These do not seem to be supporting
+%    the claim about bugs related to incompatibilities between outputs.
+%    These seem to be general citations about test flakiness or dependence.
+%    Readers will be unhappy if we make a claim but then do not support it.}\cite{Herzig:2015:EDF:2819009.2819018,Huo:2014:IOQ:2635868.2635917,ZhangJWMLEN2014,BellKMD2015,Fowler,Gyori:2015:RTD:2771783.2771793,Dan:2013:10.1007/978-3-642-39038-8_25,Vahabzadeh:2015:7332456,Sudarshan}.
 %  \todo{Citations about bugs related to this?}
@@
 So may any program that uses default formatting, such as Java's
-\<Object.toString()> which includes a memory address.
+\<Object.toString()>, which includes a memory address.
 Other nondeterministic APIs include date-and-time functions and
@@
 
+%\todo{Remove this?
+We implemented \TheDeterminismChecker as a pluggable type checker on top of the checker-framework available at \href{https://checkerframework.org/}{https://checkerframework.org/}.
+\TheDeterminismChecker is written in Java and
+%    spans approximately 1.2K lines of code.}
+
 This paper makes the following contributions:

### 2019-04-05 related.tex
related edits

@@
 in 13 classes that were identified as potential sources of flakiness. 
-This is analogous to our annotations on the JDK\@. So far, we have annotated
-928 methods across 41 classes.
 For each of the identified methods, the authors built models that
@@
 and considerable debugging effort to locate and fix the source of flakiness. \TheDeterminismChecker, in
-contrast, reports the cause of nondeterminism at compile requiring little to no debugging effort with the
+contrast, reports the cause of nondeterminism at compile time requiring little to no debugging effort with the
 caveat of producing higher false positives compared to NonDex. 
+Identifying and modeling methods with nondeterministic
+specifications is analogous to our annotations on the JDK\@. So far, we have annotated
+928 methods across 41 classes.
 %\todo{Check whether our JDK is missing anything they support - No}
@@
 version control history to detect tests that behave differently. It computes a diff of the code covered
-in the current version and the previous one. If there exists a tests case whose code coverage does not include
+in the current version and the previous one. If there exists a test case whose code coverage does not include
 this diff but still produces different results on the two versions being compared, it is flagged as being flaky.
@@
 and integrates easily with a production software. DeFlaker reported 19 previously unknown bugs
-in open source projects, 7 of which were addressed by the developers of these projects.
+in open source projects, 7 of which were addressed by the developers of these projects. DeFlaker is agnostic
+to the code under test and can therefore report flakiness arising out of concurrency which \TheDeterminismChecker cannot.
 
@@
 test flakiness. The findings in the paper suggest that most of the flakiness in tests is caused by
-the use of async away, concurrency, or due to test order dependency.~\cite{Plotkin:1993:LPP:645891.671433}
-performs an empirical study on open source benchmarks that concur with the results of~\cite{LuoHEM2014}.
+the use of async await, concurrency, or due to test order dependency.~\cite{Plotkin:1993:LPP:645891.671433}
+perform an empirical study on open source benchmarks that concur with the results of~\cite{LuoHEM2014}.
 Additionally, the authors of this paper study the effects of removing these test smells. Their results suggest that

### 2019-04-05 java-design.tex
Eliminate duplication

--- 2019-04-05 java-design.tex
@@
 code, plus 1800 lines of tests (all line measurements are non-comment,
-non-blank lines), a manual, examples, etc.
+non-blank lines), a manual, etc.
 \TheDeterminismChecker requires Java version 8.
@@
 
-Our implementation treats arrays analogously to collections:
+\TheDeterminismChecker treats arrays analogously to collections:
 the semantics and rules are similar even though the syntax varies.
@@
 client code that would produce errors at compile time.
+
 %\begin{figure}
@@
 \end{figure}
-
-\begin{itemize}
-\item
-In \theDeterminismCheckerImplementation,
-type polymorphism is handled by Java's generics mechanism, which
-\theDeterminismChecker fully supports, including class and method generics,
-inference, etc.
-Given the Java declaration
-\begin{Verbatim}
-<T> T identity(T arg) { return arg; }
-\end{Verbatim}
-the type of \<identity(arg)> is the same as the type of
-\<arg>, and a programmer can write a type such as \codeid{List<@NonDet
-  Integer>} (but see below for details about collections).
-\item
-Basetype polymorphism is enabled by writing a type qualifier on a use of a
-type variable, which overrides the type qualifier at the instantiation
-site.
-For example, a \<choose> operation on sets could be defined in Java as
-\begin{Verbatim}
-class ChooseableSet<T> implements Set<T> {
-  // return an arbitrary element from the set
-  @NonDet T choose();
-}
-\end{Verbatim}
-\item
-\todo{We do not support qualifier polymorphism on classes, only on methods.}
-Java does not provide a syntax that can be used for qualifier polymorphism,
-so we use a special type qualifier name, \<@PolyDet>.
-A qualifier-polymorphic method \<m> with signature $\forallt{\kappa} \kappa\ \<int> \times \<Det boolean> \rightarrow
-\kappa\ \<String>$ would be written in our dialect of Java as
-\begin{Verbatim}
-@PolyDet String m(@PolyDet int, @Det boolean)
-\end{Verbatim}
-Each use of \<@PolyDet> stands for a use of the qualifier variable
-$\kappa$, and there is no need to declare the qualifier variable $\kappa$.
-\TheDeterminismChecker currently supports
-qualifier
-polymorphism on methods but not on classes.
-Therefore, \<@PolyDet> may be written in methods (signatures and bodies)
-but not on fields.
-\end{itemize}
 

### 2019-04-05 related.tex
more related

@@
 
+Failing tests that are unrelated to code changes can be expensive both in terms of monetary costs and
+developer effort. \cite{Herzig:2015:EDF:2819009.2819018} propose techniques to classify tests as false alarms
+if they are known to be caused by testing infrastructure or other environment issues.~\cite{Huo:2014:IOQ:2635868.2635917}
+propose an approach that detects brittle assertions in tests by performing a taint analysis on inputs classified as 
+controlled and uncontrolled.~\cite{ZhangJWMLEN2014} investigate the effects of test independence assumption
+on other techniques such as test prioritization, selection, etc. 
+\cite{BellKMD2015}
+\cite{Gyori:2015:RTD:2771783.2771793}
+\cite{Dan:2013:10.1007/978-3-642-39038-8_25}
+\cite{Vahabzadeh:2015:7332456}
+\cite{Fowler}
+\cite{Sudarshan}
+
 \todo{A NonDet annotation:

### 2019-04-05 java-design.tex

@@
 A \<Map> is deterministic if its \<entrySet> is deterministic.
-In other words, iterating over this \<entrySet> is guaranteed to produce the same values in the exact same order across executions.
-The \<Map> interface in Java doesn't define an \<iterator()> interface. Therefore it's sound to equate the determinism of the \<Set>
-returned by \<entrySet()> to the determinism of the \<Map>. Interestingly, we found that every implementation of the \<Map> 
-interface had the following property: if the \<Set> returned by \<entrySet()> has a determinism type $\kappa$, then
-the \<Set>s returned by \<keySet()> and \<values()> also have the exact same determinism type $\kappa$.
+In other words, iterating over this \<entrySet> produces the same values in the same order across executions.
+\Cref{code-determinism} shows that the determinism qualifier on the return type of \<entrySet()> is the
+same as that on the receiver.  That is, its type (ignoring type arguments) is
+$\forallt{\kappa} \kappa\ \<Map> \rightarrow \kappa\ \<Set>$\ .
+
+We have not yet found any implementation of the \<Map>
+interface in which \<entrySet()> \<keySet()>, and \<values()> behave
+differently with respect to determinacy:  either all, or none, or deterministic.
 
@@
         HashMap} does not exist.
-    \item \<LinkedHashMap>, like \<List>, can have any of the \<@NonDet>,
-      \<@OrderNonDet>, or \<@Det> type qualifiers. Iterating over a
+    \item \<LinkedHashMap>, like \<List>, can have any of the \aNonDet,
+      \aOrderNonDet, or \aDet type qualifiers. Iterating over a
       \<LinkedHashMap> returns
-    its entries in the order of their insertion. An \<OrderNonDet
-    LinkedHashMap> can be created by calling its constructor with a
-    \<@OrderNonDet HashMap>.
+    its entries in the order of their insertion. An \aOrderNonDet\
+    \<LinkedHashMap> can be created by passing an
+    \aOrderNonDet\ \<HashMap> to its constructor, as in \<new
+    LinkedHashMap(myOndHashMap)>.
     \item \<TreeMap> can either be \<@Det> or \<@NonDet>. An \<@OrderNonDet
-      TreeMap> doesn't exist because the entries are always ordered.
+      TreeMap> doesn't exist because the entries are always sorted.
 \end{itemize}
@@
 qualifier
-polymorphism on methods but not on classes.
+polymorphism on methods but \emph{not} on classes.
 Therefore, \<@PolyDet> may be written in methods (signatures and bodies)

### 2019-04-05 related.tex

@@
 
+Nondeterminism in tests is of interest to both
+researchers and software developers alike~\cite{Fowler,Sudarshan}.
 \cite{LuoHEM2014} study 51 open source projects and analyze the most common root causes of
@@
 controlled and uncontrolled.~\cite{ZhangJWMLEN2014} investigate the effects of test independence assumption
-on other techniques such as test prioritization, selection, etc. 
-\cite{BellKMD2015}
-\cite{Gyori:2015:RTD:2771783.2771793}
-\cite{Dan:2013:10.1007/978-3-642-39038-8_25}
-\cite{Vahabzadeh:2015:7332456}
-\cite{Fowler}
-\cite{Sudarshan}
+on other techniques such as test prioritization, selection, etc. Other approaches~\cite{BellKMD2015,Gyori:2015:RTD:2771783.2771793}
+analyze test dependencies and either prevent them or use this information for other optimizations.
+\cite{Dan:2013:10.1007/978-3-642-39038-8_25,Vahabzadeh:2015:7332456} focus on differentiating bugs due to tests
+from those caused by source code.
 
-\todo{A NonDet annotation:
-  \url{http://www.swi-prolog.org/pldoc/man?section=modes}}
+%\todo{A NonDet annotation:
+%  \url{http://www.swi-prolog.org/pldoc/man?section=modes}}
 

### 2019-04-05 java-design.tex
Move a figure
Edits in secton 3

@@
 Java collections such as \<List>s and \<Set>s.
-Our implementation also supports any type that implements the \<Map>
-interface.
+
+\subsection{Java collection types}\label{sec:maps-java}
+
+A \<Map> is deterministic if its \<entrySet> is deterministic.
+In other words, iterating over this \<entrySet> produces the same values in the same order across executions.
+The determinism qualifier on the return type of \<entrySet()> is the
+same as that on the receiver (\cref{code-determinism}).
+That is, its type (ignoring type arguments) is
+$\forallt{\kappa} \kappa\ \<Map> \rightarrow \kappa\ \<Set>$\ .
+
+We have not yet found any implementation of the \<Map>
+interface in which \<entrySet()> \<keySet()>, and \<values()> behave
+differently with respect to determinacy:  either all three methods, or none, are deterministic.
+
+The most widely used \<Map> implementations have the following properties:
+\begin{itemize}
+    \item \<HashMap> is implemented in terms of a hash table, which never
+      guarantees deterministic iteration over its entries. A \codeid{@Det
+        HashMap} does not exist.
+    \item \<LinkedHashMap>, like \<List>, can have any of the \aNonDet,
+      \aOrderNonDet, or \aDet type qualifiers. Iterating over a
+      \<LinkedHashMap> returns
+    its entries in the order of their insertion. An \aOrderNonDet\
+    \<LinkedHashMap> can be created by passing an
+    \aOrderNonDet\ \<HashMap> to its constructor, as in \<new
+    LinkedHashMap(myOndHashMap)>.
+    \item \<TreeMap> can either be \<@Det> or \<@NonDet>. An \<@OrderNonDet
+      TreeMap> doesn't exist because the entries are always sorted.
+\end{itemize}
+
+\noindent
+\TheDeterminismChecker prohibits the creation of a \codeid{@Det HashMap} or an \codeid{@OrderNonDet TreeMap}.
+\Cref{fig-creation-rules} formalizes the type well-formedness rules for
+Java \<Map>s and \<Set>s.
+
+\begin{figure}
+    $\infer[\rulename{valid hashmap}]{\vdash: \|\kappa\ \<HashMap>|}{\kappa \in \{ \<OrderNonDet>, \<NonDet> \}}$
+
+    \bigskip
+
+    $\infer[\rulename{valid treemap}]{\vdash: \|\kappa\ \<TreeMap>|}{\kappa \in \{ \<Det>, \<NonDet> \}}$
+
+    \bigskip
+
+    $\infer[\rulename{valid hashset}]{\vdash: \|\kappa\ \<HashSet>|}{\kappa \in \{ \<OrderNonDet>, \<NonDet> \}}$
+
+    \bigskip
+
+    $\infer[\rulename{valid treeset}]{\vdash: \|\kappa\ \<TreeSet>|}{\kappa \in \{ \<Det>, \<NonDet> \}}$
+    \caption{Type well-formedness rules for subclasses of \<Map> and \<Set>.}
+    \label{fig-creation-rules}
+\end{figure}
+
+\subsection{Polymorphism}\label{java-polymorphism}
+
+\OurTypeSystem supports three types of polymorphism:  type
+polymorphism, basetype polymorphism, and qualifier polymorphism.
+These apply to both classes and methods.
+
+\begin{itemize}
+\item
+In \theDeterminismCheckerImplementation,
+type polymorphism is handled by Java's generics mechanism, which
+\theDeterminismChecker fully supports, including class and method generics,
+inference, etc.
+Given the Java declaration
+\begin{Verbatim}
+<T> T identity(T param) { return param; }
+\end{Verbatim}
+the type of \<identity> is $\forallt{\tau} \tau \rightarrow \tau$, and
+the type of \<identity(x)> is the same as the type of
+\<x>.  A programmer can also write a type qualifier on a type argument,
+as in \codeid{@NonDet List<@NonDet Integer>}.
+\item
+Basetype polymorphism is enabled by writing a type qualifier on a use of a
+type variable, which overrides the type qualifier at the instantiation
+site.
+For example, a \<choose> operation on sets could be defined in Java as
+\begin{Verbatim}
+interface ChooseableSet<T> implements Set<T> {
+  // return an arbitrary element from the set
+  @NonDet T choose();
+}
+\end{Verbatim}
+\item
+Java does not provide a syntax that can be used for qualifier polymorphism,
+so \theDeterminismChecker uses a special type qualifier name, \<@PolyDet>.  (\<@PolyDet> stands
+for ``polymorphic determinism qualifier''.)
+A qualifier-polymorphic method \<m> with signature
+$\forallt{\kappa} \kappa\ \<int> \times \<Det boolean>
+\rightarrow \kappa\ \<String>$ would be written as
+\begin{Verbatim}
+@PolyDet String m(@PolyDet int, @Det boolean)
+\end{Verbatim}
+Each use of \<@PolyDet> stands for a use of the qualifier variable
+$\kappa$, and there is no need to declare the qualifier variable $\kappa$.
+\TheDeterminismChecker currently supports
+qualifier
+polymorphism on methods but \emph{not} on classes.
+Therefore, \<@PolyDet> may be written in methods (signatures and bodies)
+but not on fields.
+\end{itemize}
+
+Qualifier  polymorphism is common on methods that a programmer might think
+of as deterministic.  For example, an addition method should be defined as
+
+\begin{verbatim}
+    @PolyDet int plus(@PolyDet int a, @PolyDet int b) { return a+b; }
+\end{verbatim}
+
+\noindent
+This can be used in more contexts than
+
+\begin{verbatim}
+    @Det int plus(@Det int a, @Det int b) { return a+b; }
+\end{verbatim}
+
+\noindent
+can be, as was shown in \cref{sec:basic-polymorphism}.
+
+Just as a qualifier variable $\kappa$ is written as \<@PolyDet> in Java
+source code, $\up{\kappa}$ is written as \<@PolyDet("up")>, and
+$\down{\kappa}$ is written as \<@PolyDet("down")>.  An occurrence of a
+qualifier variable that does not affect the binding of that variable
+(\cref{bindings-uses}) is written \<@PolyDet("use")>.
+
+All of our syntax is legal Java code that can be compiled with any Java 8
+or later compiler.
 
@@
             ...
-            for(Map. @Det Entry<K,V> entry : map.entrySet()) { ... }
+            for (Map. @Det Entry<K,V> entry : map.entrySet()) { ... }
         }
@@
 
-\subsection{Java collection types}\label{sec:maps-java}
-
-A \<Map> is deterministic if its \<entrySet> is deterministic.
-In other words, iterating over this \<entrySet> produces the same values in the same order across executions.
-\Cref{code-determinism} shows that the determinism qualifier on the return type of \<entrySet()> is the
-same as that on the receiver.  That is, its type (ignoring type arguments) is
-$\forallt{\kappa} \kappa\ \<Map> \rightarrow \kappa\ \<Set>$\ .
-
-We have not yet found any implementation of the \<Map>
-interface in which \<entrySet()> \<keySet()>, and \<values()> behave
-differently with respect to determinacy:  either all, or none, or deterministic.
-
-The most widely used \<Map> implementations have the following properties:
-\begin{itemize}
-    \item \<HashMap> is implemented in terms of a hash table, which never
-      guarantees deterministic iteration over its entries. A \codeid{@Det
-        HashMap} does not exist.
-    \item \<LinkedHashMap>, like \<List>, can have any of the \aNonDet,
-      \aOrderNonDet, or \aDet type qualifiers. Iterating over a
-      \<LinkedHashMap> returns
-    its entries in the order of their insertion. An \aOrderNonDet\
-    \<LinkedHashMap> can be created by passing an
-    \aOrderNonDet\ \<HashMap> to its constructor, as in \<new
-    LinkedHashMap(myOndHashMap)>.
-    \item \<TreeMap> can either be \<@Det> or \<@NonDet>. An \<@OrderNonDet
-      TreeMap> doesn't exist because the entries are always sorted.
-\end{itemize}
-
-\TheDeterminismChecker prohibits the creation of a \codeid{@Det HashMap} or an \codeid{@OrderNonDet TreeMap}.
-\Cref{fig-creation-rules} formalizes the type well-formedness rules for
-Java \<Map>s and \<Set>s.
-
-\begin{figure}
-    $\infer[\rulename{valid hashmap}]{\vdash: \|\kappa\ \<HashMap>|}{\kappa \in \{ \<OrderNonDet>, \<NonDet> \}}$
-
-    \bigskip
-
-    $\infer[\rulename{valid treemap}]{\vdash: \|\kappa\ \<TreeMap>|}{\kappa \in \{ \<Det>, \<NonDet> \}}$
-
-    \bigskip
-
-    $\infer[\rulename{valid hashset}]{\vdash: \|\kappa\ \<HashSet>|}{\kappa \in \{ \<OrderNonDet>, \<NonDet> \}}$
-
-    \bigskip
-
-    $\infer[\rulename{valid treeset}]{\vdash: \|\kappa\ \<TreeSet>|}{\kappa \in \{ \<Det>, \<NonDet> \}}$
-    \caption{Type well-formedness rules for subclasses of \<Map> and \<Set>.}
-    \label{fig-creation-rules}
-\end{figure}
-
-The JDK implements \<LinkedHashMap> as a subclass of \<HashMap>
-in order to inherit parts of \<HashMap>'s implementation.
-%\todo{Talk about this design choice.
-Josh Bloch's book: "If the code surrounding the first declaration depended on LinkedHashSet's ordering policy, then it would be incorrect to substitute HashSet for LinkedHashSet in the declaration, because HashSet makes no guarantee concerning iteration order."}
-%\todo{The above is not clear.  What are the consequences for our
-%  implementation.  What goes wrong or is hard to annotate or is surprising?
-%  I'm not sure what is the point of this paragraph.}
-\todo{From the discussion in this section, it is clear that \<LinkedHashMap> and \<HashMap> do not
-have a subtyping relationship with respect to their determinism types. We looked at the implementation of
-these two classes in the JDK and found that \<LinkedHashMap> doesn't reuse code that corresponds to 
-iterating over its keys or values with \<HashMap>.}
-
-\subsection{Polymorphism}\label{java-polymorphism}
-
-\OurTypeSystem supports three types of polymorphism:  type
-polymorphism, basetype polymorphism, and qualifier polymorphism.
-These apply to both classes and methods.
-
-\begin{itemize}
-\item
-In \theDeterminismCheckerImplementation,
-type polymorphism is handled by Java's generics mechanism, which
-\theDeterminismChecker fully supports, including class and method generics,
-inference, etc.
-Given the Java declaration
-\begin{Verbatim}
-<T> T identity(T arg) { return arg; }
-\end{Verbatim}
-the type of \<identity> is $\forallt{\tau} \tau \rightarrow \tau$, and
-the type of \<identity(arg)> is the same as the type of
-\<arg>.  A programmer can also write a type qualifier on a type argument,
-as in \codeid{@NonDet List<@NonDet Integer>}.
-\item
-Basetype polymorphism is enabled by writing a type qualifier on a use of a
-type variable, which overrides the type qualifier at the instantiation
-site.
-For example, a \<choose> operation on sets could be defined in Java as
-\begin{Verbatim}
-interface ChooseableSet<T> implements Set<T> {
-  // return an arbitrary element from the set
-  @NonDet T choose();
-}
-\end{Verbatim}
-\item
-Java does not provide a syntax that can be used for qualifier polymorphism,
-so \theDeterminismChecker uses a special type qualifier name, \<@PolyDet>.  (\<@PolyDet> stands
-for ``polymorphic determinism qualifier.)
-A qualifier-polymorphic method \<m> with signature $\forallt{\kappa} \kappa\ \<int> \times \<Det boolean> \rightarrow
-\kappa\ \<String>$ would be written in our dialect of Java as
-\begin{Verbatim}
-@PolyDet String m(@PolyDet int, @Det boolean)
-\end{Verbatim}
-Each use of \<@PolyDet> stands for a use of the qualifier variable
-$\kappa$, and there is no need to declare the qualifier variable $\kappa$.
-\TheDeterminismChecker currently supports
-qualifier
-polymorphism on methods but \emph{not} on classes.
-Therefore, \<@PolyDet> may be written in methods (signatures and bodies)
-but not on fields.
-\end{itemize}
-
-Qualifier  polymorphism is common on methods that a programmer might think
-of as deterministic.  For example, an addition method could be defined as
-
-\begin{verbatim}
-    @PolyDet int plus(@PolyDet int a, @PolyDet int b) { return a+b; }
-\end{verbatim}
-
-This can be used in more contexts than
-
-\begin{verbatim}
-    @Det int plus(@Det int a, @Det int b) { return a+b; }
-\end{verbatim}
-
-\noindent
-can be, as was shown in \cref{sec:basic-polymorphism}.
-
-Just as a qualifier variable $\kappa$ is written as \<@PolyDet> in Java
-source code, $\up{\kappa}$ is written as \<@PolyDet("up")>, and
-$\down{\kappa}$ is written as \<@PolyDet("down")>.  An occurrence of a
-qualifier variable that does not affect the binding of that variable
-(\cref{bindings-uses}) is written \<@PolyDet("use")>.
-
 \subsection{Defaulting}\label{defaulting}
@@
 \TheDeterminismChecker applies a default qualifier at each unqualified Java
-basetype (except uses of type variables, which already stand for a type that was
-defaulted at the instantiation site).
+basetype (except uses of type parameters, which already stand for a type that was
+defaulted at the instantiation site where a type argument was supplied).
 This does not change the expressivity of the type system; it merely makes
@@
 interfaces that require \<@Det> arguments or produce \<@NonDet> results.
-Otherwise, the programmer must write an explicit type qualifier.
+Otherwise, the programmer must write explicit type qualifiers in the method
+signature.
 
@@
 that take no formal parameters.  A type like $\forallt{\kappa} ()
-\rightarrow \kappa int$ does not make sense, becauses there is no way to
-instantiate the type argument $\kappa$.  Treating the type as $()
-\rightarrow \<Det> int$ permits just as many uses.
+\rightarrow \kappa\ \<int>$ does not make sense, because there is no basis
+on which to choose an
+instantiation for the type argument $\kappa$.  Treating the type as $()
+\rightarrow \<Det>\ \<int>$ permits just as many uses.
 
-The default annotation for local variables is \<@NonDet>, which is the top
-qualifier in the hierarchy.  \Cref{sec:dataflow-java} explains why.
+Types are inferred for unannotated local variables; see \cref{sec:dataflow-java}.
 
@@
 
-\todo{Relate this to defaulting.}
+\OurTypeSystem is
+flow-sensitive~\cite{Hunt:2006:FST:1111037.1111045,Adams:2011:FTR:2048066.2048105,Sui:2016:OSU:2950290.2950296}.
+That is, an expression may have
+a different type on every line of the program, based on assignments and
+side effects.  The current type must always be consistent with the declared
+type, if any.
+Consider the example below:
 
-It also performs dataflow analysis and refines the types of these locals, thereby reducing the qualifier burden 
-on the programmer. Consider the example below:
 \begin{verbatim}
-void foo(@Det int x) {
-    @NonDet int y = x;    // Type of y gets refined to @Det.
-    @Det int z = y;       // No error here.
-}
+  @NonDet int x = ...;
+                          // The type of x is @NonDet int
+  x = 42;
+                          // The type of x is @Det int
+  @Det int y = x;
+  x = random();
+                          // The type of x is @NonDet int
+  y = random();           // Error: y is declared as @Det int
 \end{verbatim}
+
+Arbitrary expressions can be given more specific types, including fields
+and pure method calls.  A type refinement is lost whenever a side effect
+might affect the value.  For example, type refinements to all fields are
+lost whenever a non-pure method is called.
+
+This flow-sensitive type refinement achieves local variable inference,
+freeing progammers from writing many local variable types.
+
+\todo{Mike doesn't understand the below.  An example would help clarify it.}
+
 In the example above, even though the local variable \codeid{y} was declared to be \codeid{NonDet}, it gets
@@
 
-With the above rules, \theDeterminismChecker will automatically annotate a local collection with type \<NonDet> and its
-type arguments will get the type qualifier \<Det>. For example, a locally declared \codeid{List} of \codeid{Strings} will be automatically annotated
-as \codeid{NonDet List<Det String>}. Recall from~\cref{collection-rules} that \theDeterminismChecker does not allow this type on
-any collection. While this is still sound, it will result in \theDeterminismChecker reporting a lot of errors because every locally declared 
-collection gets annotated with this invalid type. To avoid this situation, we make an exception to the defaulting rules for local collections
-and annotate their type parameters also with \<NonDet>.
-So a local list of Strings gets the default qualifier of \codeid{NonDet List<NonDet String>}.
-While this defaulting eliminates the possibility of automatically annotating locals collections with an invalid
-type, it could result in a high number of false positives. In the following code snippet,
-\begin{verbatim}
-void test(@Det List<@Det String> argList) {
-    List<String> localList = argList;
-}    
-\end{verbatim}
-\theDeterminismChecker would report an error at the assignment statement since the inferred type for \codeid{localList} is 
-\codeid{NonDet List<NonDet> String} and since collections types are invariant.
-We therefore recommend that programmers explicitly annotate
-all local variables that are collections to reduce false positives.
-
-\subsection{Precision improvements}\label{sec:precision-java}
+\subsection{The environment}\label{sec:environment-java}
 

### 2019-04-05 java-design.tex
Edits in section 3.6

@@
 corresponding environments.
-The user can specify input properties that must be passed on the \<java>
-command line and thus act like inputs to the program; these are also
-treated as deterministic.
+A user of \theDeterminismChecker can specify Java properties that must be passed on the \<java>
+command line and thus act like inputs to the program;
+\theDeterminismChecker treats these as deterministic.
 
-The inputs to a program are also treated as \<Det>.  That is, the type of
+The inputs to a program are also treated as deterministic.  That is, the type of
 the formal parameter to \<main> is

### 2019-04-05 java-design.tex
Spell-check

@@
 interface in which \<entrySet()> \<keySet()>, and \<values()> behave
-differently with respect to determinacy:  either all three methods, or none, are deterministic.
+differently with respect to determinism:  either all three methods, or none, are deterministic.
 
@@
 This flow-sensitive type refinement achieves local variable inference,
-freeing progammers from writing many local variable types.
+freeing programmers from writing many local variable types.
 
@@
 \<@Det String @Det []>, a deterministic array of deterministic strings.
+
+% LocalWords:  basetype NonEmpty NonDet OrderNonDet Det Iterable entrySet
+% LocalWords:  keySet LinkedHashMap myOndHashMap formedness PolyDet getenv
+% LocalWords:  expressivity getProperty

### 2019-04-05 randoop-case-study.tex
Move a figure
Spell-check

@@
 to compare, such as by removing or canonicalizing hash codes, dates, and
-other nondeterministing output (commits e15f9155--b473fd14).
+other nondeterministing output (commits e15f9155--32f72234).
 Each of these efforts was helpful, but they were not enough to solve the problem.
@@
 order-nondeterministic \<List> a \<Set>, for similar reasons:  deduplication was
-not always needed and would have increased costs.
+not always desired, and even where it was acceptable, it would have increased costs.
 
@@
 
-We annotated one package at a time, starting with the most central packges
+We annotated one package at a time, starting with the most central packages
 that are most depended upon.  Within a package, we followed a similar
@@
 
-\item \bugHashcodeOutput:
+\item[\bugHashcodeOutput]
   Diagnostic output printed a hash code for brevity.
@@
 
-\item \bugTimestampOutput:
+\item[\bugTimestampOutput]
   Diagnostic output printed a timestamp.
@@
 One is that we annotated some core components of Randoop to require all
-boolean expressions used in conditionals to be deterministic.  (This is an
+Boolean expressions used in conditionals to be deterministic.  (This is an
 option in \theDeterminismCheckerImplementation, which increases soundness;
@@
 developers had not.
+
+% LocalWords:  linters preprocess canonicalizing nondeterministing bdeebd
+% LocalWords:  e15f9155 b473fd14 deduplication c15ccbf2 ff5b4c4 eda87f src
+% LocalWords:  f72234 a8e63fb c975a9f7 e3c56 a4970 a460df97 postprocessing
+% LocalWords:  MinimizerTests SpecificationCollection findOverridden Det

### 2019-04-06 design.tex
typesetting rule

@@
 \begin{figure}
-    \todo{add line break}
-    $\infer[\rulename{set precision}]{\Gamma \vdash \|x.equals(y)| : \<Det> \ \|boolean|}{\Gamma \vdash \|x| : \<OrderNonDet> \ \|Set|\angles{ \kappa_{x} \ \beta_{x}}, \|y| : \<OrderNonDet> \ \|Set|\angles{ \kappa_{y} \ \beta_{y}} & \kappa_{x} \ \beta_{x} \nsqsubseteq \<OrderNonDet> \ \<List> & \kappa_{y} \ \beta_{y} \nsqsubseteq \<OrderNonDet> \ \<List> }$
-
+    \begin{align*}
+    \infer[\rulename{set precision}]{\Gamma \vdash \|x.equals(y)| : \<Det> \ \|boolean|}
+    {
+        &\Gamma \vdash \|x| : \<OrderNonDet> \ \|Set|\angles{ \kappa_{x} \ \beta_{x}}, \|y| : \<OrderNonDet> \ \|Set|\angles{ \kappa_{y} \ \beta_{y}} \\
+        & \kappa_{x} \ \beta_{x} \nsqsubseteq \<OrderNonDet> \ \<List>        \hspace{150pt}   \\
+        & \kappa_{y} \ \beta_{y} \nsqsubseteq \<OrderNonDet> \ \<List>     \hspace{150pt}    
+    }
+    \end{align*}
     \bigskip

### 2019-04-06 java-design.tex
Remove confusing text

@@
 
-\subsection{Dataflow analysis and type refinement}\label{sec:dataflow-java}
+\subsection{Type refinement via dataflow analysis}\label{sec:dataflow-java}
 
@@
 
-\todo{Mike doesn't understand the below.  An example would help clarify it.}
-
-In the example above, even though the local variable \codeid{y} was declared to be \codeid{NonDet}, it gets
-type refined to \codeid{Det} after the first assignment statement. As the result, the assignment of \codeid{y}
-to a \codeid{Det} variable \codeid{z} is sound. In the following example,
-\begin{verbatim}
-void foo(@NonDet Object x, @Det Object y) {
-    x = y;                // No type refinement for x.
-    @Det Object z = x;    // Error.
-}
-\end{verbatim}
-\theDeterminismChecker reports a warning at the second assignment statement of \codeid{foo}.
-Type refinement only applies to locals.
-Refining non-locals can result in side effects making the analysis unsound.
-
 \subsection{The environment}\label{sec:environment-java}

### 2019-04-06 randoop-case-study.tex
Explain fifth Randoop bug

@@
 
+\item[Moderate issues]
+  \strut
+\begin{description}
+\item[\bugHashMapOutput]
+While printing diagnostic output, Randoop iterated over a \<HashMap> in
+arbitrary order, making the output difficult to compare across different
+executions, or even to find information in.
+The class already implemented \<Comparable>, so
+the developers changed \<methodWeights.keySet()> to
+\codeid{new TreeSet<>(methodWeights.keySet()} in a \<for> loop (commit
+\todo{TO DO}).
+% This was actually introduced after the coding sprint.  Maybe some other
+% defects were too, so our criticism should be about regressions rather
+% than not finding them in the first place.
+\end{description}
+
 \item[Minor issues]

### 2019-04-06 introduction.tex
intro minor edits

@@
 to discover when a program has exhibited nondeterministic behavior on
-specific runs.  There is no need for a special JVM nor rerunning a program
+specific runs.  There is no need for a custom runtime system nor rerunning a program
 multiple times.
@@
 \begin{enumerate}
-  \item We designed a type system, named \ourTypeSystem, for expressing determinism properties.
+  \item We designed a type system for expressing determinism properties.
 
@@
   \item
-  \todo{It's a bit weird that we call this a contribution but it is never
-    explicitly discussed in the paper.  I think it's probably OK.}
+%  \todo{It's a bit weird that we call this a contribution but it is never
+%    explicitly discussed in the paper.  I think it's probably OK.}
   In a case study, we annotated several libraries, including the collection

### 2019-04-06 introduction.tex
intro citations

@@
 % ,Gyori:2015:RTD:2771783.2771793
-  tests~\cite{LuoHEM2014,nondex,deflaker,Rahman:2018:IFF:3236024.3275529}
+  tests~\cite{LuoHEM2014,nondex,deflaker,Rahman:2018:IFF:3236024.3275529,LuoHEM2014,Sudarshan}
   that sometimes pass and sometimes fail.  Flaky tests must be re-run, or
@@
   program to an original program.  This hinders debugging and maintenance,
-  and prevents use of techniques such as Delta Debugging~\cite{Zeller1999}.
+  and prevents use of techniques such as Delta Debugging~\cite{Zeller1999,YuLCZ2012}.
 %\item
@@
 at an earlier phase in the software development lifecycle would be greatly beneficial to
-developers.
+developers~\cite{briski2008minimizing}.
 

### 2019-04-06 randoop-case-study.tex
Describe false positives

@@
 the developers changed \<methodWeights.keySet()> to
-\codeid{new TreeSet<>(methodWeights.keySet()} in a \<for> loop (commit
-\todo{TO DO}).
+\codeid{new TreeSet<>(methodWeights.keySet())} in a \<for> loop (commit
+f212cc7e).
 % This was actually introduced after the coding sprint.  Maybe some other
@@
   Although the Randoop developers classified this issue as minor, it is
-  very similar to ones they fixed during their two-week coding sprint.
+  similar to ones they fixed during their two-week coding sprint.
 
@@
 
-\todo{Java 8 lambdas could help.}
+\TheDeterminismChecker issued 37 false positive warnings, or 1 every 650
+lines of code.
+
+The most common reasons (responsible for 67\% of false positive warnings)
+were:
+
+\begin{itemize}
+\item   (22\%)
+%      8 constructors guarantee all instances of this class are @Det
+  When every instance of a class is \<@Det>, \<@PolyDet> is equivalent to
+  \<@Det>, but \theDeterminismChecker does not yet recognize this fact.
+\item  (16\%)
+%      6 https://github.com/t-rasmud/checker-framework/issues/96
+  For a type that cannot be \<@OrderNonDet>, \<@PolyDet("up")> is
+  equivalent to \<@PolyDet>.
+\item (14\%)
+%      5 can't specify receiver for anonymous class
+  Java does not have syntax for specifying the type of an anonymous class.
+\item (11\%)
+%      4 actual type toString returns @PolyDet
+  A variable is declared as \<Object> which has nondeterministic \<toString>,
+  but at run time the value always has deterministic \<toString>.  For
+  example, the run-time value is always an \<Integer> or a \<String>.
+\item (5\%)
+%      2 iterating over @OrderNonDet collection to modify another
+  Iteration over an \<@OrderNonDet> collection in order to make another
+  \<@OrderNonDet> collection, but the extracted elements have type
+  \<@NonDet>.  For example, the following code is safe,
+  \theDeterminismChecker issues a warning.  Variable \<elt> has type
+  \<@NonDet>, so the call to \<add> does not typecheck.
+\begin{Verbatim}
+  void m(@OrderNonDet List<@Det String> input) {
+    @OrderNonDet List<@Det String> output = new List<>();
+    for (String elt : input) {
+      output.add(elt);
+    }
+  }
+\end{Verbatim}
+
+\end{itemize}
+
+The first two issues can be resolved by improvements to
+\theDeterminismCheckerImplementation.
+The third and fourth issues are limitations of the Java language, but by a
+local refactoring in the Randoop codebase can make the code typecheck.
+Refactoring can also eliminate the last warning.
 

### 2019-04-06 randoop-case-study.tex
Add more explanation

@@
   \<@OrderNonDet> collection, but the extracted elements have type
-  \<@NonDet>.  For example, the following code is safe,
+  \<@NonDet>.  For example, the following code is safe, but
   \theDeterminismChecker issues a warning.  Variable \<elt> has type
@@
     @OrderNonDet List<@Det String> output = new List<>();
-    for (String elt : input) {
-      output.add(elt);
+    for (String elt : input) {  // elt has type @NonDet String
+      output.add(elt);          // error: illegal argument
     }

### 2019-04-06 randoop-case-study.tex
Edits in 4.1-4.3

@@
 possible answers --- for example, in code coverage of generated tests.
-They used linters such as Error Prone, which warns if \<toString> is used on
-array objects, which don't override \<Object.toString> and therefore print a
+They used linters such as Error Prone to warn if \<toString> is used on
+objects, such as arrays, that don't override \<Object.toString> and therefore print a
 hash code which may vary from run to run.
@@
 to compare, such as by removing or canonicalizing hash codes, dates, and
-other nondeterministing output (commits e15f9155--32f72234).
+other nondeterministing output.
 Each of these efforts was helpful, but they were not enough to solve the problem.
@@
 In July 2017, the Randoop developers spent two weeks of full-time work to
-eliminate unintentional nondeterministic behavior in Randoop.
+eliminate unintentional nondeterministic behavior in Randoop (commits e15f9155--32f72234).
 Their methodology was to repeatedly
@@
  find the root cause of nondeterminism,
- and eliminate it.
+ and eliminate it (personal communication, 2019).
 Some of the nondeterminism was in libraries, such as the JDK\@.
@@
 In brief, we wrote type qualifiers in the Randoop source code to express its
-determinism specification
-We then ran
+determinism specification,
+then ran
 \theDeterminismChecker.  Each warning indicated a mismatch between the
-specification and the implementation, which we addressed by changing our
-specification or reporting a bug in the implementation.
+specification and the implementation.  We addressed each warning by changing our
+specification, reporting a bug in Randoop, or suppressing a false positive warning.
 
@@
 Sometimes after we
-annotated a subclass, we had to go back and change existing annotation on
-the supertype or elsewhere because the subclass implementation was
+annotated a subclass, we had to go back and change annotations on
+the supertype because the subclass implementation was
 inconsistent with the specification we had guessed for the supertype.
@@
 classpath passed on the command line.
-This could have led to incorrect behavior, both in Randoop's test suite and in the field,
-if a user set the environment variable.
-The developers fixed it by changing Randoop to not read the environment variable
+This can cause to incorrect behavior, both in Randoop's test suite and in the field,
+if a user sets the environment variable.
+The developers fixed the problem by changing Randoop to not read the environment variable
 (commit 330e3c56, shown in \cref{fig:randoop-bug-getenv}).
@@
   The developers may not have noticed this because their log-postprocessing
-  tools removed them from the log when doing comparisons.
+  tools removed timestamps from the log when doing comparisons.
 

### 2019-04-06 related.tex
structure related work

@@
 
-The state of the art is NonDex~\cite{nondex}.  It is a highly effective tool.
+The state of the art for detecting nondeterministic tests is NonDex~\cite{nondex}.  It is a highly effective tool.
 % https://github.com/TestingResearchIllinois/NonDex/wiki/Supported-APIs
@@
 the use of async await, concurrency, or due to test order dependency.~\cite{Plotkin:1993:LPP:645891.671433}
-perform an empirical study on open source benchmarks that concur with the results of~\cite{LuoHEM2014}.
+perform an empirical study on open source benchmarks, the results of which concur with those of~\cite{LuoHEM2014}.
 Additionally, the authors of this paper study the effects of removing these test smells. Their results suggest that
-refactoring the test code smells fixed more than half of the identified flaky tests.
+refactoring the test code smells fixed more than half of the identified flaky tests. Our approach is complementary to
+these techniques and aims to prevent nondeterminism from causing harmful effects.
 
-\cite{Jin:2012:UDR:2254064.2254075} study the effect of inefficient code sequences on the performance of software.
+The problem of inconsistencies between specifications and implementations
+is quite prevalent.~\cite{Jin:2012:UDR:2254064.2254075} study the effect of inefficient code sequences on the performance of software.
 Their findings suggest that incorrect understanding of performance specifications of APIs is a very common root cause for
@@
 \cite{Dan:2013:10.1007/978-3-642-39038-8_25,Vahabzadeh:2015:7332456} focus on differentiating bugs due to tests
-from those caused by source code.
+from those caused by source code. While we haven't fully accounted for all nondeterministic sources due
+to the environment, our technique is easily extensible by adding annotations to the corresponding classes.
 

### 2019-04-06 randoop-case-study.tex

@@
 it was a bug or a false positive warning.
+The effort would have been much easier for someone familiar with Randoop,
+and yet easier if done while code is being written and is malleable.
 
@@
   \theDeterminismChecker issues a warning.  Variable \<elt> has type
-  \<@NonDet>, so the call to \<add> does not typecheck.
+  \<@NonDet>, so the call to \<add> does not type check.
 \begin{Verbatim}
@@
 The third and fourth issues are limitations of the Java language, but by a
-local refactoring in the Randoop codebase can make the code typecheck.
+local refactoring in the Randoop codebase can make the code type check.
 Refactoring can also eliminate the last warning.
@@
 
-\todo{Report the number of annotations required, and say a few words about
-  the process and the effort required.}
+In Randoop's 24K LOC of code, we wrote
+100 \<@NonDet>,
+64 \<@OrderNonDet>,
+1391 \<@Det>,
+and 109 \<@PolyDet> annotations.
 
-\todo{Easier for someone familiar with the codebase.  Easier on new code.}
-
-It's higher than we would prefer.
+The number of \<@Det> annotations --- one per 17 lines of code --- is much
+higher than we would prefer.
+Nonetheless, it compares favorably to the
+extensive effort by the Randoop developers (\cref{sec:subject-program}),
+especially given that that \theDeterminismChecker found issues that the
+developers had not.
 
@@
 \<@PolyDet> annotation) is the default for
-formal parameter and return types.  We plan to re-do the annotation effort
-from scratch, and the number of annotations will be much lower.
+formal parameter and return types.
 
@@
 
-Even though the number of annotations is higher for our prototype
-implementation than we would prefer, it still compares favorably to the
-extensive effort by the Randoop developers (\cref{sec:subject-program}),
-especially given that that \theDeterminismChecker found issues that the
-developers had not.
+In future work, we plan to implement qualifier polymorphism for classes,
+then re-do the case study.  We estimate this will reduce the number of
+\<@Det> annotations needed by about an order of magnitude.
 
@@
 % LocalWords:  MinimizerTests SpecificationCollection findOverridden Det
+% LocalWords:  methodWeights keySet f212cc7e PolyDet OrderNonDet NonDet i7
+% LocalWords:  elt LOC

### 2019-04-06 design.tex
Remove duplication
Tweaks in 2.1-2.2

@@
 
-This section presents the key aspects of our type system,
-called \ourTypeSystem, in the context of a core calculus for an
+This section presents the key aspects of our type system
+% , called \ourTypeSystem,
+in the context of a core calculus for an
 object-oriented language.
@@
 
-In a type system,
-a \textit{type} represents a set of values.  It abstracts or restricts the
-set of possible run-time values that an expression may evaluate to.
+One way to view a type is as
+a set of values.  A type abstracts or restricts the
+set of possible run-time values that an expression may evaluate to
+and the operations that may be performed.
 A programming language provides \emph{basetypes}, such as \<Int>.
@@
 \OurTypeSystem checks all the standard typing rules including the ones
-shown in \cref{typecheck-rules-standard}.  For \ourTypeSystem, $\tau$
-represents a qualified type rather than a basetype.
+shown in \cref{typecheck-rules-standard}, which
+shows a sample of standard typing rules for
+an object-oriented programming language.
+% For \ourTypeSystem, $\tau$
+% represents a qualified type rather than a basetype.
 
@@
 As a result, a qualifier type system does not allow any values that the
-original type system does not.
+original type system does not, in the same program without qualifiers.
 However, the qualifier type system may reject more programs, and thus
@@
 Type qualifier systems are sometimes defined independently of the
-underlying type system.  Because there are interactions between the
-basetypes and the type qualifiers, defining them together improves
+underlying type system.  In \ourTypeSystem, there are interactions between the
+basetypes and the type qualifiers.  Defining them together improves
 precision, which is important in practice.
@@
 %  }
-
-\Cref{typecheck-rules-standard} shows a sample of standard typing rules for
-an object-oriented programming language.
 
@@
 the element types can be specified independently of the collection type,
-including qualifiers.
+including their qualifiers.
 

### 2019-04-06 randoop-case-study.tex
Start adding case study comments

--- 2019-04-06 randoop-case-study.tex
@@
 we applied it to the Randoop test
-generator~\ifanonymous\citepalias{randoop-tool}\else\cite{PachecoLEB2007}\fi.
-We also wrote specifications for libraries Randoop uses, such as the JDK, JUnit,
+generation tool~\ifanonymous\citepalias{randoop-tool}\else\cite{PachecoLEB2007}\fi.
+We also wrote specifications for libraries it uses, such as the JDK, JUnit,
 and others.
@@
 all the case study materials are publicly available at \myurl{\todo{url for
-    case study}} for reproducibility.
+    case study}} for reproducibility. \todo{This should include our link to
+      our annotated plume-util, which we found was required to reproduce the
+    results of the case study}
 
@@
 to compare, such as by removing or canonicalizing hash codes, dates, and
-other nondeterministing output (commits e15f9155--32f72234).
+other nondeterministing output (commits e15f9155--b473fd14).
 Each of these efforts was helpful, but they were not enough to solve the problem.
@@
 The most common causes were \<toString> routines and iteration order of sets and maps.
-The most common fixes were to change the implementation of \<toString>
+The most common fixes were to change the implementations of \<toString>
 and to use \<LinkedHashSet> and \<LinkedHashMap> or to sort collections
@@
 order-nondeterministic \<List> a \<Set>, for similar reasons:  deduplication was
-not always desired, and even where it was acceptable, it would have increased costs.
+not always needed and would have increased costs.
 
@@
 In brief, we wrote type qualifiers in the Randoop source code to express its
-determinism specification
+determinism specification.
 We then ran
@@
 
-We annotated one package at a time, starting with the most central packages
+We annotated one package at a time, starting with the most central packges
 that are most depended upon.  Within a package, we followed a similar
@@
 
-\item[Moderate issues]
-  \strut
-\begin{description}
-\item[\bugHashMapOutput]
-While printing diagnostic output, Randoop iterated over a \<HashMap> in
-arbitrary order, making the output difficult to compare across different
-executions, or even to find information in.
-The class already implemented \<Comparable>, so
-the developers changed \<methodWeights.keySet()> to
-\codeid{new TreeSet<>(methodWeights.keySet())} in a \<for> loop (commit
-f212cc7e).
-% This was actually introduced after the coding sprint.  Maybe some other
-% defects were too, so our criticism should be about regressions rather
-% than not finding them in the first place.
-\end{description}
-
 \item[Minor issues]
@@
 
-\item[\bugHashcodeOutput]
+\item \bugHashcodeOutput:
   Diagnostic output printed a hash code for brevity.
@@
   Although the Randoop developers classified this issue as minor, it is
-  similar to ones they fixed during their two-week coding sprint.
+  very similar to ones they fixed during their two-week coding sprint.
 
-\item[\bugTimestampOutput]
+\item \bugTimestampOutput:
   Diagnostic output printed a timestamp.
@@
 
-\TheDeterminismChecker issued 37 false positive warnings, or 1 every 650
-lines of code.
-
-The most common reasons (responsible for 67\% of false positive warnings)
-were:
-
-\begin{itemize}
-\item   (22\%)
-%      8 constructors guarantee all instances of this class are @Det
-  When every instance of a class is \<@Det>, \<@PolyDet> is equivalent to
-  \<@Det>, but \theDeterminismChecker does not yet recognize this fact.
-\item  (16\%)
-%      6 https://github.com/t-rasmud/checker-framework/issues/96
-  For a type that cannot be \<@OrderNonDet>, \<@PolyDet("up")> is
-  equivalent to \<@PolyDet>.
-\item (14\%)
-%      5 can't specify receiver for anonymous class
-  Java does not have syntax for specifying the type of an anonymous class.
-\item (11\%)
-%      4 actual type toString returns @PolyDet
-  A variable is declared as \<Object> which has nondeterministic \<toString>,
-  but at run time the value always has deterministic \<toString>.  For
-  example, the run-time value is always an \<Integer> or a \<String>.
-\item (5\%)
-%      2 iterating over @OrderNonDet collection to modify another
-  Iteration over an \<@OrderNonDet> collection in order to make another
-  \<@OrderNonDet> collection, but the extracted elements have type
-  \<@NonDet>.  For example, the following code is safe, but
-  \theDeterminismChecker issues a warning.  Variable \<elt> has type
-  \<@NonDet>, so the call to \<add> does not typecheck.
-\begin{Verbatim}
-  void m(@OrderNonDet List<@Det String> input) {
-    @OrderNonDet List<@Det String> output = new List<>();
-    for (String elt : input) {  // elt has type @NonDet String
-      output.add(elt);          // error: illegal argument
-    }
-  }
-\end{Verbatim}
-
-\end{itemize}
-
-The first two issues can be resolved by improvements to
-\theDeterminismCheckerImplementation.
-The third and fourth issues are limitations of the Java language, but by a
-local refactoring in the Randoop codebase can make the code typecheck.
-Refactoring can also eliminate the last warning.
+\todo{Java 8 lambdas could help.}
 
@@
 One is that we annotated some core components of Randoop to require all
-Boolean expressions used in conditionals to be deterministic.  (This is an
+boolean expressions used in conditionals to be deterministic.  (This is an
 option in \theDeterminismCheckerImplementation, which increases soundness;
@@
 developers had not.
-
-% LocalWords:  linters preprocess canonicalizing nondeterministing bdeebd
-% LocalWords:  e15f9155 b473fd14 deduplication c15ccbf2 ff5b4c4 eda87f src
-% LocalWords:  f72234 a8e63fb c975a9f7 e3c56 a4970 a460df97 postprocessing
-% LocalWords:  MinimizerTests SpecificationCollection findOverridden Det

### 2019-04-06 conclusion.tex

@@
 
-\todo{TODO}
-
 We designed a type system that helps programmers express determinism specifications
-in sequential programs. To the best of our knowledge, ours is the
-first compile time verification approach addressing the problem of nondeterminism.
+in sequential programs. 
 We implemented our type system in Java, applied it to real world 
@@
 problem of nondeterminism. We compared our tool with existing state of the art flake test detectors
-and found that \TheDeterminismChecker found all the bugs arising due to nondeterminism
+and found that \theDeterminismChecker found all the bugs arising due to nondeterminism
 in sequential programs that were used for evaluation in NonDex~\cite{nondex} and DeFlaker~\cite{deflaker}.
+To the best of our knowledge, ours is the
+first compile time verification approach addressing the problem of nondeterminism.

### 2019-04-06 paper.tex

@@
 Nondeterminism can occur even in sequential programs, for example as a
-result of iterating over the elements of a hash table.
+result of iterating over the elements of a hash table. This seemingly innocuous and
+frequently used operation can result in diverging test results
+if the test depends on iteration order for asserting correctness.
 
-We have created a type system for determinism.  A type checker warns if a
+%We have created a type system for determinism.  
+We have created a type system that can express determinism specifications
+in a program.
+A type checker warns if a
 program is nondeterministic or verifies that the program is deterministic.
@@
 restrict the typings for collections; and enhancements to polymorphism that
-improve precision when analyzing collection operations.
+improve precision when analyzing collection operations. While state of the art
+nondeterminism detection tools rely on observing runtime output, our approach
+aims to verify determinism thereby providing stronger soundness guarantees.
 
-We implemented our type system for Java.  In a case study, it found
+We implemented our type system for Java.  In a case study of a production software that spans
+approximately 24K lines of code, it found
 previously-unknown nondeterminism errors in a program that had been heavily

### 2019-04-06 related.tex

@@
 \section{Related work}\label{sec:related}
-
-\todo{TODO}
 
@@
 researchers and software developers alike~\cite{Fowler,Sudarshan}.
-\cite{LuoHEM2014} study 51 open source projects and analyze the most common root causes of
+In \cite{LuoHEM2014}, the authors studied 51 open source projects and analyzed the most common root causes of
 test flakiness. The findings in the paper suggest that most of the flakiness in tests is caused by
-the use of async await, concurrency, or due to test order dependency.~\cite{Plotkin:1993:LPP:645891.671433}
+the use of async await, concurrency, or due to test order dependency. The authors of~\cite{Plotkin:1993:LPP:645891.671433}
 perform an empirical study on open source benchmarks, the results of which concur with those of~\cite{LuoHEM2014}.
-Additionally, the authors of this paper study the effects of removing these test smells. Their results suggest that
+Additionally,~\cite{Plotkin:1993:LPP:645891.671433} studies the effects of removing these test smells. The results suggest that
 refactoring the test code smells fixed more than half of the identified flaky tests. Our approach is complementary to
@@
 The problem of inconsistencies between specifications and implementations
-is quite prevalent.~\cite{Jin:2012:UDR:2254064.2254075} study the effect of inefficient code sequences on the performance of software.
-Their findings suggest that incorrect understanding of performance specifications of APIs is a very common root cause for
-performance bugs. ~\cite{Rui:2013:180377} analyze the APIs related to security such as authorization and authentication.
+is quite prevalent.~\cite{Jin:2012:UDR:2254064.2254075} looks at the effect of inefficient code sequences on the performance of software.
+Its findings suggest that incorrect understanding of performance specifications of APIs is a very common root cause for
+performance bugs. ~\cite{Rui:2013:180377} analyzes the APIs related to security such as authorization and authentication.
 These APIs which come as a part of the SDKs provided by online providers are used by application developers to 
 build secure apps. The paper discovers that the security guarantees provided by these SDKs are violated due to
-the implicit assumptions made by the APIs that the users are unaware of. \cite{Xiao:2014:NMC:2591062.2591177}
+the implicit assumptions made by the APIs that the users are unaware of. In \cite{Xiao:2014:NMC:2591062.2591177},
+the authors
 study the effect of nondeterminism in MapReduce programs with a specific focus on nondeterminism caused by
@@
 
-\cite{Petrenko1996,Petrenko:1993:NSM:648128.761244,Savor:1997:639710,Hierons:2004:TCD:1040993.1040998} propose several techniques to test whether a
-deterministic implementation conforms to its non-deterministic finite state machine.
-\cite{Cook:2013:RNP:2491956.2491969} present an approach that can automatically verify properties
+Several techniques have been proposed to test whether a
+deterministic implementation conforms to its non-deterministic finite state machine~\cite{Petrenko1996,Petrenko:1993:NSM:648128.761244,Savor:1997:639710,Hierons:2004:TCD:1040993.1040998}.
+\cite{Cook:2013:RNP:2491956.2491969} presents an approach that can automatically verify properties
 in branching time temporal logic systems that are inherently non-deterministic.
@@
 Failing tests that are unrelated to code changes can be expensive both in terms of monetary costs and
-developer effort. \cite{Herzig:2015:EDF:2819009.2819018} propose techniques to classify tests as false alarms
+developer effort. \cite{Herzig:2015:EDF:2819009.2819018} proposes techniques to classify tests as false alarms
 if they are known to be caused by testing infrastructure or other environment issues.~\cite{Huo:2014:IOQ:2635868.2635917}
-propose an approach that detects brittle assertions in tests by performing a taint analysis on inputs classified as 
-controlled and uncontrolled.~\cite{ZhangJWMLEN2014} investigate the effects of test independence assumption
+presents an approach that detects brittle assertions in tests by performing a taint analysis on inputs classified as 
+controlled and uncontrolled.~\cite{ZhangJWMLEN2014} investigates the effects of test independence assumption
 on other techniques such as test prioritization, selection, etc. Other approaches~\cite{BellKMD2015,Gyori:2015:RTD:2771783.2771793}
 analyze test dependencies and either prevent them or use this information for other optimizations.
-\cite{Dan:2013:10.1007/978-3-642-39038-8_25,Vahabzadeh:2015:7332456} focus on differentiating bugs due to tests
+The approaches in \cite{Dan:2013:10.1007/978-3-642-39038-8_25,Vahabzadeh:2015:7332456} focus on differentiating bugs due to tests
 from those caused by source code. While we haven't fully accounted for all nondeterministic sources due

### 2019-04-06 design.tex
Fix font size
Edits in 2.3

@@
     \caption{Type well-formedness}
-\todo{What if $\beta$ has \<List> within?  Be sure everything works in that case.}
 
@@
 The element type of a \<NonDet> collection must be \<NonDet>.
-Otherwise, the type qualifier on a collection's element must be a subtype or equal to
+For other collections, the type qualifier on a collection's element must be a subtype or equal to
 the collection's type qualifier (\rulename{collection} rules of
@@
 These types are invalid:
-\begin{smaller}
 \begin{Verbatim}[commandchars=\\\{\}]
@@
 \end{Verbatim}
-\end{smaller}
 \end{minipage}
@@
 affects access, iteration, searching, etc.
-(Rules \rulename{iteration} and \rulename{next element} in \cref{fig-ordernondet-rules}.)
+(Rules \rulename{get}, \rulename{iteration}, and \rulename{next element} in \cref{fig-ordernondet-rules}.)
 \item
@@
 typical type for a list access operation, such as \<get>, is
-$$\forallt{\tau} \<List>\angles{\tau} \times \<Det Int> \rightarrow \tau$$
+$$\forallt{\tau} \<List>\angles{\tau} \times \<Int> \rightarrow \tau$$
 but this type is incorrect for \ourTypeSystem.  Each of the following is a
@@
     $
-    \forallt{\tau} \kappa \in \{\<Det>, \<NonDet>\}\ \<OrderNonDet> \ C \angles{\tau} \ \times \kappa \ \<Int> \rightarrow 
+    \forallt{\tau\ \kappa}
+    % Not necessary: invalid instantiantions are not created
+    % \in \{\<Det>, \<NonDet>\}\
+    \<OrderNonDet> \ C \angles{\tau} \ \times \kappa \ \<Int> \rightarrow 
     \<NonDet> \ \<Int>
@@
     $
-    \forallt{\kappa, \beta} \ \<OrderNonDet> \ C \angles{\kappa \ \beta} \ \times \<Det> \ \<Int> \times \kappa \ \beta_n \rightarrow 
+    \forallt{\kappa\ \beta} \ \<OrderNonDet> \ C \angles{\kappa \ \beta} \ \times \<Det> \ \<Int> \times \kappa \ \beta \rightarrow 
     \<NonDet> \ \beta
@@
 
-    \rulename{set invalid}: 
-    $
-    \forallt{\kappa, \beta} \ \<OrderNonDet> \ C \angles{\kappa \ \beta} \ \times \<NonDet> \ \<Int> \times \kappa \ \beta_n \rightarrow 
-    \|INVALID|
+%% We don't need this.  Just leaving it out means the expression has no
+%% typing and is therefore invalid.
+%     \rulename{set invalid}: 
+%     $
+%     \forallt{\kappa, \beta} \ \<OrderNonDet> \ C \angles{\kappa \ \beta} \ \times \<NonDet> \ \<Int> \times \kappa \ \beta \rightarrow 
+%     \|INVALID|
+%     $
+
+    \rulename{size}:
+    $
+    \forallt{\tau} \<OrderNonDet>\ C\angles{\tau} \rightarrow
+    \<Det> \ \<int>
     $
@@
     $
-    \bigskip
 
@@
 %    }$
-
-    \rulename{size}:
-    $
-    \forallt{\tau} \<OrderNonDet>\ C\angles{\tau} \rightarrow
-    \<Det> \ \<int>
-    $
 

### 2019-04-06 randoop-case-study.tex
Updated annotation counts

@@
 In Randoop's 24K LOC of code, we wrote
-100 \<@NonDet>,
+98 \<@NonDet>,
 64 \<@OrderNonDet>,
-1391 \<@Det>,
-and 109 \<@PolyDet> annotations.
+1083 \<@Det>,
+and 103 \<@PolyDet> annotations.
 
-The number of \<@Det> annotations --- one per 17 lines of code --- is much
+The number of \<@Det> annotations --- one per 22 lines of code --- is much
 higher than we would prefer.
@@
 The Checker Framework does not support qualifier polymorphism over
-classes.  This means that fields must be annotated as deterministic.  Our annotations in
+classes.  This means that fields must be annotated as deterministic.
+\<@Det> annotations must be added to methods and constructors that modify these
+fields, as well as any methods that call these.
+Our annotations in
 Randoop use qualifier polymorphism extensively on methods but cannot use it
@@
 then re-do the case study.  We estimate this will reduce the number of
-\<@Det> annotations needed by about an order of magnitude.
+\<@Det> annotations needed by about an order of magnitude. \todo{Do we plan on
+doing other case studies as well?}
 

### 2019-04-06 design.tex
Remove repetition

@@
 \OurTypeSystem supports parametric
-polymorphism~\cite{Abadi:1989:FIM:77350.77373,Plotkin:1993:LPP:645891.671433}
-over both classes and methods.
+polymorphism~\cite{Abadi:1989:FIM:77350.77373,Plotkin:1993:LPP:645891.671433}.
 A polymorphic abstraction (a class or method) is written and
@@
 When both are applied, it gives the effect of type polymorphism.
-Polymorphism applies to both classes and methods.
 

### 2019-04-06 design.tex
Edits on pages 6 and 7

@@
   Basetype polymorphism is rarely used on its own.  Full type polymorphism is more
-  common, even if the type is decomposed and the parts used independently,
+  common, even if the function type decomposes the type parameter and uses the parts independently,
   as in this more general type for the \<choose> function:
@@
 
+\noindent
 The body of \<length> must typecheck at every possible instantiation of its
@@
 the most specific applicable instantiation is chosen.
-Here is an example:
-
-Suppose we have two variables $\<ni> : \<NonDet Int>$ and $\<di> : \<Det Int>$.
+
+For example,
+suppose we have two variables $\<ni> : \<NonDet Int>$ and $\<di> : \<Det Int>$.
 This table shows, for several method invocations, the 
@@
 instantiation would never be the most specific applicable one.
-This is important for some of the type rules given in this paper, which
+The general rule about instantiations and type validity
+is important for some of the type rules given in this paper, which
 would otherwise need complex preconditions that recapitulate the type
@@
 $\kappa = \<NonDet>$ and $\kappa = \<Det>$.  An invocation on an
-\<OrderNonDet> collection would choose the \<NonDet> instantiation and the
+\<OrderNonDet> collection would choose the \<NonDet> instantiation, and the
 return type at that call site would be \<NonDet Int> rather than the

### 2019-04-06 paper.tex

--- 2019-04-04 paper.tex
@@
 When a program is nondeterministic, it is difficult to test and debug.
-Nondeterminism can occur even in sequential programs, for example as a
-result of iterating over the elements of a hash table.
+Nondeterminism occurs even in sequential programs: for example, as a
+result of iterating over the elements of a hash table. This seemingly innocuous and
+frequently used operation can result in diverging test results
+if the test depends on iteration order for asserting correctness.
 
-We have created a type system for determinism.  A type checker warns if a
-program is nondeterministic or verifies that the program is deterministic.
+%We have created a type system for determinism.  
+We have created a type system that expresses determinism specifications
+in a program.
 The key ideas in the type system are type qualifiers for nondeterminism,
@@
 restrict the typings for collections; and enhancements to polymorphism that
-improve precision when analyzing collection operations.
+improve precision when analyzing collection operations. While state-of-the-art
+nondeterminism detection tools rely on observing runtime output, our approach
+aims to verify determinism thereby providing stronger soundness guarantees.
 
-We implemented our type system for Java.  In a case study, it found
+We implemented our type system for Java.
+Our type checker, \theDeterminismChecker, warns if a
+program is nondeterministic or verifies that the program is deterministic.
+In a case study of a 24,000 line software project, it found
 previously-unknown nondeterminism errors in a program that had been heavily
-vetted by its developers.  In an experiment, it found all of the
-non-concurrency-related nondeterminism that was found by other tools.
+vetted by its developers,
+who were unusually concerned about nondeterminism errors.
+In an experiment, it found all of the
+non-concurrency-related nondeterminism that was found by
+state-of-the-art dynamic approaches to detecting flaky tests.
 

### 2019-04-06 randoop-case-study.tex
Addd case study comments

@@
 all the case study materials are publicly available at \myurl{\todo{url for
-    case study}} for reproducibility. \todo{This should include our link to
+    case study}} for reproducibility. \todo{This should include our
       our annotated plume-util, which we found was required to reproduce the
@@
 specification, largely from the methods it calls.  If a method $m$ calls a
-method that we had previously annotated as nondeterministic, we annotated
-$m$ as nondeterministic.
+method that we had previously annotated as returning a nondeterministic result, we annotated
+$m$ as returning a nondeterministic result.
 Sometimes after we
@@
 it was a bug or a false positive warning.
+\todo{In some cases, the number of \<@Det> annotations was especially prevalent.
+  In these cases we changed the default qualifier for that class to \<@Det>.}
 The effort would have been much easier for someone familiar with Randoop,
@@
 One use of \<HashSet> could cause a problem if a type variable's lower or upper
-bound has a type parameter that the type variable itself does not have.
+bound in the code Randoop is run on has a type parameter that the type variable itself does not have.
 This situation does occur, even in Randoop's test suite.
@@
 %      5 can't specify receiver for anonymous class
-  Java does not have syntax for specifying the type of an anonymous class.
+  Java does not have syntax for specifying the receiver type in anonymous class
+  methods.
 \item (11\%)
@@
   A variable is declared as \<Object> which has nondeterministic \<toString>,
-  but at run time the value always has deterministic \<toString>.  For
+  but at run time the subtype of \<Object> stored always has deterministic \<toString>.  For
   example, the run-time value is always an \<Integer> or a \<String>.
@@
 \theDeterminismCheckerImplementation.
-The third and fourth issues are limitations of the Java language, but by a
+The third and fourth issues are limitations of the Java language, but a
 local refactoring in the Randoop codebase can make the code type check.
-Refactoring can also eliminate the last warning.
+Refactoring can also eliminate the last warning. \todo{Can it? I thought
+  what happens when iterating over an @OrderNonDet collection is something we
+can't verify with the checker.}
 
@@
 
-In Randoop's 24K LOC of code, we wrote
+In Randoop's 24K lines of code, we wrote
 98 \<@NonDet>,

### 2019-04-06 design.tex

@@
 
-Suppose that $\<myStrings> : \ptype{OrderNonDet List}{Det String}$.
-The type of \<first(myStrings)> should be \<NonDet String>.
+Suppose that $\<myOndStrings> : \ptype{OrderNonDet List}{Det String}$.
+The type of \<first(myOndStrings)> should be \<NonDet String>.
 
@@
 would be invalid.
-At the call \<first(myStrings)>, the most specific instantiation would be
+At the call \<first(myOndStrings)>, the most specific instantiation would be
 
@@
 
-The rules prevent some invocations.  Suppose we have the following variable:
-
-$\<listOfLists> : \ptype{OrderNonDet List}{\ptype{Det List}{Det String}}$
+The given type prevents some invocations.  Suppose we have the following variable:
+
+$\<listOfLists> : \ptype{OrderNonDet List}{\lptype{Det List}{Det String}}$
 
@@
 
-\todo{TODO: Need to work on this section.}
-
-In the annotated JDK, the type of get is:
-
-\begin{Verbatim}
-    @PolyDet("up") E get(@PolyDet List<E> this, @PolyDet int index);
-\end{Verbatim}
+This section introduces the another polymorphism feature by contrasting the
+types of the \<get> and \<set> operations on lists.
 
@@
 
-$\<get> : \forallt{\kappa} \kappa\ \<List>\angles{\kappa\ \<String>} \times \kappa\ \<Int> \rightarrow \kappa\ \<String>$
-\todo{Used to be $\up{\kappa}$ for return type, but I don't think that is needed.}
-\todo{Oh, maybe it's because we need to be able to instantiate in the tye
-  parameter differently?  How do we enable that?}
-
-(The actual typing permits the qualifier on the type argument to differ
-from the type qualifier on the list, and it handles arbitrary basetypes
-including collections that may themselves be order-nondeterministic.)
-
-Instantiations are:
+$\<get> : \forallt{\kappa} \kappa\ \<List>\angles{\kappa\ \<String>} \times \kappa\ \<Int> \rightarrow \up{\kappa}\ \<String>$
+
+\noindent
+Its instantiations are:
 
@@
 
-\todo{We don't need the \<OrderNonDet> instantiation.  The other two
-  instantiations are enough.
-  We don't need the following instantiation:}
-\begin{Verbatim}
-   // instantiation: OrderNonDet List<Det String> x _ Int -> NonDet String 
-\end{Verbatim}
-
-If either the list or the index is (order) non-deterministic, then the result is
-nondeterministic:
-
-\begin{Verbatim}
-NonDet Int ni;
-   Det Int di;
-NonDet List<NonDet String> nl;
-OrderNonDet List<Det String> ol;
-Det List<Det String> dl;
-
-get(nl, ni) : NonDet String // instantiation: NonDet List<NonDet String> x NonDet Int -> NonDet String
-get(nl, di) : NonDet String
-get(ol, ni) : NonDet String
-get(ol, di) : NonDet String
-get(dl, ni) : NonDet String
-get(dl, di) : Det String    // instantiation: NonDet List<Det String> x Det Int -> Det String
-\end{Verbatim}
+\noindent
+It cannot be instantiated at $\kappa = \<OrderNonDet>$, but the other two
+instantiations are enough to ensure that if either the list or the index is (order) non-deterministic, then the result is
+nondeterministic.
+
+% \begin{Verbatim}
+% NonDet Int ni;
+%    Det Int di;
+% NonDet List<NonDet String> nl;
+% OrderNonDet List<Det String> ol;
+% Det List<Det String> dl;
+% 
+% get(nl, ni) : NonDet String // instantiation: NonDet List<NonDet String> x NonDet Int -> NonDet String
+% get(nl, di) : NonDet String
+% get(ol, ni) : NonDet String
+% get(ol, di) : NonDet String
+% get(dl, ni) : NonDet String
+% get(dl, di) : Det String    // instantiation: NonDet List<Det String> x Det Int -> Det String
+% \end{Verbatim}
 

### 2019-04-06 design.tex
Finish draft of section 2.4.3

@@
 
-This section introduces the another polymorphism feature by contrasting the
+This section introduces another polymorphism feature by contrasting the
 types of the \<get> and \<set> operations on lists.
@@
 
-$\<get> : \forallt{\kappa} \kappa\ \<List>\angles{\kappa\ \<String>} \times \kappa\ \<Int> \rightarrow \up{\kappa}\ \<String>$
-
-\noindent
+$\<get> : \forallt{\kappa} \kappa\ \<List>\angles{\down{\kappa}\ \<String>} \times \up{\kappa}\ \<Int> \rightarrow \up{\kappa}\ \<String>$
+
+\noindent
+(The actual type permits the qualifier on the type argument to differ
+from the type qualifier on the list, and it better handles
+order-nondeterminism, but this type is enough for illustration in this section.)
+
+\todo{Show actual type.}
+
 Its instantiations are:
@@
 
+$\<get> : \<OrderNonDet List>\angles{\<Det String>} \times \<NonDet Int> \rightarrow \<NonDet String>$
+
 $\<get> : \<Det List>\angles{\<Det String>} \times \<Det Int> \rightarrow \<Det String>$
@@
 \noindent
-It cannot be instantiated at $\kappa = \<OrderNonDet>$, but the other two
-instantiations are enough to ensure that if either the list or the index is (order) non-deterministic, then the result is
-nondeterministic.
+The type of \<get> ensures that if either the list or the index is (order)
+non-deterministic, then the result of the call is nondeterministic.
 
@@
 A similar typing does \emph{not} work for the \<set> function that sets an
-element of a list, because it would permit a deterministic collection to
-be side-effected in non-deterministic ways.  Consider the following
-possible type for
-
-$\<set> : \forall \forallt{\tau}$
-
-qualifier 
-on the List add method: \codeid{public void add(PolyDet ArrayList<E> this, PolyDet int index, E element)}.
+element of a list.  Consider the putative type
+
+$\<set> : \forallt{\kappa} \kappa\ \<List>\angles{\down{\kappa}\ \<String>} \times \up{\kappa}\ \<Int> \times \down{\kappa}\ \<String> \rightarrow ()$
+
+\noindent
+which has instantiations
+
+$\<set> : \<NonDet>\ \<List>\angles{\<NonDet>\ \<String>} \times \<NonDet>\ \<Int> \times \<NonDet>\ \<String> \rightarrow ()$
+
+$\<set> : \<OrderNonDet>\ \<List>\angles{\<Det>\ \<String>} \times \<NonDet>\ \<Int> \times \<Det>\ \<String> \rightarrow ()$
+
+$\<set> : \<Det>\ \<List>\angles{\<Det>\ \<String>} \times \<Det>\ \<Int> \times \<Det>\ \<String> \rightarrow ()$
+
+\noindent
 Suppose a client calls this method as follows:
-\begin{verbatim}
-@Det ArrayList<@Det String> list;
-@NonDet int random;
-@Det String str;
-list.add(random, str);
-\end{verbatim}
-This will result in the instantiation \codeid{public void add(@NonDet ArrayList<E> this, @NonDet int index, E element)} and
-the method invocation will type check. This is problematic because even though \<list> was declared to be \<Det>,
-the type checker allows the addition of an element at a \<NonDet> location violating our determinism guarantees.
-We prevent this behavior by introducing another variant of \<PolyDet>, \<PolyDet("use")> that does not affect
-type instantiations. This is especially useful in preventing methods from non-deterministically modifying the state
-of a deterministic receiver. To have this effect, the library method \<ArrayList.add> is annotated as:
-\codeid{public void add(PolyDet ArrayList<E> this, PolyDet("use") int index, E element)}.
-Since \<PolyDet("use")> has no effect on type instantiations, calling this method on a \<Det List>
-will result in the method declaration being instantiated as \codeid{public void add(@Det ArrayList<E> this, @Det int index, E element)}
-irrespective of the type of \<index>, thereby preventing undesired side-effects.
+\begin{Verbatim}[commandchars=\\\{\}]
+    Det \lptype{List}{Det String} dList = ...
+    NonDet int random = ...
+    Det String str = ...
+    set(dList, random, str)  // should be forbidden
+\end{Verbatim}
+
+\noindent
+The call should be forbidden, since on different runs it may make changes at different
+indices of a deterministic list.
+However, the call is permitted.  The middle instantiation is applicable, since
+$\<Det> \ptype{List}{Det String} \sqsubseteq \<OrderNonDet>
+\ptype{List}{Det String}$ and $\<Det> \sqsubseteq \<NonDet>$.
+
+\OurTypeSystem must prevent this behavior.  It does so via another variant
+of qualifier variables, $\use{\kappa}$, which represents a \emph{use} of
+$\kappa$ that does not affect the \emph{instantiation} of $\kappa$.
+
+Ordinarily, the polymorphic function is instantiated at the least upper
+bound of the types of all the arguments that correspond to uses of the type
+parameter.
+For example, function
+
+$f : \forallt{\kappa} \kappa\ \<Int> \times \<Det>\ \<Int> \times\kappa\ \<Int> \times \kappa\ \<Int>$
+
+\noindent
+would be instantiated at the least upper bound of the types of its first,
+third, and fourth arguments at a given call.  (If no such instantiation
+exists, or if the other arguments do not conform to the parameter types, the call does not type-check.)
+By contrast, function
+
+$f : \forallt{\kappa} \kappa\ \<Int> \times \<Det>\ \<Int> \times\use{\kappa}\ \<Int> \times \kappa\ \<Int>$
+
+\noindent
+would be instantiated at the least upper bound of the types of its first
+and forth arguments, and the third argument must conform to the
+instantiation.  However, the third argument does not affect the instantiation.
+
+This is especially useful in preventing methods from non-deterministically modifying the state
+of a deterministic receiver.
+The correct type for \<set> on \ptype{List}{String} is
+
+$\<set> : \forallt{\kappa} \kappa\ \<List>\angles{\down{\kappa}\ \<String>} \times \use{\kappa}\ \<Int> \times \down{\kappa}\ \<String> \rightarrow ()$
+
+\noindent
+and the undesired call \<set(dList, random, str)> does not type-check.
+
+(For the curious reader, the actual types of the \<get> and \<set> methods
+in the JDK are the following.
+
+\begin{Verbatim}
+    @PolyDet("up") E get(@PolyDet List<E> this, @PolyDet int index);
+    @PolyDet("up") E set(@PolyDet List<E> this, @PolyDet("use") int index, E element);
+\end{Verbatim}
+
+\noindent
+The syntax is explained in \cref{java-polymorphism}.)
 

### 2019-04-06 randoop-case-study.tex
Finish draft of section 2.4.3

--- 2019-04-06 randoop-case-study.tex
@@
 especially given that that \theDeterminismChecker found issues that the
-developers had not.
+developers did not.
+As another point of comparison,
+there are fewer type qualifiers than Java generic type arguments.
 

### 2019-04-06 introduction.tex
intro removing todo
Remove duplication
Remove todo

@@
 Nondeterminism is a serious problem for software developers and users.
-\todo{Add lots of citations in the below.}
 \begin{itemize}
@@
 % ,Gyori:2015:RTD:2771783.2771793
-  tests~\cite{LuoHEM2014,nondex,deflaker,Rahman:2018:IFF:3236024.3275529,LuoHEM2014,Sudarshan}
+  tests~\cite{LuoHEM2014,nondex,deflaker,Rahman:2018:IFF:3236024.3275529,Sudarshan}
   that sometimes pass and sometimes fail.  Flaky tests must be re-run, or
@@
 \item
-  Nondeterminism reduces users' and developers' \textbf{trust} in a program's output.
+  Nondeterminism reduces users' and developers' \textbf{trust} in a program's output.\todo{If possible, add a citation here to parallel the earlier bullets.}
 \end{itemize}
@@
 It may be surprising that nondeterminism is common even in sequential
-programs that do not flip coins.
+programs that do not flip coins.\todo{MK: This statement is very strong,
+  and is totally unjustified. I went looking for something that justified
+  it, but couldn't find anything good. I think that this is missing is a
+  core problem with the paper as-is.}
 For example, a program that iterates over a hash table
@@
 Our analysis works at compile time, giving a guarantee over every possible
-execution of the program, by contrast to unsound dynamic tools that attempt
-to discover when a program has exhibited nondeterministic behavior on
-specific runs.  There is no need for a custom runtime system nor rerunning a program
-multiple times.
+execution of the program, unlike unsound dynamic tools that attempt
+to discover when a program has exhibited nondeterministic behavior on a
+specific run.  There is no need for a custom runtime system nor
+rerunning a program multiple times---nor even running it once.
 %
@@
 %
-Our analysis is precise enough for practical use.  It permits calls to
+Our analysis is precise enough for practical use.\todo{MK: After reading
+  the eval, I don't believe this claim. Unfortunately, you can't write this
+  paper without it, so I'm not going to remove it.}
+It permits calls to
 nondeterministic APIs, and only issues a warning if they are used in ways
@@
 have been able to identify issues in real-world programs, some of which
-have been fixed by the developers. We believe that identifying and
+have been fixed by the developers. Identifying and
 resolving nondeterminism
-at an earlier phase in the software development lifecycle would be greatly beneficial to
-developers~\cite{briski2008minimizing}.
+earlier in the software development lifecycle is beneficial to
+developers, because they can avoid bugs associated with flaky tests---reducing
+costs~\cite{briski2008minimizing}.
 
-Our analysis uses three main abstractions, or approximations to run-time values:
+Our analysis uses three main abstractions:
+%, or approximations to run-time values:
+% MK: the OOPSLA audience knows what an abstraction is
 \begin{description}
@@
 \item[\<OrderNonDet>] represents collections that are guaranteed to contain
-  the same elements, albeit in possibly different iteration order.
+  the same elements but whose iteration order is non-deterministic.
 \item[\<Det>] represents values that will be the same across executions.
@@
 Our full analysis contains other features that increase
-expressiveness.
+expressiveness, permitting it to scale to programs of practical interest.
 

### 2019-04-06 paper.tex

@@
 vetted by its developers,
-who were unusually concerned about nondeterminism errors.
+who were greatly concerned about nondeterminism errors.
 In an experiment, it found all of the

### 2019-04-06 design.tex

@@
 
-\todo{Move all of this to Section 3 since the core calculus has no type
-  inference?
-  But we do want to discuss \<sort>.  So maybe just move the discussion of type inference?}
-
-\todo{The core calculus requires a qualifier on every type.}
-
-\TheDeterminismChecker performs local type inference within method bodies.
-It does not perform whole-program type inference.
-This makes separate compilation possible.
-It forces programmers to write specifications (type qualifiers) on methods,
-which is good style and valuable documentation.
-
-The local type inference is flow-sensitive.  That is, each variable may
-have a different type on every line of the program, depending on
-assignments to the variable.  The inferred type always respects any
-declared type:  that is, it is always a refinement (a subtype) of the
-declared type, if any.
-
-In addition to assignments, some method calls refine the types of their
+\todo{I feel that this section is very confusing.}
+
+Some method calls refine the types of their
 arguments.  One example is the \<sort> routine,
@@
 
-\TheDeterminismChecker has additional type refinement rules for the sorting
-methods of \<List>, \<Collections>, and \<Arrays> classes.
 \begin{figure}
@@
 \end{figure}
-As an example, if \codeid{sort()} is called on a receiver of type
-\codeid{OrderNonDet List<Det Set<Det String>>}, it will be type refined to
-\codeid{Det List<Det Set<Det String>>}. However, it is not permitted to
-call \<sort> on a receiver of type \codeid{OrderNonDet List<OrderNonDet
-  Set<Det String>>}.  The rule is not applicable, because doing so would
+
+For example, if \codeid{sort()} is called on a receiver of type
+\ptype{OrderNonDet List}{\lptype{Det Set}{Det String}}, it will be type refined to
+\ptype{Det List}{\lptype{Det Set}{Det String}}. However, it is not permitted to
+call \<sort> on a receiver of type \ptype{OrderNonDet List}{\lptype{OrderNonDet
+  Set}{Det String}}.  The rule is not applicable, because doing so would
 instantiate invalid types.

### 2019-04-06 java-design.tex

@@
 
+\todo{
+\OurTypeSystem performs local type inference within method bodies.
+It does not perform whole-program type inference.
+This makes separate compilation possible.
+It forces programmers to write specifications (type qualifiers) on methods,
+which is good style and valuable documentation.
+
+The local type inference is flow-sensitive.  That is, each variable may
+have a different type on every line of the program, depending on
+assignments to the variable.  The inferred type always respects any
+declared type:  that is, it is always a refinement (a subtype) of the
+declared type, if any.
+}
+
 \OurTypeSystem is

### 2019-04-06 design.tex
Edits in 2.6

@@
 \item
-  Basetype polymorphism lets the basetype vary indepentently of the
+  Basetype polymorphism lets the basetype vary independently of the
   qualifier, which might be fixed or might be polymorphic.
@@
 order-nondeterminism, but this type is enough for illustration in this section.)
-
-\todo{Show actual type.}
 
@@
 
-\subsection{Maps}\label{maps}
+\subsection{Maps and sets}\label{maps}
 A map is a collection of key--value pairs.
@@
 
-\subsection{Improving precision}\label{precision}
-
-\todo{Sets vs lists.  Set equality is independent of order.}
-An order nondeterministic list does not provide any guarantee on the
-equality as it is dependent on the iteration order
-
-We annotated the \codeid{equals()} method of the \<Set> class in the JDK as
-\begin{verbatim}
-PolyDet("up") boolean equals(PolyDet Set<E> this, PolyDet Object o)
-\end{verbatim}
-While this is sound, it is imprecise in the case of \codeid{OrderNonDet
-  Set}s.\todo{Explain why.}
- We added the rule shown in \cref{fig-precision-rules} to our checker to improve precision.
+\subsubsection{Improving precision for equality and map lookup}\label{precision}
+
+List equality is dependent on iteration order, but set equality is not.
+Comparing two objects of type \ptype{OrderNonDet List}{Det String} yields a \<NonDet>
+result:  depending on the execution, the lists might or might not be in the
+same order.  However, 
+comparing two objects of type \ptype{OrderNonDet Set}{Det String} yields a
+\<Det> result:  always the same on every execution.
+
 \begin{figure}
@@
     {
-        &\Gamma \vdash \|x| : \<OrderNonDet> \ \|Set|\angles{ \kappa_{x} \ \beta_{x}}, \|y| : \<OrderNonDet> \ \|Set|\angles{ \kappa_{y} \ \beta_{y}} \\
-        & \kappa_{x} \ \beta_{x} \nsqsubseteq \<OrderNonDet> \ \<List>        \hspace{150pt}   \\
-        & \kappa_{y} \ \beta_{y} \nsqsubseteq \<OrderNonDet> \ \<List>     \hspace{150pt}    
+        &\Gamma \vdash \|x| : \<OrderNonDet> \ \|Set|\angles{ \kappa_{x} \ \beta_{x}}, \|y| : \<OrderNonDet> \ \|Set|\angles{ \kappa_{y} \ \beta_{y}} \hspace{10pt}\\
+        & \kappa_{x} \ \beta_{x} \nsqsubseteq \<OrderNonDet> \ \<List> 
+        \qquad \kappa_{y} \ \beta_{y} \nsqsubseteq \<OrderNonDet> \ \<List>     \hspace{30pt}
     }
@@
 
-\todo{
-  In the JDK, the \<equals()> method of \<Set> is annotated as
-
-@PolyDet("up") boolean equals(@PolyDet Set<E> this, @PolyDet Object o);
-
-While this is sound, it is imprecise.
-For example, consider the code below:
-
-    @OrderNonDet Set<@Det Integer> a; \\
-    @OrderNonDet Set<@Det Integer> b; \\
-    a.equals(b);
-
-Here, the return type of \<equals()> gets the annotation \<@NonDet>.
-But it would be more precise to annotate it as \<@Det>.
-To achieve this precision, the checker annotates the return type
-of \<equals()> called on a receiver of type \<Set> as \<@Det> if the
-following conditions are true:
-\begin{itemize}
-    \item the receiver does not have {@code List} or its subtype as a type parameter,
-    \item the argument to equals() is also an {@code @OrderNonDet Set}, and
-    \item the argument to equals() also does not have {@code List} or its subtype as a
-    type parameter
-\end{itemize}
-}
-
-According to this rule, if the \<Set> receiver has type \codeid{@OrderNonDet Set<@Det List>} and the argument to the \codeid{equals()} method is
-also of the same type, the return type gets the qualifier \<Det>. Without this precision rule, the return type would have been instantiated as \<NonDet> which
-is sound but imprecise. This rule is careful enough to not annotate the return type of \<equals()> as \<Det> if a type argument to either the
-receiver or the argument happens to be an \codeid{OrderNonDet List}.
+More specifically, rule \rulename{set precision} of
+\cref{fig-precision-rules} expresses that the return type of \<equals()>
+is \<Det> if both arguments have type \<OrderNonDet Set> and nether argument
+has {@code List} or a subtype within its type argument.
+Without this rule, the type would be \<NonDet> which is sound but imprecise.
 
 Similarly, the rule \rulename{map precision}\ in \cref{fig-precision-rules}
-improves precision for \<OrderNonDet> receivers.
+improves precision when calling \<get> on an \<OrderNonDet> receiver.
 
@@
 % LocalWords:  LinkedHashMap TreeMap HashSet LinkedHashSet LinkedHashSet's
+% LocalWords:  basetype ddlist typechecked typecheck ni di

### 2019-04-06 randoop-case-study.tex

@@
 classpath passed on the command line.
-This can cause to incorrect behavior, both in Randoop's test suite and in the field,
+This can cause incorrect behavior, both in Randoop's test suite and in the field,
 if a user sets the environment variable.
@@
 extensive effort by the Randoop developers (\cref{sec:subject-program}),
-especially given that that \theDeterminismChecker found issues that the
+especially given that \theDeterminismChecker found issues that the
 developers did not.

### 2019-04-06 experiments.tex
expts edit

@@
 We investigated the tests covering code that \theDeterminismChecker reported as nondeterministic.
-\bugHashSet and \bugClasspath have corresponding test cases whereas \bugTimestampOutput and 
-\bugTimestampOutput do not.
+\bugHashSet and \bugClasspath have corresponding test cases whereas the remainder of the bugs do not.
 As NonDex is a \textit{flaky test} detection tool, detecting \bugTimestampOutput and \bugTimestampOutput 
@@
     \end{subfigure}    
-    \caption{Errors detected by \TheDeterminismChecker.}
+    \caption{Errors detected by \TheDeterminismChecker in NonDex benchmarks.}
     \label{fig:nondex-source}
@@
 warning at the assignment, because \<getDeclaredMethods> returns \<@Det
-Field @Det []>, which is an order-nondeterministic array of deterministic
+Field @OrderNonDet []>, which is an order-nondeterministic array of deterministic
 \<Field>s.
@@
 case that treated an array returned by \<Class.getDeclaredConstructors()>
-as deterministic.
+as deterministic. This is erroneous because \<getDeclaredConstructors> returns an order nondeterministic array.

### 2019-04-06 randoop-case-study.tex

@@
 generator~\ifanonymous\citepalias{randoop-tool}\else\cite{PachecoLEB2007}\fi.
-We also wrote specifications for libraries Randoop uses, such as the JDK, JUnit,
-and others.
+We also wrote specifications for libraries Randoop uses, such as the JDK,
+JUnit, plume-util, and others.
 Like \theDeterminismCheckerImplementation,
 all the case study materials are publicly available at \myurl{\todo{url for
-    case study}} for reproducibility. \todo{This should include our
-      our annotated plume-util, which we found was required to reproduce the
-    results of the case study}
+    case study}} for reproducibility.
 
@@
 it was a bug or a false positive warning.
-\todo{In some cases, the number of \<@Det> annotations was especially prevalent.
-  In these cases we changed the default qualifier for that class to \<@Det>.}
+When the number of \<@Det> annotations in a file was overwhelming,
+we changed the default qualifier for that class to \<@Det>.  (Users
+  can control defaulting on a file-by-file and method-by-method basis.)
 The effort would have been much easier for someone familiar with Randoop,
@@
 
-\TheDeterminismChecker issued 37 false positive warnings, or 1 every 650
+\TheDeterminismChecker issued 37 false positive warnings, or 1 for every 650
 lines of code.
@@
 local refactoring in the Randoop codebase can make the code type check.
-Refactoring can also eliminate the last warning. \todo{Can it? I thought
-  what happens when iterating over an @OrderNonDet collection is something we
-can't verify with the checker.}
+Refactoring can also eliminate the last warning, for example by using the
+higher-order \<map()> function instead of a loop; alternately,
+\theDeterminismChecker could be enhanced to pattern-match some loops.
 

### 2019-04-06 introduction.tex
Some edits in related work

@@
 %
-Our analysis is precise enough for practical use.\todo{MK: After reading
-  the eval, I don't believe this claim. Unfortunately, you can't write this
-  paper without it, so I'm not going to remove it.}
-It permits calls to
+% Our analysis is precise enough for practical use.\todo{MK: After reading
+%   the eval, I don't believe this claim. Unfortunately, you can't write this
+%   paper without it, so I'm not going to remove it.}
+Our analysis permits calls to
 nondeterministic APIs, and only issues a warning if they are used in ways

### 2019-04-06 related.tex
some more related

@@
 NonDex uses a hand-crafted list of 47 methods (25 unique method names)
-in 13 classes that were identified as potential sources of flakiness. 
+in 13 classes as potential sources of flakiness. 
 For each of the identified methods, the authors built models that
@@
 and considerable debugging effort to locate and fix the source of flakiness. \TheDeterminismChecker, in
-contrast, reports the cause of nondeterminism at compile time requiring little to no debugging effort with the
-caveat of producing higher false positives compared to NonDex. 
-Identifying and modeling methods with nondeterministic
-specifications is analogous to our annotations on the JDK\@. So far, we have annotated
-928 methods across 41 classes.
+contrast, reports the cause of nondeterminism at compile time requiring
+little to no debugging effort.  However, \theDeterminismChecker requires
+much more upfront effort, and it produce false positive warnings.
+NonDex's approach of identifying and modeling methods with nondeterministic
+specifications is analogous to our library specifications. So far, we have annotated
+928 methods across 41 classes in the JDK, in addition to other libraries.
 %\todo{Check whether our JDK is missing anything they support - No}
@@
 
-DeFlaker~\cite{deflaker} is another approach for flake test detection. It relies on
-version control history to detect tests that behave differently. It computes a diff of the code covered
+DeFlaker~\cite{deflaker} is another approach for flaky test detection. It
+relies on a
+version control history. It computes a diff of the code covered
 in the current version and the previous one. If there exists a test case whose code coverage does not include
@@
 This approach doesn't require JVM modifications
-and integrates easily with a production software. DeFlaker reported 19 previously unknown bugs
+and integrates easily with production software. DeFlaker reported 19 previously unknown bugs
 in open source projects, 7 of which were addressed by the developers of these projects. DeFlaker is agnostic
-to the code under test and can therefore report flakiness arising out of concurrency which \TheDeterminismChecker cannot.
+to the code under test and can therefore report flakiness arising out of concurrency, which \theDeterminismChecker cannot.
 
@@
 researchers and software developers alike~\cite{Fowler,Sudarshan}.
-In \cite{LuoHEM2014}, the authors studied 51 open source projects and analyzed the most common root causes of
-test flakiness. The findings in the paper suggest that most of the flakiness in tests is caused by
+Luo et al.~\cite{LuoHEM2014} studied 51 open source projects and analyzed the most common root causes of
+test flakiness. The findings suggest that most of the flakiness in tests is caused by
 the use of async await, concurrency, or due to test order dependency. The authors of~\cite{Plotkin:1993:LPP:645891.671433}
@@
 in branching time temporal logic systems that are inherently non-deterministic.
+While there has been research on type systems that provide compile time determinism guarantees for parallel programs~\cite{Bocchino:2009:TES:1640089.1640097,Bocchino:2011:SND:1926385.1926447},
+there hasn't been a similar effort for sequential programs.
 

### 2019-04-06 design.tex

@@
 \<OrderNonDet> may only be written on collections and maps.
-A map is a dictionary or associative array, such as a hash table.
+A map is a dictionary or an associative array, such as a hash table.
 \OurTypeSystem largely treats a map as a collection of key--value pairs.
@@
 affects access, iteration, searching, etc.
-(Rules \rulename{get}, \rulename{iteration}, and \rulename{next element} in \cref{fig-ordernondet-rules}.)
+(As shown by types of methods \rulename{get}, \rulename{iterator}, \rulename{next} in \cref{fig-ordernondet-rules})
 \item
@@
 arguments are \<Det>,
-and our typing must have that as a special case, which is the first line above.
+and our typing must have that as a special case, which is the last line above.
 
 \begin{figure}
-\todo{Give the types for get and set, which are more basic list/array operations,
-  at the beginning before iteration.}
 
@@
 
-    \rulename{iteration}: 
+    \rulename{iterator}: 
     $
@@
 
-    \rulename{next element}:
+    \rulename{next}:
     $
@@
 
-    \rulename{hasnext element}:
+    \rulename{hasnext}:
     $
@@
     These are partial types indicating the methods' behavior on
-    \<OrderNonDet> arguments.  Full types are given in \cref{polymorphism}.}
-\todo{These should be method names, not rule names, right?  The figure is
-  just giving the types of methods.}
+    \<OrderNonDet> arguments.}
+%\todo{These should be method names, not rule names, right?  The figure is
+%  just giving the types of methods.}
 \label{fig-ordernondet-rules}
@@
 
-$\<length>\ : \<NonDet>\ \<String> \rightarrow \<NonDet> \<Int>$
-
-$\<length>\ :\<Det>\ \<String> \rightarrow \<Det> \<Int>$
+$\<length>\ : \<NonDet>\ \<String> \rightarrow \<NonDet> \ \<Int>$
+
+$\<length>\ :\<Det>\ \<String> \rightarrow \<Det> \  \<Int>$
 
@@
   Expression & instantiation & type \\
-\<plus(ni, ni)> & $\<NonDet>\ \<String> \rightarrow \<NonDet> \<Int>$ & \<NonDet> \\
-\<plus(di, di)> & $\<Det>\ \<String> \rightarrow \<Det> \<Int>$ & \<Det> \\
-\<plus(ni, di)> & $\<NonDet>\ \<String> \rightarrow \<NonDet> \<Int>$ & \<NonDet> \\
-\<plus(di, ni)> & $\<NonDet>\ \<String> \rightarrow \<NonDet> \<Int>$ & \<NonDet> \\
+\<plus(ni, ni)> & $\<NonDet>\ \<String> \rightarrow \<NonDet> \ \<Int>$ & \<NonDet> \\
+\<plus(di, di)> & $\<Det>\ \<String> \rightarrow \<Det> \ \<Int>$ & \<Det> \\
+\<plus(ni, di)> & $\<NonDet>\ \<String> \rightarrow \<NonDet> \ \<Int>$ & \<NonDet> \\
+\<plus(di, ni)> & $\<NonDet>\ \<String> \rightarrow \<NonDet> \ \<Int>$ & \<NonDet> \\
 \end{tabular}

### 2019-04-06 related.tex
Add todo for Rashmi
More about Bocchino

@@
 
+\todo{I don't see how the first two (especially the first) citations in the
+  next paragraph are related to our work.  Make the connection explicit, or
+  remove them.}
+
 The problem of inconsistencies between specifications and implementations
@@
 \cite{Cook:2013:RNP:2491956.2491969} presents an approach that can automatically verify properties
-in branching time temporal logic systems that are inherently non-deterministic.
-While there has been research on type systems that provide compile time determinism guarantees for parallel programs~\cite{Bocchino:2009:TES:1640089.1640097,Bocchino:2011:SND:1926385.1926447},
-there hasn't been a similar effort for sequential programs.
+in branching time temporal logic systems that are inherently
+non-deterministic.
+Bocchino et
+al.~\cite{Bocchino:2009:TES:1640089.1640097,Bocchino:2011:SND:1926385.1926447}
+provide a type-and-effect system that provides compile-time determinism
+guarantees for parallel programs, with a focus on barrier removal,
+reasoning about interference and thread interleavings.
+We believe our work to be the first verification approach
+aimed at ensuring determinism in sequential programs.
 
@@
 presents an approach that detects brittle assertions in tests by performing a taint analysis on inputs classified as 
-controlled and uncontrolled.~\cite{ZhangJWMLEN2014} investigates the effects of test independence assumption
+controlled and uncontrolled.~\cite{ZhangJWMLEN2014} investigates the
+effects of the test independence assumption
 on other techniques such as test prioritization, selection, etc. Other approaches~\cite{BellKMD2015,Gyori:2015:RTD:2771783.2771793}

### 2019-04-06 paper.tex
Edits in abstract

@@
 nondeterminism detection tools rely on observing runtime output, our approach
-aims to verify determinism thereby providing stronger soundness guarantees.
+aims to verify determinism at compile time, thereby providing stronger soundness guarantees.
 
@@
 program is nondeterministic or verifies that the program is deterministic.
-In a case study of a 24,000 line software project, it found
+In a case study of a 24,000-line software project, it found
 previously-unknown nondeterminism errors in a program that had been heavily
@@
 %% comma separated list
-\keywords{\todo{keyword1}, \todo{keyword2}, \todo{keyword3}}  %% \keywords are mandatory in final camera-ready submission
+\keywords{nondeterminism, type system, verification, specification, hash table}  %% \keywords are mandatory in final camera-ready submission
 

### 2019-04-06 conclusion.tex
Edits in conclusion

--- 2019-03-31 conclusion.tex
@@
 
-\todo{TODO}
-
-\todo{\TheDeterminismChecker found errors that the
-  Randoop developers had missed, despite spending over 100 hours on the
-  problem of nondeterminism.}
+We designed a type system that helps programmers express determinism specifications
+in sequential programs. 
+We implemented our type system in Java, applied it to real world 
+software, and found it to be effective in finding previously unknown determinism bugs. 
+Our tool \theDeterminismChecker found errors that the
+Randoop developers had missed, despite spending
+% over 100 hours
+extensive effort on the
+problem of nondeterminism. We compared our tool with existing state of the art flaky test detectors
+and found that \theDeterminismChecker found all the bugs arising due to nondeterminism
+in sequential programs that were used for evaluation in NonDex~\cite{nondex} and DeFlaker~\cite{deflaker}.
+To the best of our knowledge, ours is the
+first compile-time verification approach addressing the problem of
+nondeterminism in sequential programs.

### 2019-04-06 design.tex
sec 2 edits

@@
 
-\paragraph{Discussion of the type of \<length>}
+\paragraph{Discussion of the type of \<first>}
 
@@
 A similar typing does \emph{not} work for the \<set> function that sets an
-element of a list.  Consider the putative type
+element of a list.  Consider the mutative type
 
@@
 \cref{fig-precision-rules} expresses that the return type of \<equals()>
-is \<Det> if both arguments have type \<OrderNonDet Set> and nether argument
+is \<Det> if both arguments have type \<OrderNonDet Set> and neither argument
 has {@code List} or a subtype within its type argument.

### 2019-04-06 experiments.tex
Some edits in related work
Edits in NonDex and DeFlaker experiments

@@
 We ran NonDex on the same commit of Randoop as we used for the case study
-of \cref{sec:randoop-case-study}.  This version contains all the nondeterminism bugs that
+of \cref{sec:randoop-case-study}.  This version contains all
+\numRandoopBugs nondeterminism bugs that
 \theDeterminismChecker found.
@@
 %  to account for the additional bugs.}
+
 We investigated the tests covering code that \theDeterminismChecker reported as nondeterministic.
@@
         \end{verbatim}
-        \caption{Annotated source code from ActionGenerator.}
+        \caption{Nondeterministic source code from ActionGenerator.}
         \label{code-actiongenerator}
@@
 \<fields.entrySet()> (which is itself \<@OrderNonDet>) yields \<@NonDet>
-\<Entry> values..
+\<Entry> values.
 
@@
 \TheDeterminismChecker is capable of detecting nondetermism in test
-cases, as some of these examples are.
+cases, as some of the NonDex examples are.
 We wrote specifications for the
@@
 projects built with Maven, but Randoop
-uses Gradle for its build system. We plan to convert Randoop's build system to Maven and
-perform this case study.
+uses Gradle for its build system.
+%% Don't make promises: they give reviewers an excuse to reject the paper
+%% as not a completed research project.
+% We plan to convert Randoop's build system to Maven and perform this case study.
 %Ideally, we would have 
@@
 Since network operations are nondeterministic, we annotated the method in \<nutz> that
-returns this response as \<NonDet> which lead \TheDeterminismChecker to report an error at the print statement. 
-\TheDeterminismChecker was able to find the source of flakiness in \<checkstyle>. This bug was in a test
+returns this response as \<NonDet> which led \TheDeterminismChecker to report an error at the print statement. 
+\TheDeterminismChecker also found the source of flakiness in \<checkstyle>. This bug was in a test
 case that treated an array returned by \<Class.getDeclaredConstructors()>
-as deterministic. This is erroneous because \<getDeclaredConstructors> returns an order nondeterministic array.
+as deterministic. This is erroneous because \<getDeclaredConstructors> returns an order-nondeterministic array.

### 2019-04-06 design.tex
design edits

@@
 \OurTypeSystem largely treats a map as a collection of key--value pairs.
-Both collections and maps may be \<Det>, \<OrderNonDet>, or \<NonDet>, and
-the element types can be specified independently of the collection type,
-including their qualifiers.
+Both collections and maps may be \<Det>, \<OrderNonDet>, or \<NonDet>.
+The basetypes of their elements can be specified independently of the collection basetypes.
+However, an element type qualifier must be a subtype of the collection type qualifier.
+The next section discusses type well-formedness in detail.
+%and
+%the element types can be specified independently of the collection type,
+%including their qualifiers.
 
@@
 
-\todo{Can we come up with more examples of interesting functions to
-  describe here?  Or put them in section 3.}
+%\todo{Can we come up with more examples of interesting functions to
+%  describe here?  Or put them in section 3.}
 
@@
 
-\todo{I feel that this section is very confusing.}
+%\todo{I feel that this section is very confusing.}
 
@@
 is \<Det> if both arguments have type \<OrderNonDet Set> and neither argument
-has {@code List} or a subtype within its type argument.
+has \code{@OrderNonDet List} or a subtype within its type argument.
 Without this rule, the type would be \<NonDet> which is sound but imprecise.

### 2019-04-06 design.tex
Fix LaTeX error

@@
 is \<Det> if both arguments have type \<OrderNonDet Set> and neither argument
-has \code{@OrderNonDet List} or a subtype within its type argument.
+has \codeid{@OrderNonDet List} or a subtype within its type argument.
 Without this rule, the type would be \<NonDet> which is sound but imprecise.

### 2019-04-06 related.tex
related todo

@@
 
-\todo{I don't see how the first two (especially the first) citations in the
-  next paragraph are related to our work.  Make the connection explicit, or
-  remove them.}
+%\todo{I don't see how the first two (especially the first) citations in the
+%  next paragraph are related to our work.  Make the connection explicit, or
+%  remove them.}
 
 The problem of inconsistencies between specifications and implementations
-is quite prevalent.~\cite{Jin:2012:UDR:2254064.2254075} looks at the effect of inefficient code sequences on the performance of software.
-Its findings suggest that incorrect understanding of performance specifications of APIs is a very common root cause for
+is quite prevalent. Other researchers have explored ways of reducing the gap between implicit
+specifications in APIs and the assumptions made by their clients in a variety of problem domains.
+The findings in~\cite{Jin:2012:UDR:2254064.2254075} suggest that incorrect understanding of performance specifications of 
+APIs is a very common root cause for
 performance bugs. ~\cite{Rui:2013:180377} analyzes the APIs related to security such as authorization and authentication.

### 2019-04-06 introduction.tex
intro todos

@@
 \item
-  Nondeterminism reduces users' and developers' \textbf{trust} in a program's output.\todo{If possible, add a citation here to parallel the earlier bullets.}
+  Nondeterminism reduces users' and developers' \textbf{trust} in a program's output~\cite{Tene,Shah}.
 \end{itemize}
@@
 It may be surprising that nondeterminism is common even in sequential
-programs that do not flip coins.\todo{MK: This statement is very strong,
-  and is totally unjustified. I went looking for something that justified
-  it, but couldn't find anything good. I think that this is missing is a
-  core problem with the paper as-is.}
+programs that do not flip coins.
+%\todo{MK: This statement is very strong,
+%  and is totally unjustified. I went looking for something that justified
+%  it, but couldn't find anything good. I think that this is missing is a
+%  core problem with the paper as-is.}
 For example, a program that iterates over a hash table

### 2019-04-06 design.tex
edits

@@
 The type of \<get> ensures that if either the list or the index is (order)
-non-deterministic, then the result of the call is nondeterministic.
+nondeterministic, then the result of the call is nondeterministic.
 
@@
 
-This is especially useful in preventing methods from non-deterministically modifying the state
+This is especially useful in preventing methods from nondeterministically modifying the state
 of a deterministic receiver.

### 2019-04-06 related.tex
edits

@@
 Several techniques have been proposed to test whether a
-deterministic implementation conforms to its non-deterministic finite state machine~\cite{Petrenko1996,Petrenko:1993:NSM:648128.761244,Savor:1997:639710,Hierons:2004:TCD:1040993.1040998}.
+deterministic implementation conforms to its nondeterministic finite state machine~\cite{Petrenko1996,Petrenko:1993:NSM:648128.761244,Savor:1997:639710,Hierons:2004:TCD:1040993.1040998}.
 \cite{Cook:2013:RNP:2491956.2491969} presents an approach that can automatically verify properties
 in branching time temporal logic systems that are inherently
-non-deterministic.
+nondeterministic.
 Bocchino et

### 2019-04-06 design.tex
section 2 typos

@@
     \item \<OrderNonDet> indicates that the expression is a collection or
-      map that contains the same elements in every execution, but possibly
+      a map that contains the same elements in every execution, but possibly
       in a different order.
@@
 would be instantiated at the least upper bound of the types of its first
-and forth arguments, and the third argument must conform to the
+and fourth arguments, and the third argument must conform to the
 instantiation.  However, the third argument does not affect the instantiation.
@@
 
-It might seem that the notion of an order-nondeterministic set or map is
+It might seem that the notion of a deterministic set or map is
 nonsensical, since the set and map specifications make no promises about

### 2019-04-06 java-design.tex
Edits in case study and java-design
java design typos

--- 2019-04-06 java-design.tex
@@
 qualifier variable that does not affect the binding of that variable
-(\cref{bindings-uses}) is written \<@PolyDet("use")>.
+($\use{\kappa}$ in \cref{bindings-uses}) is written \<@PolyDet("use")>.
 
@@
 
+Although \theDeterminismChecker performs local type inference within method bodies,
+it does not perform whole-program type inference.
+This makes separate compilation possible.
+It forces programmers to write specifications (type qualifiers) on methods,
+which is good style and valuable documentation.
+
 \subsection{The environment}\label{sec:environment-java}

### 2019-04-06 related.tex
final edits

@@
 little to no debugging effort.  However, \theDeterminismChecker requires
-much more upfront effort, and it produce false positive warnings.
+much more upfront effort, and it produces false positive warnings.
 NonDex's approach of identifying and modeling methods with nondeterministic
@@
 al.~\cite{Bocchino:2009:TES:1640089.1640097,Bocchino:2011:SND:1926385.1926447}
-provide a type-and-effect system that provides compile-time determinism
+present a type-and-effect system that provides compile-time determinism
 guarantees for parallel programs, with a focus on barrier removal,

### 2019-06-05 conclusion.tex
Use external bibliography when possible

@@
 and found that \theDeterminismChecker found all the bugs arising due to nondeterminism
-in sequential programs that were used for evaluation in NonDex~\cite{nondex} and DeFlaker~\cite{deflaker}.
+in sequential programs that were used for evaluation in NonDex~\cite{ShiGLM2016} and DeFlaker~\cite{BellLHEYM2018}.
 To the best of our knowledge, ours is the

### 2019-06-05 experiments.tex
final edits
Use external bibliography when possible

@@
 
-The state of the art in flaky test detection is NonDex~\cite{nondex}.
+The state of the art in flaky test detection is NonDex~\cite{ShiGLM2016}.
 \Cref{sec:related} explains how NonDex works. This section compares the
@@
 \bugHashSet and \bugClasspath have corresponding test cases whereas the remainder of the bugs do not.
-As NonDex is a \textit{flaky test} detection tool, detecting \bugTimestampOutput and \bugTimestampOutput 
+As NonDex is a \textit{flaky test} detection tool, detecting bugs that don't have corresponding tests 
 is outside of its scope.
@@
     \end{tabular}
-\caption{Souces of flakiness known to NonDex~\cite{nondex}.}
+\caption{Souces of flakiness known to NonDex~\cite{ShiGLM2016}.}
 \label{fig:flaky-sources}
@@
 Its authors ran NonDex on 195 open-source projects, and NonDex found flaky tests in
-21 of them~\cite{nondex}.
+21 of them~\cite{ShiGLM2016}.
 The authors also reported the sources of flakiness
@@
 \section{Comparison to Deflaker}
-Deflaker~\cite{deflaker}, like NonDex, reports tests that could be flaky. 
+Deflaker~\cite{BellLHEYM2018}, like NonDex, reports tests that could be flaky. 
 We were unable to run DeFlaker on Randoop, because Deflaker works with
@@
 the paper was written. The authors reported 19 of these tests, out of which 7 were addressed by the maintainers of
-those projects~\cite{deflaker}.
+those projects~\cite{BellLHEYM2018}.
 We ran \theDeterminismChecker on the part of each of these

### 2019-06-05 introduction.tex
edits
Use external bibliography when possible

@@
   correct behaviors.  Test oracles that are too strict lead to flaky
-% ,Gyori:2015:RTD:2771783.2771793
-  tests~\cite{LuoHEM2014,nondex,deflaker,Rahman:2018:IFF:3236024.3275529,Sudarshan}
+% ,GyoriSHM2015
+  tests~\cite{LuoHEM2014,ShiGLM2016,BellLHEYM2018,Rahman:2018:IFF:3236024.3275529,Sudarshan}
   that sometimes pass and sometimes fail.  Flaky tests must be re-run, or
@@
 %    These seem to be general citations about test flakiness or dependence.
-%    Readers will be unhappy if we make a claim but then do not support it.}\cite{Herzig:2015:EDF:2819009.2819018,Huo:2014:IOQ:2635868.2635917,ZhangJWMLEN2014,BellKMD2015,Fowler,Gyori:2015:RTD:2771783.2771793,Dan:2013:10.1007/978-3-642-39038-8_25,Vahabzadeh:2015:7332456,Sudarshan}.
+%    Readers will be unhappy if we make a claim but then do not support it.}\cite{Herzig:2015:EDF:2819009.2819018,Huo:2014:IOQ:2635868.2635917,ZhangJWMLEN2014,BellKMD2015,Fowler,GyoriSHM2015,Dan:2013:10.1007/978-3-642-39038-8_25,Vahabzadeh:2015:7332456,Sudarshan}.
 %  \todo{Citations about bugs related to this?}
@@
 approaches that identify flaky test cases.
-NonDex~\cite{nondex} uses a modified JVM that returns different results on different
+NonDex~\cite{ShiGLM2016} uses a modified JVM that returns different results on different
 executions, for a few key JDK methods with loose specifications.  Running a
@@
 methods.
-%NonDex~\cite{nondex} is a tool that manually\todo{This is confusing:  the
+%NonDex~\cite{ShiGLM2016} is a tool that manually\todo{This is confusing:  the
 %  tool can't manually identify.  Do you mean that the developers manually
@@
 %identifies 
-DeFlaker~\cite{deflaker} looks at a range of commit versions
+DeFlaker~\cite{BellLHEYM2018} looks at a range of commit versions
 of a code, and marks a test as flaky if it doesn't execute any modified code but still fails in the newer version. These techniques
@@
 \item[\<OrderNonDet>] represents collections that are guaranteed to contain
-  the same elements but whose iteration order is non-deterministic.
+  the same elements but whose iteration order is nondeterministic.
 \item[\<Det>] represents values that will be the same across executions.

### 2019-06-05 related.tex
Use external bibliography when possible

@@
 
-The state of the art for detecting nondeterministic tests is NonDex~\cite{nondex}.  It is a highly effective tool.
+The state of the art for detecting nondeterministic tests is
+NonDex~\cite{ShiGLM2016}.  It is a highly effective tool.
 % https://github.com/TestingResearchIllinois/NonDex/wiki/Supported-APIs
@@
 
-DeFlaker~\cite{deflaker} is another approach for flaky test detection. It
+DeFlaker~\cite{BellLHEYM2018} is another approach for flaky test detection. It
 relies on a
@@
 test flakiness. The findings suggest that most of the flakiness in tests is caused by
-the use of async await, concurrency, or due to test order dependency. The authors of~\cite{Plotkin:1993:LPP:645891.671433}
+the use of async await, concurrency, or due to test order dependency. The authors of~\cite{PlotkinA1993}
 perform an empirical study on open source benchmarks, the results of which concur with those of~\cite{LuoHEM2014}.
-Additionally,~\cite{Plotkin:1993:LPP:645891.671433} studies the effects of removing these test smells. The results suggest that
+Additionally,~\cite{PlotkinA1993} studies the effects of removing these test smells. The results suggest that
 refactoring the test code smells fixed more than half of the identified flaky tests. Our approach is complementary to
@@
 effects of the test independence assumption
-on other techniques such as test prioritization, selection, etc. Other approaches~\cite{BellKMD2015,Gyori:2015:RTD:2771783.2771793}
+on other techniques such as test prioritization, selection, etc. Other approaches~\cite{BellKMD2015,GyoriSHM2015}
 analyze test dependencies and either prevent them or use this information for other optimizations.

### 2019-06-12 oopsla2019-rebuttal.txt
plain text rebuttal

We thank the reviewers for their constructive feedback.
We are glad that they are enthusiastic about our choice of problem and our approach.
For the future, we will address the reviewers’ two key points:  a proof of soundness (using a different formalization) and expanded case studies.
The remainder of this response explains other points.

Applying our tool to more benchmarks:
We have applied our determinism checker to all the benchmarks of NonDex (21 in total as described in Section 5.2) and three benchmarks from DeFlaker (these were benchmarks for which the authors reported the bug to the maintainers; more details are in Section 6.1).

“Was NonDex applied to Randoop? Could it have found the bugs that Garçon
spotted?”

We did run NonDex on Randoop (Section 5.1).  NonDex was unable to find the bugs that Garçon spotted.

*) section 5.1: Of the 5 tests, 3 are out of scope for NonDex due to not being triggered by the test suite and 1 is not modeled by NonDex (the Classpath bug). What about the fifth bug?

The fifth bug could have been found by NonDex (it is in scope and executed), but NonDex did not find it.

Why use Randoop for a case study?
As explained in Section 4.1 and in Randoop’s documentation, Randoop generates the same tests when run on the same inputs (unless the program under test is itself deterministic).  It is a misconception that Randoop is nondeterministic.

A user can pass in a different seed for the pseudo-random number generator (for instance, to run Randoop twice to obtain different test suites), but that seed is an input and defaults to zero.

Limits to Expressivity:
The type well-formedness rules for nested types do not limit expressivity (also see Section 2.1). 
For instance, 
a) We do not allow “Det List<NonDet Integer>”. According to our semantics of types on collections, this type indicates a “List” that upon iteration would result in the same values of elements in the same order, but the contained elements are nondeterministic which is contradictory. Any collection that has “NonDet” elements has to be “NonDet” in order to be consistent with our definition of a collection type.
b) Similarly, a “NonDet List<Det Integer>” is not more expressive than a “NonDet List<NonDet Integer>” because the collection type here says that there is no guarantee either on the order of iteration or the element values.

Unsoundness caused by turning off warnings on conditionals:
Our type system does not allow nondeterminism on conditionals. However, our tool implementation enables the user to enable or disable this behavior.  This can be useful to limit tool output when first applying the type-checker to a legacy program.

“consider extending dynamic tools such as NonDex so as to interpret Garçon's annotation?”

That’s an interesting suggestion. This heuristic will still not be able to guarantee soundness because NonDex cannot guarantee that a nondeterministic execution will manifest in its reruns. In our experiments, NonDex could not detect nondeterminism that our tool discovered.  However, the hybrid approach could potentially expose some determinism bugs that are not reported when the warnings are disabled.

Generality of our technique:
Our technique applies to any statically-typed object-oriented language, requiring only engineering effort. We chose Java because of its support for type annotations and to enable direct comparison with previous work, which also used Java.

Overheads of false positives and Annotations:
“Another choice is to inspect the code for known Java idioms that lead to non-determinism.”

Manual examination is much more work in the long run, because it must be repeated after every code change.  It is easier to annotate once and have a verification tool run on every subsequent compile.

The Randoop authors tried manual examination, and our tool found errors despite their effort.

Note that a user doesn't want to forbid all uses of nondeterminism.  Nondeterminism is often acceptable in a certain part of the program, such as within a given module, behind an abstraction barrier, or for certain values.  What matters is how the nondeterminism is used and where it can flow, and this cannot be captured by a lexical or local analysis.  People are not good at performing this sort of analysis manually.

Hard-coding a limited number of Java idioms would restrict the guarantees to just those idioms and would make the approach inextensible.  It might apply only to clients of the standard library but not to user-defined classes, for example.

Special typing rules for Sets and maps:
“Are the standard libraries annotated with types that may be exploited by client code but that is trusted w.r.t. the implementation of the library”

There is no type loophole, and clients cannot exploit map and set types.  When clients implement sets or maps, the clients use the map and set types, and they are fully checked, never trusted. The typechecker sometimes issues false positive warnings for an algorithm that is beyond its capabilities, especially for low-level implementations such as some in the standard library.  These must be verified in some other way, such as by manual inspection.

Clarification on type refinement rules:
The rules in Section 2.5 apply for refining the types, not their validity. The argument of “sort” can be of any type. We refine the type of the receiver only when it is order-nondeterministic and all its type arguments are deterministic.
Sorting a nondeterministic collection yields a nondeterministic collection and does not refine the receiver type.

Imprecision of Floating point operations:
Our type system operates on source code and cannot be cognizant of reordering operations performed by compiler optimizations.  Imprecision due to floating-point reordering is not nondeterminism and is not within scope for our paper.

Specific Reviewer Questions:

Review #408B:

*) l 826: What do you mean by "When every instance of a class is @Det"?

When defining a class, a programmer can specify that all instances are deterministic.

Suppose a method “foo” has type “@PolyDet foo (@PolyDet X arg)”. This method must type check for every instantiation of “@PolyDet”. However, if class X has type “@Det class X”, it means that every valid instance of X will be “@Det”. So it is sufficient that “foo” type checks for “@Det”. Our checker currently doesn’t recognize this and issues false warnings as a result.

Review #408C:

*) How does a programmer reason about buggy annotations?

If a programmer writes a wrong annotation, it will be caught during compilation by our type checker.  There is no type loophole.
Consider a field.  If the annotation is too strict (@Det when it should be @NonDet), then assignments to the field will fail.  If the annotation is too loose (@NonDet when it should be @Det), then uses of the field will fail, such as attempts to pass the field to a method that requires a @Det argument.

As a minor point, your example said “labels an operation as Det while in fact, it is NonDet”, but our system is a type system where types are labeled (qualified), not an effect system where operations are labeled.

Review #408D:

*) line 124: What does it mean that “$\beta$” lacks a top-level type?

A type in our type system consists of two components: a type qualifier and a basetype. ““$\beta$” lacks a top-level type” means that “$\beta$” doesn’t have a type qualifier (Explained in Section 2.1).

*) line 141: Does “different values” mean non-equal values (incl. objects) or non-identical values (objects)? … I’d like to see a discussion of how Java’s spec for equals (including for HashMaps and HashSets) interacts with your type rules (and soundness)!

Yes, equality is determined by the equals() method. When a class overrides the equals() method, this more specific equals() method establishes determinism guarantees.  This applies to HashSets and HashMaps as well.

*) line 142: I think it would be indicated to say what OrderNonDet means in terms of identity or equality of the collection object.

OrderNonDet has no effect on identity or equality.  It indicates the nondeterministic iteration order of a collection.

*) line 192: How do NonDet collections come about? As elements of OrderNonDet collections? Why, then, does not the type of every generically typed member of a generic type need to be NonDet? E.g., is

class C<T> { T f; }
NonDet C<Det Object> c = …

legal?

NonDet collections could be constructed as a result of iterating over other Nondet or OrderNonDet collections.  They can also be constructed from random values (or values that are not known to be deterministic).
“NonDet C<Det Object>” is illegal according to our type well-formedness rules (Section 2.3). For generic types, our default upper bound is “NonDet”

*) line 228 and following: I am confused. Is not the problem using an indet index? Also, AFAIK, neither Collection nor Map offer indexed access. If you need to cater for a type hole that List could introduce, how about other (present and future) subtypes of Collection?

The problem is aliasing where a “Nondet List<Det Int>” and a “Det List<Det Int>” both point to the same “List” instance. Mutating a “NonDet” location in the “NonDet List<Det Int>” causes the determinism guarantees to break for the “Det List<Det Int>”. Disallowing “NonDet List<Det Int>” ensures that the aliasing these two types isn’t possible (Because of Java’s invariant subtyping rules). 
While Collection and Map do not offer indexed accesses, not allowing “NonDet Collection<Det ...>” or “NonDet Map<Det ...>” has no impact on the expressivity of valid types.

*) line 269 and following: This may be my ignorance, but I do not understand the meaning of the “in this case” phrases, when the preceding statement is universally quantified over $\kappa$.

Our writing wasn’t clear here. Consider the first type of “get”. We mean that if the return type of “get” was “NonDet $beta$”, then “$kappa$” was instantiated as “NonDet”.

*) lines 305 and 309: what do you mean by “common”? Frequent in your experiments?

“common” as in common for any type system.

*) line 563: I find the overriding of type qualifiers provided at the use (instantiation) site, at the definition site, ad hoc. It appears there are no rules (restrictions) for this overriding? How does this affect soundness?

A type variable represents any type. The type qualifier of its lower bound is the bottom type and that of its upper bound is the top type. Overriding this with a specific type qualifier at instantiation cannot allow a type that isn’t otherwise allowed. So this does not affect soundness.

Review #408E

*) You also then reveal that this system is flow sensitive. How did you do that? Please give more details.

We built our checker on top of the Checker Framework.  Any type system built with it is automatically flow-sensitive. It has a built-in dataflow framework.

### 2019-06-24 design.tex
Use external bibliography when possible
Add ToDos.

@@
 \ptype{NonDet List}{Det String}.
+\todo{This isn't true.  Using standard typing rules, the type of an array index
+expression, "ni" in this example, doesn't have to be related to the array
+element type. The Determinism Checker adds a rule for this.}
+\todo{We also need to add a section about non-standard typing rules. We
+implemented non-standard rules for both fields and arrays.}
+\todo{ Also, we should point out that nondet collection types are not related
+to order non det collection types.  But order non det collection types are a
+supertype of det collection types.  We use this fact elsewhere in the paper,
+but don't explicitly state it.}
 
@@
 \OurTypeSystem supports parametric
-polymorphism~\cite{Abadi:1989:FIM:77350.77373,Plotkin:1993:LPP:645891.671433}.
+polymorphism~\cite{AbadiPG1989,PlotkinA1993}.
 A polymorphic abstraction (a class or method) is written and
@@
 $\kappa = \<NonDet>$ and $\kappa = \<Det>$.  An invocation on an
-\<OrderNonDet> collection would choose the \<NonDet> instantiation, and the
+\<OrderNonDet> collection would choose the \<NonDet> instantiation,
+\todo{Because OrderNonDet collection is not a subtype of NonDet collections,
+this isn't true.  It just would not be legal to call the method with an
+OrderNonDet collection.  Any time @PolyDet is on both a collection type and a
+non-collection type, then the method can't be called on an OrderNonDet
+collection.  This is also true when @PolyDet("use") is used.}
+and the
 return type at that call site would be \<NonDet Int> rather than the
@@
 
+\todo{Because NonDet collections are not related to OrderNonDet collections, this is not the most specific instantiation. Without @PolyDet("up"), first can't be called on an OrderNonDet collection.}
 \noindent making the type of the expression \<NonDet String> as desired.
@@
 \noindent
-Then the invocation \<first(listOfLists)> is illegal.
+Then the invocation \<first(listOfLists)> is illegal. \todo{I don't understand why this is illegal, but \<first(myOndStrings)> isn't}
 Its type would be \ptype{NonDet List}{Det String}, which is an invalid
@@
 \begin{Verbatim}
-    @PolyDet("up") E get(@PolyDet List<E> this, @PolyDet int index);
+    @PolyDet("up") E get(@PolyDet List<E> this, @PolyDeti("up") int index);
     @PolyDet("up") E set(@PolyDet List<E> this, @PolyDet("use") int index, E element);
@@
 The syntax is explained in \cref{java-polymorphism}.)
+
+\todo{Add a section for @PolyDet("upDet") which should be renmaed to @PolyDet("shuffle")}
 
@@
 
+\todo{\<shuffle> should only be called on an @NonDet Collection or an
+OrderNonDet collection that has no alaises that are Det collections.}
 As another example, it is only sometimes permitted to call \<shuffle> on a receiver of
@@
 has \codeid{@OrderNonDet List} or a subtype within its type argument.
+\todo{The type argument must be Det for any base type.}
 Without this rule, the type would be \<NonDet> which is sound but imprecise.
@@
 improves precision when calling \<get> on an \<OrderNonDet> receiver.
+\todo{I don't understand why the return type isn't just PolyDet("up")?}
 

### 2019-06-24 java-design.tex
Use external bibliography when possible
Add ToDos.

@@
 Therefore, \<@PolyDet> may be written in methods (signatures and bodies)
-but not on fields.
+but not on fields. \todo{PolyDet on fields is now supported.}
 \end{itemize}
@@
     \end{verbatim}
-    \caption{Example: Error detected by \theDeterminismChecker in \<scribe-java>~\cite{nondex}.}
+    \caption{Example: Error detected by \theDeterminismChecker in \<scribe-java>~\cite{ShiGLM2016}.}
 %    \todo{Use real examples of bugs from case studies, throughout.  Artificial
@@
 \<@Det String @Det []>, a deterministic array of deterministic strings.
+\todo{Add a method declaration annotation EntryPoint that changes the type of all parameters to Det.}
 
